### 2020-06-11 paper/rei.tex
added rei.tex, rei.pdf

\documentclass[prd,twocolumn,amsmath,amssymb,floatfix,superscriptaddress,nofootinbib]{revtex4-1}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{epsfig}
\usepackage{color}
\usepackage{natbib}
\usepackage{textcase}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{xstring}
\usepackage{graphicx}
\usepackage[utf8]{inputenc} 
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{epstopdf}
\epstopdfsetup{update}
\DeclareGraphicsExtensions{.ps, .png}
\epstopdfDeclareGraphicsRule{.ps}{pdf}{.pdf}{ps2pdf -dEPSCrop -dNOSAFER #1 \OutputFile} 
\usepackage{dcolumn} 
\usepackage{multirow}
\usepackage{appendix}
\usepackage{footnote}
\usepackage{tabularx,ragged2e,booktabs}
\usepackage[normalem]{ulem}
\usepackage{float}
\restylefloat{table}

\newcommand{\Omegamzero}{\Omega_{{\rm m,0}}}
\newcommand{\Rbar}{$\bar{R}$}
\newcommand{\lsc}{\mathcal{L}}
\newcommand{\rhom}{\rho_{\rm m}}
\newcommand{\Mpch}{\mbox{Mpc}/h}
\newcommand{\iMpch}{h/\mbox{Mpc}}
\newcommand{\Msun}{M_\odot}
\newcommand{\Mv}{M_{\rm v}}

\newcommand{\refsec}[1]{section~\ref{sec:#1}}
\newcommand{\refeq}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\refFig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\curv}{{\cal R}}
\newcommand{\xef}{x_e^{\rm fid}}
\newcommand{\xmax}{x_e^{\rm max}}
\newcommand{\zmax}{z_{\rm max}}
\newcommand{\zmin}{z_{\rm min}}
\newcommand{\xemin}{x_e^{\rm min}}

\newcommand{\ra}{\rightarrow}
\def\max{_{\mathrm{max}}}
\def\lsim{\mathrel{\raise.3ex\hbox{$$<$$\kern-.75em\lower1ex\hbox{$\sim$}}}}
\def\gsim{\mathrel{\raise.3ex\hbox{$$>$$\kern-.75em\lower1ex\hbox{$\sim$}}}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}

\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}

\newcommand{\wh}[1]{\textcolor{blue}{#1}}
\newcommand{\ch}[1]{\textcolor{red}{#1}}

\def\mnras{Mon.\ Not.\ R.\ Astron.\ Soc.\ }
\definecolor{darkgreen}{cmyk}{0.85,0.2,1.00,0.2} 
\definecolor{purple}{cmyk}{0.5,1.0,0,0} 
\def\physrep{Phys.~Rep.}

\definecolor{ultramarine}{rgb}{0.07, 0.04, 0.56}
\definecolor{cadmiumgreen}{rgb}{0.0, 0.42, 0.24}
\definecolor{indigo(dye)}{rgb}{0.0, 0.25, 0.42}
\usepackage[linktocpage=true]{hyperref}
\hypersetup{
colorlinks=true,
citecolor=ultramarine,
linkcolor=cadmiumgreen,
urlcolor=indigo(dye),
pdfauthor={},
pdftitle={},
pdfsubject={}
}

\begin{document}

\title{Reionization Planck 2018 ...}

\author{Chen Heinrich}\email{chenhe@caltech.edu}
\affiliation{$Jet\ Propulsion\ Laboratory,\ California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
\affiliation{$California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}

\author{Wayne Hu}
\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago Illinois 60637}
\affiliation{Department of Astronomy \& Astrophysics,
 University of Chicago, Illinois 60637}

\begin{abstract}

...

\end{abstract}
\pacs{}

\maketitle

\section{Outline (in progress)}

\begin{enumerate}
	\item{Intro}
	\item{Background}
		\begin{itemize}
			\item{Reionization Principal Components}	
			\item{Kernel Density Estimate}
		\end{itemize}
	\item{Planck 2018 PC results:\\
		- can discuss discrepancy here on high-redshift tau($>$15) constraints. }
	\item{Effective Likelihood}
		\begin{itemize}
			\item{Code Description}
			\item{Examples - one and two parameter models}
		\end{itemize}
	\item{Discussion}

\end{enumerate}

\section{Introduction}
\label{sec:intro}

\begin{figure}[t]
\includegraphics[width=0.40\textwidth]{plots/plot_tau_gtz.pdf}
\caption{...
}
\label{fig:}
\end{figure}

\begin{figure}
\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
\caption{...
}
\label{fig:}
\end{figure}

\bibliography{rei.bib}

\end{document}

### 2020-06-23 results/notes.tex

--- 2020-06-11 paper/rei.tex
@@
 
-\section{Outline (in progress)}
-
-\begin{enumerate}
-	\item{Intro}
-	\item{Background}
-		\begin{itemize}
-			\item{Reionization Principal Components}	
-			\item{Kernel Density Estimate}
-		\end{itemize}
-	\item{Planck 2018 PC results:\\
-		- can discuss discrepancy here on high-redshift tau($>$15) constraints. }
-	\item{Effective Likelihood}
-		\begin{itemize}
-			\item{Code Description}
-			\item{Examples - one and two parameter models}
-		\end{itemize}
-	\item{Discussion}
-
-\end{enumerate}
-
 \section{Introduction}
@@
 
+Trying to collect all of our tests here.\\
+
+List of tests:\\
+-
+
 \begin{figure}[t]
-\includegraphics[width=0.40\textwidth]{plots/plot_tau_gtz.pdf}
+\includegraphics[width=0.40\textwidth]{../paper/plots/plot_tau_gtz.pdf}
 \caption{...
@@
 \begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+\includegraphics[width=0.40\textwidth]{python_kde_test/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
 \caption{...

### 2020-06-23 results/notes.tex

@@
 List of tests:\\
--
+\begin{itemize}
+    \item {Direct MCMC chains
+        \begin{enumerate}
+            \item {PC zmax = 30}
+            \item {PC zmax = 50}
+            \item{tanh tau}
+            \item{tanh highz model}
+            \item{tanh highz model: fixed tau_lo, MCMC tau_hi ?}
+        \end{enumerate}
+    }
+
+    \item {Direct MCMC chains}
+
+\end{itemize}
+
+\begin{figure}[t]
+\includegraphics[width=0.40\textwidth]{results/cosmomc_runs/pl18_plots_zmax50/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax50_pliklite_post_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{Comparing zmax = 50 PC chains using Planck 2018 original lowE vs srollv2 likelihood.
+}
+\label{fig:}
+\end{figure}
 

### 2020-06-23 results/notes.tex
Updates from Overleaf

@@
             \item{tanh highz model}
-            \item{tanh highz model: fixed tau_lo, MCMC tau_hi ?}
+        \end{enumerate}
+    }
+    \item{best tests
+        \begin{enumerate}
+            \item {from tanh highz 2-parameter model we found that for tau hi > 0.02, we get ML is tau hi = 0.0206, tau lo = 0.0414 corresponding to zre = 6.29 and xe hi = 0.0748.
+This model has a -loglike = 499.5458 against the full ML 498.2425 which gives a difference of 2.6 in chi2.
+(full ML has tau lo = 0.0543, tau hi = 0.00542 corresponding to zre = 7.68 and xe hi = 0.0214.
+}
+            \item{Using above informatiion, we fixed tau hi = 0.02, and ran a best-fit searching for best tau lo and found tau lo = 0.045 (c.f.
+0.0414 from chains), together with the fixed tau hi = 0.02, corresponds to zre = 6.67 and xe min = 0.073.
+The likelihood here is 499.2914, a difference of 2.1 (instead of 2.6) in chi2.}
         \end{enumerate}
@@
 
-\begin{figure}[t]
+\begin{figure}
+\includegraphics[width=0.40\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{Comparing zmax = 50 PC chains using Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
 \includegraphics[width=0.40\textwidth]{results/cosmomc_runs/pl18_plots_zmax50/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax50_pliklite_post_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
-\caption{Comparing zmax = 50 PC chains using Planck 2018 original lowE vs srollv2 likelihood.
+\caption{Comparing zmax = 50 PC chains using Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.40\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax50_pliklite_post.pdf}
+\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE).
+Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
+}
+\label{fig:tau_gtz_zmax_30_vs_50_simall_EE}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.40\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{Same as Fig.~\ref{fig:tau_gtz_zmax_30_vs_50_simall_EE}, but for the srollv2 likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
 }

### 2020-06-23 results/notes.tex
Updates from Overleaf

--- 2020-06-23 results/notes.tex
@@
-\documentclass[prd,twocolumn,amsmath,amssymb,floatfix,superscriptaddress,nofootinbib]{revtex4-1}
+\documentclass[prd,amsmath,amssymb,floatfix,superscriptaddress,nofootinbib]{revtex4-1}
 \usepackage{bm}
@@
 
-\title{Reionization Planck 2018 ...}
+\title{Reionization Planck 2018 Notes}
 
@@
 List of tests:\\
--
+\begin{itemize}
+    \item {Direct MCMC chains
+        \begin{enumerate}
+            \item {PC zmax = 30}
+            \item {PC zmax = 50}
+            \item{tanh tau}
+            \item{two-parameter model with taulo and tauhi}
+        \end{enumerate}
+    }
+    \item{best tests
+        \begin{enumerate}
+            \item {from tanh highz 2-parameter model we found that for tau hi $>$ 0.02, we get ML is tau hi = 0.0206, tau lo = 0.0414 corresponding to zre = 6.29 and xe hi = 0.0748.
+This model has a -loglike = 499.5458 against the full ML 498.2425 which gives a difference of 2.6 in chi2.
+(full ML has tau lo = 0.0543, tau hi = 0.00542 corresponding to zre = 7.68 and xe hi = 0.0214.
+}
+            \item{Using above informatiion, we fixed tau hi = 0.02, and ran a best-fit searching for best tau lo and found tau lo = 0.045 (c.f.
+0.0414 from chains), together with the fixed tau hi = 0.02, corresponds to zre = 6.67 and xe min = 0.073.
+The likelihood here is 499.2914, a difference of 2.1 (instead of 2.6) in chi2.}
+        \end{enumerate}
+        }
 
-\begin{figure}[t]
-\includegraphics[width=0.40\textwidth]{../paper/plots/plot_tau_gtz.pdf}
-\caption{...
+\end{itemize}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
 }
@@
 \begin{figure}
-\includegraphics[width=0.40\textwidth]{python_kde_test/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
-\caption{...
+\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax50/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax50_pliklite_post_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{PC chains for zmax = 50. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax50_pliklite_post.pdf}
+\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE).
+Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
+}
+\label{fig:tau_gtz_zmax_30_vs_50_simall_EE}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{Same as Fig.~\ref{fig:tau_gtz_zmax_30_vs_50_simall_EE}, but for the srollv2 likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:}
+\end{figure}
+
+%Optional
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
+\caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
+}
+\label{fig:}
+\end{figure}
+
+ \begin{figure}
+\includegraphics[width=0.65\textwidth]{   results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_xe_full.pdf}
+\caption{PC zmax = 30 vs tanh chains with plik lite and srollv2, with ML model from two-parameter model for tau hi restricted to $>$ 0.02 (ML model: tau hi = 0.0206, tau lo = 0.0414).
+}
+\label{fig:two_parameter_model_ML}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{  results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_PC_proj.pdf}
+\caption{Same as Fig.~\ref{fig:two_parameter_model_ML} except the blue curve is a PC projection instead of the full xe(z) of the model.
+}
+\label{fig:}
+\end{figure}
+
+ \begin{figure}
+\includegraphics[width=0.65\textwidth]{    results/cosmomc_runs/two_parameter_model/plot_1D_tau_hi_pl18_tanh_highz_fixed_taulo_test2_run1.pdf
+}
+\caption{
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{python_kde_test/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+\caption{KDE tests in python for the tanh model with various f = 0.1, 0.12, 0.14.
+It is probably safest to use any of these, so given the smaller f the better, f = 0.1 is probably the best choice here.
 }

### 2020-06-23 results/notes.tex
Updates from Overleaf

@@
 
-\newcommand{\Omegamzero}{\Omega_{{\rm m,0}}}
-\newcommand{\Rbar}{$\bar{R}$}
-\newcommand{\lsc}{\mathcal{L}}
-\newcommand{\rhom}{\rho_{\rm m}}
-\newcommand{\Mpch}{\mbox{Mpc}/h}
-\newcommand{\iMpch}{h/\mbox{Mpc}}
-\newcommand{\Msun}{M_\odot}
-\newcommand{\Mv}{M_{\rm v}}
+\newcommand{\tauhi}{$\tau_{\rm hi}\,$}
+\newcommand{\taulo}{$\tau_{\rm lo}\,$}
 
@@
 
-\author{Chen Heinrich}\email{chenhe@caltech.edu}
-\affiliation{$Jet\ Propulsion\ Laboratory,\ California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
-\affiliation{$California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
-
+\author{Chen Heinrich}
 \author{Wayne Hu}
-\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago Illinois 60637}
-\affiliation{Department of Astronomy \& Astrophysics,
- University of Chicago, Illinois 60637}
-
-\begin{abstract}
-
-...
-
-\end{abstract}
-\pacs{}
 
@@
             \item {PC zmax = 50}
-            \item{tanh tau}
-            \item{two-parameter model with taulo and tauhi}
+            \item{Tanh model}
+            \item{Two-parameter model with \taulo and \tauhi}
         \end{enumerate}
     }
-    \item{best tests
+    \item{Best-fit tests
         \begin{enumerate}
-            \item {from tanh highz 2-parameter model we found that for tau hi $>$ 0.02, we get ML is tau hi = 0.0206, tau lo = 0.0414 corresponding to zre = 6.29 and xe hi = 0.0748.
-This model has a -loglike = 499.5458 against the full ML 498.2425 which gives a difference of 2.6 in chi2.
-(full ML has tau lo = 0.0543, tau hi = 0.00542 corresponding to zre = 7.68 and xe hi = 0.0214.
+            \item {From the chains for the 2-parameter model, we found that if we restrict \tauhi $> 0.02$, we get that the ML model is: 
+            \begin{itemize}
+                \item {\tauhi = 0.0206}
+                \item {\taulo = 0.0414} 
+                \item {$z_{\rm re}$ = 6.29} 
+                \item {$x_{e, \mathrm{hi}}$ = 0.0748}
+                \item {-loglike = 499.5458} 
+            \end{itemize}
+            There is a difference of $\Delta \chi^2 = 2.6$, compared against the ML model from the full chain:
+            \begin{itemize}
+                \item {\tauhi = 0.00542}
+                \item {\taulo = 0.0543} 
+                \item {$z_{\rm re}$ = 7.68} 
+                \item {$x_{e, \mathrm{hi}}$ = 0.0214}
+                \item {-loglike = 498.2425} 
+            \end{itemize}
              }
-            \item{Using above informatiion, we fixed tau hi = 0.02, and ran a best-fit searching for best tau lo and found tau lo = 0.045 (c.f.
-0.0414 from chains), together with the fixed tau hi = 0.02, corresponds to zre = 6.67 and xe min = 0.073.
-The likelihood here is 499.2914, a difference of 2.1 (instead of 2.6) in chi2.}
+            \item{Using above information, we ran a best-fit search fixing \tauhi = 0.02 to find the best-fit model:
+            \begin{itemize}
+                \item {\tauhi = 0.02 (fixed)}
+                \item {\taulo = 0.045 (c.f. 0.0414 from chains)} 
+                \item {$z_{\rm re}$ = 6.67} 
+                \item {$x_{e, \mathrm{hi}}$ = 0.073}
+                \item {-loglike = 499.2914} 
+            \end{itemize} 
+            giving a difference of $\Delta \chi^2 = 2.1$  (c.f. 2.6 above).}\\
+            Note: I need to update the plots with xe(z) to this model.
+            \item{We also ran a chain fixing \taulo = 0.04, but I forgot what this was for ...}
+        \end{enumerate}
+        }
+
+    \item {KDE tests
+        \begin{enumerate}
+            \item {The tanh model: direct MCMC, vs python (f = 0.10, 0.12, 0.14) and fortran version (0.14) - coming up}
+            \item {The 2-parameter model: direct MCMC vs fortran version (f = 0.10, 0.12, 0.14) - coming up}
         \end{enumerate}
@@
  \begin{figure}
+\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_2D.pdf}
+\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: 2D plot of \tauhi and \taulo.
+}
+\label{fig:two_parameter_model_2D_plot}
+\end{figure}
+
+ \begin{figure}
 \includegraphics[width=0.65\textwidth]{   results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_xe_full.pdf}
-\caption{PC zmax = 30 vs tanh chains with plik lite and srollv2, with ML model from two-parameter model for tau hi restricted to $>$ 0.02 (ML model: tau hi = 0.0206, tau lo = 0.0414).
+\caption{PC zmax = 30 vs tanh chains with plik lite and srollv2, with ML model from two-parameter model for \tauhi restricted to $>$ 0.02 (ML model: \tauhi = 0.0206, \taulo = 0.0414).
 }
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{    results/cosmomc_runs/two_parameter_model/plot_1D_tau_hi_pl18_tanh_highz_fixed_taulo_test2_run1.pdf
-}
-\caption{
-}
+\includegraphics[width=0.65\textwidth]{   results/cosmomc_runs/two_parameter_model/plot_1D_tau_hi_pl18_tanh_highz_fixed_taulo_test2_run1.pdf}
+\caption{Chains for two-parameter model with fixed \taulo = 0.04: 1D distribution of \tauhi (not sure what this is for anymore...).}
 \label{fig:}

### 2020-06-24 results/notes.tex
Updates from Overleaf

@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\includegraphics[width=0.65\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
 \caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax50/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax50_pliklite_post_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\includegraphics[width=0.65\textwidth]{results/direct_mcmc/pl18_plots_zmax50/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax50_pliklite_post_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
 \caption{PC chains for zmax = 50. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax50_pliklite_post.pdf}
+\includegraphics[width=0.65\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax50_pliklite_post.pdf}
 \caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE).
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\includegraphics[width=0.65\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
 \caption{Same as Fig.~\ref{fig:tau_gtz_zmax_30_vs_50_simall_EE}, but for the srollv2 likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
+\includegraphics[width=0.65\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
 \caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{results/cosmomc_runs/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_2D.pdf}
-\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: 2D plot of \tauhi and \taulo.
+\includegraphics[width=0.65\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
+\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
 }
@@
  \begin{figure}
-\includegraphics[width=0.65\textwidth]{   results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_xe_full.pdf}
+\includegraphics[width=0.65\textwidth]{   results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_xe_full.pdf}
 \caption{PC zmax = 30 vs tanh chains with plik lite and srollv2, with ML model from two-parameter model for \tauhi restricted to $>$ 0.02 (ML model: \tauhi = 0.0206, \taulo = 0.0414).
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{  results/cosmomc_runs/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_PC_proj.pdf}
+\includegraphics[width=0.65\textwidth]{  results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_tanh_post_pliklite_srollv2_with_added_two_parameter_ML_PC_proj.pdf}
 \caption{Same as Fig.~\ref{fig:two_parameter_model_ML} except the blue curve is a PC projection instead of the full xe(z) of the model.
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{   results/cosmomc_runs/two_parameter_model/plot_1D_tau_hi_pl18_tanh_highz_fixed_taulo_test2_run1.pdf}
+\includegraphics[width=0.65\textwidth]{   results/direct_mcmc/two_parameter_model/fixed_taulo_chains/plot_1D_tau_hi_pl18_tanh_highz_fixed_taulo_test2_run1.pdf}
 \caption{Chains for two-parameter model with fixed \taulo = 0.04: 1D distribution of \tauhi (not sure what this is for anymore...).}
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{python_kde_test/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+\includegraphics[width=0.65\textwidth]{python_kde/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
 \caption{KDE tests in python for the tanh model with various f = 0.1, 0.12, 0.14.

### 2020-06-25 results/notes.tex
Updates from Overleaf

@@
 \caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
+The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
+The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
 }
@@
 \caption{PC zmax = 30 vs tanh chains with plik lite and srollv2, with ML model from two-parameter model for \tauhi restricted to $>$ 0.02 (ML model: \tauhi = 0.0206, \taulo = 0.0414).
+(Note: I need to update the xe(z) curve to be that of the best-fit model in the best-fit search with tau_hi = 0.02 fixed.)
 }
@@
 
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/direct_mcmc_vs_python_kde_and_cosmomc_kde_tanh_tau.png}
+\caption{LATEST (to be debugged) -- Added cosmomc KDE version of tau posterior computed with f = 0.14, against python versions with f = 0.1, 0.12, 0.14.
+They do not agree ...
+(Note: need to make sure python version had the same covariance matrix, and length of chain ...)
+}
+\label{fig:}s
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/direct_mcmc_vs_python_kde_and_cosmomc_kde_tanh_tau.png}
+\caption{LATEST (to be debugged) -- Added cosmomc KDE version of tau posterior computed with f = 0.14, against python versions with f = 0.1, 0.12, 0.14.
+They do not agree ...
+(Note: need to make sure python version had the same covariance matrix, and length of chain ...)
+}
+\label{fig:}s
+\end{figure}
+
 \bibliography{rei.bib}

### 2020-06-25 results/notes.tex
Updates from Overleaf

@@
 They do not agree ...
-(Note: need to make sure python version had the same covariance matrix, and length of chain ...)
+(Note: need to make sure python version had the same covariance matrix, and length of chain ...
+should probably rerun python version actually.)
 }
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{cosmomc_kde/direct_mcmc_vs_python_kde_and_cosmomc_kde_tanh_tau.png}
-\caption{LATEST (to be debugged) -- Added cosmomc KDE version of tau posterior computed with f = 0.14, against python versions with f = 0.1, 0.12, 0.14.
-They do not agree ...
-(Note: need to make sure python version had the same covariance matrix, and length of chain ...)
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_tanh_highz_test2_run1_vs_relike_tanh_highz_test1_run1_f0p14_tri.png}
+\caption{LATEST (to be debugged) -- 2-parameter model 2D plot from cosmomc KDE with f = 0.14 vs direct MCMC.
+They don't agree for low \taulo, in particular there is a strip around \tauhi = 0.02 that seems to have large likelihood (Note: perhaps I should pick a single point in that region to check that PC projection works fine and see if it is something with the PC chain when doing the KDE ...
+Also, there is a \taulo lower bound = 0.02 used just like for the direct MCMC, but giving getdist that limit doesn't help.
+We need this lower bound because otherwise the conversion \taulo, \tauhi to xemin and zre doesn't converge.)
 }
-\label{fig:}s
+\label{fig:}
 \end{figure}

### 2020-07-02 results/notes.tex
Updates from Overleaf

--- 2020-06-25 results/notes.tex
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{cosmomc_kde/direct_mcmc_vs_python_kde_and_cosmomc_kde_tanh_tau.png}
-\caption{LATEST (to be debugged) -- Added cosmomc KDE version of tau posterior computed with f = 0.14, against python versions with f = 0.1, 0.12, 0.14.
-They do not agree ...
-(Note: need to make sure python version had the same covariance matrix, and length of chain ...)
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_tanh_highz_test2_run1_vs_relike_tanh_highz_test1_run1_f0p14_tri.png}
+\caption{version 1 (has bug that \taulo is not limited to where PCs can model z>6): 2-parameter model 2D plot from cosmomc KDE with f = 0.14 vs direct MCMC.
 }
-\label{fig:}s
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_tanh_highz_test2_run1_vs_relike_tanh_highz_test1_run1_f0p14_taulo_prior_0p0415_tri.png}
+\caption{version 2 (added prior \taulo $\leq 0.0415$ for the KDE version, f = 0.14). Direct MCMC chain with the same prior is still running.
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_tanh_highz_test2_run1_vs_relike_tanh_highz_test1_run1_f0p14_and_f0p20_taulo_prior_0p0415_tri.png}
+\caption{version 3, added a curve with f = 0.20 (blue).
+}
+\label{fig:}
 \end{figure}

### 2020-07-02 results/notes.tex
Updates from Overleaf

@@
 }
-\label{fig:}s
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_and_python_kde.png}
+\caption{DEBUGGING - updated the KDE tanh results to have tanh dz = 0.5. Also added in faint blue lines Gaussian and convolved Gaussian with $\sigma \rightarrow \sigma$.
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{ cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_kde.png}
+\caption{DEBUGGING - tanh dz = 0.5 for all curves. Same as above, but only keeping some curves for clarity.)
+}
+\label{fig:}
 \end{figure}

### 2020-07-06 results/notes.tex

@@
 \begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_and_python_kde_double_checked_python.png}
+\caption{DEBUGGING - updated the python curves with the same covariance matrix, chains, etc (also confirmed that tanh dz = 0.5). 
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
 \includegraphics[width=0.65\textwidth]{ cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_kde.png}
 \caption{DEBUGGING - tanh dz = 0.5 for all curves. Same as above, but only keeping some curves for clarity.)
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_and_python_kde_with_grid_test.png}
+\caption{DEBUGGING - tanh dz = 0.5 for all curves. ...)
 }

### 2020-07-07 results/notes.tex
Updates from Overleaf

@@
 \includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_and_python_kde.png}
-\caption{DEBUGGING - updated the KDE tanh results to have tanh dz = 0.5. Also added in faint blue lines Gaussian and convolved Gaussian with $\sigma \rightarrow \sigma$.
+\caption{DEBUGGING - updated the KDE tanh results to have tanh dz = 0.5. Also added in faint blue lines Gaussian and convolved Gaussian with $\sigma \rightarrow \sigma \sqrt(1+f)$.
 }
@@
 \includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_and_python_kde_with_grid_test.png}
-\caption{DEBUGGING - tanh dz = 0.5 for all curves. ...)
+\caption{NEW - added blue circles which are evaluating the Fortran KDE code on a grid of 100 values of $\tau$.
+This shows that the extra smoothing in the red curve (Fotran MCMC code) vs green (python grid evaluations) are \textbf{likely due to the smoothing used in getdist to make the red curve}.
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_vs_mock_gaussian_fp10_fp14.png
+}
+\caption{NEW - comparing f = 0.10 and 0.14 in python KDE (latest version) against an example of gaussian smoothing with 0.10, 0.14 on a symmetric Gaussian function.
 }

### 2020-07-09 results/notes.tex
Updates from Overleaf

@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_vs_mock_gaussian_fp10_fp14.png
-}
-\caption{NEW - comparing f = 0.10 and 0.14 in python KDE (latest version) against an example of gaussian smoothing with 0.10, 0.14 on a symmetric Gaussian function.
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_1D_direct_vs_fortran_vs_mock_gaussian_fp10_fp14.png}
+\caption{NEW - comparing f = 0.10 (orange dotted) and 0.14 (green dotted) in python KDE (latest version) against an example of gaussian smoothing with f = 0.10 (orange dashed) and 0.14 (blue dashed) on a Gaussian function (blue solid).
 }

### 2020-07-21 results/notes.tex

@@
 
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_tanh_highz_test2_run1_vs_relike_tanh_highz_test1_run1_f0p14_and_f0p20_taulo_prior_0p0415_tri.png}
+\label{fig:}
+\end{figure}}
+\caption{NEW: Direct MCMC without prior (gray) vs with prior (red) vs KDE fortran with same prior (\taulo \geq 0.0415) and $f = 0.14$.
+} 
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+\label{fig:}
+\end{figure}}
+\caption{Testing \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
+The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l) = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1) $.
+} 
+\label{fig:taulo_prior_test_cl}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+\label{fig:}
+\end{figure}}
+\caption{Full $x_e(z)$ of the model (\taulo, \tauhi) = (0.04, 0.02) shown in the Fig.~\ref{taulo_prior_test_cl}. 
+} 
+\label{fig:taulo_prior_test_xez}
+\end{figure}
+
 \bibliography{rei.bib}

### 2020-07-21 results/notes.tex

--- 2020-07-09 results/notes.tex
@@
             \item {The 2-parameter model: direct MCMC vs fortran version (f = 0.10, 0.12, 0.14) - coming up}
+        \end{enumerate}
+    }
+
+    \item {Preparation for manuscript
+        \begin{enumerate}
+            \item {Testing \taulo prior (\taulo $\geq 0.04$) - Figs~\ref{fig:taulo_prior_test_cl} and~\ref{fig:tau_lo_prior_test_xe}: $\chi^2 = 0.03$ for $l\in[2, 100]$ between $C_l^{EE}$ from full vs PC projected $x_e(z)$ for the model (\taulo, \tauhi) = (0.04, 0.02).}
+            \item {Plots to remake for paper: 1) }
         \end{enumerate}
@@
 
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/pl18_tanh_highz_test2_run1_vs_relike_tanh_highz_test1_run1_f0p14_and_f0p20_taulo_prior_0p0415_tri.png}
+\label{fig:}
+\caption{NEW: Direct MCMC without prior (gray) vs with prior (red) vs KDE fortran with same prior (\taulo \geq 0.0415) and $f = 0.14$.
+} 
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_gauss_smoothed_tanh_std_0p0067.png}
+\label{fig:}
+\caption{Gaussian convolving the direct MCMC 1D distribution by hand with f = 0.14 where the original standard deviation is set to 0.0067.
+This corresponds to the right side of asymmetric distribution from the original MCMC $\tau = 0.0591 + 0.0055 - 0.0067$.
+} 
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_gauss_smoothed_tanh_std_0p0055.png}
+\label{fig:}
+\caption{Gaussian convolving the direct MCMC 1D distribution by hand with f = 0.14 where the original standard deviation is set to 0.0055.
+This corresponds to the left side of asymmetric distribution from the original MCMC $\tau = 0.0591 + 0.0055 - 0.0067$.
+} 
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
+The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$.
+} 
+\label{fig:taulo_prior_test_cl}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
+\caption{Testing \taulo prior -- Full $x_e(z)$ of the model (\taulo, \tauhi) = (0.04, 0.02) shown in the Fig.~\ref{fig:taulo_prior_test_cl}. 
+}
+\label{fig:tau_lo_prior_test_xe}
+\end{figure}
+
 \bibliography{rei.bib}

### 2020-07-30 results/notes.tex

@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_gauss_smoothed_tanh_std_0p0067.png}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_gauss_smoothed_tanh_std_0p0067.png}
 \label{fig:}
 \caption{Gaussian convolving the direct MCMC 1D distribution by hand with f = 0.14 where the original standard deviation is set to 0.0067.
-This corresponds to the right side of asymmetric distribution from the original MCMC $\tau = 0.0591 + 0.0055 - 0.0067$.
+This corresponds to the right side of asymmetric distribution $\tau = 0.0591 + 0.0055 - 0.0067$ from the original MCMC.
 } 
@@
 \begin{figure}
-\includegraphics[width=0.65\textwidth]{getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_gauss_smoothed_tanh_std_0p0055.png}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_gauss_smoothed_tanh_std_0p0055.png}
 \label{fig:}
 \caption{Gaussian convolving the direct MCMC 1D distribution by hand with f = 0.14 where the original standard deviation is set to 0.0055.
-This corresponds to the left side of asymmetric distribution from the original MCMC $\tau = 0.0591 + 0.0055 - 0.0067$.
+This corresponds to the left side of asymmetric distribution $\tau = 0.0591 + 0.0055 - 0.0067$ from the original MCMC.
 } 
 \end{figure}
+
+\begin{figure}
+\includegraphics[width=0.65\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_python_kde_f_0p1_vs_0p14.png}
+\label{fig:}
+\caption{Comparing f = 0.10 and f = 0.14 lines from python KDE - to see that f = 0.14 is indeed more smoothed than f = 0.10 as expected.
+} 
+\end{figure}
+
+results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_1600.png
+
+results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_800.png
+
+results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_100.png
+
+results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_50.png
 

### 2020-08-06 results/notes.tex
Updates from Overleaf

@@
 
-results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_1600.png
-
-results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_800.png
-
-results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_100.png
-
-results/cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_50.png
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_1600.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_800.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_100.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/getdist_smoothing_test/pl18_relike_tanh_tau_dz0p5_1D_smooth_scale_1D_2_num_bins_1D_50.png}
+\label{fig:}
+\caption{Changing num\_bins = 1600 for the Fortran/cosmomc KDE w/ f = 0.14, which recovers close to the python KDE line w/ the same f. Lowering num\_bins increases smoothing.
+} 
+\end{figure}
 

### 2020-08-06 paper/rei.tex

@@
 
-\section{Outline (in progress)}
-
-\begin{enumerate}
-	\item{Intro}
-	\item{Background}
-		\begin{itemize}
-			\item{Reionization Principal Components}	
-			\item{Kernel Density Estimate}
-		\end{itemize}
-	\item{Planck 2018 PC results:\\
-		- can discuss discrepancy here on high-redshift tau($>$15) constraints. }
-	\item{Effective Likelihood}
-		\begin{itemize}
-			\item{Code Description}
-			\item{Examples - one and two parameter models}
-		\end{itemize}
-	\item{Discussion}
-
-\end{enumerate}
-
 \section{Introduction}
 \label{sec:intro}
+
+\section{Background}
+
+\subsection{Reionization Principal Components}
+\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
+
+\section{Planck 2018 PC Results}
+\label{sec:results}
+
+\subsection{Constraints on Principle Components}
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{results/pc_results/plot_mj_triangle_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite.pdf}
+\caption{PC chains for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+}
+\label{fig:plot_mjs}
+\end{figure}
+
+[Plot: Insert m1-m2 plot about priors]
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
+\caption{m1-m2 plane for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+}
+\label{fig:plot_m1m2}
+\end{figure}
+
+[Plot: Insert m1-m2 plot to show Gaussian covariance does not work]
+
+[Text: recap our appendix from Planck 2015]
+
+[Text: compare with Planck 2015 our PC results]
+
+[Text: statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.]
+
+[Text: make a point that tanh misses high-z again, compare to PCs using the tau($>z$) plot tanh vs PCs.]
+
+[Table: insert table of mean and covariance matrix for the PC components?]
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
+.
+}
+\label{fig:}
+\end{figure}
+
+%Optional
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
+\caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
+}
+\label{fig:}
+\end{figure}
 
@@
 
+\subsection{A note on priors}
+
+[Text: compare with Planck official results]
+
+[Text: comment that Planck tau(15,30) constraints are too stringent (flex knots). Compare to ours. Point out why.]
+
+[Text: compare with Planck official results, PCs (add comments on priors), FlexKnots]
+
+[*Text (maybe in another section): make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck.
+This was for higher tau, so transition happens at higher redshift.
+Now tau is much lower.]
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Effective Likelihood}
+\label{sec:effective_likelihood}
+
+\subsection{Code description}
+\label{sec:code}
+
+\subsection{Example 1: tanh model}
+\label{sec:example1}
+
+\subsection{Example 2: high-z model}
+\label{sec:example2}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
+\caption{Testing \taulo prior -- Full $x_e(z)$ of the model (\taulo, \tauhi) = (0.04, 0.02) shown in the Fig.~\ref{fig:taulo_prior_test_cl}. 
+}
+\label{fig:tau_lo_prior_test_xe}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
+The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$. } 
+\label{fig:taulo_prior_test_cl}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
+\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
+The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
+The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
+}
+\label{fig:two_parameter_model_2D_plot}
+\end{figure}
+
+\section{Conclusion}
+\label{sec:conclusion}
+
 \bibliography{rei.bib}
 
+\appendix
+
+\section{Outline (in progress)}
+
+\begin{enumerate}
+	\item{Intro}
+	\item{Background}
+		\begin{itemize}
+			\item{Reionization Principal Components}	
+			\item{Kernel Density Estimate}
+		\end{itemize}
+	\item{Planck 2018 PC results:\\
+		- can discuss discrepancy here on high-redshift tau($>$15) constraints.\\
+		- compare with our own 2015 PC results
+		- zmax = 30 vs 50}
+	\item{Effective Likelihood}
+		\begin{itemize}
+			\item{Code Description}
+			\item{Examples - one and two parameter models}
+		\end{itemize}
+	\item{Discussion}
+
+\end{enumerate}
+
 \end{document}

### 2020-08-22 paper/rei.tex

@@
 
+\begin{widetext}
+
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{results/pc_results/plot_mj_triangle_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite.pdf}
+\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite.pdf}
 \caption{PC chains for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
@@
 
+\end{widetext}
+
 [Plot: Insert m1-m2 plot about priors]
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
+\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
 \caption{m1-m2 plane for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.

### 2020-09-11 paper/rei.tex

@@
 
+The cosmic microwave background (CMB) has entered an era of precision cosmology.
+The measurements from the \textit{Planck} satellite agree well with predictions from the standard $\Lambda$CDM model describing the initial conditions and evolution of perturbations in the Universe.
+While many components of the standard cosmology model is well-understood, the details of the process of reionization remain however one of the most uncertain pieces.
+Its uncertainty impacts other inferences such as the primordial power spectrum amplitude, through which it will become the dominant source of uncertainty in the future for inferring the sum of neutrino masses from gravitational lensing measurements of the CMB.
+It also has implications for the inferences on cosmic acceleration through the growth of structure.
+[add more refs here, possibly restructure this paragraph.]
+
+Typically, the impact of reionization on the primary CMB fluctuations has been modeled a steplike transition in the global ionization history parametrized by its total Thomson optical depth.
+This model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization beforehand.
+However, the shape of the reionization bump in the large-angle CMB E-mode polarization carries more information on the coarse grained evolution of the ionization history than just an effective transition redshift, which can be used to constrain a variety of more general models.
+
+In fact, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, in which a few principal components describe all of the physical ionization model space with respect to their impact on the large-angle E-mode power spectrum.
+For Planck 2015 data release, this method uncovered a component of high-redshift ionization that would have been missed by a simple steplike model~\cite{Heinrich:2016ojb}.
+It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.
+~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}.
+
+In the latest official Planck 2018 release~\cite{Aghanim:2018eyx}, the PC method was also used to probe ionization at high-redshift, whose significance was reduced since the Planck 2015 release, largely due to the reduction of systematics at large-scales~\cite{Aghanim:2018eyx, Heinrich:2018btc}.
+The FlexKnot method was also used, which could also capture more general ionization history by defining
+``knots" in redshifts and allowing the amplitude of the ionization fraction to vary at those points. [need to polish this paragraph.]
+% Planck intermediate results on reionization history: \cite{Adam:2016hgk}
+
+Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019.
+This new likelihood, called $\textsc{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from XX to XX.
+
+The Planck final release will be the best full-sky survey from space available in the next decade.
+It is therefore important to fully extract all information available to us in the data.
+In light of the $\textsc{srollv2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
+Because of the completeness property of the PCs, the PC constraints can then be turned into an effective likelihood that is useful for assessing the CMB likelihood of any reionization model out to $\zmax$ = 30.
+This is exactly what we did following the development in Ref.~\cite{Heinrich:2016ojb}.
+The code is publicly available on GitHub\footnote{[give link]}.
+[comment on speed].
+[comment on joint likelihood].
+
+We have verified with our PC results that there is no hint of ionization beyond $z>30$.
+But for the high-z ionization below $z=30$, we found a less stringent constraint on the optical depth in the redshift range 15 to 30, $\tau(15, 30)$ that is XX times more less stringent than the Planck official results [make sure to say its flex knot or PC]~\cite{Aghanim:2018eyx}.
+Furthermore we comment on the smaller differences in our PC results, which are due to the difference in the choice of priors used in Planck (following Ref.
+~\cite{Millea:2018bko}), that leads to [.. add conclusion here ...], as shown already in Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data.
+
+The paper is structured as follows.
+We describe in section~\ref{sec:background} the background on reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
+In section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \textsc{srollv2} likelihood, and compare with previous results in literature.
+In section~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the original one-step model in 1).
+We conclude finally in section
+~\ref{sec:conclusion}.
+
 \section{Background}
+\label{sec:background}
 
 \subsection{Reionization Principal Components}
+
+[reword and shorten, refer to previous papers]
+
+We briefly summarize the principal component technique for the complete characterization
+of reionization constraints from the large angle $C_\ell^{EE}$ polarization spectrum
+as introduced in Ref.~\cite{Hu:2003gh} and implemented in \cite{Mortonson:2008rx,Mortonson:2007hq}.
+
+We parametrize the ionization fraction relative to fully ionized hydrogen
+$x_e(z)$ into its
+principal components  with respect to the $E$-mode polarization of the 
+CMB~\citep{Hu:2003gh}:
+\begin{equation}
+x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z).
+\label{eq:mmutoxe}
+\end{equation}
+Here $m_a$ are the PC amplitudes and $S_{a}(z)$
+are the eigenfunctions of the  Fisher information matrix for $x_e(z)$ in a given range  $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements,
+and
+$\xef(z)$ is the fiducial model around which the Fisher matrix is computed.  
+In practice, we discretize the redshift space to $\delta z= 0.25$, well beyond the
+resolution limit of CMB observables, and assume linear interpolation between points to form the
+continuous functions $S_a(z)$.
+The
+components are rank ordered by their Fisher-estimated variances and  in practice
+the first 5 components carry all the information in $C_\ell^{EE}$ to the cosmic variance limit
+\cite{Hu:2003gh}.  In this work, we therefore truncate the PC expansion and retain 5 $m_a$
+parameters to describe reionization.  We take $z_{\rm min}=6$ to be consistent with Ly$\alpha$
+forest constraints (e.g. \cite{Becker:2015lua}) and $z_{\rm max}=30$.
+
+In this truncated representation, the 5 PC decomposition is {\it not} complete in the ionization history itself.   
+Instead it is a complete representation of the {\it observable} impact on $C_\ell^{EE}$ of
+any given $x_e^{\rm true}(z)$ through its projection onto the 5 PC basis
+\begin{equation}
+m_{a}=
+  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin} .
+\label{eq:xetommu}
+\end{equation}
+When reconstructed through Eq.~(\ref{eq:mmutoxe}) with 5 PCs, $x_e(z) \ne x_e^{\rm true}(z)$, even 
+though it models the observables to high precision.   The PC analysis therefore
+is a forward tool to infer constraints on all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ with a single analysis, not an inverse tool that reconstructs the ionization history.
+
+We use a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh} to compute the CMB power spectra given the ionization history.  
+Because CAMB integrates the Boltzmann equation by parts, it requires a smooth ionization 
+history for numerical stability, whereas the $S_a(z)$ are continuous but not smooth.    Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
+Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.   This does not affect our results in a statistically significant way.  
+However for consistency,
+when integrating the ionization history to form the cumulative
+Thomson optical depth 
+\begin{equation}
+\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)} ,
+\label{eq:cumtau}
+\end{equation}
+ we employ the smoothed $x_e$ which
+formally has support beyond the bounds. 
+ We include this small correction by integrating
+slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
+
+For the fiducial ionization history, we take $\xef=0.15$ for $6<z<30$ on the $\delta z=0.25$ spaced discrete points in the interval with linear interpolation in between,
+ in order to let the PC amplitudes
+fluctuate the ionization history both up and down without entering the unphysical region 
+$x_e <0$,  where the number is chosen to give a reasonable $\tau(0,z_{\rm max})$.  It is important to note that the PCs allow
+arbitrarily large deviations from the fiducial model, and so this choice does not bias
+results.    For $z\ge 30$ we assume $x_e$ follows the ionization history from
+recombination. For $z\le 6$ we  assume fully ionized hydrogen and singly ionized helium
+ \begin{equation}
+ x_e = 1+ f_{\rm He},
+ \end{equation}
+ for $z \gtrsim z_{\rm He}$ and doubly ionized helium
+  \begin{equation}
+x_e =  1+ 2f_{\rm He},
+ \end{equation}
+for $z \lesssim z_{\rm He}$,
+where
+\beq
+f_{\rm He} = \frac{n_{\rm He}}{n_{\rm H}} = \frac{m_{\rm H}}{m_{\rm He}}\frac{Y_p}{1-Y_p}
+\eeq
+is the ratio of the helium to hydrogen number density.
+We  take the helium mass fraction $Y_p$ to be consistent with big bang nucleosynthesis given the baryon density. 
+  Following CAMB, we take this helium reionization transition to be mediated by
+ a tanh function in redshift centered at  $z_{\rm He} = 3.5$ \cite{Becker:2010cu} with width $\Delta z=0.5$.
+
+In Fig.~\ref{fig:reion_models}, we show the fiducial ionization history (thick blue) and contrast it with
+the standard approach of CAMB (thin black) that takes hydrogen and singly ionized helium reionization to be given by
+the tanh form
+ \begin{equation}
+x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+ \label{eqn:tanh}
+ \end{equation}
+ with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$.
+We take here $z_*= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
+Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
+ Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an 
+ excellent representation of the polarization power spectrum.
+
+ We also show in Fig.~\ref{fig:reion_models} the cumulative optical depth from $z$ to $z_{\rm max}$.    Although the
+two models have comparable total optical depth $\tau(0,z_{\rm max})$ the fiducial model 
+receives much of its contribution from high redshift.   Note also that although we do not 
+allow for uncertainties in helium reionization, its entire impact is a small correction on an already small contribution to $\tau$.
+Furthermore, although the reconstruction of $\tau(z,z_{\rm max})$ using 5 PCs
+is still imperfect (black dashed line), it is much better than $x_e(z)$ as it is more closely related to the polarization
+observables.    The PC reconstruction smooths out sharp transitions in the cumulative
+optical depth but gives an accurate representation of the high and low redshift contributions
+of the model.   In the analysis below, in order to compare exactly the same statistic $\tau(z,z_{\rm max})$ between models, we always employ the PC reconstructed version.
+
+In Fig.~\ref{fig:reion_basis} we show the 5 PC ionization functions $S_a(z)$ which allow
+observationally complete variations around the fiducial model.   We also show  the cumulative
+optical depth $\tau(z,z_{\rm max})$ for a unit amplitude $m_a$ in each mode.
+The lowest-variance eigenmode $S_1$ adjusts the total optical depth, mainly from the
+high redshift end.   The $S_2$ mode allows a redistribution of the optical depth between
+high redshift and low redshift.   The higher modes allow finer adjustments in the redshift
+distribution of the optical depth and carry very little total optical depth $\tau(0,z_{\rm max})$.
+
+For the Planck data set, most of the information in the ionization history is carried by the
+first two modes and therefore relates to the amount of high vs.~low redshift optical depth.   
+We keep all 5 PCs  for completeness in representing the observable
+impact of a given ionization history and to marginalize uncertainties that they introduce.
+
+In fact, PCs with no prior constraints on the mode amplitudes $m_a$
+allow deviations that are unphysical $x_e<0$ and $x_e >x_e^{\rm max}$.  With a truncation at 
+5 PCs, physicality cannot be strictly enforced since the missing eigenmodes, while irrelevant
+for the observables, can restore physicality of a model.   We follow Ref.~\cite{Mortonson:2008rx} in placing necessary but not sufficient conditions for physicality  
+\begin{equation}
+\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
+\end{equation}
+where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
+\begin{equation}
+m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
+\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
+\label{eq:individualprior}
+\end{equation}
+In principle $x_e^{\rm max}=1+ 2 f_{\rm He}$ but in practice for compatibility with the ionization state  at $z_{\rm min}$ we
+take $x_e^{\rm max} =1+f_{\rm He}$.   Since this prior is applied very conservatively as a necessary condition, we in fact still allow
+ionization fractions out beyond the true maximum with this prior.
+
+For analyzing any physical model  of hydrogen reionization, $x_e^{\rm true}(z)$, using our 5 PC decomposition from
+Eq.~(\ref{eq:xetommu}), the priors are noninformative
+as they are automatically satisfied.  We apply them mainly for visualizing whether the data 
+ constrain the mode amplitudes significantly better than the physicality bounds.
+
 \subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
+
+[to be shortened and reworded]
+
+In practice, the MCMC only provides a sample of the PC posterior, represented by
+discrete elements in the chain with parameter values $\mathbf{m}_i = \{m_1, \ldots, m_5\}$
+and multiplicities $w_i$,
+whereas a given model produces a continuum of values for $\mathbf{m}$.   
+Given the samples, we approximate the effective likelihood with a kernel
+density estimator of the form 
+\beq
+{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N} w_i K_f(\mathbf{m}-\mathbf{m}_i),
+\eeq
+where $N$ is the total number of elements in the chain and the overall normalization is arbitrary. Here $K_f$ is a smoothing kernel that
+makes the function estimate continuous at the expense of artificially broadening the distribution.
+We choose the shape of $K_f$ to be a multivariate Gaussian of zero mean and 
+covariance $f\mathbf{C}$ where $\mathbf{C}$ is the $m_a$ 
+covariance matrix estimated from the
+chain from Table~\ref{tab:PC_stats} and Eq.~(\ref{eq:cov}).    For a Gaussian posterior, the effect of smoothing
+is to increase the covariance by $1+f$ or the errors by approximately $1+f/2$.   
+To minimize the amount of smoothing required to capture the behavior of models
+in the tail of the distribution like the standard tanh model, we oversample the posterior by running  the 
+chain past normal convergence requirements for a total of $N\approx 1.4 \times 10^6$ chain elements.
+In practice we choose $f=0.14$.  Note also that we employ the full chain without physicality priors since the smoothing kernel transfers information across these boundaries.
+
+To illustrate and test this approach, we  use the effective likelihood to compare constraints on
+the 
+standard tanh model. In this case the model parameter is the total $\tau$ 
+given the tanh ionization history $x_e(z; \tau)$.   We can then construct the posterior probability of
+$\tau$ as usual via Bayes' theorem and the effective likelihood given $\mathbf{m}(\tau)$
+\begin{equation}
+P(\tau| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\tau) \right] P(\tau).
+\end{equation}
+To match the MCMC analysis of the standard tanh model we take flat priors $P(\tau)=1$.   
+For the conversion between the ionization history and $\tau$ we take the cosmological parameters
+of the ML tanh model.
+
+In Fig.~\ref{fig:tanh_tau_posterior}, we compare the posterior probabilities from the direct MCMC
+analysis and the effective PC likelihood.   The distributions match to much better than $1\sigma$
+in their means and widths.    This is a fairly stringent test on the method given that 
+tanh models live in the tails of the PC posterior.  
+ The cutoff at low $\tau$ in the effective likelihood method
+ simply reflects the $z_{\rm min}$ restriction for the PCs which assumes hydrogen ionization occurred at $z>6$ as is observationally the case.
+
+It is also interesting to compare these results to an even simpler effective likelihood.  In
+Fig.~\ref{fig:tanh_tau_posterior} (shaded region), we also show the result of approximating
+ the $m_a$ posterior as a multivariate Gaussian with mean $\bar m_a$
+ and covariance $\mathbf{C}$
+ \begin{equation}
+ {\cal L}_{\rm G}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^5 | \mathbf{C}| }}
+.
+ \label{eq:gaussian}
+ \end{equation}
+ For an extremely fast but approximate effective likelihood and for models near the peak
+ of the distribution, the Gaussian approximation may suffice.
+
+Validated on the tanh model, our effective likelihood technique allows a rapid exploration of other models without the need for
+ a separate MCMC analysis.
 
@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite.pdf}
-\caption{PC chains for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
+\caption{PC chains plane for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
 }
-\label{fig:plot_mjs}
+\label{fig:plot_mjs_2018_vs_2015}
 \end{figure}
@@
 
-[Plot: Insert m1-m2 plot about priors]
-
 \begin{figure}
 \includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
-\caption{m1-m2 plane for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+\caption{m1-m2 plane for zmax = 30 using Planck 2018 original lowE vs srollv2 likelihood. [add tanh trajectory]
 }
@@
 
-[Plot: Insert m1-m2 plot to show Gaussian covariance does not work]
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
+\caption{[placeholder, maybe not] need m1-m2 plane for zmax = 30 using Planck 2018 vs 2015.
+}
+\label{fig:plot_m1m2}
+\end{figure}
+
+[Plot: Add Gaussian covariance to m1-m2 plot, add legend, shrink x-axis]
+
+[Plot: Insert m1-m2 plot with priors boxes (?)]
+
+ \begin{figure}
+          \includegraphics[width=0.9\columnwidth]{paper/plots/plot_rotated_box_flat_tau_prior_fac_0p8.pdf}
+          \caption
+          {[Placeholder right now (Planck 2015 plot)] Priors on the $m_1-m_2$ parameter space: the original HMH16 prior (inset square, solid lines) and a prior that is flat in
+          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) whose sides are aligned with lines of
+          constant $\tau_{12}$ (light gray lines).   The HMH16 prior allows more parameter space at high vs low $\tau_{12}$ but this
+          is not relevant since the data constraints (ellipses) exclude this region.
+In the allowed region, the only difference is that the HMH16 prior clips the low $m_1$ edge of the allowed region due to physicality and the assumption that reionization occurs at
+          $z\ge 6$. [add in text the comparison of priors, and how tau is not biased, and for KDE physical models determine the priors]}  \label{fig:prior_box}
+\end{figure}
 
@@
 
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+[Plot: add plot for tau(>z) contours for the 2 parameter model vs PC; direct integration first]
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015).
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
 \caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
@@
 
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
 \caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
@@
 %Optional
-\begin{figure}
+\begin{figure}[ht]
 \includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
@@
 
-\begin{figure}[t]
+\begin{figure}
 \includegraphics[width=0.40\textwidth]{plots/plot_tau_gtz.pdf}
@@
 
-\subsection{A note on priors}
+\subsection{Comparison with Previous Results}
 
@@
 
+[to be reworded]
+In Fig.~\ref{fig:reion_models}, we show the fiducial ionization history (thick blue) and contrast it with
+the standard approach of CAMB (thin black) that takes hydrogen and singly ionized helium reionization to be given by
+the tanh form
+ \begin{equation}
+x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+ \label{eqn:tanh}
+ \end{equation}
+ with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$.
+We take here $z_*= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
+Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
+ Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an 
+ excellent representation of the polarization power spectrum.
+
 \subsection{Example 2: high-z model}
 \label{sec:example2}
+[rewording needed]
+Our example is a two-step model 
+ \bea
+x_e^{\rm true}&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
+&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
+ \label{eqn:tanh_highz}
+ \eea
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$, with $\Delta z_1 = 0.25$ instead of the usual $\Delta z_1 = 0.5$,
+to provide sharper distinctions between the two steps.
+We choose the second step to have $z_{\rm t}=28$  and  $\Delta z_2 = 1.0$  to illustrate below
+how the 5 PC analysis with $\zmax=30$ represents the same model.  Here $x_e^{\rm rec}$ is the ionization history from recombination only.
+The canonical tanh model is essentially recovered in the limit $\xemin$ approaches the negligible $x_e^{\rm rec}$.  Therefore the double
+step model adds a single parameter to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$.   We show an example
+of the two-step model in Fig.~\ref{fig:two_step_model}.
 
@@
 }
-\label{fig:tau_lo_prior_test_xe}
+\label{fig:two_step_model}
 \end{figure}
@@
     \item{Intro}
+        \begin{enumerate}
+            \item 1st paragraph: CMB, reionization, why important
+            \item discuss modeling of ionization history in CMB inference, tanh vs PCs (fail to capture high-z; PC captures complete parameter space), mention flex knots/other methods; cite previous PC works.
+            \item give context about Planck final results; latest srollv2 likelihood, important to harness all information available, PC allows us to do that and turn into an effective likelihood.
+            \item this is what we did: we obtain PC results for Planck 2018, and turn PC chains into a effective likelihood for ionization history: highlight key points like fast inference, entire model space up to zmax 30, code publicly available.
+            \item we also compared to Planck official results, which is over stringent at high z, which gives additional motivations for using this likelihood; verified no hint of ionization at z>30; state any additional results.
+            \item break down of sections.
+        \end{enumerate}
+
 	\item{Background}

### 2020-09-11 results/notes.tex

@@
 
+\begin{widetext}
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite.pdf}
+\caption{PC chains for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+}
+\label{fig:plot_mjs}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
+\caption{PC chains plane for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+}
+\label{fig:plot_mjs_2018_vs_2015}
+\end{figure}
+
+\end{widetext}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
+\caption{m1-m2 plane for zmax = 30 using Planck 2018 original lowE vs srollv2 likelihood.
+}
+\label{fig:plot_m1m2}
+\end{figure}
+
+Tests done for tanh high-z model:
+
+Problem:
+1) For the smaller tanh width at low z, now set to be dz = 0.015(1+z), the model with parameters taulo = 4.0, tauhi = 0.0 gives Cl = NaN while applying smoothing or using tauhi = 1e-4 gives sensible Cls.
+For the original fixed dz = 0.25, this was not problem.
+
+Observations and solution:
+1) It is automatically set that xemin = 0.0 if tauhi < some threshold. 
+2) xemin = 0.0 gives Cl = NaN because of precision issues.
+3) In fact, xemin < approximately 1e-18 would also give Cl = NaN.
+4) Instead of setting xemin to 0.0 for tauhi below some threshold, we let iterative search converge by itself.
+
+Additional fixes:
+1) We set xemin to sign(xemin) * 1e-7 if |xemin| < 1e-7. 
+2) We start the xemin search between -0.1 and 100 instead of 0.0 and 100, to allow for negative values of xemin to be searched. We checked that setting xemin = -1e-18 does not cause a problem.
+3) Smoothing capability is added to smooth xe(z), but this is not used for it is too slow.
+
+The above changes are implemented in the branch chen/debug\_cl\_nan.
+
+-----
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test3_run1_vs_relike_tanh_highz_test1_run1_f0p10_taulo_prior_0p0415_0p10_0p14_0p20_tri_from_getdist_chains_smooth_scale_1D_-1_num_bins_1D_100.png}
+\caption{
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\caption{
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/plot_tanh_highz_direct_mcmc_with_taulo_priors.png}
+\caption{
+}
+\label{fig:}
+\end{figure}
+
 \bibliography{rei.bib}

### 2020-10-09 paper/rei.tex
Updates from Overleaf

--- 2020-08-22 paper/rei.tex
@@
 
+The cosmic microwave background (CMB) has entered an era of precision cosmology.
+The measurements from the \textit{Planck} satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution.
+While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces.
+Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude.
+Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB; and will also have implications for inferring cosmic acceleration through the growth of structure.
+[add more refs here]
+
+Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history, with the step location parameterized by the total Thomson optical depth induced.
+This steplike model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition.
+However, through the shape of the reionization bump induced in the CMB E-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained.
+
+In fact, to extract the most information possible from this E-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum.
+This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}.
+It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}[also cite our inflation papers?].
+
+In a re-analysis of the Planck 2015 data with PCs~\cite{Heinrich:2016ojb}, a component of the high-redshift ionization that would have been missed by a simple steplike model~\cite{Heinrich:2016ojb} was uncovered.
+In the latest official Planck 2018 release~\cite{Aghanim:2018eyx}, the PC method was also adopted to probe ionization at high-redshift, whose significance was reduced since the Planck 2015 release, largely due to the reduction of systematics at large-scales~\cite{Aghanim:2018eyx, Heinrich:2018btc}.
+In addition to PCs, the FlexKnot method was employed, which is also able to capture general ionization histories by varying the number of ``knots" in redshifts and the amplitude of the ionization fraction at these knots.
+% Planck intermediate results on reionization history: \cite{Adam:2016hgk}
+
+Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019.
+This new likelihood, called $\textsc{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from XX to XX.
+[fill in numbers]
+
+Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data.
+In light of the $\textsc{srollv2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
+Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.
+
+The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].
+
+Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release.
+But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots [fill in number and make sure to say its flex knot or PC] in Ref.~\cite{Aghanim:2018eyx}.
+Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
+~\cite{Millea:2018bko} and ends up penalizing lower optical depth models.
+[quote difference in tau(15, 30) or total tau possibly attributed to this treatment].
+%We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data
+
+The paper is structured as follows.
+We first describe in section~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
+Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \textsc{srollv2} likelihood, and compare with previous results in literature.
+Then in section~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
+Finally, we summarize our results and conclude in section~\ref{sec:conclusion}.
+
 \section{Background}
+\label{sec:background}
+
+The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}.
+We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}.
+We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.
 
 \subsection{Reionization Principal Components}
+\label{sec:PC}
+
+We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
+%
+\begin{equation}
+x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
+\label{eq:mmutoxe}
+\end{equation}
+%
+where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model.
+We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements
+%
+\beq
+F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
+\eeq
+%
+where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
+We have rank-ordered the PCs from low to high variance, so that for the range $z_{min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.
+\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
+In the work that follows, we therefore truncate the sum at $a = 1 ..
+n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.
+
+Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
+given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
+\begin{equation}
+m_{a}=
+  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
+\label{eq:xetommu}
+\end{equation}
+%
+where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision.
+We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
+
+[rephrase this paragraph:]
+``For the Planck data set, most of the information in the ionization history is carried by the
+first two modes and therefore relates to the amount of high vs.~low redshift optical depth. We keep all 5 PCs  for completeness in representing the observable
+impact of a given ionization history and to marginalize uncertainties that they introduce."
+
+We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$.
+Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
+For $z<6$, we follow camb to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width  $\Delta z = 0.5$.
+
+We also put constraints on the cumulative Thomson optical depth derived parameters 
+\begin{equation}
+\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
+\label{eq:cumtau}
+\end{equation}
+where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.
+Because the optical depth is more directly related to the CMB observables than the ionization history itself, it allows us to make meaningful comparisons between the PC results and those with a specific prior in model space such as those admitting a particular functional form.
+
+For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
+ \begin{equation}
+x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+ \label{eqn:tanh}
+ \end{equation}
+ where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
+and has an implicit prior in model space that excludes finite ionization above its transition redshift.
+
+The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.
+When showing results, however, we do impose a physicality prior derived from the condition $0<x_e<x_e^{\rm max}$ following Ref.~\cite{Mortonson:2008rx}
+%
+\begin{equation}
+\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
+\end{equation}
+where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
+\begin{equation}
+m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
+\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
+\label{eq:individualprior}
+\end{equation}
+%
+Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
+They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
+
+%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
+% We include this small correction by integrating
+% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
+   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
+%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.
+
 \subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
+\label{sec:KDE}
+
+The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood.
+In the following, we briefly recap the kernel density estimate technique used here to build the likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+
+The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ and posterior values for $i = 1$...$N_{\rm sample}$.
+Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}.
+Since $\mathbf{m}$ could take any continuous values, we approximate its effective likelihood with a kernel density estimate
+\beq
+{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
+\eeq
+where the overall normalization is arbitrary, and we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero.
+We take the covariance of the Gaussian to be $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller 1.
+
+[I'm here].
+For a Gaussian posterior, the effect of smoothing
+is to increase the covariance by $1+f$ or the errors by approximately $1+f/2$.   
+To minimize the amount of smoothing required to capture the behavior of models
+in the tail of the distribution like the standard tanh model, we oversample the posterior by running  the 
+chain past normal convergence requirements for a total of $N\approx 1.4 \times 10^6$ chain elements.
+In practice we choose $f=0.14$.  Note also that we employ the full chain without physicality priors since the smoothing kernel transfers information across these boundaries.
+
+To illustrate and test this approach, we  use the effective likelihood to compare constraints on
+the 
+standard tanh model. In this case the model parameter is the total $\tau$ 
+given the tanh ionization history $x_e(z; \tau)$.   We can then construct the posterior probability of
+$\tau$ as usual via Bayes' theorem and the effective likelihood given $\mathbf{m}(\tau)$
+\begin{equation}
+P(\tau| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\tau) \right] P(\tau).
+\end{equation}
+To match the MCMC analysis of the standard tanh model we take flat priors $P(\tau)=1$.   
+For the conversion between the ionization history and $\tau$ we take the cosmological parameters
+of the ML tanh model.
+
+In Fig.~\ref{fig:tanh_tau_posterior}, we compare the posterior probabilities from the direct MCMC
+analysis and the effective PC likelihood.   The distributions match to much better than $1\sigma$
+in their means and widths.    This is a fairly stringent test on the method given that 
+tanh models live in the tails of the PC posterior.  
+ The cutoff at low $\tau$ in the effective likelihood method
+ simply reflects the $z_{\rm min}$ restriction for the PCs which assumes hydrogen ionization occurred at $z>6$ as is observationally the case.
+
+It is also interesting to compare these results to an even simpler effective likelihood.  In
+Fig.~\ref{fig:tanh_tau_posterior} (shaded region), we also show the result of approximating
+ the $m_a$ posterior as a multivariate Gaussian with mean $\bar m_a$
+ and covariance $\mathbf{C}$
+ \begin{equation}
+ {\cal L}_{\rm G}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^5 | \mathbf{C}| }}
+.
+ \label{eq:gaussian}
+ \end{equation}
+ For an extremely fast but approximate effective likelihood and for models near the peak
+ of the distribution, the Gaussian approximation may suffice.
+
+Validated on the tanh model, our effective likelihood technique allows a rapid exploration of other models without the need for
+ a separate MCMC analysis.
 
@@
 
+We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principle component technique.
+
 \subsection{Constraints on Principle Components}
+
+We use the official Planck likelihoods \textsc{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \textsc{lowl} for the low-$\ell$ $TT$ throughout the paper.
+(We have tested that our results do not change when we switch to the full version \textsc{plik full} where the foregrounds are marginalized over in the likelihood.) For the low-$\ell$ $EE$, we use the third-party released \textsc{srollv2} likelihood for our official result.
+Compared to the official version released by Planck \textsc{lowE}, the new \textsc{srollv2} likelihood was made with an improved foreground cleaning method [cite], and leads to a factor of two improvement in $\tau$ in the steplike model.
+We will compare the PC results from these likelihoods later in this section.
+
+[I'm here.]
+
+In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the constraints in the $m_a-m_b$ plane of Planck 2015 vs Planck 2018. [Describe changes] [Mention Gaussian covariance]
+
+\begin{itemize}
+    \item compare with Planck 2015 our PC results; Fig.~\ref{fig:plot_mjs_2018_vs_2015} shows the PC constraints of Planck 2015 vs Planck 2018. [Describe changes] [Mention Gaussian covariance]
+    \item insert table of mean and covariance matrix for the PC components?
+    \item make a point that tanh misses high-z again, compare to PCs using the tau($>z$) plot tanh vs PCs.
+    \item statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.\\
+\end{itemize}
+
+\subsection{Comparison with Previous Results}
+
+\begin{itemize}
+    \item compare with Planck official results
+    \item comment that Planck tau(15,30) constraints are too stringent (flex knots). Compare to ours. Point out why.
+    \item{ Comments on priors (see below).}
+    \item (*maybe in  another section): make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck.
+This was for higher tau, so transition happens at higher redshift.
+Now tau is much lower.
+\end{itemize}
+
+Recap our appendix from Planck 2015 on the issue of prior:
+\begin{itemize}
+    \item {describe tau prior issue raised by MB18: claimed that prior not flat in tau so bias PC; proposed to flatten by multiplying the inverse of prior point-by-point in the PC space; claim that this flattening removes a shift between tanh and PC results.
+}
+    \item {summarize that we pointed it out the flaw in the logic in appendix of \cite{Heinrich:2018btc}.}
+    \item{reproduce a few tests here: 1) tau12 posterior (no need to show again, point to appendix); 2) show prior boxes in m1-m2 plane (done in Fig.4) - repeat the length of lines of constant tau12 within the box describe the prior distribution.
+Make the point that though the line rises, the Planck data is more informative than priors in m1, m2 makes no impact on the posterior.
+However when inverting the prior, introduces a bias in tau.}
+    \item{show test of explicitly constructing a prior that's locally flat in m1-m2, show posterior almost doesn't change.
+(only change is low-z enhancement, due to prior cut off at $z>6$.
+between our two priors, things don't change as much (might want to reproduce this result for this Planck 2018 results.
+And marius version too?}
+    \item{inversion method should not be used on future dataset, since data is more informative than prior.}
+\end{itemize}
 
@@
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite.pdf}
-\caption{PC chains for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
-}
-\label{fig:plot_mjs}
+\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
+\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
+We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \textsc{srollv2}.
+The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
+regions in the $m_a-m_b$ planes in the lower triangle.
+The physicality priors on the ionization history used in this work are delineated by the box boundaries.
+[TODO: update legend]}
+\label{fig:plot_mjs_2018_vs_2015}
 \end{figure}
@@
 
-[Plot: Insert m1-m2 plot about priors]
-
 \begin{figure}
 \includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
-\caption{m1-m2 plane for zmax = 30 using Planck 2018 srollv2 likelihood vs Planck 2015.
+\caption{The best-constrained $z_{\rm max} = 30$ PC plane, $m_1-m_2$.
+We compare Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: the official \textsc{lowE} likelihood vs.
+the third-party released \textsc{srollv2} likelihood (zoomed-in from Fig.~\ref{fig:plot_mjs_2018_vs_2015}).
+[Comments on shift m1-m2 plane].
+[TODO: Add Gaussian covariance, add legend, shrink x-axis]
 }
@@
 
-[Plot: Insert m1-m2 plot to show Gaussian covariance does not work]
-
-[Text: recap our appendix from Planck 2015]
-
-[Text: compare with Planck 2015 our PC results]
-
-[Text: statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.]
-
-[Text: make a point that tanh misses high-z again, compare to PCs using the tau($>z$) plot tanh vs PCs.]
-
-[Table: insert table of mean and covariance matrix for the PC components?]
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
+\caption{[placeholder, maybe not] need m1-m2 plane for zmax = 30 using Planck 2018 vs 2015.
+}
+\label{fig:plot_m1m2}
+\end{figure}
 
  \begin{figure}
-\includegraphics[width=0.48\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+          \includegraphics[width=0.9\columnwidth]{paper/plots/plot_rotated_box_flat_tau_prior_fac_0p8.pdf}
+          \caption
+          {[Placeholder right now (Planck 2015 plot)] Priors on the $m_1-m_2$ parameter space: the original HMH16 prior (inset square, solid lines) and a prior that is flat in
+          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) whose sides are aligned with lines of
+          constant $\tau_{12}$ (light gray lines).   The HMH16 prior allows more parameter space at high vs low $\tau_{12}$ but this
+          is not relevant since the data constraints (ellipses) exclude this region.
+In the allowed region, the only difference is that the HMH16 prior clips the low $m_1$ edge of the allowed region due to physicality and the assumption that reionization occurs at
+          $z\ge 6$. [add in text the comparison of priors, and how tau is not biased, and for KDE physical models determine the priors]}  \label{fig:prior_box}
+\end{figure}
+
+[Plot: add plot for tau(>z) contours for the 2 parameter model vs PC; direct integration first]
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015).
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
 \caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
@@
 
-\subsection{A note on priors}
-
-[Text: compare with Planck official results]
-
-[Text: comment that Planck tau(15,30) constraints are too stringent (flex knots). Compare to ours. Point out why.]
-
-[Text: compare with Planck official results, PCs (add comments on priors), FlexKnots]
-
-[*Text (maybe in another section): make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck.
-This was for higher tau, so transition happens at higher redshift.
-Now tau is much lower.]
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+In Fig.~\ref{fig:reion_models}, we show the fiducial ionization history (thick blue) and contrast it with
+the standard approach of CAMB (thin black) that takes hydrogen and singly ionized helium reionization to be given by
+the tanh form
+ \begin{equation}
+x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+ \label{eqn:tanh}
+ \end{equation}
+ with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$.
+We take here $z_*= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
+Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
+ Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an 
+ excellent representation of the polarization power spectrum.
+
 \subsection{Example 2: high-z model}
 \label{sec:example2}
+[rewording needed]
+Our example is a two-step model 
+ \bea
+x_e^{\rm true}&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
+&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
+ \label{eqn:tanh_highz}
+ \eea
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$, with $\Delta z_1 = 0.25$ instead of the usual $\Delta z_1 = 0.5$,
+to provide sharper distinctions between the two steps.
+We choose the second step to have $z_{\rm t}=28$  and  $\Delta z_2 = 1.0$  to illustrate below
+how the 5 PC analysis with $\zmax=30$ represents the same model.  Here $x_e^{\rm rec}$ is the ionization history from recombination only.
+The canonical tanh model is essentially recovered in the limit $\xemin$ approaches the negligible $x_e^{\rm rec}$.  Therefore the double
+step model adds a single parameter to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$.   We show an example
+of the two-step model in Fig.~\ref{fig:two_step_model}.
 

### 2020-10-09 results/notes.tex
Updates from Overleaf

@@
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test3_run1_vs_relike_tanh_highz_test1_run1_f0p10_taulo_prior_0p0415_0p10_0p14_0p20_tri_from_getdist_chains_smooth_scale_1D_-1_num_bins_1D_100.png}
-\caption{
-}
-\label{fig:}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-\caption{
+\includegraphics[width=0.4\textwidth]{results/cosmomc_kde/taulo_prior_test/relike_tanh_highz_f0p14_taulo_prior_0p04_with_derived_parameters_zre_greater_than_6p25_tri.png}
+\caption{Illustration of the bug in the tanh high-z model, introduced before running chains w/ taulo prior 0.04 instead of 0.0415.
 }
@@
 \includegraphics[width=0.4\textwidth]{cosmomc_kde/plot_tanh_highz_direct_mcmc_with_taulo_priors.png}
-\caption{
+\caption{Previous plot of direct MCMC w/ various \taulo priors showing that the MCMC runs look fine and inferring bug is on the KDE side of the code.
+} 
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\caption{After bug fix: direct MCMC ($z_{\rm re} > 6.1$) vs KDE ($z_{\rm re} > 6.1$) done by hand) everything is in good agreement!
+This is applying the bug fix on the latest code, i.e.
+w/ new definition of tanh high-z model that adopts a smaller $dz = 0.015 (1+z)$ at the low-z transition and boundary point for \taulo and \tauhi defined at $z_{\rm re} + 1 \times dz$ instead of $z_{\rm re} + 2 \times dz$ and various other minor changes (e.g.
+$|x_{e,\rm min}|<10^{-7}$ is sent to $x_{e,\rm min} = \mathrm{sgn}(x_{e,\rm min}) \times 10^{-7}$, $\tau$ convergence change, etc.) —
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{results/cosmomc_kde/getdist_smoothing_test/pl18_tanh_highz_test3_run1_vs_relike_tanh_highz_test1_run1_f0p10_taulo_prior_0p0415_0p10_0p14_0p20_tri_from_getdist_chains_smooth_scale_1D_-1_num_bins_1D_100.png}
+\caption{Previous plot of direct MCMC vs KDE for the old tanh high-z model (less sharp transition, etc) for various f = 0.10, 0.14, 0.20.
+(CH: Same test to be done for the after bug fixed version shown in the previous plot.)
 }

### 2020-10-12 paper/rei.tex
Updates from Overleaf

@@
 The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood.
-In the following, we briefly recap the kernel density estimate technique used here to build the likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
 
-The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ and posterior values for $i = 1$...$N_{\rm sample}$.
+The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
 Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}.
-Since $\mathbf{m}$ could take any continuous values, we approximate its effective likelihood with a kernel density estimate
+Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
 \beq
@@
 \eeq
-where the overall normalization is arbitrary, and we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero.
-We take the covariance of the Gaussian to be $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller 1.
+where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
 
-[I'm here].
-For a Gaussian posterior, the effect of smoothing
-is to increase the covariance by $1+f$ or the errors by approximately $1+f/2$.   
-To minimize the amount of smoothing required to capture the behavior of models
-in the tail of the distribution like the standard tanh model, we oversample the posterior by running  the 
-chain past normal convergence requirements for a total of $N\approx 1.4 \times 10^6$ chain elements.
-In practice we choose $f=0.14$.  Note also that we employ the full chain without physicality priors since the smoothing kernel transfers information across these boundaries.
+For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$.
+To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx XX \times 10^{X}$ [to fill] chain samples.
+For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
+Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.
 
-To illustrate and test this approach, we  use the effective likelihood to compare constraints on
-the 
-standard tanh model. In this case the model parameter is the total $\tau$ 
-given the tanh ionization history $x_e(z; \tau)$.   We can then construct the posterior probability of
-$\tau$ as usual via Bayes' theorem and the effective likelihood given $\mathbf{m}(\tau)$
+Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
 \begin{equation}
-P(\tau| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\tau) \right] P(\tau).
+P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
 \end{equation}
-To match the MCMC analysis of the standard tanh model we take flat priors $P(\tau)=1$.   
-For the conversion between the ionization history and $\tau$ we take the cosmological parameters
-of the ML tanh model.
+we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time.
+We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the tanh model and a two-parameter model allowing for high-redshift ionization.
 
-In Fig.~\ref{fig:tanh_tau_posterior}, we compare the posterior probabilities from the direct MCMC
-analysis and the effective PC likelihood.   The distributions match to much better than $1\sigma$
-in their means and widths.    This is a fairly stringent test on the method given that 
-tanh models live in the tails of the PC posterior.  
- The cutoff at low $\tau$ in the effective likelihood method
- simply reflects the $z_{\rm min}$ restriction for the PCs which assumes hydrogen ionization occurred at $z>6$ as is observationally the case.
+%Therefore, the PC-based effective likelihood allows one to easily explore different models of ionization history while providing a much faster convergence in the MCMC analysis.
 
-It is also interesting to compare these results to an even simpler effective likelihood.  In
-Fig.~\ref{fig:tanh_tau_posterior} (shaded region), we also show the result of approximating
- the $m_a$ posterior as a multivariate Gaussian with mean $\bar m_a$
- and covariance $\mathbf{C}$
- \begin{equation}
- {\cal L}_{\rm G}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^5 | \mathbf{C}| }}
-.
- \label{eq:gaussian}
- \end{equation}
- For an extremely fast but approximate effective likelihood and for models near the peak
- of the distribution, the Gaussian approximation may suffice.
-
-Validated on the tanh model, our effective likelihood technique allows a rapid exploration of other models without the need for
- a separate MCMC analysis.
+%In what follows, we also show a even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$:
+%\begin{equation}
+% {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
+% \label{eq:gaussian}
+% \end{equation}
+% This likelihood may suffice for models near the peak of the distribution, but would not be accurate for sampling less-constrained models or in the tails of distributions.
 

### 2020-10-12 paper/rei.tex
Updates from Overleaf

@@
 Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019.
-This new likelihood, called $\textsc{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from XX to XX.
+This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from XX to XX.
 [fill in numbers]
@@
 Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data.
-In light of the $\textsc{srollv2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
+In light of the $\texttt{srollv2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
 Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.
@@
 We first describe in section~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
-Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \textsc{srollv2} likelihood, and compare with previous results in literature.
+Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature.
 Then in section~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
@@
 we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time.
-We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the tanh model and a two-parameter model allowing for high-redshift ionization.
+We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
 
@@
 
-We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principle component technique.
+We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principle components.
 
@@
 
-We use the official Planck likelihoods \textsc{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \textsc{lowl} for the low-$\ell$ $TT$ throughout the paper.
-(We have tested that our results do not change when we switch to the full version \textsc{plik full} where the foregrounds are marginalized over in the likelihood.) For the low-$\ell$ $EE$, we use the third-party released \textsc{srollv2} likelihood for our official result.
-Compared to the official version released by Planck \textsc{lowE}, the new \textsc{srollv2} likelihood was made with an improved foreground cleaning method [cite], and leads to a factor of two improvement in $\tau$ in the steplike model.
-We will compare the PC results from these likelihoods later in this section.
+We use the official Planck likelihoods [add citation] \texttt{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper.
+We have tested that our results do not change if in lieu of \texttt{plik\_lite} we used the full \texttt{plik\_full} likelihood in which the foreground parameters have not been marginalized over.
+For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{srollv2} likelihood [add citation] in our official PC results and the effective likelihood code.
+In comparison with the official Planck-released \texttt{lowE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].
 
-[I'm here.]
+In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
+The box boundaries correspond to the physicality priors shown in Eq.~\ref{eq:individualprior} and include all physical models along with some unphysical models.
+Results using the Planck 2018 data are shown in blue.
+While all five PC amplitudes are being constrained by Planck, the first two are particularly well-constrained.
+Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, which contributes positively to the total optical depth, leading to a reduced optical depth.
 
-In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the constraints in the $m_a-m_b$ plane of Planck 2015 vs Planck 2018. [Describe changes] [Mention Gaussian covariance]
+[I'm here] \\
 
-\begin{itemize}
-    \item compare with Planck 2015 our PC results; Fig.~\ref{fig:plot_mjs_2018_vs_2015} shows the PC constraints of Planck 2015 vs Planck 2018. [Describe changes] [Mention Gaussian covariance]
-    \item insert table of mean and covariance matrix for the PC components?
-    \item make a point that tanh misses high-z again, compare to PCs using the tau($>z$) plot tanh vs PCs.
-    \item statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.\\
+[Should save this for later?
+Or expand to explain more clearly, use a whole paragraph] This reduction is partly attributed to better foreground removal at large angular scales in the $E$-mode polarization data, which would mimic a high redshift ionization component [cite planck papers].
+
+Make these plots and add text:
+- mention Gaussian covariance \\
+- mention table of covariance and mean\\
+- compare in m1-m2 plane the two low-l likelihoods (and perhaps Planck 2015 vs Planck 2018)
+- make a point that tanh misses high-z again\\
+- compare to PCs using the tau($>z$) plot tanh vs PCs.\\
+- statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.\\
 \end{itemize}
+
+\begin{widetext}
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
+\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
+We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}.
+The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
+regions in the $m_a-m_b$ planes in the lower triangle.
+The physicality priors on the ionization history used in this work are delineated by the box boundaries.
+[TODO: update legend]}
+\label{fig:plot_mjs_2018_vs_2015}
+\end{figure}
+
+\end{widetext}
+
+\begin{table}[b]
+\centering
+\caption{[Placeholder, need to fill in real numbers] PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
+\label{tab:PC_stats}
+\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
+\hline
+
+			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
+		\\ \hline
+$m_1$ 
+	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
+$m_2$ 
+	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
+$m_3$ 
+	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
+$m_4$  
+	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
+$m_5$ 
+	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ \hline
+\end{tabular}
+\end{table}
 
@@
 
-\begin{widetext}
-
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
-\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
-We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \textsc{srollv2}.
-The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
-regions in the $m_a-m_b$ planes in the lower triangle.
-The physicality priors on the ionization history used in this work are delineated by the box boundaries.
-[TODO: update legend]}
-\label{fig:plot_mjs_2018_vs_2015}
-\end{figure}
-
-\end{widetext}
-
 \begin{figure}
@@
 \caption{The best-constrained $z_{\rm max} = 30$ PC plane, $m_1-m_2$.
-We compare Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: the official \textsc{lowE} likelihood vs.
-the third-party released \textsc{srollv2} likelihood (zoomed-in from Fig.~\ref{fig:plot_mjs_2018_vs_2015}).
+We compare Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: the official \texttt{lowE} likelihood vs.
+the third-party released \texttt{srollv2} likelihood (zoomed-in from Fig.~\ref{fig:plot_mjs_2018_vs_2015}).
 [Comments on shift m1-m2 plane].

### 2020-10-16 paper/rei.tex
Updates from Overleaf

@@
 
+[Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
+
 Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
@@
 
-%In what follows, we also show a even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$:
-%\begin{equation}
-% {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
-% \label{eq:gaussian}
-% \end{equation}
-% This likelihood may suffice for models near the peak of the distribution, but would not be accurate for sampling less-constrained models or in the tails of distributions.
-
 \section{Planck 2018 PC Results}
@@
 For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{srollv2} likelihood [add citation] in our official PC results and the effective likelihood code.
-In comparison with the official Planck-released \texttt{lowE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].
+In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].
+
+\begin{figure*}
+\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
+\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
+We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}.
+The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
+regions in the $m_a-m_b$ planes in the lower triangle.
+The physicality priors on the ionization history used in this work are delineated by the box boundaries.
+[TODO: update legend] [add tau trajectories] [add prior boundaries]}
+\label{fig:plot_mjs_2018_vs_2015}
+\end{figure*}
+%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
 
@@
 While all five PC amplitudes are being constrained by Planck, the first two are particularly well-constrained.
-Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, which contributes positively to the total optical depth, leading to a reduced optical depth.
+Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.
 
-[I'm here] \\
-
-[Should save this for later?
-Or expand to explain more clearly, use a whole paragraph] This reduction is partly attributed to better foreground removal at large angular scales in the $E$-mode polarization data, which would mimic a high redshift ionization component [cite planck papers].
-
-Make these plots and add text:
-- mention Gaussian covariance \\
-- mention table of covariance and mean\\
-- compare in m1-m2 plane the two low-l likelihoods (and perhaps Planck 2015 vs Planck 2018)
-- make a point that tanh misses high-z again\\
-- compare to PCs using the tau($>z$) plot tanh vs PCs.\\
-- statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.\\
-\end{itemize}
-
-\begin{widetext}
+The trajectories shown in black lines are the steplike models of instantaneous reionization allowed by the Planck 2015 data with the arrow pointing in the direction of increasing optical depth.
+Clearly, these steplikes models are only a part of the allowed physical model space.
+By construction, they miss any high-redshift ionization component by assuming neglibile ionization levels before the transition redshift $z_{\rm re}$.
 
 \begin{figure}
-\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
-\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
-We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}.
-The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
-regions in the $m_a-m_b$ planes in the lower triangle.
-The physicality priors on the ionization history used in this work are delineated by the box boundaries.
-[TODO: update legend]}
-\label{fig:plot_mjs_2018_vs_2015}
+\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wGaussEllipse.png}
+\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, $m_1-m_2$ plane, comparing Planck 2018 (blue) vs 2015 results (red).
+The black solid line indicates the steplike models with the arrow pointing in the direction of increasing $\tau$.
+The dashed lines indicate the contours corresponding a Gaussian likelihood (Eq.~\ref{eq:gaussian}) using the PC covariance from the Planck 2018 chains.
+[add Gaussian covariance legend?] [add tau trajectories here?]
+}
+\label{fig:plot_m1m2_2015_vs_2018}
 \end{figure}
-
-\end{widetext}
 
@@
 \end{table}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
+\caption{Same as Fig.~\ref{fig:plot_mjs_2018_vs_2015} but comparing the Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the official \texttt{simall\_EE} likelihood and the \texttt{srollv2} likelihood which has better foreground removal (shown throughout this work).
+[Comments on shift m1-m2 plane].
+}
+\label{fig:plot_m1m2_lowE_vs_srollv2}
+\end{figure}
+
+We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area.
+[Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
+
+In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}:
+\begin{equation}
+ {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
+ \label{eq:gaussian}
+ \end{equation}
+ This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement.
+The Gaussian likelihood is even faster than the kernel density estimate, and may be used to sample well-constrained models near the peak of the PC distributions, while becoming possibly less accurate near the tails of the distributions.
+
+In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}.
+The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
+[Might be nice to overlay the tanh trajectories, no 1 sigma, 2 sigma business].
+
+\subsection{Implications for optical depth}
+
+We continue the comparison in the space of cumulative optical depth $\tau(z, \zmax)$, which is a better quantity for visualizing and comparing PC results than the ionization history $x_e$ itself because of it is the quantity most closely related to observables.
+Recall that these truncated PCs provide only accurate descriptions of the models in terms of their \textit{observable} impact, and are not tools for reconstructing the precise ionization history.
+
+In Fig.~\ref{fig:plot_taugtz_2015_vs_2018}, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines).
+...
+
+In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE likelihoods.
+
+In Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}, we compare the PC results against the steplike model.
+It is clear that the tanh model does not allow for high redshift ionization by construction.
+The total optical depth allowed by the PC chains sampling the entire physical model space is actually slightly higher due to the contributions from these higher redshifts.
+[do we want PC version with the physicality priors actually?]\\
+
+We show however, using a separate PC chains with $\zmax = 50$, that there is no evidence for optical depth at higher redshifts than 30.
+In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the contours in $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines), and that their total optical depth constraints are practically identical:
+\beq
+\tau_{\rm PC, \zmax=30} = XX \pm XX,
+\eeq
+and
+\beq
+\tau_{\rm PC, \zmax=50} = XX \pm XX.
+\eeq
+We have an upper limit on $\tau(15, \zmax)$ of 
+\beq
+\tau(z, \zmax)_{\rm PC, \zmax=30} < XX \; (95\% CL)
+\eeq
+and 
+\beq
+\tau(z, \zmax)_{\rm PC, \zmax=50} < XX \; (95\% CL)
+\eeq
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015).
+}
+\label{fig:plot_taugtz_2015_vs_2018}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:plot_taugtz_lowE_vs_srollv2}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
+\caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
+}
+\label{fig:plot_taugtz_PC_vs_tanh}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
+.
+}
+\label{fig:plot_taugtz_zmax30_vs_zmax50}
+\end{figure}
 
@@
  \begin{figure}
-\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
-\caption{The best-constrained $z_{\rm max} = 30$ PC plane, $m_1-m_2$.
-We compare Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: the official \texttt{lowE} likelihood vs.
-the third-party released \texttt{srollv2} likelihood (zoomed-in from Fig.~\ref{fig:plot_mjs_2018_vs_2015}).
-[Comments on shift m1-m2 plane].
-[TODO: Add Gaussian covariance, add legend, shrink x-axis]
-}
-\label{fig:plot_m1m2}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
-\caption{[placeholder, maybe not] need m1-m2 plane for zmax = 30 using Planck 2018 vs 2015.
-}
-\label{fig:plot_m1m2}
-\end{figure}
-
- \begin{figure}
           \includegraphics[width=0.9\columnwidth]{paper/plots/plot_rotated_box_flat_tau_prior_fac_0p8.pdf}
@@
 [Plot: add plot for tau(>z) contours for the 2 parameter model vs PC; direct integration first]
-
-\begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
-\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015).
-}
-\label{fig:}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
-\caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
-}
-\label{fig:}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
-\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
-Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
-.
-}
-\label{fig:}
-\end{figure}
-
-%Optional
-\begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
-\caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
-}
-\label{fig:}
-\end{figure}
 

### 2020-10-22 paper/rei.tex
Updates from Overleaf

--- 2020-10-12 paper/rei.tex
@@
 \author{Chen Heinrich}\email{chenhe@caltech.edu}
-\affiliation{$Jet\ Propulsion\ Laboratory,\ California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
-\affiliation{$California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
+\affiliation{California Institute of Technology, Pasadena, California 91109, USA}
+\affiliation{Jet Propulsion Laboratory, California Institute of Technology, Pasadena, California 91109, USA}
 
@@
 We first describe in section~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
-Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \textsc{srollv2} likelihood, and compare with previous results in literature.
-Then in section~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
-Finally, we summarize our results and conclude in section~\ref{sec:conclusion}.
+Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and k
 
-\section{Background}
-\label{sec:background}
+In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}.
+The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
+[Might be nice to overlay the tanh trajectories, no 1 sigma, 2 sigma business].
 
-The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}.
-We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}.
-We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.
+\subsection{Implications for optical depth}
 
-\subsection{Reionization Principal Components}
-\label{sec:PC}
+We continue the comparison in the space of cumulative optical depth $\tau(z, \zmax)$, which is a better quantity for visualizing and comparing PC results than the ionization history $x_e$ itself because of it is the quantity most closely related to observables.
+Recall that these truncated PCs provide only accurate descriptions of the models in terms of their \textit{observable} impact, and are not tools for reconstructing the precise ionization history.
 
-We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
-%
-\begin{equation}
-x_e(z)=\xef(z)+\sum_{a}m_{a}S_{a}(z),
-\label{eq:mmutoxe}
-\end{equation}
-%
-where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model.
-We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements
-%
+In Fig.~\ref{fig:plot_taugtz_2015_vs_2018}, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines).
+...
+
+In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE$ likelihoods. [Do we need this?]
+
+In Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}, we compare the PC results against the steplike model.
+It is clear that the tanh model does not allow for high redshift ionization by construction.
+The total optical depth allowed by the PC chains sampling the entire physical model space is actually slightly higher due to the contributions from these higher redshifts.
+[do we want PC version with the physicality priors actually?]\\
+
+We show however, using a separate PC chains with $\zmax = 50$, that there is no evidence for optical depth at higher redshifts than 30.
+In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the contours in $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines), and that their total optical depth constraints are practically identical:
 \beq
-F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
+\tau_{\rm PC,\, \zmax=30} = XX \pm XX,
 \eeq
-%
-where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
-We have rank-ordered the PCs from low to high variance, so that for the range $z_{min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.
-\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
-In the work that follows, we therefore truncate the sum at $a = 1 ..
-n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.
+and
+\beq
+\tau_{\rm PC,\, \zmax=50} = XX \pm XX.
+\eeq
+We have an upper limit on $\tau(15, \zmax)$ of 
+\beq
+\tau(z, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})
+\eeq
+and 
+\beq
+\tau(z, \zmax)_{\rm PC,\, \zmax=50} < XX \; (95\% \mathrm{CL})
+\eeq
 
-Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
-given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
-\begin{equation}
-m_{a}=
-  \int _{\zmin}^{\zmax} dz\, \frac{S_{a}(z) [x_e^{\rm true}(z)-\xef(z)]}{\zmax-\zmin},
-\label{eq:xetommu}
-\end{equation}
-%
-where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision.
-We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015).
+}
+\label{fig:plot_taugtz_2015_vs_2018}
+\end{figure}
 
-[rephrase this paragraph:]
-``For the Planck data set, most of the information in the ionization history is carried by the
-first two modes and therefore relates to the amount of high vs.~low redshift optical depth. We keep all 5 PCs  for completeness in representing the observable
-impact of a given ionization history and to marginalize uncertainties that they introduce."
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
+\caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+}
+\label{fig:plot_taugtz_lowE_vs_srollv2}
+\end{figure}
 
-We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$.
-Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
-For $z<6$, we follow camb to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width  $\Delta z = 0.5$.
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
+\caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
+}
+\label{fig:plot_taugtz_PC_vs_tanh}
+\end{figure}
 
-We also put constraints on the cumulative Thomson optical depth derived parameters 
-\begin{equation}
-\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
-\label{eq:cumtau}
-\end{equation}
-where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.
-Because the optical depth is more directly related to the CMB observables than the ionization history itself, it allows us to make meaningful comparisons between the PC results and those with a specific prior in model space such as those admitting a particular functional form.
-
-For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
- \begin{equation}
-x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
- \label{eqn:tanh}
- \end{equation}
- where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
-and has an implicit prior in model space that excludes finite ionization above its transition redshift.
-
-The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.
-When showing results, however, we do impose a physicality prior derived from the condition $0<x_e<x_e^{\rm max}$ following Ref.~\cite{Mortonson:2008rx}
-%
-\begin{equation}
-\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
-\end{equation}
-where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
-\begin{equation}
-m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
-\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
-\label{eq:individualprior}
-\end{equation}
-%
-Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
-They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
-
-%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
-% We include this small correction by integrating
-% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
-   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
-%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.
-
-\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
-\label{sec:KDE}
-
-The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood.
-In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
-
-The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
-Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}.
-Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
-\beq
-{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
-\eeq
-where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
-
-For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$.
-To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx XX \times 10^{X}$ [to fill] chain samples.
-For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
-Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.
-
-Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
-\begin{equation}
-P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
-\end{equation}
-we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time.
-We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the tanh model and a two-parameter model allowing for high-redshift ionization.
-
-%Therefore, the PC-based effective likelihood allows one to easily explore different models of ionization history while providing a much faster convergence in the MCMC analysis.
-
-%In what follows, we also show a even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$:
-%\begin{equation}
-% {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
-% \label{eq:gaussian}
-% \end{equation}
-% This likelihood may suffice for models near the peak of the distribution, but would not be accurate for sampling less-constrained models or in the tails of distributions.
-
-\section{Planck 2018 PC Results}
-\label{sec:results}
-
-We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principle component technique.
-
-\subsection{Constraints on Principle Components}
-
-We use the official Planck likelihoods \textsc{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \textsc{lowl} for the low-$\ell$ $TT$ throughout the paper.
-(We have tested that our results do not change when we switch to the full version \textsc{plik full} where the foregrounds are marginalized over in the likelihood.) For the low-$\ell$ $EE$, we use the third-party released \textsc{srollv2} likelihood for our official result.
-Compared to the official version released by Planck \textsc{lowE}, the new \textsc{srollv2} likelihood was made with an improved foreground cleaning method [cite], and leads to a factor of two improvement in $\tau$ in the steplike model.
-We will compare the PC results from these likelihoods later in this section.
-
-[I'm here.]
-
-In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the constraints in the $m_a-m_b$ plane of Planck 2015 vs Planck 2018. [Describe changes] [Mention Gaussian covariance]
-
-\begin{itemize}
-    \item compare with Planck 2015 our PC results; Fig.~\ref{fig:plot_mjs_2018_vs_2015} shows the PC constraints of Planck 2015 vs Planck 2018. [Describe changes] [Mention Gaussian covariance]
-    \item insert table of mean and covariance matrix for the PC components?
-    \item make a point that tanh misses high-z again, compare to PCs using the tau($>z$) plot tanh vs PCs.
-    \item statement that zmax = 50 and zmax = 30 are consistent, there is no evidence for high-redshift ionization for $z>30$.\\
-\end{itemize}
+\begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
+Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
+.
+}
+\label{fig:plot_taugtz_zmax30_vs_zmax50}
+\end{figure}
 
@@
 
+Outline
 \begin{itemize}
@@
 
-\begin{widetext}
-
-\begin{figure}
-\includegraphics[width=0.7\textwidth]{results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf}
-\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
-We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \textsc{srollv2}.
-The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
-regions in the $m_a-m_b$ planes in the lower triangle.
-The physicality priors on the ionization history used in this work are delineated by the box boundaries.
-[TODO: update legend]}
-\label{fig:plot_mjs_2018_vs_2015}
-\end{figure}
-
-\end{widetext}
-
-\begin{figure}
-\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
-\caption{The best-constrained $z_{\rm max} = 30$ PC plane, $m_1-m_2$.
-We compare Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: the official \textsc{lowE} likelihood vs.
-the third-party released \textsc{srollv2} likelihood (zoomed-in from Fig.~\ref{fig:plot_mjs_2018_vs_2015}).
-[Comments on shift m1-m2 plane].
-[TODO: Add Gaussian covariance, add legend, shrink x-axis]
-}
-\label{fig:plot_m1m2}
-\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.4\textwidth]{results/pc_results/plot_m1_m2_pl18_pc_zmax30_pliklite_srollv2_vs_pl18_pc_zmax30_pliklite_wTauTrajectory.pdf}
-\caption{[placeholder, maybe not] need m1-m2 plane for zmax = 30 using Planck 2018 vs 2015.
-}
-\label{fig:plot_m1m2}
-\end{figure}
-
  \begin{figure}
@@
 
+------ \textbf{Some Numbers} ------ \\
+
+\textbf{Planck official results 2015, 2016, 2018} \\
+
+\begin{itemize}
+
+    \item Planck 2015 results (LFI): $\tau = 0.067 \pm 0.022$ (68\%) \\
+
+    \item Planck 2016  intermediate results (LFI): $\tau = 0.055 \pm 0.009$ (68\%) \\
+
+    \item Planck 2018 results: $\tau = 0.0506 \pm 0.0086$ (68\%) \\
+
+\end{itemize}
+
+\textbf{Planck 2018 paper} \\
+
+Here lowE means lowE data only, fixing all other cosmological parameters including $A_s e^{-2\tau}$\\
+
+\begin{itemize}
+
+\item $\tau = 0.0519+0.0030-0.0079$ (lowE; flat $\tau$ prior; TANH)(68\%); \\
+
+\item $\tau = 0.0504+0.0050
+-0.0079$ (lowE; flat $\tau$ prior; FlexKnot)(68\%); \\
+
+\item $\tau = 0.0487+0.0038
+−0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).
+
+\end{itemize}
+
+\textbf{Upper limit on high-redshift optical depth} (only given for the FlexKnot method):
+
+$\tau(15, 30) < 0.006$ (lowE, flat $\tau(15, 30)$, FlexKnot); \\
+
+$\tau(15, 30) < 0.007$ (lowE, flat knot, FlexKnot).\\
+
+Statements to address:
+
+\begin{enumerate}
+    \item {\textbf{Statement on prior in Planck 2018}:  ``Heinrich & Hu (2018) construct a prior that is uniform on τ, but which increases the allowed unphysical parameter space and is chosen a posteriori.
+Here we instead use the flat prior constructed by the procedure described in Millea & Bouchet (2018) and Handley & Millea (2019), which does not admit extra unphysical models and gives the most generic prior that leaves the prior on $\tau$ uniform."}
+
+    \item {Is this statement on tanh in Planck 2018 paper correct?} `` The TANH result gives slightly higher optical depth than the others, which is primarily driven by the fixed duration of reionization assumed."} \\
+
+    \item{\textbf{Other comments on our PC prior giving higher optical depth}: ``The PCA result is slightly lower, and is partly affected by the imperfect physicality priors that allow unphysical negative ionization fractions." and ``Millea \& Bouchet showed that the majority of this (2$\sigma$) preference disappeared when using the lower-noise Planck HFI SimLow likelihood (intermediate results reduction of systematics), with an additional sub-dominant effect due to the choice of prior."}
+
+\end{enumerate}
+
+How likelihoods and systematics evolved over time:
+
+\begin{enumerate}
+    \item 
+    \item {Planck 2018 paper attribute the 2018 upper limit on $\tau(15, 30)$ being \textbf{about 3 times more} stringent than the intermediate results by Millea \& Bouchet to the $SimLow \rightarrow SimAll$ where there was better control of systematics in HFI polarization.}
+\end{enumerate}
+
+[high redshift going away should be attributed to better control of systematics in HFI polarization data (changes in the SimAll likelihood compared to SimLow) [cite Planck 2018 results cp]]
+
 \end{document}

### 2020-10-29 paper/rei.tex

@@
 \beq
-\tau(z, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})
+\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})
 \eeq
@@
 \beq
-\tau(z, \zmax)_{\rm PC,\, \zmax=50} < XX \; (95\% \mathrm{CL})
+\tau(15, \zmax)_{\rm PC,\, \zmax=50} < XX \; (95\% \mathrm{CL})
 \eeq
@@
 
-Outline
-\begin{itemize}
-    \item compare with Planck official results
-    \item comment that Planck tau(15,30) constraints are too stringent (flex knots). Compare to ours. Point out why.
-    \item{ Comments on priors (see below).}
-    \item (*maybe in  another section): make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck.
-This was for higher tau, so transition happens at higher redshift.
-Now tau is much lower.
-\end{itemize}
+The Planck collaboration has released constraints on the total optical depth $\tau$ with a steplike model parameterized by the tanh function, which has evolved from $\tau = 0.067 \pm 0.022$ (Planck 2015) to $\tau = 0.055 \pm 0.009$ (Planck 2016), then to $\tau = 0.0506 \pm 0.0086$ (Planck 2018) [cite papers].
+The major improvement comes from clean-up of systematics in the High Frequency Instrument (HFI) data.
+
+In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for a steplike model, $\tau = 0.0504^{+0.0050}
+_{-0.0079}$ for the FlexKnot method and $\tau = 0.0487^{+0.0038}_{−0.0081}$ for the PC method with a different prior than that used in this paper.
+
+They also found the upper limit on high-redshift optical depth to be: $\tau(15, 30) < 0.006$ for the FlexKnot method with a flat $\tau(15, 30)$ prior or $<0.007$ with a flat knot prior.
+The PC result for the upper limit was not computed.
+
+Here we show that we find a less stringent upper limit using the PC method $\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})$. [explain why FlexKnot could give more stringent results.]
+
+\subsection{A note on priors}
+
+In Ref.~\cite{Millea:2018bko} (hereafter MB18)  a concern was raised that the physicality prior we take for the PC analysis in Ref.~\cite{Heinrich:2016ojb} correspond to a particular shape of prior that is not flat in of $\tau$ and would bias the PC result.
+The authors also proposed a method for flattening the prior by multiplying its inverse at each point in the multidimensional PC space.
+They claim that the choice of the physicality prior used here and in Ref.~\cite{Heinrich:2016ojb} contributed to the a large fraction of the shift in the $\tau$ posterior between the steplike and PC method.
+
+In Planck 2018 paper on cosmological parameters, the shift between the steplike model and the PC result in $\tau$ was attributed mostly to the reduction of HFI systematics, with subdominant contribution from priors.
+We therefore would like to clarify the impacts of the priors we choose and illustrate explicitly that there is no bias with this choice, whereas the flattening method proposed in MB18 is actually at risk of creating a non-flat prior and biasing results.
 
@@
 
-[Plot: add plot for tau(>z) contours for the 2 parameter model vs PC; direct integration first]
+[Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first]
 
@@
 \end{figure}
+
+\begin{itemize}
+    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higheroptical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
+Then transition to the following point}
+    \item {Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.}
+\end{itemize}
 

### 2020-11-04 paper/rei.tex
Updates from Overleaf

--- 2020-10-16 paper/rei.tex
@@
 Finally, we summarize our results and conclude in section~\ref{sec:conclusion}.
-
 \section{Background}
 \label{sec:background}
-
 The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}.
@@
 We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.
-
 \subsection{Reionization Principal Components}
 \label{sec:PC}
-
 We begin by parametrizing $x_e(z)$, the ionization fraction relative to the fully ionized hydrogen at redshift $z$, into its principal components $S_{a}(z)$ with respect to the CMB $E$-mode polarization:
@@
 n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.
-
 [Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
-
 Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
@@
 We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
-
 [rephrase this paragraph:]
@@
 impact of a given ionization history and to marginalize uncertainties that they introduce."
-
 We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$.
@@
 Because the optical depth is more directly related to the CMB observables than the ionization history itself, it allows us to make meaningful comparisons between the PC results and those with a specific prior in model space such as those admitting a particular functional form.
-
 For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
@@
 They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
-
 %we employ the smoothed $x_e$ which formally has support beyond the bounds. 
@@
 %Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.   
-
 \subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
 \label{sec:KDE}
-
 The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood.
 In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
-
 The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
@@
 where the overall normalization is arbitrary, and where we have chosen the smoothing kernel $K_f$ to be a  multivariate Gaussian with mean zero and covariance $f\mathbf{C}$, where $\mathbf{C}$ is the $N_{\rm PC} \times N_{\rm PC}$ covariance matrix estimated from the PC chains (see Table~\ref{tab:PC_stats}) and $f$ is a fraction smaller than 1.
-
 For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$.
@@
 Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.
-
 Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
@@
 We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
-
 %Therefore, the PC-based effective likelihood allows one to easily explore different models of ionization history while providing a much faster convergence in the MCMC analysis.
-
 \section{Planck 2018 PC Results}
 \label{sec:results}
-
 We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principle components.
-
 \subsection{Constraints on Principle Components}
-
 We use the official Planck likelihoods [add citation] \texttt{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper.
@@
 In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].
-
 \begin{figure*}
@@
 %results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
-
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
@@
 Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.
-
 The trajectories shown in black lines are the steplike models of instantaneous reionization allowed by the Planck 2015 data with the arrow pointing in the direction of increasing optical depth.
@@
 By construction, they miss any high-redshift ionization component by assuming neglibile ionization levels before the transition redshift $z_{\rm re}$.
-
 \begin{figure}
@@
 \end{figure}
-
 \begin{table}[b]
@@
 \end{table}
-
 \begin{figure}
@@
 \end{figure}
-
 We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area.
 [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
-
 In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}:
@@
 
-In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE likelihoods.
+In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE$ likelihoods. [Do we need this?]
 
@@
 \beq
-\tau_{\rm PC, \zmax=30} = XX \pm XX,
+\tau_{\rm PC,\, \zmax=30} = XX \pm XX,
 \eeq
@@
 \beq
-\tau_{\rm PC, \zmax=50} = XX \pm XX.
+\tau_{\rm PC,\, \zmax=50} = XX \pm XX.
 \eeq
@@
 \beq
-\tau(z, \zmax)_{\rm PC, \zmax=30} < XX \; (95\% CL)
+\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})
 \eeq
@@
 \beq
-\tau(z, \zmax)_{\rm PC, \zmax=50} < XX \; (95\% CL)
+\tau(15, \zmax)_{\rm PC,\, \zmax=50} < XX \; (95\% \mathrm{CL})
 \eeq
@@
 
-\begin{itemize}
-    \item compare with Planck official results
-    \item comment that Planck tau(15,30) constraints are too stringent (flex knots). Compare to ours. Point out why.
-    \item{ Comments on priors (see below).}
-    \item (*maybe in  another section): make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck.
-This was for higher tau, so transition happens at higher redshift.
-Now tau is much lower.
-\end{itemize}
+The Planck collaboration has released constraints on the total optical depth $\tau$ with a steplike model parameterized by the tanh function, which has evolved from $\tau = 0.067 \pm 0.022$ (Planck 2015) to $\tau = 0.055 \pm 0.009$ (Planck 2016), then to $\tau = 0.0506 \pm 0.0086$ (Planck 2018) [cite papers].
+The major improvement comes from clean-up of systematics in the High Frequency Instrument (HFI) data.
 
-Recap our appendix from Planck 2015 on the issue of prior:
-\begin{itemize}
-    \item {describe tau prior issue raised by MB18: claimed that prior not flat in tau so bias PC; proposed to flatten by multiplying the inverse of prior point-by-point in the PC space; claim that this flattening removes a shift between tanh and PC results.
-}
-    \item {summarize that we pointed it out the flaw in the logic in appendix of \cite{Heinrich:2018btc}.}
-    \item{reproduce a few tests here: 1) tau12 posterior (no need to show again, point to appendix); 2) show prior boxes in m1-m2 plane (done in Fig.4) - repeat the length of lines of constant tau12 within the box describe the prior distribution.
-Make the point that though the line rises, the Planck data is more informative than priors in m1, m2 makes no impact on the posterior.
-However when inverting the prior, introduces a bias in tau.}
-    \item{show test of explicitly constructing a prior that's locally flat in m1-m2, show posterior almost doesn't change.
-(only change is low-z enhancement, due to prior cut off at $z>6$.
-between our two priors, things don't change as much (might want to reproduce this result for this Planck 2018 results.
-And marius version too?}
-    \item{inversion method should not be used on future dataset, since data is more informative than prior.}
-\end{itemize}
+In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for a steplike model, $\tau = 0.0504^{+0.0050}
+_{-0.0079}$ for the FlexKnot method and $\tau = 0.0487^{+0.0038}_{−0.0081}$ for the PC method with a different prior than that used in this paper.
 
+They also found the upper limit on high-redshift optical depth to be: $\tau(15, 30) < 0.006$ for the FlexKnot method with a flat $\tau(15, 30)$ prior or $<0.007$ with a flat knot prior.
+The PC result for the upper limit was not computed.
+
+Here we show that we find a less stringent upper limit using the PC method $\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})$. [explain why FlexKnot could give more stringent results.]
+
+\subsection{A note on priors}
+
+In Ref.~\cite{Millea:2018bko} (hereafter MB18)  a concern was raised that the physicality prior we take for the PC analysis in Ref.~\cite{Heinrich:2016ojb} implies a non-flat prior in $\tau$ and introduces bias to the PC result.
+The authors also proposed a method to flatten the $\tau$ prior by multiplying its inverse at each point in the multidimensional PC space.
+They claim that adopting the physicality prior used here and in Ref.~\cite{Heinrich:2016ojb} contributed to the a large fraction of the shift in the $\tau$ posterior between the steplike and PC method in the Planck 2015 data.
+
+In the Planck 2018 paper on cosmological parameters, the reduction of optical depth from the PC results seen in the Planck 2018 data compared to the Planck 2015 data was attributed to the reduction of HFI systematics, with subdominant contribution from priors.
+We therefore would like to clarify the impacts of the priors we choose and illustrate explicitly that there is no bias with this choice, whereas the flattening method proposed in MB18 is actually at risk of creating a non-flat prior and biasing results.
+
+In the following, we will use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily.
+We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
+\beq
+\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
+\eeq
+where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
+Because most of the contributions to $\tau$ comes from the first two components as shown in Fig.
+7 of Ref.~\cite{Heinrich:2018btc} in the appendix, we can approximate it as
+\beq
+\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
+\eeq
+We also show it here in Fig.~\ref{fig:tau12} where the posterior of $\tau_{12}$ and $\tau$ (with five PCs) are compared for the Planck 2018 data.
+
+In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
+The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior.
+The physicality prior upon projection would correspond to a linearly rising prior in $\tau_{12}$ over the region where the posterior has support.
+One would be eager to conclude that this produces a biased constraint on $\tau_{12}$.
+However, because the Planck data is more constraining than the prior in $m_1$ and $m_2$, (the ellipse occupying a much smaller area in one part of the prior box), the value of the prior at high $\tau_{12}$ becomes irrelevant for the posterior.
+
+Using the procedure proposed by Ref.~\cite{Millea:2018bko}, inverting the prior value at each point in the PC space, would actually up-weight the high $\tau_{12}$ regions that are not supported by data, while down-weighting the regions that are.
+This produces an actual bias in the $\tau_{12}$ posterior toward lower optical depth.
+
+We compare the three priors in Fig.~\ref{fig:tau12_posterior_compare_priors} by showing their corresponding $\tau_{12}$ posteriors.
+Compared to the explicitly flat prior (dashed box in Fig.~\ref{fig:prior_box}) as the ground truth, the posterior from using the physicality prior shifts negligibly.
+The main difference is a slightly lower posterior at very low $\tau_{12}$, which is due to the fact that the physicality constraints restricts hydrogen to be fully ionized at $z \leq 6$ (part of the ellipse is outside of the box in Fig.~\ref{fig:prior_box}.
+A similar effect would be observed between two steplike models with and without restricting the reionization redshift to $z>6$.
+In comparison, inverting the prior to make it flat globally ends up biasing the tau posterior.
+[show in figure?]
+
+Therefore we conclude that the prior inversion procedure proposed in Ref.~\cite{Millea:2018bko} should not be used where data is more constraining than prior, which includes the Planck 2015, Planck 2018 and possible future datasets.
  \begin{figure}
-          \includegraphics[width=0.9\columnwidth]{paper/plots/plot_rotated_box_flat_tau_prior_fac_0p8.pdf}
+          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_plot_rotated_box_flat_tau_prior_fac_0.8.png}
           \caption
@@
 In the allowed region, the only difference is that the HMH16 prior clips the low $m_1$ edge of the allowed region due to physicality and the assumption that reionization occurs at
-          $z\ge 6$. [add in text the comparison of priors, and how tau is not biased, and for KDE physical models determine the priors]}  \label{fig:prior_box}
+          $z\ge 6$. [add in text the comparison of priors, and how tau is not biased, and for KDE physical models determine the priors]} 
+          \label{fig:prior_box}
 \end{figure}
 
-[Plot: add plot for tau(>z) contours for the 2 parameter model vs PC; direct integration first]
+\begin{itemize}
+    \item{Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first}
+    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higheroptical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
+Then transition to the following point}
+    \item {Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.}
+\end{itemize}
 
@@
 
-\subsection{Code description}
-\label{sec:code}
+[One paragraph intro to effective likelihood code, and overview of comparison given in this section].
 
@@
 \label{sec:example1}
+
+\begin{itemize}
+    \item Show posteriors of tanh model from direct MCMC vs KDE. Give timing improvement. Comment on differences. State f = 0.14 used.
+    \item (here?) Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
+\end{itemize}
 
@@
 \label{sec:example2}
+
+\begin{itemize}
+    \item Define 2 step model (record differences compared to previous paper)
+    \item Show 2D posterior plots between direct MCMC and KDE
+    \item Note timing differences (quote number of samples to convergence if needed)
+    \item Comment on any differences. Note f = 0.14 (any additional study of varying f). 
+\end{itemize}
+
 [rewording needed]
@@
 
+\begin{itemize}
+    \item Summarize what we did in one paragraph.
+    \item Give main conclusions in bullet format
+    \item main results from PCs: PC constraints, tau, $tau(>z)$.  
+    \item comparison w/ Planck 
+    \item note on priors
+    \item likelihood code, comparisons
+    \item other interesting points to discuss and add 
+    \item future outlook
+\end{itemize}
+
 \bibliography{rei.bib}

### 2020-11-18 paper/rei.tex

@@
 
-\newcommand{\Omegamzero}{\Omega_{{\rm m,0}}}
-\newcommand{\Rbar}{$\bar{R}$}
-\newcommand{\lsc}{\mathcal{L}}
-\newcommand{\rhom}{\rho_{\rm m}}
-\newcommand{\Mpch}{\mbox{Mpc}/h}
-\newcommand{\iMpch}{h/\mbox{Mpc}}
-\newcommand{\Msun}{M_\odot}
-\newcommand{\Mv}{M_{\rm v}}
-
 \newcommand{\refsec}[1]{section~\ref{sec:#1}}
@@
 \newcommand{\refFig}[1]{Fig.~\ref{fig:#1}}
-\newcommand{\curv}{{\cal R}}
+
 \newcommand{\xef}{x_e^{\rm fid}}
@@
 \newcommand{\zmin}{z_{\rm min}}
+\newcommand{\zre}{z_{\rm re}}
 \newcommand{\xemin}{x_e^{\rm min}}
+\newcommand{\lsc}{\mathcal{L}}
+\newcommand{\tauhi}{\tau_{\rm hi}}
+\newcommand{\taulo}{\tau_{\rm lo}}
 
@@
 
-\title{Reionization Planck 2018 ...}
+\title{Effective Reionization Likelihood from Planck 2018 Data (?)}
 
@@
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
-For $z<6$, we follow camb to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width  $\Delta z = 0.5$.
+For $z<6$, we follow CAMB to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$.
 
@@
 For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$.
-To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx XX \times 10^{X}$ [to fill] chain samples.
-For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
-Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.
+To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples.
+% 1.06 cut w/ physicality prior
+% 1.072 no cut
+For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient. Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
 Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
@@
 \beq
-\tau_{\rm PC,\, \zmax=30} = XX \pm XX,
+\tau_{\rm PC,\, \zmax=30} = 0.0619^{+0.0056}_{-0.0068}\;\; (\textsc{srollv2}),
 \eeq
@@
 \beq
-\tau_{\rm PC,\, \zmax=50} = XX \pm XX.
+\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\textsc{srollv2}).
 \eeq
-We have an upper limit on $\tau(15, \zmax)$ of 
+These total optical depth constraints have slightly higher central values and smaller error bars compared to those obtained using the original Planck-released \textsc{lowE} likelihood
 \beq
-\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})
+\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\textsc{lowE}),
 \eeq
@@
 \beq
-\tau(15, \zmax)_{\rm PC,\, \zmax=50} < XX \; (95\% \mathrm{CL})
+\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\textsc{lowE}).
+\eeq
+[more comments here]
+%
+
+The upper limit on $\tau(15, \zmax)$ are 
+\beq
+\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\textsc{srollv2}),
+\eeq
+[double check on full chain (1015) this is from 0930]
+and 
+\beq
+\tau(15, \zmax)_{\rm PC,\, \zmax=50} < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\textsc{srollv2}).
 \eeq
@@
 
-In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
+In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints (ellipse) in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
 The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior.
@@
 Compared to the explicitly flat prior (dashed box in Fig.~\ref{fig:prior_box}) as the ground truth, the posterior from using the physicality prior shifts negligibly.
-The main difference is a slightly lower posterior at very low $\tau_{12}$, which is due to the fact that the physicality constraints restricts hydrogen to be fully ionized at $z \leq 6$ (part of the ellipse is outside of the box in Fig.~\ref{fig:prior_box}.
-A similar effect would be observed between two steplike models with and without restricting the reionization redshift to $z>6$.
-In comparison, inverting the prior to make it flat globally ends up biasing the tau posterior.
-[show in figure?]
+The main difference is a slightly reduced posterior values at very low $\tau_{12}$, which is due to the physicality condition and the restriction that hydrogen reionization be completed by $z = 6$.
+This is manifested as the data constraint ellipses being partially outside of the solid prior box in Fig.~\ref{fig:prior_box}.
+A similar effect would be observed for the steplike models with or without the  restrictions that reionization occurs at $z>6$.
+We also show that the results between the two priors are similar for the $\tau(15, \zmax)$ statistics: $\tau_{12}(15, \zmax) > XX $(95\% C.L.) for the flat $\tau_{12}$ prior vs $\tau_{12}(15, \zmax) > XX $ (95\% C.L.) for the physicality prior.
 
-Therefore we conclude that the prior inversion procedure proposed in Ref.~\cite{Millea:2018bko} should not be used where data is more constraining than prior, which includes the Planck 2015, Planck 2018 and possible future datasets.
+This is a small difference in comparison to the bias introduced by multiplying the posterior with the inverse of the $\tau_{12}$ prior.
+[show in figure?] [add a sentence to reiterate that the shift between PC and steplike model seen in the Planck 2015 is not due to the use of physicality prior].
+Furthermore, the bias caused by using the inverse prior method in the Planck 2018 data is on the order of $\Delta \tau \sim 10^{-X}$, which could partially [or fully] account for the difference between FlexKnots and PC results seen in Ref.~\cite{Aghanim:2018eyx} on the total optical depth.
+
+[Discussion of why FlexKnots is too stringent for $\tau_{12}(15, \zmax)$. Not sure how to do yet. Where should this go? Maybe not prior related.]
+
+We conclude that the prior inversion procedure proposed in Ref.~\cite{Millea:2018bko} should not be used where data is more constraining than prior, which includes the Planck 2015, Planck 2018 and possible future datasets.
+
  \begin{figure}
           \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_plot_rotated_box_flat_tau_prior_fac_0.8.png}
-          \caption
-          {[Placeholder right now (Planck 2015 plot)] Priors on the $m_1-m_2$ parameter space: the original HMH16 prior (inset square, solid lines) and a prior that is flat in
-          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) whose sides are aligned with lines of
-          constant $\tau_{12}$ (light gray lines).   The HMH16 prior allows more parameter space at high vs low $\tau_{12}$ but this
-          is not relevant since the data constraints (ellipses) exclude this region.
-In the allowed region, the only difference is that the HMH16 prior clips the low $m_1$ edge of the allowed region due to physicality and the assumption that reionization occurs at
-          $z\ge 6$. [add in text the comparison of priors, and how tau is not biased, and for KDE physical models determine the priors]} 
+          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
+          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) which are aligned with lines of constant $\tau_{12}$ (light gray lines).
+Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
+The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.}
           \label{fig:prior_box}
@@
 
-\begin{itemize}
-    \item{Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first}
-    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higheroptical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
-Then transition to the following point}
-    \item {Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.}
-\end{itemize}
+% NEXT:
 
-\begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/plot_tau_gtz.pdf}
-\caption{...
-}
-\label{fig:}
-\end{figure}
+% - fill in all numbers: tau(15, 30) upper bounds, store in table or spreadsheet in git; covariance matrix (store in git too); other numbers mentioned in the paper.
 
-\begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
-\caption{...
-}
-\label{fig:}
-\end{figure}
+% - talk w/ Wayne, finalize decisions, and make polished plots as needed.
+
+% - write abstract, find title and give the text another pass to refine.
 
@@
 
-[One paragraph intro to effective likelihood code, and overview of comparison given in this section].
+As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$.
+This code is downloadable on Github at \url{} [give URL].
+The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_ TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.
+
+To obtain posteriors of model parameters from the corresponding Planck likelihoods, simply specify the functional form for $x_e(z)$ in the code, and run a MCMC analysis using the extension we provide to a MCMC sampler (currently supporting CosmoMC).
+Note that the likelihood cannot distinguish the effects of the model parameters that impact $z \sim 6$, where the Universe is assumed to be fully ionized.
+Furthermore, transitions faster than $\Delta z \sim 0.1$ between $6 < z < 6.25$ may not be fully captured because of how PC functions are discretized with $\delta z = 0.25$ and are linearly interpolated within this bin size.
+[review this statement] To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
+Finally, you may choose to run the chains at a different cosmological model than that of the default Planck 2018 best-fit model, provided it is close enough.
+
+We demonstrate successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has an one additional parameter to model the high-$z$ ionization as a plateau in $x_e$ before the transition to full ionization happens in the standard tanh model.
 
@@
 
+\begin{figure}
+\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+\caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?].
+The results agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
+}
+\label{fig:tanh_}
+\end{figure}
+
+The steplike model used as the standard approach in CAMB describes the hydrogen and singly ionized helium reionization as a tanh function:
+
+ \begin{equation}
+x_e(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} + x_e^{\rm rec},
+ \label{eqn:tanh_1D}
+ \end{equation}
+
+ with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$. Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
+ \beq
+ f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
+ \eeq
+ is the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.
+In addition, the doubly ionized helium reionization is parameterized by an additional tanh function in redshift centered at $z_{\rm He} = 3.5$ with width $\Delta z = 0.5$.
+
+In \reffig{tanh_1D} we show the posterior distribution of the $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$.
+The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
+We expect the smoothing to be at roughly the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
+
 \begin{itemize}
-    \item Show posteriors of tanh model from direct MCMC vs KDE. Give timing improvement. Comment on differences. State f = 0.14 used.
     \item (here?) Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
+    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
+Then transition to the following point}
 \end{itemize}
 
-In Fig.~\ref{fig:reion_models}, we show the fiducial ionization history (thick blue) and contrast it with
-the standard approach of CAMB (thin black) that takes hydrogen and singly ionized helium reionization to be given by
-the tanh form
- \begin{equation}
-x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
- \label{eqn:tanh}
- \end{equation}
- with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$.
-We take here $z_*= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
+ %We take here $z_{\rm re}= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
 Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
- Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an 
- excellent representation of the polarization power spectrum.
+ %Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an excellent representation of the polarization power spectrum.
 
@@
 
-\begin{itemize}
-    \item Define 2 step model (record differences compared to previous paper)
-    \item Show 2D posterior plots between direct MCMC and KDE
-    \item Note timing differences (quote number of samples to convergence if needed)
-    \item Comment on any differences. Note f = 0.14 (any additional study of varying f). 
-\end{itemize}
-
-[rewording needed]
-Our example is a two-step model 
- \bea
-x_e^{\rm true}&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
-&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
- \label{eqn:tanh_highz}
- \eea
-where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$, with $\Delta z_1 = 0.25$ instead of the usual $\Delta z_1 = 0.5$,
-to provide sharper distinctions between the two steps.
-We choose the second step to have $z_{\rm t}=28$  and  $\Delta z_2 = 1.0$  to illustrate below
-how the 5 PC analysis with $\zmax=30$ represents the same model.  Here $x_e^{\rm rec}$ is the ionization history from recombination only.
-The canonical tanh model is essentially recovered in the limit $\xemin$ approaches the negligible $x_e^{\rm rec}$.  Therefore the double
-step model adds a single parameter to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$.   We show an example
-of the two-step model in Fig.~\ref{fig:two_step_model}.
-
 \begin{figure}
 \includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
-\caption{Testing \taulo prior -- Full $x_e(z)$ of the model (\taulo, \tauhi) = (0.04, 0.02) shown in the Fig.~\ref{fig:taulo_prior_test_cl}. 
+\caption{Ionization history $x_e(z)$ of the two-step model $(\taulo, \tauhi) = (0.04, 0.02)$ for the hydrogen and singly ionized helium.
 }
@@
 
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
-\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
-The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$. } 
-\label{fig:taulo_prior_test_cl}
+The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
+ \bea
+x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
+&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
+ \label{eqn:tanh_highz}
+ \eea
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ as in the first example, but now we choose $\Delta z_1 = 0.015\, (1+z_{\rm re})$ instead of the usual $\Delta z_1 = 0.5$, to provide sharper distinctions between the two steps.
+We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
+We recover the standard tanh model in the limit when $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
+Therefore this model adds a single parameter $\xemin$ to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$. We show an example
+of this two-step model in Fig.~\ref{fig:two_step_model}.
+
+\begin{figure}[t]
+\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\caption{Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
+}
+\label{fig:two_parameter_model_2D}
 \end{figure}
 
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
-\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
+To obtain posterior distributions with flat $\tau$ priors, we convert the model to be parameterized by low and high-redshift optical depth, $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$, 
+each defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ respectively.
+We choose the split to be $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
+We sample in the $(\taulo, \tauhi)$ parameter space and the corresponding ($\zre, \xemin$) is found through an iterative search similar to that used for finding $\zre$ given total $\tau$ in the canonical tanh model.
+The priors are chosen to be $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where $\tau_{\rm max} = 0.35$, and $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth for both the effective and full likelihood runs,
+except that we have chosen $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. 
+Note that $\taulo \lesssim 0.038$ roughly corresponds to $\zre < 6$ in the canonical tanh model.
+For the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
+So to make the comparison apples-to-apples, we further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the effective reionization likelihood chains and the full likelihood chains.
+
+In ~\reffig{two_parameter_model_2D}, we compare the posteriors of $\taulo$ and $\tauhi$ obtained with these two likelihood codes, which show good agreement.
+Again we used $f = 0.14$ here for the KDE computation.
+[add timing results here?] Note that the truncation of the 2D constraint ellipses on the lower end of $\taulo$ is curved cut rather than a straight line, because $\zre$ cut we performed corresponds to a different value of $\taulo$ for a different $\tauhi$.
+
+The marginalized 1D constraints are $\taulo = XX \pm XX$, $\tauhi = XX \pm XX$ for both runs.
+The 95\% C.L.
+upper limit on $\tauhi$ is $\tauhi < XX$ is consistent with the $\tau(15, 30) < XX$ results we got from the PC constraints.
+[is it?
+double check] This model of course do not capture the larger space of physical models with high-$z$ ionization like the FlexKnot or PC methods would, but consists of a subset of models for which the high-redshift optical depth results must still be consistent with the constraints from these more general methods.
+
+To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}.
+Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts.
+According to the FlexKnots constraints in Ref.~\cite{}, however, this model would have been ruled out given that $\tau(15, 30) < 0.007$ (95\% C.L.) with the FlexKnots results.
+[provide comment ...; any further discussion on literature results with FlexKnots?]
+
+[add a conclusive paragraph or sentence]
+
+%Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first?
+
+\begin{figure}[t]
+\includegraphics[width=0.450\textwidth]{plots/plot_tau_gtz.pdf}
+\caption{Same as \reffig{plot_taugtz_PC_vs_tanh}, but superimposed with the cumulative optical depth $\tau(z, \zmax)$ of the best-fit two-step model.
+The two-step model is a toy model that adds a high-$z$ ionization plateau to the standard tanh model with one additional parameter.
+Compared to the cumulative $\tau$ distributions from PCs, the best-fit step-model model is within the 95\% upper limit of the cumulative $\tau$ constraints at all redshifts.
+}
+\label{fig:plot_taugtz_two_step_best_fit}
+\end{figure}
+
+%\begin{figure}
+%\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
+%\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
 The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
@@
 }
-\label{fig:two_parameter_model_2D_plot}
-\end{figure}
+%\label{fig:two_parameter_model_2D_plot}
+%\end{figure}
+
+%\begin{figure}
+%\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+%\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
+%The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$. } 
+%\label{fig:taulo_prior_test_cl}
+%\end{figure}
 

### 2020-11-19 paper/rei.tex
Updates from Overleaf

@@
 \newcommand{\taulo}{\tau_{\rm lo}}
+\newcommand{\sample}{{\rm sample}}
 
@@
 For $z<6$, we follow CAMB to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$.
+\wh{Probably need some brief words about MCMC sampling of the posteriors of $m_a$ jointly with $\Lambda$CDM parameters here so that the KDE section makes sense.
+I moved the optical depth stuff to a following subsection to clarify role, and order of operations, of the physicality prior.}
 
-We also put constraints on the cumulative Thomson optical depth derived parameters 
-\begin{equation}
-\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
-\label{eq:cumtau}
-\end{equation}
-where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.
-Because the optical depth is more directly related to the CMB observables than the ionization history itself, it allows us to make meaningful comparisons between the PC results and those with a specific prior in model space such as those admitting a particular functional form.
-For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
- \begin{equation}
-x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
- \label{eqn:tanh}
- \end{equation}
- where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
-and has an implicit prior in model space that excludes finite ionization above its transition redshift.
-
-The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.
-When showing results, however, we do impose a physicality prior derived from the condition $0<x_e<x_e^{\rm max}$ following Ref.~\cite{Mortonson:2008rx}
-%
-\begin{equation}
-\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
-\end{equation}
-where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
-\begin{equation}
-m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
-\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
-\label{eq:individualprior}
-\end{equation}
-%
-Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
-They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
-%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
-% We include this small correction by integrating
-% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
-   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
-%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.   
 \subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
 \label{sec:KDE}
-The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood.
-In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+\wh{The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data.}  In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
 The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
@@
 We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
-%Therefore, the PC-based effective likelihood allows one to easily explore different models of ionization history while providing a much faster convergence in the MCMC analysis.
+
+\subsection{\wh{Cumulative Optical Depth}}
+
+\wh{Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth}
+\begin{equation}
+\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
+\label{eq:cumtau}
+\end{equation}
+where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.
+The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of
+$C_\ell^{EE}$ -- the higher the redshift, the
+larger the relative contribution at higher $\ell$.
+
+Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
+\wh{decide whether to introduce models one-by-one or together - we could postpone the step model discussion until later without interrupting the flow}
+For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
+ \begin{equation}
+x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+ \label{eqn:tanh}
+ \end{equation}
+ where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
+ excludes finite ionization above its transition redshift by definition.
+
+The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.
+\wh{The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
+unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.}
+Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
+ we do impose a physicality prior
+ by truncating the posteriors of $m_a$
+ {\it after} obtaining them from Markov Chain Monte Carlo sampling, following Ref.~\cite{Mortonson:2008rx}
+%
+\begin{equation}
+\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
+\end{equation}
+where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
+\begin{equation}
+m_a^{\pm} = \int_{z_{\rm min}}^{z_{\rm max} } dz \frac{S_a(z)[x_e^{\rm max} -2 x_e^{\rm fid}(z)]
+\pm x_e^{\rm max} | S_a(z)|}{2(z_{\rm max}-z_{\rm min})}.
+\label{eq:individualprior}
+\end{equation}
+%
+Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
+\wh{In other words, no physical model is excluded by these priors.}
+% I think we don't impose the prior at all
+% for KDE 
+%
+%They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
+
+%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
+% We include this small correction by integrating
+% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
+   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
+%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.
+
+\wh{Think about whether the refutation of Marius' prior and should go here.  If not then we probably want a forward pointing sentence here about impact of priors in practice.}
+
 \section{Planck 2018 PC Results}
 \label{sec:results}
-We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principle components.
-\subsection{Constraints on Principle Components}
+We now present the complete reionization constraints obtained from the Planck 2018 likelihoods using the principal components.
+\subsection{Constraints on Principal Components}
 We use the official Planck likelihoods [add citation] \texttt{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper.
@@
 \includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
-\caption{Constraints from Planck data on the amplitudes of the five reionization principle components that describe all physical ionization models up to $z_{\rm max} = 30$.
+\caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$.
 We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}.
@@
 The physicality priors on the ionization history used in this work are delineated by the box boundaries.
-[TODO: update legend] [add tau trajectories] [add prior boundaries]}
+[TODO: update legend] [add tau trajectories] [add prior boundaries] \wh{I wonder if for pedagogy we should show the full range and just put a box around the physical region for the $m_1$'s?}}
 \label{fig:plot_mjs_2018_vs_2015}
@@
 \end{figure}
+
 \begin{table}[b]
 \centering
-\caption{[Placeholder, need to fill in real numbers] PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
+\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
 \label{tab:PC_stats}
@@
 		\\ \hline
+
+$m_1$ & -0.119 & -0.098 & 1.000 & 0.657 & 0.208 & 0.094 & 0.067 \\ 
+$m_2$ & -0.013 & 0.010 & 0.657 & 1.000 & -0.048 & 0.273 & 0.142 \\ 
+$m_3$ & -0.115 & -0.065 & 0.208 & -0.048 & 1.000 & 0.064 & 0.018 \\ 
+$m_4$ & -0.025 & 0.085  & 0.094 & 0.273 & 0.064 & 1.000 & 0.195 \\ 
+$m_5$ & 0.118 & 0.072  & 0.067 & 0.142 & 0.018 & 0.195 & 1.000 \\
+
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[b]
+\centering
+\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
+\label{tab:PC_stats}
+\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
+\hline
+
+			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
+		\\ \hline
+
 $m_1$ 
@@
 $m_5$ 
-	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ \hline
+	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ 
+\hline
 \end{tabular}
 \end{table}
+
 \begin{figure}
@@
 The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
-[Might be nice to overlay the tanh trajectories, no 1 sigma, 2 sigma business].
+\wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}  [Might be nice to overlay the tanh trajectories, no 1 sigma, 2 sigma business].
+
+\wh{I wonder if the order should be reversed and talk about the model testing and KDE first, then the model independent cumulative optical depth?
+That way the model stuff could all be introduced here and simulaneously compared with the fiducal results?
+The flow could be that we first show more high z optical depth is allowed in the two step model but not FlexKnot and then talk about it being a model independent PC result, concluding that we are more robust than FlexKnot?
+Of course I'm assuming that the 2 step model shows this...}
 

### 2020-11-25 paper/rei.tex
Updates from Overleaf

@@
 In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].
+
 \begin{figure*}
-\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
+\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
 \caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$.
@@
 The physicality priors on the ionization history used in this work are delineated by the box boundaries.
-[TODO: update legend] [add tau trajectories] [add prior boundaries] \wh{I wonder if for pedagogy we should show the full range and just put a box around the physical region for the $m_1$'s?}}
+[TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$.
+make point that tanh ML is now within the the center of PC constraints]}
 \label{fig:plot_mjs_2018_vs_2015}
@@
 %results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
+%
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
@@
 By construction, they miss any high-redshift ionization component by assuming neglibile ionization levels before the transition redshift $z_{\rm re}$.
+
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wGaussEllipse.png}
-\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, $m_1-m_2$ plane, comparing Planck 2018 (blue) vs 2015 results (red).
-The black solid line indicates the steplike models with the arrow pointing in the direction of increasing $\tau$.
-The dashed lines indicate the contours corresponding a Gaussian likelihood (Eq.~\ref{eq:gaussian}) using the PC covariance from the Planck 2018 chains.
-[add Gaussian covariance legend?] [add tau trajectories here?]
+\includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper.
 }
@@
 
-\begin{figure}
-\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.png}
-\caption{Same as Fig.~\ref{fig:plot_mjs_2018_vs_2015} but comparing the Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the official \texttt{simall\_EE} likelihood and the \texttt{srollv2} likelihood which has better foreground removal (shown throughout this work).
-[Comments on shift m1-m2 plane].
-}
-\label{fig:plot_m1m2_lowE_vs_srollv2}
-\end{figure}
 We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area.

### 2020-12-03 paper/rei.tex
Before rearranging order of text to be PC results->KDE->model independent results.
Updates from Overleaf

@@
 
-$m_1$ & -0.119 & -0.098 & 1.000 & 0.657 & 0.208 & 0.094 & 0.067 \\ 
-$m_2$ & -0.013 & 0.010 & 0.657 & 1.000 & -0.048 & 0.273 & 0.142 \\ 
-$m_3$ & -0.115 & -0.065 & 0.208 & -0.048 & 1.000 & 0.064 & 0.018 \\ 
-$m_4$ & -0.025 & 0.085  & 0.094 & 0.273 & 0.064 & 1.000 & 0.195 \\ 
-$m_5$ & 0.118 & 0.072  & 0.067 & 0.142 & 0.018 & 0.195 & 1.000 \\
+$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & -0.208 & -0.094 & 0.067 \\
+$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & -0.273 & 0.142 \\
+$m_3$ & $$0.078 & 0.098 & -0.208 & 0.048 & 1.000 & 0.064 & -0.018 \\
+$m_4$ & -0.077 & 0.131 & -0.094 & -0.273 & 0.064 & 1.000 & -0.195 \\
+$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & -0.018 & -0.195 & 1.000 \\
 
@@
 [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
-In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}:
+In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}.
+
 \begin{equation}
@@
 
-\subsection{Implications for optical depth}
+\section{Effective Likelihood}
+\label{sec:effective_likelihood}
+
+As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$.
+This code is downloadable on Github at \url{} [give URL].
+The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_ TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.
+
+To obtain posteriors of model parameters from the corresponding Planck likelihoods, simply specify the functional form for $x_e(z)$ in the code, and run a MCMC analysis using the extension we provide to a MCMC sampler (currently supporting CosmoMC).
+Note that the likelihood cannot distinguish the effects of the model parameters that impact $z \sim 6$, where the Universe is assumed to be fully ionized.
+Furthermore, transitions faster than $\Delta z \sim 0.1$ between $6 < z < 6.25$ may not be fully captured because of how PC functions are discretized with $\delta z = 0.25$ and are linearly interpolated within this bin size.
+[review this statement] To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
+Finally, you may choose to run the chains at a different cosmological model than that of the default Planck 2018 best-fit model, provided it is close enough.
+
+We demonstrate successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has an one additional parameter to model the high-$z$ ionization as a plateau in $x_e$ before the transition to full ionization happens in the standard tanh model.
+
+\subsection{Example 1: tanh model}
+\label{sec:example1}
+
+\begin{figure}
+\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+\caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?].
+The results agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
+}
+\label{fig:tanh_}
+\end{figure}
+
+The steplike model used as the standard approach in CAMB describes the hydrogen and singly ionized helium reionization as a tanh function:
+
+ \begin{equation}
+x_e(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} + x_e^{\rm rec},
+ \label{eqn:tanh_1D}
+ \end{equation}
+
+ with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$. Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
+ \beq
+ f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
+ \eeq
+ is the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.
+In addition, the doubly ionized helium reionization is parameterized by an additional tanh function in redshift centered at $z_{\rm He} = 3.5$ with width $\Delta z = 0.5$.
+
+In \reffig{tanh_1D} we show the posterior distribution of the $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$.
+The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
+We expect the smoothing to be at roughly the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
+
+\begin{itemize}
+    \item (here?) Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
+    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
+Then transition to the following point}
+\end{itemize}
+
+ %We take here $z_{\rm re}= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
+Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
+ %Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an excellent representation of the polarization power spectrum.
+
+\subsection{Example 2: high-z model}
+\label{sec:example2}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
+\caption{Ionization history $x_e(z)$ of the two-step model $(\taulo, \tauhi) = (0.04, 0.02)$ for the hydrogen and singly ionized helium.
+}
+\label{fig:two_step_model}
+\end{figure}
+
+The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
+ \bea
+x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
+&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
+ \label{eqn:tanh_highz}
+ \eea
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ as in the first example, but now we choose $\Delta z_1 = 0.015\, (1+z_{\rm re})$ instead of the usual $\Delta z_1 = 0.5$, to provide sharper distinctions between the two steps.
+We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
+We recover the standard tanh model in the limit when $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
+Therefore this model adds a single parameter $\xemin$ to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$. We show an example
+of this two-step model in Fig.~\ref{fig:two_step_model}.
+
+\begin{figure}[t]
+\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\caption{Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
+}
+\label{fig:two_parameter_model_2D}
+\end{figure}
+
+To obtain posterior distributions with flat $\tau$ priors, we convert the model to be parameterized by low and high-redshift optical depth, $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$, 
+each defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ respectively.
+We choose the split to be $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
+We sample in the $(\taulo, \tauhi)$ parameter space and the corresponding ($\zre, \xemin$) is found through an iterative search similar to that used for finding $\zre$ given total $\tau$ in the canonical tanh model.
+The priors are chosen to be $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where $\tau_{\rm max} = 0.35$, and $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth for both the effective and full likelihood runs,
+except that we have chosen $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. 
+Note that $\taulo \lesssim 0.038$ roughly corresponds to $\zre < 6$ in the canonical tanh model.
+For the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
+So to make the comparison apples-to-apples, we further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the effective reionization likelihood chains and the full likelihood chains.
+
+In ~\reffig{two_parameter_model_2D}, we compare the posteriors of $\taulo$ and $\tauhi$ obtained with these two likelihood codes, which show good agreement.
+Again we used $f = 0.14$ here for the KDE computation.
+[add timing results here?] Note that the truncation of the 2D constraint ellipses on the lower end of $\taulo$ is curved cut rather than a straight line, because $\zre$ cut we performed corresponds to a different value of $\taulo$ for a different $\tauhi$.
+
+The marginalized 1D constraints are $\taulo = XX \pm XX$, $\tauhi = XX \pm XX$ for both runs.
+The 95\% C.L.
+upper limit on $\tauhi$ is $\tauhi < XX$ is consistent with the $\tau(15, 30) < XX$ results we got from the PC constraints.
+[is it?
+double check] This model of course do not capture the larger space of physical models with high-$z$ ionization like the FlexKnot or PC methods would, but consists of a subset of models for which the high-redshift optical depth results must still be consistent with the constraints from these more general methods.
+
+To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}.
+Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts.
+According to the FlexKnots constraints in Ref.~\cite{}, however, this model would have been ruled out given that $\tau(15, 30) < 0.007$ (95\% C.L.) with the FlexKnots results.
+[provide comment ...; any further discussion on literature results with FlexKnots?]
+
+[add a conclusive paragraph or sentence]
+
+%Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first?
+
+\begin{figure}[t]
+\includegraphics[width=0.450\textwidth]{plots/plot_tau_gtz.pdf}
+\caption{Same as \reffig{plot_taugtz_PC_vs_tanh}, but superimposed with the cumulative optical depth $\tau(z, \zmax)$ of the best-fit two-step model.
+The two-step model is a toy model that adds a high-$z$ ionization plateau to the standard tanh model with one additional parameter.
+Compared to the cumulative $\tau$ distributions from PCs, the best-fit step-model model is within the 95\% upper limit of the cumulative $\tau$ constraints at all redshifts.
+}
+\label{fig:plot_taugtz_two_step_best_fit}
+\end{figure}
+
+%\begin{figure}
+%\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
+%\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
+The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
+The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
+}
+%\label{fig:two_parameter_model_2D_plot}
+%\end{figure}
+
+%\begin{figure}
+%\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+%\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
+%The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$. } 
+%\label{fig:taulo_prior_test_cl}
+%\end{figure}
+
+\section{Implications for optical depth}
+
+\subsection{Cumulative optical depth}
 
@@
 
-\section{Effective Likelihood}
-\label{sec:effective_likelihood}
-
-As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$.
-This code is downloadable on Github at \url{} [give URL].
-The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_ TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.
-
-To obtain posteriors of model parameters from the corresponding Planck likelihoods, simply specify the functional form for $x_e(z)$ in the code, and run a MCMC analysis using the extension we provide to a MCMC sampler (currently supporting CosmoMC).
-Note that the likelihood cannot distinguish the effects of the model parameters that impact $z \sim 6$, where the Universe is assumed to be fully ionized.
-Furthermore, transitions faster than $\Delta z \sim 0.1$ between $6 < z < 6.25$ may not be fully captured because of how PC functions are discretized with $\delta z = 0.25$ and are linearly interpolated within this bin size.
-[review this statement] To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
-Finally, you may choose to run the chains at a different cosmological model than that of the default Planck 2018 best-fit model, provided it is close enough.
-
-We demonstrate successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has an one additional parameter to model the high-$z$ ionization as a plateau in $x_e$ before the transition to full ionization happens in the standard tanh model.
-
-\subsection{Example 1: tanh model}
-\label{sec:example1}
-
-\begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
-\caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?].
-The results agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
-}
-\label{fig:tanh_}
-\end{figure}
-
-The steplike model used as the standard approach in CAMB describes the hydrogen and singly ionized helium reionization as a tanh function:
-
- \begin{equation}
-x_e(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} + x_e^{\rm rec},
- \label{eqn:tanh_1D}
- \end{equation}
-
- with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$. Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
- \beq
- f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
- \eeq
- is the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.
-In addition, the doubly ionized helium reionization is parameterized by an additional tanh function in redshift centered at $z_{\rm He} = 3.5$ with width $\Delta z = 0.5$.
-
-In \reffig{tanh_1D} we show the posterior distribution of the $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$.
-The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
-We expect the smoothing to be at roughly the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
-
-\begin{itemize}
-    \item (here?) Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
-    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
-Then transition to the following point}
-\end{itemize}
-
- %We take here $z_{\rm re}= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
-Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
- %Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an excellent representation of the polarization power spectrum.
-
-\subsection{Example 2: high-z model}
-\label{sec:example2}
-
-\begin{figure}
-\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
-\caption{Ionization history $x_e(z)$ of the two-step model $(\taulo, \tauhi) = (0.04, 0.02)$ for the hydrogen and singly ionized helium.
-}
-\label{fig:two_step_model}
-\end{figure}
-
-The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
- \bea
-x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
-&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
- \label{eqn:tanh_highz}
- \eea
-where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ as in the first example, but now we choose $\Delta z_1 = 0.015\, (1+z_{\rm re})$ instead of the usual $\Delta z_1 = 0.5$, to provide sharper distinctions between the two steps.
-We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
-We recover the standard tanh model in the limit when $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
-Therefore this model adds a single parameter $\xemin$ to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$. We show an example
-of this two-step model in Fig.~\ref{fig:two_step_model}.
-
-\begin{figure}[t]
-\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-\caption{Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
-}
-\label{fig:two_parameter_model_2D}
-\end{figure}
-
-To obtain posterior distributions with flat $\tau$ priors, we convert the model to be parameterized by low and high-redshift optical depth, $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$, 
-each defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ respectively.
-We choose the split to be $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
-We sample in the $(\taulo, \tauhi)$ parameter space and the corresponding ($\zre, \xemin$) is found through an iterative search similar to that used for finding $\zre$ given total $\tau$ in the canonical tanh model.
-The priors are chosen to be $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where $\tau_{\rm max} = 0.35$, and $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth for both the effective and full likelihood runs,
-except that we have chosen $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. 
-Note that $\taulo \lesssim 0.038$ roughly corresponds to $\zre < 6$ in the canonical tanh model.
-For the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
-So to make the comparison apples-to-apples, we further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the effective reionization likelihood chains and the full likelihood chains.
-
-In ~\reffig{two_parameter_model_2D}, we compare the posteriors of $\taulo$ and $\tauhi$ obtained with these two likelihood codes, which show good agreement.
-Again we used $f = 0.14$ here for the KDE computation.
-[add timing results here?] Note that the truncation of the 2D constraint ellipses on the lower end of $\taulo$ is curved cut rather than a straight line, because $\zre$ cut we performed corresponds to a different value of $\taulo$ for a different $\tauhi$.
-
-The marginalized 1D constraints are $\taulo = XX \pm XX$, $\tauhi = XX \pm XX$ for both runs.
-The 95\% C.L.
-upper limit on $\tauhi$ is $\tauhi < XX$ is consistent with the $\tau(15, 30) < XX$ results we got from the PC constraints.
-[is it?
-double check] This model of course do not capture the larger space of physical models with high-$z$ ionization like the FlexKnot or PC methods would, but consists of a subset of models for which the high-redshift optical depth results must still be consistent with the constraints from these more general methods.
-
-To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}.
-Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts.
-According to the FlexKnots constraints in Ref.~\cite{}, however, this model would have been ruled out given that $\tau(15, 30) < 0.007$ (95\% C.L.) with the FlexKnots results.
-[provide comment ...; any further discussion on literature results with FlexKnots?]
-
-[add a conclusive paragraph or sentence]
-
-%Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first?
-
-\begin{figure}[t]
-\includegraphics[width=0.450\textwidth]{plots/plot_tau_gtz.pdf}
-\caption{Same as \reffig{plot_taugtz_PC_vs_tanh}, but superimposed with the cumulative optical depth $\tau(z, \zmax)$ of the best-fit two-step model.
-The two-step model is a toy model that adds a high-$z$ ionization plateau to the standard tanh model with one additional parameter.
-Compared to the cumulative $\tau$ distributions from PCs, the best-fit step-model model is within the 95\% upper limit of the cumulative $\tau$ constraints at all redshifts.
-}
-\label{fig:plot_taugtz_two_step_best_fit}
-\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
-%\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
-The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
-The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
-}
-%\label{fig:two_parameter_model_2D_plot}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
-%\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
-%The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$. } 
-%\label{fig:taulo_prior_test_cl}
-%\end{figure}
-
 \section{Conclusion}

### 2020-12-03 results/notes.tex
Modified notes; changed section order in paper

@@
 
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_oversampled_1D_smooth_scale_1D_2_num_bins_1D_1600_post_bug_fix.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_oversampled_1D_smooth_scale_1D_2_num_bins_1D_800_post_bug_fix.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_oversampled_1D_smooth_scale_1D_2_num_bins_1D_100_post_bug_fix.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_relike_tanh_tau_dz0p5_oversampled_1D_smooth_scale_1D_2_num_bins_1D_50_post_bug_fix.png}
+\label{fig:}
+\caption{Oversampled tanh tau posterior after fixing bug and with even more samples than previously.
+Changing num\_bins = 1600, 800, 100, 50 for the Fortran/cosmomc KDE w/ f = 0.14, which entirely recovers (for num\_bins = 100) close to the python KDE line w/ the same f which is computed as a an evaluation of the effective likelihood at 100 $\tau$ values rather than MCMC sampling.
+This shows that the discrepancy seen previously between the two was only due to smoothing applied by getdist on the MCMC samples when very few samples were present in the chains.
+} 
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.7\textwidth]{ cosmomc_kde/pl18_relike_tanh_tau_dz0p5_oversampled_1D_smooth_scale_1D_2_num_bins_1D_100_post_bug_fix_wGauss.png}
+\caption{Same as previous figure for num\_bins = 100, but adding also the posterior obtained from the Gaussian likelihood using directly the covariance and mean of the PC chains.
+}
+\label{fig:}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/plot_ellipse.png}
+\caption{Examining the variation in the ellipse area expected when a mock covariance is multiplied by 1.0, 1.14 and 1.28.
+}
+\label{fig:}
+\end{figure}
+
 \bibliography{rei.bib}

### 2020-12-03 results/notes.tex
Updates from Overleaf

@@
 \begin{figure}
+    \centering
+    \includegraphics[width=1.0\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
+    \caption{Caption}
+    \label{fig:plot_triangle_wGauss}
+\end{figure}
+
+\bibliography{rei.bib}
+
+\begin{figure}
 \includegraphics[width=0.7\textwidth]{ cosmomc_kde/pl18_relike_tanh_tau_dz0p5_oversampled_1D_smooth_scale_1D_2_num_bins_1D_100_post_bug_fix_wGauss.png}
@@
 
-\bibliography{rei.bib}
+\begin{figure}
+
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run1_f0p28_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.4\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run1_f0p28_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run4_f0p18_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+\includegraphics[width=0.4\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run3_f0p10_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+\label{fig:}
+\caption{Top 2x2 panels: Automated getdist smoothing (left column) and hand tuned smoothing (right column) for f = 0.14 and f = 0.28.
+Bottom 2 panels: f = 0.18 and f = 0.1 for the same hand tuned smoothing settings (slightly aggressive binning).
+} 
+\end{figure}
+
+%\begin{figure}
+%    \centering
+%    \includegraphics{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0393_tri.png}
+%    \caption{$\taulo$ prior imposed at 0.0393, giving a straight cut in the 2D constraint, rather than the rounded zre = 6.1 prior.}
+%    \label{fig:my_label}
+%\end{figure}
 

### 2020-12-03 paper/rei.tex

@@
 \begin{figure*}
-\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
+%\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
+\includegraphics[width=1.0\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
+
 \caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$.
@@
 \label{fig:plot_taugtz_two_step_best_fit}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.40\textwidth]{plots/pl18_tanh_highz_kde_f0p14_tau_z_to_zstart_plot_tau_gtz_dz0p1.pdf}
+\caption{
+}
+\label{fig:plot_taugtz_two_step_contours}
 \end{figure}

### 2020-12-05 paper/rei.tex

@@
 %\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
-\includegraphics[width=1.0\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
+\includegraphics[width=0.8\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
 
@@
 }
-\label{fig:tanh_}
+\label{fig:tanh}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
+\caption{Tanh chains with plik\_full\_TTTEEE for the high-l likelihood. [remove PC zmax curve; add PC filtered version]
+}
+\label{fig:plot_taugtz_PC_vs_tanh}
 \end{figure}
@@
 
-\begin{figure}[t]
-\includegraphics[width=0.450\textwidth]{plots/plot_tau_gtz.pdf}
-\caption{Same as \reffig{plot_taugtz_PC_vs_tanh}, but superimposed with the cumulative optical depth $\tau(z, \zmax)$ of the best-fit two-step model.
+%PC w/ ML model from 2parameter model (tauhi>0.02)
+%\begin{figure}[t]
+%\includegraphics[width=0.450\textwidth]{plots/plot_tau_gtz.pdf}
+%\caption{Same as \reffig{plot_taugtz_PC_vs_tanh}, but superimposed with the cumulative optical depth $\tau(z, \zmax)$ of the best-fit two-step model.
 The two-step model is a toy model that adds a high-$z$ ionization plateau to the standard tanh model with one additional parameter.
-Compared to the cumulative $\tau$ distributions from PCs, the best-fit step-model model is within the 95\% upper limit of the cumulative $\tau$ constraints at all redshifts.
-}
-\label{fig:plot_taugtz_two_step_best_fit}
-\end{figure}
+Compared to the cumulative $\tau$ distributions from PCs, the best-fit two-step model is within the 95\% upper limit of the cumulative $\tau$ constraints at all redshifts.
+%}
+%\label{fig:plot_taugtz_two_step_best_fit}
+%\end{figure}
 
@@
 \includegraphics[width=0.40\textwidth]{plots/pl18_tanh_highz_kde_f0p14_tau_z_to_zstart_plot_tau_gtz_dz0p1.pdf}
-\caption{
+\caption{The 68\% and 95\% C.L.
+constraints on the cumulative optical depth $\tau(z, \zmax)$ for the two-step model, which allows for high-redshift ionization up to $\zmax =30$.
+note that a physical model with $\tau(15, \zmax) = 0.02$ is still allowed at the 95\% C.L.
+whereas the Planck FlexKnot analysis would have ruled it out with their $\tau(15, \zmax) < 0.007$ 95\% upper limit.
+[add not PC filtered version]
 }
@@
 \label{fig:plot_taugtz_lowE_vs_srollv2}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
-\caption{PC zmax = 30 vs tanh chains with plik\_full\_TTTEEE for the high-l likelihood.
-}
-\label{fig:plot_taugtz_PC_vs_tanh}
 \end{figure}

### 2020-12-24 paper/rei.tex
Added text after restructuring

@@
 \label{sec:KDE}
-\wh{The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data.}  In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data.
+In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
 The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
@@
 
-\subsection{\wh{Cumulative Optical Depth}}
+An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
+%
+\begin{equation}
+ {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
+ \label{eq:gaussian}
+ \end{equation}
+%
+We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.
 
-\wh{Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth}
+\subsection{Cumulative Optical Depth}
+
+Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth
 \begin{equation}
@@
 Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
-\wh{decide whether to introduce models one-by-one or together - we could postpone the step model discussion until later without interrupting the flow}
-For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
- \begin{equation}
-x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
- \label{eqn:tanh}
- \end{equation}
- where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
- excludes finite ionization above its transition redshift by definition.
-
+%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
+% \begin{equation}
+%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+% \label{eqn:tanh}
+% \end{equation}
+% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
+% excludes finite ionization above its transition redshift by definition.
 The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.
-\wh{The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
-unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.}
+The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
+unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.
 Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
@@
 Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
-\wh{In other words, no physical model is excluded by these priors.}
+In other words, no physical model is excluded by these priors.
 % I think we don't impose the prior at all
 % for KDE 
-%
 %They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
@@
 
-\wh{Think about whether the refutation of Marius' prior and should go here.  If not then we probably want a forward pointing sentence here about impact of priors in practice.}
+\wh{want a forward pointing sentence here about impact of priors in practice.}
 
@@
 
-$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & -0.208 & -0.094 & 0.067 \\
-$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & -0.273 & 0.142 \\
-$m_3$ & $$0.078 & 0.098 & -0.208 & 0.048 & 1.000 & 0.064 & -0.018 \\
-$m_4$ & -0.077 & 0.131 & -0.094 & -0.273 & 0.064 & 1.000 & -0.195 \\
-$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & -0.018 & -0.195 & 1.000 \\
+$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
+$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
+$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$0.018 \\
+$m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
+$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\
 
@@
 [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
-In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to an even simpler effective likelihood, which approximates the $m_a$ posterior as a multivariate Gaussian with mean $\bar{\bf m}$ and covariance $\bf C$ (also shown in Table~\ref{tab:PC_stats}.
+In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to the simpler Gaussian likelihood of Eq.~\ref{eq:gaussian}.
+The exact values of the PC mean and $\bar{\bf m}$ and covariance $\bf C$ are shown in Table~\ref{tab:PC_stats}.
 
-\begin{equation}
- {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
- \label{eq:gaussian}
- \end{equation}
  This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement.
-The Gaussian likelihood is even faster than the kernel density estimate, and may be used to sample well-constrained models near the peak of the PC distributions, while becoming possibly less accurate near the tails of the distributions.
+The Gaussian likelihood is even faster than the kernel density estimate.
+As shall see in the next section for two explicit examples, the Gaussian likelihood shows generally good agreement with the full KDE result, but is unable to capture, as expected given its symmetric nature, any skewness in the PC distribution that may affect parameter constraints.
 
@@
 The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
-\wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}  [Might be nice to overlay the tanh trajectories, no 1 sigma, 2 sigma business].
-
-\wh{I wonder if the order should be reversed and talk about the model testing and KDE first, then the model independent cumulative optical depth?
-That way the model stuff could all be introduced here and simulaneously compared with the fiducal results?
-The flow could be that we first show more high z optical depth is allowed in the two step model but not FlexKnot and then talk about it being a model independent PC result, concluding that we are more robust than FlexKnot?
-Of course I'm assuming that the 2 step model shows this...}
+\wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}.
 
@@
 As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$.
-This code is downloadable on Github at \url{} [give URL].
-The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_ TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.
+This code is available on GitHub at \url{} [give URL].
+The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.
 
@@
 \begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+\includegraphics[width=0.40\textwidth]
+%{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
+{plots/plot_tanh_gaussian_MCMC.png}
 \caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?].
@@
 The steplike model used as the standard approach in CAMB describes the hydrogen and singly ionized helium reionization as a tanh function:
-
+%
  \begin{equation}
@@
  \end{equation}
-
- with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$. Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
+ %
+ with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$.
+Instead of using $\Delta z = 0.5$ as in the standard setting by camb, we adopt a smaller width $\Delta z = 0.015(1+z)$.
+Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
  \beq
@@
 
-In \reffig{tanh_1D} we show the posterior distribution of the $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$.
+In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$.
+Because the PC method cannot accurately represent models that are not fully ionized below $z = 6$, we apply a prior $\zre < 6.1$ to both chains given the transition width of about $0.015(1+z) \approx 0.1$ at $z = 6$.
 The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
-We expect the smoothing to be at roughly the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
+We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
+Indeed, the KDE result $\tau = 0.0588 \pm 0.0066$ has a negligibly lower mean ($0.05\sigma$) and a 6.5\% larger standard deviation (as expected) than the full likelihood result $\tau = 0.0591 \pm 0.0062$;.
 
-\begin{itemize}
-    \item (here?) Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.
-    \item{Might also want to address this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
-Then transition to the following point}
-\end{itemize}
+Instead of the effective likelihood, we can also use its Gaussian approximation shown in Eq.~\ref{eq:gaussian}.
+The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain.
+[add timing here]
+%and amounts to one evaluation of the multi-variate Gaussian function per sample in the reionization model parameter space. 
+We obtain for the tanh model $tau = 0.0594 \pm 0.0063$.
+While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two.
+This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution.
+As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ if we consider the 68\% lower and upper limits, whereas the full KDE result $\tau = 0.0588_{-0.0072}^{+0.0060}$ is able to capture the skewness in the distribution $\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.
 
- %We take here $z_{\rm re}= 9.85$, corresponding the chain maximum likelihood (ML) model ($\tau = 0.0765$) from \S \ref{sec:MCMC},  for illustrative purposes.
-Projected onto 5 PCs and resummed into $x_e(z)$, Eq.~(\ref{eq:mmutoxe}) yields a poor reconstruction of the ionization history itself.
- %Nonetheless as we shall see in Fig.~\ref{fig:clee}, the PC decomposition provides an excellent representation of the polarization power spectrum.
+\ch{[double check all these numbers again, write script to output in a table automatically]}.
+
+[Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.]
+
+[Note sure if it is related to this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
+Then transition to the following point]
+
+We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integration the exact tanh model in the full likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains.
+[describe more once plot is made].
+From these constraints it is also clear that the tanh model does not allow for high redshift ionization by construction, while the PC chains sampling the entire space of physical models would still allow for some high-$z$ ionization in the Planck 2018 data, as we shall see later in in Section~\ref{sec:results}.
 
@@
  \eea
-where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ as in the first example, but now we choose $\Delta z_1 = 0.015\, (1+z_{\rm re})$ instead of the usual $\Delta z_1 = 0.5$, to provide sharper distinctions between the two steps.
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{\rm re})$ just as in the first example for a sharper step.
 We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
-We recover the standard tanh model in the limit when $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
-Therefore this model adds a single parameter $\xemin$ to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$. We show an example
+In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$.
+We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
+We show an example
 of this two-step model in Fig.~\ref{fig:two_step_model}.
@@
 
-To obtain posterior distributions with flat $\tau$ priors, we convert the model to be parameterized by low and high-redshift optical depth, $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$, 
-each defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ respectively.
-We choose the split to be $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
-We sample in the $(\taulo, \tauhi)$ parameter space and the corresponding ($\zre, \xemin$) is found through an iterative search similar to that used for finding $\zre$ given total $\tau$ in the canonical tanh model.
-The priors are chosen to be $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where $\tau_{\rm max} = 0.35$, and $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth for both the effective and full likelihood runs,
-except that we have chosen $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. 
-Note that $\taulo \lesssim 0.038$ roughly corresponds to $\zre < 6$ in the canonical tanh model.
-For the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
-So to make the comparison apples-to-apples, we further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the effective reionization likelihood chains and the full likelihood chains.
+In practice, we sample the model in  $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$ space to obtain posterior distributions with flat $\tau$ priors.
+They are defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
+[double check this too.]
+Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given total optical depth $\tau$.
+For both the full likelihood and effective likelihood runs, we adopt the priors $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$.
+%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
+We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains. Here we used again $f = 0.14$ in the KDE computation.
 
-In ~\reffig{two_parameter_model_2D}, we compare the posteriors of $\taulo$ and $\tauhi$ obtained with these two likelihood codes, which show good agreement.
-Again we used $f = 0.14$ here for the KDE computation.
-[add timing results here?] Note that the truncation of the 2D constraint ellipses on the lower end of $\taulo$ is curved cut rather than a straight line, because $\zre$ cut we performed corresponds to a different value of $\taulo$ for a different $\tauhi$.
+In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.
+They agree very well.
+Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
+This is why the cut on the 2D constraint ellipses at low $\taulo$ is of a curved shape rather than a straight cut, because the $\zre$ prior corresponds to a different value of $\taulo$ for a different $\tauhi$.
 
-The marginalized 1D constraints are $\taulo = XX \pm XX$, $\tauhi = XX \pm XX$ for both runs.
+[add timing results here?]
+
+[comment on effective likelihood w/ KDE vs Gaussian likelihood?]
+
+The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$.
 The 95\% C.L.
-upper limit on $\tauhi$ is $\tauhi < XX$ is consistent with the $\tau(15, 30) < XX$ results we got from the PC constraints.
-[is it?
-double check] This model of course do not capture the larger space of physical models with high-$z$ ionization like the FlexKnot or PC methods would, but consists of a subset of models for which the high-redshift optical depth results must still be consistent with the constraints from these more general methods.
+upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
 
-To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}.
+The Planck 2018 paper~\cite{Aghanim:2018eyx} also constrained $\tau(15, 30)$ in a model-independent way using the FlexKnot method, in which $x_e(z_i)$ at the $i$-th knots are varied as free parameters as well as the number of total knots.
+They obtained $\tau(15, 30) < 0.007$ (95\% C.L.), which is not consistent with the results we obtain for the two-parameter models.
+In principle, the FlexKnot method allows for all possible physical ionization histories that increase monotonically with redshifts, and would include the two-parameter model described here.
+But it seems that the FlexKnot method gave more stringent upper limits on high-redshift optical depth in the analysis of Ref.~\cite{} than actually allowed by the data.
+
+In Ref.~\cite{Aghanim:2018eyx}, the authors also suggested that the PC method has limitations due to the fact that it includes some unphysical models in addition to all the physical models.
+The authors gave this reason for not using PCs to constrain reionization histories.
+We would like to argue that this claim is not well founded.
+
+In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this 2-parameter model.
+In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ integrated over the exact physical models in the full likelihood chains themselves, as well as those obtained from the 5-PC representation of those physical models in the effective likelihood chains.
+Because the truncated 5 PCs may represent non-physical models (higher PCs can restore the physicality of the model), negative $\tau$ values are allowed at some redshifts, which is of course, unphysical.
+Nevertheless, the full vs truncated PC chains give similar 95\% upper limits on the high-redshift optical depth, $\tau(15, 30) < XX$ and $\tau(15, 30) < XX$ respectively.
+We have showed that the discrepancy between the PC-based effective likelihood results and the FlexKnot results are not due to the inclusion of non-physical models.
+
+With this in mind, we will now turn to the model-independent constraints derived from the PC analysis.
+
+%\ch{Could also mention best-fit model below as well.}
+%To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}.
 Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts.
-According to the FlexKnots constraints in Ref.~\cite{}, however, this model would have been ruled out given that $\tau(15, 30) < 0.007$ (95\% C.L.) with the FlexKnots results.
-[provide comment ...; any further discussion on literature results with FlexKnots?]
 
-[add a conclusive paragraph or sentence]
-
-%Plot: add plot for $\tau(>z)$ contours for the 2 parameter model vs PC; direct integration first?
+[add plot for $\tau(>z)$ contours for the 2 parameter model for direct MCMC vs KDE chains]
 
@@
 
-\section{Implications for optical depth}
+\section{Model-Independent Results for Optical Depth}
 
-\subsection{Cumulative optical depth}
+So far no priors on the PC chains have been used.
+In constructing the kernel density estimate of Section~\ref{sec:effective_likelihood} we used PC chains without priors; and when using the effective likelihood for model testing, it would always be restricted to a particular set of physical models like in the two examples we showed above.
+To derive model-independent constraints in this section, however, we will now impose the physicality priors of Eq.~\ref{eq:individualprior}.
 
-We continue the comparison in the space of cumulative optical depth $\tau(z, \zmax)$, which is a better quantity for visualizing and comparing PC results than the ionization history $x_e$ itself because of it is the quantity most closely related to observables.
-Recall that these truncated PCs provide only accurate descriptions of the models in terms of their \textit{observable} impact, and are not tools for reconstructing the precise ionization history.
+Before we show the cumulative optical depth constraints from PCs, let us first make a note about the impact of the physicality prior chosen for this analysis and clarify some wrong claims against it in the literature.
 
-In Fig.~\ref{fig:plot_taugtz_2015_vs_2018}, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines).
-...
+\subsection{A note on priors}
+\label{sec:note_on_priors}
 
-In Fig.~\ref{fig:plot_taugtz_lowE_vs_srollv2}, we compare the Planck 2018 results with the two different low-$\ell$ $EE$ likelihoods. [Do we need this?]
+As mentioned in Section~\ref{sec:background}, the priors of Eq.~\ref{eq:individualprior} are necessary but not sufficient conditions for physicality.
+Therefore, while retaining all physical models, they also includes some unphysical models.
+The authors of Ref.~\cite{Millea:2018bko} (hereafter MB18) raised a concern that these PC priors implies a non-flat prior in $\tau$ and claimed that it biases the PC result of  Ref.~\cite{Heinrich:2016ojb} and accounts for most of the shift between the tanh and PC $\tau$ posteriors.
+In Ref.~\cite{Heinrich:2018btc}, we showed that the use of our physicality prior does not cause bias for data that is constraining, which includes the Planck 2015 dataset.
+On the other hand, we showed in Ref.~\cite{Heinrich:2016ojb} that if one adopts the method proposed in Ref.~\cite{Millea:2018bko}.
+in an attempt to flatten the $\tau$ prior (by multiplying its inverse at each point in the multidimensional PC space), one would actually end up with a biased $\tau$ constraint for the Planck 2015 data.
 
-In Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}, we compare the PC results against the steplike model.
-It is clear that the tanh model does not allow for high redshift ionization by construction.
-The total optical depth allowed by the PC chains sampling the entire physical model space is actually slightly higher due to the contributions from these higher redshifts.
-[do we want PC version with the physicality priors actually?]\\
+We briefly recap the argument below.
+Let us use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily.
+We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
+\beq
+\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
+\eeq
+where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
+Because most of the contributions to $\tau$ comes from the first two components (see also Fig.
+7 of Ref.~\cite{Heinrich:2018btc} in the Appendix), we can approximate it as
+\beq
+\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
+\eeq
 
-We show however, using a separate PC chains with $\zmax = 50$, that there is no evidence for optical depth at higher redshifts than 30.
-In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the contours in $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines), and that their total optical depth constraints are practically identical:
+In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints (ellipse) in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
+The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior.
+The physicality prior upon projection would correspond to a linearly rising prior in $\tau_{12}$ over the region where the posterior has support.
+One would be eager to conclude that this produces a biased constraint on $\tau_{12}$.
+However, because the Planck data is more constraining than the prior in $m_1$ and $m_2$, (the ellipse occupying a much smaller area in one part of the prior box), the value of the prior at high $\tau_{12}$ becomes irrelevant for the posterior.
+However, if one tries to mitigate for the $\tau_{12}$ prior by multiplying the inverse prior value at each point in the PC space over the area where data is more constraining than the prior, then one would bias the posterior toward lower $\tau_{12}$ in this case.
+
+What is then the true impact of this physicality prior we have imposed.
+We have established that the orientation of the prior box in the $m_1-m_2$ plane does not matter for this dataset since the data is more constraining than prior.
+However, the location of the data constraint does matter, as the prior box currently cuts off half of the ellipse in the Planck 2018 data (a larger amount than in Planck 2015).
+The consequence of this cut in the $\tau$ constraints essentially amounts to excluding models not fully ionized models by $\zre \sim 6$, as consistent with observational constraints.
+
+The other impact of using the physicality prior is that some unphysical models are included inside the prior.
+We reiterate that we do not adopt this prior for the KDE and unphysical models do not impact results when actually testing physical models using the effective likelihood code.
+We will show in the next section that their impact on the model-independent $\tau$ results are also minimal.
+
+[add a sentence to reiterate that the shift between PC and steplike model seen in the Planck 2015 is not due to the use of physicality prior? and some other Marius/Planck paper points?]
+
+[restore some text about evolution of planck data and HFI systematics causing PC-tanh shifts in Planck 2015 data?
+And address the claim: ``In the Planck 2018 paper on cosmological parameters, the reduction of optical depth from the PC results seen in the Planck 2018 data compared to the Planck 2015 data was attributed to the reduction of HFI systematics, with subdominant contribution from prior."]
+%They also proposed a method to flatten the $\tau$ prior by multiplying its inverse at each point in the multidimensional PC space.
+The authors claim that adopting the physicality priors in Ref.~\cite{Heinrich:2016ojb} contributed to a large fraction of the shift in the $\tau$ posterior between the tanh and the PC results in the Planck 2015 data.
+
+ \begin{figure}
+          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_plot_rotated_box_flat_tau_prior_fac_0.8.png}
+          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
+          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) which are aligned with lines of constant $\tau_{12}$ (light gray lines).
+Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
+The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.}
+          \label{fig:prior_box}
+\end{figure}
+
+\subsection{Cumulative optical depth constraints}
+
+Let us now constrain the cumulative optical depth $\tau(z, \zmax)$ using the PC results.
+The optical depth is a better quantity than the ionization history $x_e$ for visualizing and comparing PC results because it is much more closely related to observables.
+Recall that these truncated PCs provide an accurate description of the models only in terms of their \textit{observable} impact, and are not meant to be used to reconstruct the exact ionization history.
+
+In Fig.~\ref{fig:plot_taugtz_2015_vs_2018} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines).
+In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.
+
+% with this prior you still have that $\tau(>z)$ plot includes unphysical models (however, most likely total tau doesn't change, and if any bias from unphysical models, it would lower high-z tau).\\
+
+We also show using a separate PC chains with $\zmax = 50$ that there is no evidence for optical depth at $z>30$. 
+In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that 68\% and 95\% C.L.
+contours of $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines).
+Furthermore, their total optical depth constraints are practically identical:
 \beq
@@
 \eeq
-These total optical depth constraints have slightly higher central values and smaller error bars compared to those obtained using the original Planck-released \textsc{lowE} likelihood
+
+Compared the \textsc{lowE} likelihood results, 
 \beq
@@
 \eeq
-[more comments here]
+the \textsc{srollv2} error bars on PC $\tau$ was reduced by about 20\%, and the central values rose by about half a sigma. [wanna comment more here?]
 %
 
-The upper limit on $\tau(15, \zmax)$ are 
+For the high-redshift optical depth, we also obtain consistent upper limits
 \beq
@@
 \eeq
-[double check on full chain (1015) this is from 0930]
+[double check on full chain (from 10/15) this is from 09/30.]
 and 
@@
 \includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
-\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015).
+\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015). [To be created and merged with the next plot as a single figure with two panels.]
 }
@@
 
-\subsection{Comparison with Previous Results}
+\subsection{Comparison to previous results}
 
@@
 
-In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for a steplike model, $\tau = 0.0504^{+0.0050}
-_{-0.0079}$ for the FlexKnot method and $\tau = 0.0487^{+0.0038}_{−0.0081}$ for the PC method with a different prior than that used in this paper.
+In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for the tanh model, $\tau = 0.0504^{+0.0050}
+_{-0.0079}$ for the FlexKnot method (with flat $\tau$ prior) and $\tau = 0.0487^{+0.0038}_{−0.0081}$ for the PC method but using the prior inversion procedure discussed Section~\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
+[do we want to check whether the prior inversion procedure could cause a shift at this level?
+do we want to evaluate this statement?]
 
@@
 
-Here we show that we find a less stringent upper limit using the PC method $\tau(15, \zmax)_{\rm PC,\, \zmax=30} < XX \; (95\% \mathrm{CL})$. [explain why FlexKnot could give more stringent results.]
+Compared to our PC results ($\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\% \mathrm{CL})$), the FlexKnot upper limits are too stringent.
+The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in Section~\ref{sec:example2} with a concrete example of the two-step model.
+We showed that for this chosen physical model, the $\tau(15, 30)$ upper limits derived from the full Planck likelihood without using any PCs were already not compatible with the FlexKnot upper limits, while the PC results were compatible.
+So the discrepancy cannot be attributed to the prior choice in the PC results.
+It also seems that the inclusion of unphysical model would in any case shift the optical depth at high-redshift toward more negative values, which would decrease rather than increase the upper limit.
 
-\subsection{A note on priors}
+[just noticed that our tau(15, 30) $<$ 0.023 result is with srollv2, and flexknot tau(15,30) $<$ 0.007 is with lowE likelihood.
+In our analysis lowE to sroll2 had a 0.004 increase in total tau.
+I should run the tau(15, 30) constraint on the lowE chain as well just to double check.]
 
-In Ref.~\cite{Millea:2018bko} (hereafter MB18)  a concern was raised that the physicality prior we take for the PC analysis in Ref.~\cite{Heinrich:2016ojb} implies a non-flat prior in $\tau$ and introduces bias to the PC result.
-The authors also proposed a method to flatten the $\tau$ prior by multiplying its inverse at each point in the multidimensional PC space.
-They claim that adopting the physicality prior used here and in Ref.~\cite{Heinrich:2016ojb} contributed to the a large fraction of the shift in the $\tau$ posterior between the steplike and PC method in the Planck 2015 data.
+[Need to address Planck2018 cosmology paper claim: ``The PCA result is slightly lower, and is partly affected by the imperfect physicality priors that allow unphysical negative ionization fractions."]
 
-In the Planck 2018 paper on cosmological parameters, the reduction of optical depth from the PC results seen in the Planck 2018 data compared to the Planck 2015 data was attributed to the reduction of HFI systematics, with subdominant contribution from priors.
-We therefore would like to clarify the impacts of the priors we choose and illustrate explicitly that there is no bias with this choice, whereas the flattening method proposed in MB18 is actually at risk of creating a non-flat prior and biasing results.
-
-In the following, we will use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily.
-We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
-\beq
-\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
-\eeq
-where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
-Because most of the contributions to $\tau$ comes from the first two components as shown in Fig.
-7 of Ref.~\cite{Heinrich:2018btc} in the appendix, we can approximate it as
-\beq
-\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
-\eeq
-We also show it here in Fig.~\ref{fig:tau12} where the posterior of $\tau_{12}$ and $\tau$ (with five PCs) are compared for the Planck 2018 data.
-
-In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints (ellipse) in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
-The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior.
-The physicality prior upon projection would correspond to a linearly rising prior in $\tau_{12}$ over the region where the posterior has support.
-One would be eager to conclude that this produces a biased constraint on $\tau_{12}$.
-However, because the Planck data is more constraining than the prior in $m_1$ and $m_2$, (the ellipse occupying a much smaller area in one part of the prior box), the value of the prior at high $\tau_{12}$ becomes irrelevant for the posterior.
-
-Using the procedure proposed by Ref.~\cite{Millea:2018bko}, inverting the prior value at each point in the PC space, would actually up-weight the high $\tau_{12}$ regions that are not supported by data, while down-weighting the regions that are.
-This produces an actual bias in the $\tau_{12}$ posterior toward lower optical depth.
-
-We compare the three priors in Fig.~\ref{fig:tau12_posterior_compare_priors} by showing their corresponding $\tau_{12}$ posteriors.
-Compared to the explicitly flat prior (dashed box in Fig.~\ref{fig:prior_box}) as the ground truth, the posterior from using the physicality prior shifts negligibly.
-The main difference is a slightly reduced posterior values at very low $\tau_{12}$, which is due to the physicality condition and the restriction that hydrogen reionization be completed by $z = 6$.
-This is manifested as the data constraint ellipses being partially outside of the solid prior box in Fig.~\ref{fig:prior_box}.
-A similar effect would be observed for the steplike models with or without the  restrictions that reionization occurs at $z>6$.
-We also show that the results between the two priors are similar for the $\tau(15, \zmax)$ statistics: $\tau_{12}(15, \zmax) > XX $(95\% C.L.) for the flat $\tau_{12}$ prior vs $\tau_{12}(15, \zmax) > XX $ (95\% C.L.) for the physicality prior.
-
-This is a small difference in comparison to the bias introduced by multiplying the posterior with the inverse of the $\tau_{12}$ prior.
-[show in figure?] [add a sentence to reiterate that the shift between PC and steplike model seen in the Planck 2015 is not due to the use of physicality prior].
-Furthermore, the bias caused by using the inverse prior method in the Planck 2018 data is on the order of $\Delta \tau \sim 10^{-X}$, which could partially [or fully] account for the difference between FlexKnots and PC results seen in Ref.~\cite{Aghanim:2018eyx} on the total optical depth.
-
-[Discussion of why FlexKnots is too stringent for $\tau_{12}(15, \zmax)$. Not sure how to do yet. Where should this go? Maybe not prior related.]
-
-We conclude that the prior inversion procedure proposed in Ref.~\cite{Millea:2018bko} should not be used where data is more constraining than prior, which includes the Planck 2015, Planck 2018 and possible future datasets.
-
- \begin{figure}
-          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_plot_rotated_box_flat_tau_prior_fac_0.8.png}
-          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
-          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) which are aligned with lines of constant $\tau_{12}$ (light gray lines).
-Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
-The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.}
-          \label{fig:prior_box}
-\end{figure}
-
-% NEXT:
-
-% - fill in all numbers: tau(15, 30) upper bounds, store in table or spreadsheet in git; covariance matrix (store in git too); other numbers mentioned in the paper.
-
-% - talk w/ Wayne, finalize decisions, and make polished plots as needed.
-
-% - write abstract, find title and give the text another pass to refine.
+[Also, they only used lowE likelihood, fixed all other cosmological parameters (including $A_s e^{-2\tau}$). Wondering if that does anything to the difference between our results]
 

### 2021-01-07 paper/rei.tex

@@
 where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
-We have rank-ordered the PCs from low to high variance, so that for the range $z_{min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.
-\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
+We have rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
 In the work that follows, we therefore truncate the sum at $a = 1 ..
-n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ and $n_{\rm{PC}} = 7$ for $z_{\rm max} = 50$.
+n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis.
+We also test its robustness by taking 
+ for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.
 [Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
@@
 We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
-[rephrase this paragraph:]
-``For the Planck data set, most of the information in the ionization history is carried by the
-first two modes and therefore relates to the amount of high vs.~low redshift optical depth. We keep all 5 PCs  for completeness in representing the observable
-impact of a given ionization history and to marginalize uncertainties that they introduce."
+For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.
+However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.
+
 We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$.
 Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
-For $z<6$, we follow CAMB to assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$.
-\wh{Probably need some brief words about MCMC sampling of the posteriors of $m_a$ jointly with $\Lambda$CDM parameters here so that the KDE section makes sense.
-I moved the optical depth stuff to a following subsection to clarify role, and order of operations, of the physicality prior.}
+For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).
 
-\subsection{Effective Likelihood from PCs -- Kernel Density Estimate}
+We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using COSMOMC\footnote{COSMOMC: \url{http://cosmologist.info/cosmomc}} of the relevant Planck likelihood discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density
+$\Omega_c h^2$, effective angular sound horizon $\theta_{\rm MC}$, scalar power spectrum log amplitude $\ln (10^{10} A_s)$ and
+its tilt $n_s$.  We take flat uninformative
+priors in $m_a$ but discuss additional priors imposed by
+removing demonstrably  unphysical ionization histories in \S \ref{sec:cumulative}.
+
+\subsection{Effective PC Likelihood}
 \label{sec:KDE}
@@
 The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
-Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~\ref{eq:xetommu}.
-Since $\mathbf{m}$ could take any continuous value, we approximate its effective likelihood with a kernel density estimate
+Given any physical ionization history $x_e(z)$, we first obtain its PC representation $\mathbf{m}$ using Eq.~(\ref{eq:xetommu}).
+Since $\mathbf{m}$ can take any continuous value, we approximate its effective likelihood with a kernel density estimate
 \beq
@@
 % 1.072 no cut
-For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient. Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
+For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
+%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors. 
 Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
 \begin{equation}
-P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
+P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
 \end{equation}
-we demonstrate a successful recovery of the posterior distributions from direct MCMC analyses (using the Planck likelihood and varying all cosmological parameters) with significantly less computational time.
-We show these results in sections~\ref{sec:example1} and~\ref{sec:example2} for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
 
-An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
+This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
+For comparison, an even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
 %
@@
 %
-We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.
+In section~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
+for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
+
+%
+%
+%
+%We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.
 
 \subsection{Cumulative Optical Depth}
+\label{sec:cumulative}
 
-Although reionization PCs are mainly a tool for model testing, as they do not reconstruct the ionization history $x_e(z)$ itself, they do provide model-independent constraints on  the cumulative Thomson optical depth
+Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
 \begin{equation}
@@
 % excludes finite ionization above its transition redshift by definition.
-The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no prior constraints on the mode amplitudes $m_a$ are imposed.
-The one subtlety when placing constraints on the cumulative optical depth is that the analysis also therfore allows
-unphysical ionization fractions where $0<x_e<x_e^{\rm max}$  is not satisfied.
+The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed.
+The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
+unphysical ionization fractions where $0 \le x_e \le x_e^{\rm max}$  is not satisfied.
 Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
@@
 \end{equation}
+\wh{comment on the sum prior?  we don't seem to talk about it later - is it applied and is it always weaker in practice?}
+\ch{I think it is always weaker, let me double check.}
 %
-Note that the original $0<x_e<x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
-In other words, no physical model is excluded by these priors.
-% I think we don't impose the prior at all
-% for KDE 
-%They are however noninformative priors when the PC constraints are being used for evaluating the likelihood of a series of physical models, which we will now describe in the next section.
-
-%we employ the smoothed $x_e$ which formally has support beyond the bounds. 
-% We include this small correction by integrating
-% slightly past $z_{\rm max}$ in practice.  Here $n_{\rm H}(0)$ is the hydrogen number density at $z=0$.
-   %Consequently we smooth the ionization history in Eq.~(\ref{eq:mmutoxe}) with a 
-%Gaussian in $\mathrm{ln}(1+z)$ of width $\sigma_{\mathrm{ln}(1+z)} = 0.015$.
-
-\wh{want a forward pointing sentence here about impact of priors in practice.}
+Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
+In other words, no physical model is excluded by these priors but some unphysical models are included.
+In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against direct constraints for two example models.
 
@@
 For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{srollv2} likelihood [add citation] in our official PC results and the effective likelihood code.
-In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood had improved foreground cleaning, which enabled a factor of XX improvement in $\tau$ just in the steplike model [cite].
+In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood improves foreground cleaning,
+which enables stronger reionization constraints, especially for the low multipole moments that are important for the low redshift constraints.
+
+% can add a comment about impact 
+%d a factor of XX improvement in $\tau$ just in the steplike model [cite].
 
@@
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
-The box boundaries correspond to the physicality priors shown in Eq.~\ref{eq:individualprior} and include all physical models along with some unphysical models.
+The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional
+physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior.
 Results using the Planck 2018 data are shown in blue.
-While all five PC amplitudes are being constrained by Planck, the first two are particularly well-constrained.
+While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained.
 Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.
-The trajectories shown in black lines are the steplike models of instantaneous reionization allowed by the Planck 2015 data with the arrow pointing in the direction of increasing optical depth.
+\wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
+We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
+In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}
+
+In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show
+the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
+from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher \ch{[lower?]} total optical depth.
+\ch{[Or maybe you meant Planck 2015 to simall EE?]}
+
+\wh{possibly point to a figure below with $\tau_{12}$ in}
+
+%\wh{possibly move the discussion of trajectories into the model section}
+%The trajectories shown in black lines are the steplike models of instantaneous reionization allowed by the Planck 2015 data with the arrow pointing in the direction of increasing optical depth.
 Clearly, these steplikes models are only a part of the allowed physical model space.
@@
 \caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper.
+\wh{some of these figure versions look like low res bit maps - final versions should be vector graphics and look uniform in style between figures.} \ch{got it, will polish for sure.}
 }
@@
 [Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
-In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show using the PC chain covariance $\bf C$ (black dashed lines) what corresponds to the simpler Gaussian likelihood of Eq.~\ref{eq:gaussian}.
-The exact values of the PC mean and $\bar{\bf m}$ and covariance $\bf C$ are shown in Table~\ref{tab:PC_stats}.
 
- This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement.
+In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.
+The visual agreement with the full posteriors is remarkably good, especially in the well constrained components.
+This suggests that the simple Gaussian  likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models.
+However we shall see that a quantitative comparison shows that there is enough skewness in the posteriors that the KDE likelihood
+still performs noticeably better.  \wh{revisit this 
+statement once we firm up the comparison, e.g. what is noticeably better, means, 95\% confidence intervals, far tails?} \ch{yup, will do.}
+
+% This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement.
 The Gaussian likelihood is even faster than the kernel density estimate.
@@
 
-In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}.
+%In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}.
 The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
@@
 
-\section{Effective Likelihood}
+\section{Model Testing with PCs}
 \label{sec:effective_likelihood}
 
-As described in ~\refsec{KDE}, we turn the PC constraints from the Planck 2018 into an effective likelihood code to constrain arbitrary models of ionization history between $6 \leq z \leq \zmax$.
-This code is available on GitHub at \url{} [give URL].
-The initial release is offered for the Planck 2018 $\textsc{plik\_lite\_TTTTTE} + \textsc{lowl} + \textsc{srollv2}$ likelihood combination with $\zmax = 30$.
+We now turn to constraining reionization models with PCs using the
+effective likelihood approach described in~\refsec{KDE}.  The benefit of this approach is that
+complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ can now be obtained without reanalyzing the Planck data. 
+This effective likelihood code is available on GitHub at \url{} [give URL].
+The initial release corresponds to our fiducial constraints derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihood combination with $\zmax = 30$.
 
-To obtain posteriors of model parameters from the corresponding Planck likelihoods, simply specify the functional form for $x_e(z)$ in the code, and run a MCMC analysis using the extension we provide to a MCMC sampler (currently supporting CosmoMC).
-Note that the likelihood cannot distinguish the effects of the model parameters that impact $z \sim 6$, where the Universe is assumed to be fully ionized.
-Furthermore, transitions faster than $\Delta z \sim 0.1$ between $6 < z < 6.25$ may not be fully captured because of how PC functions are discretized with $\delta z = 0.25$ and are linearly interpolated within this bin size.
-[review this statement] To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
-Finally, you may choose to run the chains at a different cosmological model than that of the default Planck 2018 best-fit model, provided it is close enough.
+To use this approach, one simply specifies the functional form for $x_e(z)$ in the code, which then is converted to the PC parameters $m_a$ for the effective likelihood evaluation assuming a \ch{fiducial} $\Lambda$CDM cosmology.
+To obtain reionization model parameter posteriors, one samples this likelihood with any desired priors.   
+We have also provided an MCMC sampler for models of high dimension (currently in CosmoMC). Note that the model parameters or priors must be arranged to explicitly satisfy 
+full ionization below $z=6$. To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
 
-We demonstrate successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has an one additional parameter to model the high-$z$ ionization as a plateau in $x_e$ before the transition to full ionization happens in the standard tanh model.
+Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has one additional parameter modeling a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.
 
@@
  with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$.
-Instead of using $\Delta z = 0.5$ as in the standard setting by camb, we adopt a smaller width $\Delta z = 0.015(1+z)$.
+Instead of using $\Delta z = 0.5$ as in the standard setting by camb, we adopt a smaller width $\Delta z = 0.015(1+z)$ so as to better capture the required full ionization below $z=6$.
 Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
@@
 
-In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which all six $\Lambda$CDM parameters were varied including $\tau$.
-Because the PC method cannot accurately represent models that are not fully ionized below $z = 6$, we apply a prior $\zre < 6.1$ to both chains given the transition width of about $0.015(1+z) \approx 0.1$ at $z = 6$.
+The $\tanh$ model is parameterized by its total optical depth $\tau$ for which we take a flat prior 
+with an additional constraint that $z_{\rm re}\ge 6.1$ so as to better satisfy the required $z_{\rm min}=6$ given $\Delta z$.
+In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the trajectory that the $\tanh$ model takes in the PC space.
+Notice that this trajectory passes near the center of all constraints for Planck 2018 but only through the tails for some of the Planck 2015 constraints.
+Correspondingly, whereas the $\tanh$ model was marginally disfavored with Planck 2015 data that is not the case with Planck 2018 data.
+
+In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which the five other $\Lambda$CDM parameters were also varied using the same priors for the $\tanh$ model.
+%including $\tau$. 
+%Because the PC method cannot accurately represent models that are not fully ionized below $z = 6$, we apply a prior $\zre < 6.1$ to both chains given the transition width of about $0.015(1+z) \approx 0.1$ at $z = 6$.
 The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
 We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
-Indeed, the KDE result $\tau = 0.0588 \pm 0.0066$ has a negligibly lower mean ($0.05\sigma$) and a 6.5\% larger standard deviation (as expected) than the full likelihood result $\tau = 0.0591 \pm 0.0062$;.
+Indeed, the KDE result $\tau = 0.0588 \pm 0.0066$ has a negligibly lower mean ($0.05\sigma$) and a 6.5\% larger standard deviation (as expected) than the full likelihood result $\tau = 0.0591 \pm 0.0062$.
 
-Instead of the effective likelihood, we can also use its Gaussian approximation shown in Eq.~\ref{eq:gaussian}.
+Instead of the effective likelihood, we can alternately use its Gaussian approximation shown in Eq.~(\ref{eq:gaussian}).
 The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain.
@@
 %and amounts to one evaluation of the multi-variate Gaussian function per sample in the reionization model parameter space. 
-We obtain for the tanh model $tau = 0.0594 \pm 0.0063$.
+We obtain for the tanh model $\tau = 0.0594 \pm 0.0063$.
 While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two.
@@
 
-In practice, we sample the model in  $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$ space to obtain posterior distributions with flat $\tau$ priors.
-They are defined as in Eq.~\refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
+In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.
+They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
 [double check this too.]
@@
 
-In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.
-They agree very well.
+In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$. They agree very well. 
+\wh{Not sure if we need this comment - was more important when the width was larger - also not sure if this is the explanation for the visible effect - getdist is smoothing things and that might be more important.}
 Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
@@
 The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$.
-The 95\% C.L.
-upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
+\wh{will be interesting to see if the numbers are more consistent with the $f$ smoothing than the visual plot is - that might help us see if getdist is responsible for extra smoothing that hides things.}
+The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
 
@@
 
-So far no priors on the PC chains have been used.
+\wh{
+While the PC approach to constraining reionization should mainly be used for 
+testing wide classes of models efficiently and with their own physically motivated priors, it can also be employed to extract model-independent constraints on the total and cumulative optical depth.
+As discussed in \S \ref{sec:cumulative}, since the PC amplitudes are not intrinisically required
+to produce physically sensible ionization histories where $0 \le x_e(z) \le x_e^{\rm max}$, this requires imposing physicality priors using Eq.~\ref{eq:individualprior}.
+In addition these flat range-bound priors on the PC amplitudes $m_a$, imply implicit priors on the optical depth.
+Finally our fiducial analysis has an explicit prior that reionization occured
+at $z<z_{\rm max}=30$.   In this section we present these model-independent results and discuss their robustness to these explicit and implicit priors.
+}
+
+%So far no priors on the PC chains have been used.
 In constructing the kernel density estimate of Section~\ref{sec:effective_likelihood} we used PC chains without priors; and when using the effective likelihood for model testing, it would always be restricted to a particular set of physical models like in the two examples we showed above.
@@
 
-Before we show the cumulative optical depth constraints from PCs, let us first make a note about the impact of the physicality prior chosen for this analysis and clarify some wrong claims against it in the literature.
+%Before we show the cumulative optical depth constraints from PCs, let us first make a note about the impact of the physicality prior chosen for this analysis and clarify some wrong claims against it in the literature.
 
-\subsection{A note on priors}
+\subsection{Total optical depth}
 \label{sec:note_on_priors}
 
-As mentioned in Section~\ref{sec:background}, the priors of Eq.~\ref{eq:individualprior} are necessary but not sufficient conditions for physicality.
-Therefore, while retaining all physical models, they also includes some unphysical models.
+Using Eqs.~(\ref{eq:mmutoxe}) and (\ref{eq:cumtau}), we can derive constraints on the total optical
+depth $\tau=\tau(0,z_{\rm max)$ with the physicality priors of Eq.~\ref{eq:individualprior}.
+These are shown in Fig.~XX \wh{figure could include $\tau$, $\tau_{12}$ and possibly $\tau_{12}$ without physicality priors?} and correspond to
+\beq
+\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068},
+\eeq  
+\wh{comment on comparison to tanh tau, flexknot etc.}
+Notice that the PC mean, while still compatible with the tanh 
+tau derived constraints, is almost $0.5\sigma$ higher...
+
+The difference between these results come down to implicit and explicit priors.
+Results based on the tanh model requires that reionization happens suddenly at $z_{\rm re}$ and is maintained thereafter.
+Therefore, the model has a strong implicit prior that a given a total optical depth $\tau$ comes from $z\lesssim z_{\rm re}$ regardless of whether the data allow or prefer
+ionization at higher $z$.  While this prior may be well motivated
+theoretically \cite{XX} (sims that give prompt reionization), it should be distinguished from constraints that are data driven.  
+This was especially apparent with the Planck 2015 data where
+the PC constraints gave $\tau=XX\pm YY$ which differed substantially from the tanh $\tau=XX\pm YY$.  As we shall discuss
+below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift.
+
+PC constraints on the total optical depth do not suffer from assuming a specific class of ionization histories as they parameterize any $x_e(z)$ from $z_{\rm min}=6$ to $z_{\rm max}=30$.
+They do of course come with their own priors namely the flat
+range-bound ones from Eq.~\eqref{eq:individualprior}.
+These have in fact caused some confusion in the literature when interpreted in terms of the total optical depth \cite{Millea:2018bko}.
+Here we first review the impact of flatness in PC space as discussed extensively in Ref.~\cite{Heinrich:2018btc} and then highlight the more important role of physicality priors with Planck 2018 data as compared with
+2015 data.
+
+ We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
+\beq
+\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
+\eeq
+where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
+The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
+the first two components
+\beq
+\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
+\eeq
+%
+% Therefore, while retaining all physical models, they also includes some unphysical models.
 The authors of Ref.~\cite{Millea:2018bko} (hereafter MB18) raised a concern that these PC priors implies a non-flat prior in $\tau$ and claimed that it biases the PC result of  Ref.~\cite{Heinrich:2016ojb} and accounts for most of the shift between the tanh and PC $\tau$ posteriors.
@@
 in an attempt to flatten the $\tau$ prior (by multiplying its inverse at each point in the multidimensional PC space), one would actually end up with a biased $\tau$ constraint for the Planck 2015 data.
-
-We briefly recap the argument below.
+%
+%We briefly recap the argument below.
 Let us use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily.
 We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
-\beq
-\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
-\eeq
-where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
+%\beq
+%\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
+%\eeq
+%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
 Because most of the contributions to $\tau$ comes from the first two components (see also Fig.
 7 of Ref.~\cite{Heinrich:2018btc} in the Appendix), we can approximate it as
+%\beq
+%\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
+%\eeq
+Due to this linearity, flat priors in $m_1$ and $m_2$ correspond to
+flat priors in $\tau_{12}$ for any linear trajectory through the space.   For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015}.   Even without the explicit 
+construction of a flat tanh $\tau$ prior in \S \ref{sec:example1}, 
+the implicit prior from $m_1,m_2$ is already nearly flat.
+
+Ref.~\cite{Millea:2018bko} raised a potential objection regarding more general models.  As illustrated in Fig.~\ref{fig:prior_box}, lines of constant $\tau_{12}$ are rotated compared with the $m_1$,
+$m_2$ axis.
+Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
+for each of its values, e.g. at $\tau_{12}=0.162$ \wh{might be helpful to draw in the figure the $\tau_{12}$ or rotated rectangle that intersects the
+bottom left corner the square instead, so we can clearly state that at any lower $\tau_{12}$ there is less parameter space}
+than at $\tau_{12} = XX$.   Ref.~\cite{Millea:2018bko} suggest 
+inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in \cite{} (planck paper).
+First notice that this inversion would in fact then produce a non-flat prior on $\tau$ for the tanh model
+that disfavors high values.   This is because
+the additional parameter space allowed by the range-bound $m_a$ prior is excluded by the functional form of the model even though the reweighting occurs globally and includes the tanh trajectory as well.
+
+More generally, to the extent that the data constrain $m_1$ and $m_2$ better than the flat range-bound priors, these priors become irrelevant and the reweighting of Ref.~\cite{Millea:2018bko} becomes erroneously motivated.
+To see this, consider a range-bound prior that is rotated to be oriented along constant $\tau_{12}$ (see Fig.~\ref{fig:prior_box}, blue dashed rectangle).
+Had the original prior (blue square) been shifted to slightly lower
+$m_1$ and encompassed all of the 95\% CL allowed
+region the two priors would yield  identical constraints at the
+95\% CL.   This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would still change the original prior but not the
+rotated one.
+
+The real impact of our range-bound $m_1,m_2$ priors comes from the fact that with the Planck 2018 data, the physicality prior on $m_1$
+removes nearly half of space that would otherwise be allowed by the
+data at 95\% CL.
+This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).
+\wh{Maybe add to the $\tau$ vs $\tau_{12}$ figure a version that does not have this $\tau_{12}$ cut.
+Discuss more if the difference significant}.
+
+As discussed in \S \ref{sec:cumulative}, our range-bound priors 
+are conservative in the sense that all the models that are excluded are definitely unphysical but some of the models that are included
+may also be unphysical since their physicality depends on higher order PC components that are explicitly truncated in the analysis.
+On the other hand the tanh model provides an explicitly physical example whose trajectory runs nearly parallel to the our physicality prior and so we do not expect the bias incurred by
+including these models to be large.   In any case, using PCs to 
+test explicitly physical models as in \S \ref{sec:example1} and
+\ref{sec:example2} cannot be affected by our conservative physicality
+cut.
+
+Finally, we check that our results are robust to 
+ extending the PC
+range to
+$z_{\rm max}=50$ which produces negligibly different results
 \beq
-\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
+\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{srollv2}).
 \eeq
+Reverting the Planck 2018 likelihood to \texttt{lowE} at $z_{\rm max}=30$ would give
+\beq
+\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{lowE}).
+\eeq
+As with the tanh model, the \texttt{srollv2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.
 
-In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints (ellipse) in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
+%models that 
+%In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints (ellipse) in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
 The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior.
@@
 
-What is then the true impact of this physicality prior we have imposed.
+%What is then the true impact of this physicality prior we have imposed.
 We have established that the orientation of the prior box in the $m_1-m_2$ plane does not matter for this dataset since the data is more constraining than prior.
@@
 
-The other impact of using the physicality prior is that some unphysical models are included inside the prior.
+%The other impact of using the physicality prior is that some unphysical models are included inside the prior.
 We reiterate that we do not adopt this prior for the KDE and unphysical models do not impact results when actually testing physical models using the effective likelihood code.
@@
 
-[add a sentence to reiterate that the shift between PC and steplike model seen in the Planck 2015 is not due to the use of physicality prior? and some other Marius/Planck paper points?]
+%[add a sentence to reiterate that the shift between PC and steplike model seen in the Planck 2015 is not due to the use of physicality prior? and some other Marius/Planck paper points?]
 
-[restore some text about evolution of planck data and HFI systematics causing PC-tanh shifts in Planck 2015 data?
+%[restore some text about evolution of planck data and HFI systematics causing PC-tanh shifts in Planck 2015 data?
 And address the claim: ``In the Planck 2018 paper on cosmological parameters, the reduction of optical depth from the PC results seen in the Planck 2018 data compared to the Planck 2015 data was attributed to the reduction of HFI systematics, with subdominant contribution from prior."]
@@
 
-\subsection{Cumulative optical depth constraints}
+\subsection{Cumulative optical depth}
 
-Let us now constrain the cumulative optical depth $\tau(z, \zmax)$ using the PC results.
-The optical depth is a better quantity than the ionization history $x_e$ for visualizing and comparing PC results because it is much more closely related to observables.
-Recall that these truncated PCs provide an accurate description of the models only in terms of their \textit{observable} impact, and are not meant to be used to reconstruct the exact ionization history.
+The differences between model dependent and model independent approaches are larger for the allowed cumulative optical depth $\tau(z, \zmax)$ at high $z$
+compared to the total optical depth of the previous section.
+The main results are again related to the $m_1$ and $m_2$ PC constraints. 
+As discussed in Ref.~\cite{Heinrich:2016ojb}, raising $m_1$ increases the  optical depth at all $z$ but with a larger weighting at high $z$ whereas
+raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
+Consequently, as shown in Fig.~\ref{fig:prior_box}, the second aspect of PC constraints beyond the well constrained total optical depth, is a positive
+correlation between the two components that represents the high to low $z$ 
+contribution to the total optical depth.  Any finer detail in the ionization history is much more poorly constrained.
 
-In Fig.~\ref{fig:plot_taugtz_2015_vs_2018} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines).
-In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.
+In Fig.~\ref{fig:plot_taugtz_2015_vs_2018} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines) using all 5 PCs.
+\wh{remind the reader about impact of PC filtering smoothing out features in redshift - refer to the tanh examples to illustrate}
+In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.   Notice first the dramatic improvement in the
+constraint at high-$z$.  Whereas the Planck 2015 allowed and even preferred
+finite optical depth at $z\gtrsim 15$, Planck 2018 no longer does....
+On the other hand, some optical depth is still allowed 
+\beq
+\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}),
+\eeq
+in contrast with the tanh model where it is strictly forbidden in the direct analysis and in the PC filtered effective analysis only enters due to the smoothness of the first few PCs in redshift space.
+\wh{contrast this with Flexknot and talk about compatibility with example 2.}
 
@@
 contours of $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines).
-Furthermore, their total optical depth constraints are practically identical:
-\beq
-\tau_{\rm PC,\, \zmax=30} = 0.0619^{+0.0056}_{-0.0068}\;\; (\textsc{srollv2}),
-\eeq
-and
-\beq
-\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\textsc{srollv2}).
-\eeq
 
-Compared the \textsc{lowE} likelihood results, 
-\beq
-\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\textsc{lowE}),
-\eeq
-and
-\beq
-\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\textsc{lowE}).
-\eeq
-the \textsc{srollv2} error bars on PC $\tau$ was reduced by about 20\%, and the central values rose by about half a sigma. [wanna comment more here?]
+%Furthermore, their total optical depth constraints are practically identical: %
+%\beq
+%\tau_{\rm PC,\, \zmax=30} = 0.0619^{+0.0056}_{-0.0068}\;\; (\texttt{srollv2}),
+%\eeq
+%and
+%\beq
+%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{srollv2}).
+%\eeq
+%
+%Compared the \texttt{lowE} likelihood results, 
+%\beq
+%\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{lowE}),
+%\eeq
+%and
+%\beq
+%\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\texttt{lowE}).
+%\eeq
+%the \texttt{srollv2} error bars on PC $\tau$ was reduced by about 20\%, and the central values rose by about half a sigma. [wanna comment more here?]
 %
@@
 For the high-redshift optical depth, we also obtain consistent upper limits
+%\beq
+%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{srollv2}),
+%\eeq
+%[double check on full chain (from 10/15) this is from 09/30.]
+%and 
 \beq
-\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\textsc{srollv2}),
-\eeq
-[double check on full chain (from 10/15) this is from 09/30.]
-and 
-\beq
-\tau(15, \zmax)_{\rm PC,\, \zmax=50} < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\textsc{srollv2}).
+\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}).
 \eeq
@@
 
+\section{Conclusion}
+\label{sec:conclusion}
+
+\begin{itemize}
+    \item Summarize what we did in one paragraph.
+    \item Give main conclusions in bullet format
+    \item main results from PCs: PC constraints, tau, $tau(>z)$.  
+    \item comparison w/ Planck 
+    \item note on priors
+    \item likelihood code, comparisons
+    \item other interesting points to discuss and add 
+    \item future outlook
+\end{itemize}
+
+\bibliography{rei.bib}
+
+\appendix
+
 \subsection{Comparison to previous results}
+\wh{integrate this into the model sections}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Conclusion}
-\label{sec:conclusion}
-
-\begin{itemize}
-    \item Summarize what we did in one paragraph.
-    \item Give main conclusions in bullet format
-    \item main results from PCs: PC constraints, tau, $tau(>z)$.  
-    \item comparison w/ Planck 
-    \item note on priors
-    \item likelihood code, comparisons
-    \item other interesting points to discuss and add 
-    \item future outlook
-\end{itemize}
-
-\bibliography{rei.bib}
-
-\appendix
 

### 2021-01-07 results/notes.tex

--- 2020-12-03 results/notes.tex
@@
 
-\bibliography{rei.bib}
+\begin{figure}
+%
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run1_f0p28_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run1_f0p28_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+%
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run4_f0p18_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run4_f0p18_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+%
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+%
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run3_f0p10_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.35\textwidth]{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run3_f0p10_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_2_num_bins_1D_20_smooth_scale_2D_2_num_bins_2D_25.png}
+%
+\label{fig:}
+\caption{Left: Automated getdist smoothing; Right: hand tuned smoothing (slightly aggressive binning).
+} 
+\end{figure}
+
+%\begin{figure}
+%    \centering
+%    \includegraphics{cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0393_tri.png}
+%    \caption{$\taulo$ prior imposed at 0.0393, giving a straight cut in the 2D constraint, rather than the rounded zre = 6.1 prior.}
+%    \label{fig:my_label}
+%\end{figure}
 

### 2021-01-07 results/notes.tex

@@
 
+\begin{figure}
+%
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run1_f0p28_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.3_num_bins_1D_100_smooth_scale_2D_0.3_num_bins_2D_100.png}
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run1_f0p28_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.5_num_bins_1D_100_smooth_scale_2D_0.5_num_bins_2D_100.png}
+%
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run4_f0p18_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.3_num_bins_1D_100_smooth_scale_2D_0.3_num_bins_2D_100.png}
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run4_f0p18_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.5_num_bins_1D_100_smooth_scale_2D_0.5_num_bins_2D_100.png}
+%
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.3_num_bins_1D_100_smooth_scale_2D_0.3_num_bins_2D_100.png}
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.5_num_bins_1D_100_smooth_scale_2D_0.5_num_bins_2D_100.png}
+%
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run3_f0p10_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.3_num_bins_1D_100_smooth_scale_2D_0.3_num_bins_2D_100.png}
+\includegraphics[width=0.35\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test9_run3_f0p10_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri_smooth_scale_1D_0.5_num_bins_1D_100_smooth_scale_2D_0.5_num_bins_2D_100.png}
+%
+\label{fig:}
+\caption{Left: getdist smooth\_scale\_1D = smooth\_scale\_2D = 0.3; Right: same as left but with 0.5.
+} 
+\end{figure}
+
 \end{document}

### 2021-01-14 paper/rei.tex

@@
 \sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
+\label{eq:prior_sum}
 \end{equation}
@@
 \end{equation}
+\ch{It turns out that in practice the prior on the sum of squares in Eq.~\ref{eq:prior_sum} is weaker than the individual priors in Eq.~\ref{eq:individualprior} for the Planck 2018 data.}
 \wh{comment on the sum prior?  we don't seem to talk about it later - is it applied and is it always weaker in practice?}
-\ch{I think it is always weaker, let me double check.}
+\ch{CH: The sum prior is applied at all times; I also checked  (for the zmax = 30 chains this year) that it is weaker than the prior on individual m's just as in Planck 2015.}
 %
@@
 
-\ch{[double check all these numbers again, write script to output in a table automatically]}.
+\ch{[double check all these numbers again (some chains are longer now)]}.
 

### 2021-01-20 paper/rei.tex

@@
 \begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/pl18_tanh_highz_kde_f0p14_tau_z_to_zstart_plot_tau_gtz_dz0p1.pdf}
+\includegraphics[width=0.40\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
 \caption{The 68\% and 95\% C.L.
@@
  \begin{figure}
-          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_plot_rotated_box_flat_tau_prior_fac_0.8.png}
+          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau_prior_fac_0.8.pdf}
           \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in

### 2021-01-27 paper/rei.tex

@@
 \begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
-\caption{[placeholder] PC chains for zmax = 30. Planck 2018 srollv2 likelihood vs Planck 2015). [To be created and merged with the next plot as a single figure with two panels.]
+\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2.pdf}
+\caption{Cumulative optical depth distributions from PC chains for zmax = 30 Planck 2015 vs 2018 simall\_EE vs 2018 srollv2 likelihood. 
 }
-\label{fig:plot_taugtz_2015_vs_2018}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_post_0930_and_pl18_pc_zmax30_pliklite_srollv2_0930.pdf}
-\caption{PC chains for zmax = 30. Planck 2018 original lowE vs srollv2 likelihood (plik\_lite\_TTTEEE + lowl + simall\_EE vs plik\_lite\_TTTEEE + lowl + sroll2\_EE).
-}
-\label{fig:plot_taugtz_lowE_vs_srollv2}
+\label{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2}
 \end{figure}
@@
 \end{document}
+
+% Notes:
+% tau1 = (taulo + tauhi) --> (zre, xemin) --> xe(z) ==> tau2 ==> mjs --> tau3(mjs)
+% taulo + tauhi = tau_desired

### 2021-02-03 paper/rei.tex
Updates from Overleaf

@@
 Finally, we summarize our results and conclude in section~\ref{sec:conclusion}.
+
+%
+%
+%
+%
+%
+
 \section{Background}
@@
 We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.
+
+%
+%
 \subsection{Reionization Principal Components}
@@
 removing demonstrably  unphysical ionization histories in \S \ref{sec:cumulative}.
+
+ \begin{figure}
+          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
+          \caption{Reionization PCs for $\zmax = 30$.} 
+          \label{fig:xe}
+\end{figure}
+
+%
+%
+%
+%
 
@@
 [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$.
-make point that tanh ML is now within the the center of PC constraints]}
+make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)]}
 \label{fig:plot_mjs_2018_vs_2015}

### 2021-02-03 paper/rei.tex

@@
 \begin{figure}[ht]
+\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
+\caption{Cumulative optical depth distributions from PC chains for zmax = 30 Planck 2018 srollv2 likelihood, with vs without physicality prior. 
+}
+\label{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2}
+\end{figure}
+
+\begin{figure}[ht]
 \includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2.pdf}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_pliklite_srollv2_0930_and_pl18_pc_zmax50_pliklite_srollv2.pdf}
+\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50.pdf}
 \caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).

### 2021-02-13 paper/rei.tex

@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/direct_mcmc/pl18_plots_zmax30/plot_pub_tau_gtz_dz_0p1_pl18_pc_zmax30_plikfull_and_pl18_tanh_post_plikfull.pdf}
-\caption{Tanh chains with plik\_full\_TTTEEE for the high-l likelihood. [remove PC zmax curve; add PC filtered version]
+\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
+\caption{Cumulative optical depth for the tanh model from direct MCMC chains, KDE chians and Gaussian likelihood chains. 
 }
@@
 \label{fig:two_step_model}
+\end{figure}
+
+\begin{figure}[t]
+\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\caption{Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
+}
+\label{fig:two_parameter_model_2D}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
+\caption{The 68\% and 95\% C.L.
+constraints on the cumulative optical depth $\tau(z, \zmax)$ for the two-step model, which allows for high-redshift ionization up to $\zmax =30$.
+note that a physical model with $\tau(15, \zmax) = 0.02$ is still allowed at the 95\% C.L.
+whereas the Planck FlexKnot analysis would have ruled it out with their $\tau(15, \zmax) < 0.007$ 95\% upper limit.
+[add not PC filtered version]
+}
+\label{fig:plot_taugtz_two_step_contours}
 \end{figure}
@@
 
-\begin{figure}[t]
-\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-\caption{Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
-}
-\label{fig:two_parameter_model_2D}
-\end{figure}
-
 In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.
@@
 %\end{figure}
-
-\begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
-\caption{The 68\% and 95\% C.L.
-constraints on the cumulative optical depth $\tau(z, \zmax)$ for the two-step model, which allows for high-redshift ionization up to $\zmax =30$.
-note that a physical model with $\tau(15, \zmax) = 0.02$ is still allowed at the 95\% C.L.
-whereas the Planck FlexKnot analysis would have ruled it out with their $\tau(15, \zmax) < 0.007$ 95\% upper limit.
-[add not PC filtered version]
-}
-\label{fig:plot_taugtz_two_step_contours}
-\end{figure}
 

### 2021-02-13 paper/rei.tex

@@
 \begin{figure}[t]
-\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-\caption{Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
+%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
+\includegraphics[width=0.47\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
+\caption{[Need to figure out dashed vs solid line convention in plots].
+Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
 }
@@
 \begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
+\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
 \caption{Cumulative optical depth distributions from PC chains for zmax = 30 Planck 2018 srollv2 likelihood, with vs without physicality prior. 

### 2021-03-04 paper/rei.tex

@@
 
-...
+We publish Relike, an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC).
+This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is tested.
+We tested the code on two example models
+%: 1) the tanh model with a single transition redshift, and 2) a two-parameter model with two transitions allowing for a high-redshift ionization plateau. 
+which showed good agreement with directly 
+%The Relike chains converged 300$\times$ and 100$\times$ faster for model 1 and 2 respectively, and showed good agreement compared to directly
+sampling the original Planck likelihoods.
+We also find that contrary to the Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation.
+We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler.
+We expect the code release to dramatically reduce the time for testing individual ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
+%We expect the code release to be useful for individual model testing as the user would not need to modify CAMB every time a different ionization history model is needed; the effective likelihood code also makes it easier to perform joint constraints of the Planck data with other reionization data such as 21cm, luminosity functions, star formation rate measurements, etc.
+Finally, we derive model-independent constraints: $\tau_{\rm PC} = ...$ and $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.).
+The upper limits on high-redshift optical depth is at least a factor of 2 looser those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.) using a  two-parameter toy model without using PCs.
 
@@
 %
+ \begin{figure}
+          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
+          \caption{Reionization PCs for $\zmax = 30$.} 
+          \label{fig:xe}
+\end{figure}
+
 \subsection{Reionization Principal Components}
@@
 where $m_a$ are the PC amplitudes and $\xef(z)$ is the fiducial model.
-We obtain the PCs $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements
+We obtain the PC basis functions $S_{a}(z)$ as eigenfunctions of the Fisher information matrix for $x_e(z)$ in a given range $z_{\rm min}<z<z_{\rm max}$ from cosmic variance limited $C_\ell^{EE}$ measurements
 %
 \beq
-F_{ij} = \sum_l \frac{1}{\sigma_l^2} \frac{\partial C_l^{EE}}{\partial x_e(z_i)}\frac{\partial C_l^{EE}}{\partial x_e(z_j)} = \sum_a S_a(z_i) \sigma_a^{-2} S_a(z_j),
+F_{ij} = \sum_l \wh{\frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)}}= \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
 \eeq
@@
 where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
-We have rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
+The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
+ Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
+
+We then rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
 In the work that follows, we therefore truncate the sum at $a = 1 ..
-n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis.
-We also test its robustness by taking 
- for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.
-[Maybe insert reference to a plot of all five PCs (we might refer to its shapes later)]
-Note that the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
+n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S XX by  taking
+ for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}
+
+ In Fig.~\ref{fig:xe} (top), we show the corresponding fiducial basis functions $S_a(z)$.  
+Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
+This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
+As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
 given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
@@
 
-We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}, where the PCs were discretized at $\delta z = 0.25$, and computed at the fiducial model $x_e^{\rm fid} = 0.15$.
-Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the $\zmax = 30$ PCs to be computed with Planck 2015 rather than WMAP best-fit parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
-For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).
+We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
+For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a transition width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).
 
@@
 
- \begin{figure}
-          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
-          \caption{Reionization PCs for $\zmax = 30$.} 
-          \label{fig:xe}
-\end{figure}
-
 %
@@
 The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data.
-In the following, we briefly recap the kernel density estimate technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+In the following, we briefly recap the kernel density estimate (KDE) technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
 The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
@@
 where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
+\wh{Possibly add a few words on the PC version and refer to the
+bottom panel of Fig.~\ref{fig:xe}}
 The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
@@
 Therefore, when determining cumulative optical depth constraints as opposed to using reionization PCs as a tool for testing models,
- we do impose a physicality prior
+ we do impose a  prior
  by truncating the posteriors of $m_a$
@@
 \end{equation}
-\ch{It turns out that in practice the prior on the sum of squares in Eq.~\ref{eq:prior_sum} is weaker than the individual priors in Eq.~\ref{eq:individualprior} for the Planck 2018 data.}
-\wh{comment on the sum prior?  we don't seem to talk about it later - is it applied and is it always weaker in practice?}
-\ch{CH: The sum prior is applied at all times; I also checked  (for the zmax = 30 chains this year) that it is weaker than the prior on individual m's just as in Planck 2015.}
-%
-Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for physicality.
+In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
+% I think it can't depend on the data but does depend on the truncation
+%the Planck 2018 data \wh{and
+our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.}
+
+%
+Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.
 In other words, no physical model is excluded by these priors but some unphysical models are included.
@@
 regions in the $m_a-m_b$ planes in the lower triangle.
-The physicality priors on the ionization history used in this work are delineated by the box boundaries.
+The physicality priors on the ionization history used in this work are delineated by the box boundaries \wh{and the unshaded region for $m_1$}.
 [TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$.
-make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)]}
+make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)] \wh{double check relative sign on the definition of $m_i$ between 2015-18 - seems like some correlations have flipped, e.g.
+$m_4$.} }
 \label{fig:plot_mjs_2018_vs_2015}
@@
 the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
-from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher \ch{[lower?]} total optical depth.
-\ch{[Or maybe you meant Planck 2015 to simall EE?]}
+from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher total optical depth.
 
@@
 \caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper.
-\wh{some of these figure versions look like low res bit maps - final versions should be vector graphics and look uniform in style between figures.} \ch{got it, will polish for sure.}
 }
@@
 Using Eqs.~(\ref{eq:mmutoxe}) and (\ref{eq:cumtau}), we can derive constraints on the total optical
-depth $\tau=\tau(0,z_{\rm max)$ with the physicality priors of Eq.~\ref{eq:individualprior}.
+depth $\tau=\tau(0,z_{\rm max})$ with the physicality priors of Eq.~\ref{eq:individualprior}.
 These are shown in Fig.~XX \wh{figure could include $\tau$, $\tau_{12}$ and possibly $\tau_{12}$ without physicality priors?} and correspond to
 \beq
-\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068},
+\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068} \;\; (\texttt{srollv2}),
+\label{eq:tau_pc_zmax30}
 \eeq  
@@
 construction of a flat tanh $\tau$ prior in \S \ref{sec:example1}, 
-the implicit prior from $m_1,m_2$ is already nearly flat.
+the implicit prior from $m_1,m_2$ is already nearly flat. \ch{CH: need to ask a question about this.}
 
@@
 Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
-for each of its values, e.g. at $\tau_{12}=0.162$ \wh{might be helpful to draw in the figure the $\tau_{12}$ or rotated rectangle that intersects the
-bottom left corner the square instead, so we can clearly state that at any lower $\tau_{12}$ there is less parameter space}
-than at $\tau_{12} = XX$.   Ref.~\cite{Millea:2018bko} suggest 
-inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in \cite{} (planck paper).
+for each of its values, e.g. at $\tau_{12}=0.018$
+than at $\tau_{12} = 0.125$.   Ref.~\cite{Millea:2018bko} suggest 
+inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the \ch{Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}}.
 First notice that this inversion would in fact then produce a non-flat prior on $\tau$ for the tanh model
@@
 To see this, consider a range-bound prior that is rotated to be oriented along constant $\tau_{12}$ (see Fig.~\ref{fig:prior_box}, blue dashed rectangle).
+\ch{This rotated prior in $m_1-m_2$ space corresponds to a flat prior in $\tau_{12}$ space.}
 Had the original prior (blue square) been shifted to slightly lower
@@
 95\% CL.   This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would still change the original prior but not the
-rotated one.
+rotated one.  \ch{CH: will try rephrasing.}
 
 The real impact of our range-bound $m_1,m_2$ priors comes from the fact that with the Planck 2018 data, the physicality prior on $m_1$
-removes nearly half of space that would otherwise be allowed by the
+removes nearly half of the space that would otherwise be allowed by the
 data at 95\% CL.
@@
 On the other hand the tanh model provides an explicitly physical example whose trajectory runs nearly parallel to the our physicality prior and so we do not expect the bias incurred by
-including these models to be large.   In any case, using PCs to 
+including these models to be large. \ch{CH: sorry, what do we mean by these models here?}  In any case, using PCs to 
 test explicitly physical models as in \S \ref{sec:example1} and
@@
 
-Finally, we check that our results are robust to 
+Finally, we check that our results (Eq.~\ref{eq:tau_pc_zmax30}) are robust to
  extending the PC
 range to
-$z_{\rm max}=50$ which produces negligibly different results
+$z_{\rm max}=50$ which is negligibly different
 \beq
@@
 
-In Fig.~\ref{fig:plot_taugtz_2015_vs_2018} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines) using all 5 PCs.
+\ch{[CH: Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior} needs to go somewhere]}
+% with this prior you still have that $\tau(>z)$ plot includes unphysical models (however, most likely total tau doesn't change, and if any bias from unphysical models, it would lower high-z tau).\\
+
+In Fig.~\ref{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines) using all 5 PCs.
 \wh{remind the reader about impact of PC filtering smoothing out features in redshift - refer to the tanh examples to illustrate}
@@
 
-% with this prior you still have that $\tau(>z)$ plot includes unphysical models (however, most likely total tau doesn't change, and if any bias from unphysical models, it would lower high-z tau).\\
-
 We also show using a separate PC chains with $\zmax = 50$ that there is no evidence for optical depth at $z>30$. 
-In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that 68\% and 95\% C.L.
+In Fig.~\ref{fig:plot_taugtz_zmax30_vs_zmax50}, we show that the 68\% and 95\% C.L.
 contours of $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines).
@@
 }
-\label{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2}
+\label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
 \end{figure}
@@
 
+In conclusion, we used constraints on reionization principal components with respect to the large-scale $EE$ data to produce a fast and accurate effective likelihood for the final release of Planck 2018 data.
+In lieu of the official $\texttt{simallEE}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{srollv2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
+Our effective likelihood code package is called $\textsc{Relike}$ [to be changed] and is available publicly on Github.
+This code is beneficial for the community for the purpose of testing \textit{any} global ionization history between $6 < z < 30$ with the Planck data without having to modify CAMB which can be very cumbersome to do.
+The convergence time to reach the Gilman-Rubin criteria $R_{-1} < 0.01$ is about 300 times less with the Relike code than with a traditional MCMC sampling of the Planck likelihood for the one-parameter canonical tanh model.
+
+To test this effective Planck likelihood code, we used two examples: 1) a canonical tanh model with a single transition redshift, and 2) a two-parameter model consisting of two tanh transitions allowing in addition for a high-redshift ionization plateau compared to the first example.
+We demonstrated accuracy of the Relike code for both examples by comparing the constraints on the reionization model parameters with those obtained from directly sampling the full Planck likelihood and using a modified CAMB code.
+We also showed great agreement between the confidence levels of the cumulative optical depth evolution between the 5-PC representation and the exact ionization model.
+
+Besides conducting explicit model testing with the Relike code, we also obtained model-independent constraints on the reionization history using the PC chains themselves.
+We summarize these results below.
+
 \begin{itemize}
-    \item Summarize what we did in one paragraph.
-    \item Give main conclusions in bullet format
-    \item main results from PCs: PC constraints, tau, $tau(>z)$.  
-    \item comparison w/ Planck 
-    \item note on priors
-    \item likelihood code, comparisons
+
+    \item {We obtained the constraint on the total optical depth $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ (68\% C.L.) 
+    using PCs and the Planck 2018 likelihoods using the \texttt{plik\_lite\_TTTEEE+lowl+srollv2} likelihood combination.}
+
+    \item {The better foreground cleaned \texttt{srollv2} likelihood allowed to tighten the PC constraint on $\tau$ and shifted it toward slightly higher values when compared with the \texttt{plik\_lite\_TTTEEE+lowl+lowE} result  $\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}$ (68\% C.L.).}
+
+    \item {We checked the robustness of our $\zmax=30$ results by performing a PC analysis of the Planck data with $\zmax = 50$.
+The results are changed negligibly: $\tau(15, 50) < 0.022 $ (95\% C.L.) in comparison with $\tau(15, 30) < 0.023$ (95\% C.L.).}
+
+     \item {Though our high-redshift optical depth constraints are looser than those from the Planck 2018 cosmological parameter paper $\tau(15, 30) < 0.007$ (95\% C.L.) obtained from another model-independent method called FlexKnot, we have tested using explicitly the tanh high-$z$ toy model (without using any PCs) that an $\tau(15, 30) < 0.02$ (95\% C.L.) [get more precise numbers] is still allowed (see Fig.~\ref{fig:plot_taugtz_two_step_contours}), which is more consistent with our PC results.}
+
+\end{itemize}
+
+We clarified some confusion raised in the literature on the effects of applying a range-bound prior on the PC amplitude during the analysis, and demonstrated that these priors, though they include some unphysical ionization models, do not affect the total $\tau$ constraint and the high-redshift optical depth upper limits significantly.
+In any case, as shown in Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior} removing more unphysical models would only loosen the high-$z$ optical depth upper limit, in contrary to what was claimed in Ref.~\cite{Millea:2018bko}.
+The shift from a higher to lower total $\tau$ from Planck 2015 to Planck 2018 in any model-independent is however due to the use of XX data as demonstrated by Ref.~\cite{Millea:2018bko}, in relation to better systematics reduction on the low multipoles $l<10$ [check number]~\cite{Heinrich:2018btc}.
+[cite also a Planck paper].
+
+\begin{itemize}
     \item other interesting points to discuss and add 
-    \item future outlook
+    \item future outlook (mention joint analyses)
 \end{itemize}

### 2021-03-04 paper/rei.tex

@@
           \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
-          \caption{Reionization PCs for $\zmax = 30$.} 
+          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
+We assume full ionized Universe at $z < 6$.
+\textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC.}
           \label{fig:xe}
@@
 \beq
-F_{ij} = \sum_l \wh{\frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)}}= \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
+F_{ij} = \sum_l \frac{2 l+1}{2} \frac{\partial \ln C_l^{EE}}{\partial x_e(z_i)}\frac{\partial \ln C_l^{EE}}{\partial x_e(z_j)} = \sum_a \frac{ S_a(z_i) S_a(z_j)}{\sigma_a^2},
 \eeq
@@
 
- In Fig.~\ref{fig:xe} (top), we show the corresponding fiducial basis functions $S_a(z)$.  
+ In Fig.~\ref{fig:xe}
+ %
+ \footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
+ %
+ (top), we show the corresponding fiducial basis functions $S_a(z)$.  
 Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
@@
 regions in the $m_a-m_b$ planes in the lower triangle.
-The physicality priors on the ionization history used in this work are delineated by the box boundaries \wh{and the unshaded region for $m_1$}.
-[TODO: update legend] [Add text: The black solid line indicates the steplike models varying $\tau$.
-make point that tanh ML is now within the the center of PC constraints][say that Planck 2015 point is chain ML, whereas Plnack 2018 point is from minimizer, also is a different tanh model (w/ dz = 0.015(1+z), not dz = 0.5)] \wh{double check relative sign on the definition of $m_i$ between 2015-18 - seems like some correlations have flipped, e.g.
-$m_4$.} }
+The physicality priors on the ionization history used in this work are delineated by the box boundaries and the unshaded region for $m_1$.
+The black solid line indicates the tanh models with varying $\tau$.
+The red and black crosses indicate the best-fit tanh models from Planck 2015 and 2018 respectively.
+Note that the best-fit tanh model has moved to a lower $\tau$ in Planck 2018, and in particular, is now within the 68\% C.L.
+contours of the constraints.
+\wh{double check relative sign on the definition of $m_i$ between 2015-18 - seems like some correlations have flipped, e.g. $m_4$.} }
 \label{fig:plot_mjs_2018_vs_2015}
@@
 %
-In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
+In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$
+%
+\footnote{We The Planck 2015 best-fit is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Plnack 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $dz = 0.015(1+z)$ instead of $dz = 0.5$ for the Planck 2015 best-fit.
+Using $dz = 0.5$ for Planck 2018 moves the cross negligible on this plot.}.
+%
 The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
@@
 \includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $\zmax=30$ PC plane, the $m_1-m_2$ plane, comparing Planck 2015 (gray) vs Planck 2018 results using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) vs the \texttt{srollv2} likelihood (blue), the latter had better foreground removal and is the one we show as our fiducial results throughout this paper.
+\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane.
+We compare the Planck 2015 (gray) to the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue).
+The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper.
 }
@@
 {plots/plot_tanh_gaussian_MCMC.png}
-\caption{Posterior of total optical depth $\tau$: full likelihood (solid) vs effective likelihood (dotted) with f = 0.14 [remove more curves?].
-The results agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
+\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the full Planck likelihoods (solid), our effective likelihood in the kernel density estimate mode with $f = 0.14$ (dotted), and in the Gaussian approximation mode (dashed) \ch{tune plot and line styles}.
+The KDE mode agree well with the full results, while giving a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
+The Gaussian likelihood works well too, with no extra broadening of the posterior, but a negligible shift in the mean ($\sim XX\%$ of the error in $\tau$).
+This posterior is symmetric and does not capture the small skewness found in the full and KDE chains.
+\ch{might need to rephrase.}
 }
@@
 \includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
-\caption{Cumulative optical depth for the tanh model from direct MCMC chains, KDE chians and Gaussian likelihood chains. 
-}
+\caption{Comparison of the 68\% and 95\% C.L.
+constraints on the cumulative optical depth in the tanh model (described in \refsec{example1}) between the full Planck likelihoods (shaded bands), our effective likelihoods in the KDE mode (blue solid) and Gaussian mode (black dashed).
+The effective likelihoods results oscillate around zero at high-z because of the truncated representation of the models at 5 PCs (enough to represent any observable effects in the data).
+While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth.}
+%double check plot, upper and lower limits seem to cross?
 \label{fig:plot_taugtz_PC_vs_tanh}
@@
 \includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
-\caption{Ionization history $x_e(z)$ of the two-step model $(\taulo, \tauhi) = (0.04, 0.02)$ for the hydrogen and singly ionized helium.
+\caption{Example ionization history $x_e(z)$ for the high-$z$ model of \refsec{example2}.
+In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$.
+This is a toy model constructed for illustration purposes only.
+Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (XX, XX)$, as well as  a model at the 95\% C.L.
+$(\taulo, \tauhi) = (XX, XX)$ (see also Fig.~\ref{two_parameter_model_2}).
+\ch{Include 2nd helium in the plot.}
 }
@@
 
+%
+%
+
 \begin{figure}[t]
@@
 \includegraphics[width=0.47\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
-\caption{[Need to figure out dashed vs solid line convention in plots].
-Marginalized posteriors for $\tauhi$ and $\taulo$ in the two-step model described in \refssec{example2}: effective reionization likelihood with f = 0.14 (red) agrees very well with the full likelihood results (blue).
+\caption{\ch{Need to fix this plot.}   Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L.
+contours for the two-step model described in \refsec{example2} between different likelihoods: The full Planck likelihoods (shaded), the effective likelihoods in the KDE mode with f = 0.14 (blue solid) and in the Gaussian approximation mode (black dashed).
+The results agree well with each other.
+As in Fig.~\ref{fig:tanh}, both the KDE and Gaussian modes agree very well with the full likelihood results, with the KDE  contours being slightly broader as expected.
 }
@@
 \includegraphics[width=0.5\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
-\caption{The 68\% and 95\% C.L.
+\caption{Comparison of the 68\% and 95\% C.L.
 constraints on the cumulative optical depth $\tau(z, \zmax)$ for the two-step model, which allows for high-redshift ionization up to $\zmax =30$.
-note that a physical model with $\tau(15, \zmax) = 0.02$ is still allowed at the 95\% C.L.
-whereas the Planck FlexKnot analysis would have ruled it out with their $\tau(15, \zmax) < 0.007$ 95\% upper limit.
-[add not PC filtered version]
+Note that a physical model with $\tau(15, \zmax) = 0.02$ is still allowed at the 95\% C.L.
+whereas it is ruled with the Planck official results using the FlexKnot method $\tau(15, \zmax) < 0.007$.
+Here again, the KDE and Gaussian likelihoods agree well with the full likelihoods, with better 95\% C.L.
+limits on the total optical depth.
+The truncation at 5 PCs causes wiggles at higher redshifts, but these cause negligible differences at most redshifts.
 }
@@
           \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
-          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) which are aligned with lines of constant $\tau_{12}$ (light gray lines).
+          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) \ch{CH: [why is this flat in m1 and m2?]} which are aligned with lines of constant $\tau_{12}$ (light gray lines).
 Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
@@
 \includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
-\caption{Cumulative optical depth distributions from PC chains for zmax = 30 Planck 2018 srollv2 likelihood, with vs without physicality prior. 
+\caption{Comparison of the 68\% and 95\% C.L.
+constraints on the cumulative optical depth $\tau(z, \zmax)$ from PC chains with (shaded) and without (black) applying the physicality prior.
+While the lower limits on $\tau(z, \zmax)$ are lowered when more unphysical models are included, the upper limits do not vary significantly (slightly lowered).
+The total $\tau$ distribution is broadened only toward the lower $\tau$.
+\ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.}
 }
@@
 \includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2.pdf}
-\caption{Cumulative optical depth distributions from PC chains for zmax = 30 Planck 2015 vs 2018 simall\_EE vs 2018 srollv2 likelihood. 
+\caption{Comparison of the 68\% and 95\% C.L.
+constraints on the cumulative optical depth with different Planck likelihoods.
+\textit{Top}: The Planck 2015 vs Planck 2018 likelihoods, the latter including the official low-$\ell$ $EE$ likelihood $\texttt{simallEE}$.
+\textit{Bottom}: The Planck 2018 likelihoods using $\texttt{simallEE}$ for the low-$\ell$ $EE$ likelihood vs \texttt{srollv2} which has better foreground cleaning.
+Both yield consistent high-redshift constraints, whereas the \texttt{srollv2} likelihood tightens the lower limit on the low-$z$ contribution, resulting on a tighter lower limit and higher mean for the total $\tau$.
 }
@@
 \includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50.pdf}
-\caption{Comparing zmax = 30 and 50 PC chains using Planck 2018 original lowE likelihood (plik\_lite\_TTTEEE + lowl + sroll2\_EE).
-Note that the zmax = 30 uses 5 PCs, whereas the zmax = 50 chains uses 7 PCs.
-.
+\caption{Comparison of the 68\% and 95\% C.L.
+constraints on the cumulative optical depth from PC chains with $\zmax = 30$ vs $\zmax = 50$ using the Planck 2018 likelihoods.
+The constraints on the cumulative high-redshift optical depth are shown to be robust to $\zmax$.
+Both the low-redshift constraints and the high-redshift upper limits agree well.
 }

### 2021-03-12 paper/rei.tex

@@
 
-\section{Background}
+\section{PC Technique}
 \label{sec:background}
@@
           \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
-We assume full ionized Universe at $z < 6$.
-\textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC.}
+We assume fully ionized Universe at $z < 6$.
+\textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC.
+\ch{plot done}}
           \label{fig:xe}
@@
 We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
-For the Planck data set, most of the information on the ionization history comes from the first two modes which probe the amount of low vs.\ high redshift optical depth.
+For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.
 However all 5 PCs for $z_{\rm max}=30$ carry some constraint, with the higher modes controlling finer variations in redshift.
@@
 where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
-\wh{Possibly add a few words on the PC version and refer to the
+\wh{Possibly add a few words on the PC version, list $\tau_a = \tau_a(z_{\rm min},z_{\rm max})$ in an equation or table, and refer to the
 bottom panel of Fig.~\ref{fig:xe}}
@@
 %the Planck 2018 data \wh{and
-our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.}
+our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.
 
@@
 contours of the constraints.
-\wh{double check relative sign on the definition of $m_i$ between 2015-18 - seems like some correlations have flipped, e.g. $m_4$.} }
+\ch{plot done}
+}
 \label{fig:plot_mjs_2018_vs_2015}
 \end{figure*}
-%results/pc_results/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite.pdf
+%
 %
@@
 
+\begin{table}[b]
+\centering
+\caption{$\tau_a$}
+\label{tab:PC_stats}
+%\begin{tabular}{|r | r r r r r r|}
+\begin{tabular}{|c | c | c | c | c | c|}
+\hline
+$\tau_{\rm fid}$ & $\tau_1$  & $\tau_2$ & $\tau_3$ & $\tau_4$ & $\tau_5$ \\ \hline
+0.08626 & 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
+\hline
+\end{tabular}
+\end{table}
+% all digits without flipping m3, m4:
+%tau_fid 0.08625526676478058
+%taumj [ 0.29403315 -0.11227135 -0.04505736  0.0189787   0.00960184
 \begin{table}[b]
@@
 Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (XX, XX)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (XX, XX)$ (see also Fig.~\ref{two_parameter_model_2}).
+$(\taulo, \tauhi) = (XX, XX)$ (see also Fig.~\ref{two_parameter_model_2}  \wh{place x's on the models plotted in that plot}).
 \ch{Include 2nd helium in the plot.}
@@
 The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
-}
+%}
 %\label{fig:two_parameter_model_2D_plot}
@@
 \eeq
+\wh{possibly redefine
+\beq
+\tau_{12} =  m_1 \tau_1 + m_2 \tau_2 +
+\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) \approx \tau.
+\eeq
+where the term in parentheses is a fixed quantity (refer to eqns/table where tau's and mbars are defined) whose purpose is to remove offsets due to the arbitrary choice of fiducial model.
+}
 %
@@
 Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
-for each of its values, e.g. at $\tau_{12}=0.018$
-than at $\tau_{12} = 0.125$.   Ref.~\cite{Millea:2018bko} suggest 
+for each of its values, e.g. at  left corners (Fig.~\ref{fig:prior_box} solid square or lowest $m_1$) of the range $\tau_{12}=0.018$
+than at $\tau_{12} = 0.125$ \wh{revise numbers} and the implied prior increases linearly from zero between the two.
+   Ref.~\cite{Millea:2018bko} suggest 
 inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the \ch{Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}}.
@@
 \wh{Maybe add to the $\tau$ vs $\tau_{12}$ figure a version that does not have this $\tau_{12}$ cut.
-Discuss more if the difference significant}.
+Discuss more if the difference significant - also mention that even in the direct tanh constraints, the impact of taking $z\gtrsim 6$ has the effect of removing some of the low $\tau$ region.}.
 
 As discussed in \S \ref{sec:cumulative}, our range-bound priors 
-are conservative in the sense that all the models that are excluded are definitely unphysical but some of the models that are included
+are conservative in the sense that all the cases that are excluded are definitely unphysical but some of the cases that are included
 may also be unphysical since their physicality depends on higher order PC components that are explicitly truncated in the analysis.
-On the other hand the tanh model provides an explicitly physical example whose trajectory runs nearly parallel to the our physicality prior and so we do not expect the bias incurred by
-including these models to be large. \ch{CH: sorry, what do we mean by these models here?}  In any case, using PCs to 
+On the other hand the tanh model provides an explicitly physical example whose trajectory runs close  and nearly parallel to the our physicality prior in the allowed region and so we do not expect the bias incurred by
+including these possibly \wh{unphysical cases} to be large. 
+%\ch{CH: sorry, what do we mean by these models here?}  
+In any case, using PCs to 
 test explicitly physical models as in \S \ref{sec:example1} and
@@
           \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
-          both $\tau_{12}$ and $m_1-m_2$ (rotated rectangle, dashed lines) \ch{CH: [why is this flat in m1 and m2?]} which are aligned with lines of constant $\tau_{12}$ (light gray lines).
+           $\tau_{12}$ and \wh{range-bound flat - also if we change $\tau_{12}$ definitions change the numbers in the labels here} in $m_1-m_2$ (rotated rectangle, dashed lines) 
+           %\ch{CH: [why is this flat in m1 and m2?]} 
+           % flat in 2D or locally flat
+           which are aligned with lines of constant $\tau_{12}$ (light gray lines).
 Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
@@
           \label{fig:prior_box}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
+\caption{$\tau_{12,new}$}
+}
+\label{fig:tau12}
 \end{figure}
@@
 }
+% no prior: 95\% C.L.  = [-0.01935756790924691, 0.021029019067727237]
+% with prior: {'z': 15, 'mean': 0.00694422689213187, 'std': 0.007568921831067864, 'limit68': [-0.0007215571045808687, 0.014684826477572082], 'limit95': [-0.006316156744039184, 0.02284538046347211]}
+%
+%}
 \label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
@@
 
-In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for the tanh model, $\tau = 0.0504^{+0.0050}
-_{-0.0079}$ for the FlexKnot method (with flat $\tau$ prior) and $\tau = 0.0487^{+0.0038}_{−0.0081}$ for the PC method but using the prior inversion procedure discussed Section~\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
+In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for the tanh model,
+$\tau = 0.0504^{+0.0050}_{-0.0079}$ 
+for the FlexKnot method (with flat $\tau$ prior) and 
+$\tau = 0.0487^{+0.0038}_{-0.0081}$
+for the PC method but using the prior inversion procedure
+discussed Section~\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
 [do we want to check whether the prior inversion procedure could cause a shift at this level?
@@
 \item $\tau = 0.0487+0.0038
-−0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).
+-0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).
 
@@
 \begin{enumerate}
-    \item {\textbf{Statement on prior in Planck 2018}:  ``Heinrich & Hu (2018) construct a prior that is uniform on τ, but which increases the allowed unphysical parameter space and is chosen a posteriori.
-Here we instead use the flat prior constructed by the procedure described in Millea & Bouchet (2018) and Handley & Millea (2019), which does not admit extra unphysical models and gives the most generic prior that leaves the prior on $\tau$ uniform."}
-
-    \item {Is this statement on tanh in Planck 2018 paper correct?} `` The TANH result gives slightly higher optical depth than the others, which is primarily driven by the fixed duration of reionization assumed."} \\
+    \item {\textbf{Statement on prior in Planck 2018}:  ``Heinrich \& Hu (2018) construct a prior that is uniform on $\tau$, but which increases the allowed unphysical parameter space and is chosen a posteriori.
+Here we instead use the flat prior constructed by the procedure described in Millea \& Bouchet (2018) and Handley \& Millea (2019), which does not admit extra unphysical models and gives the most generic prior that leaves the prior on $\tau$ uniform."}
+
+    \item {Is this statement on tanh in Planck 2018 paper correct?} ``The TANH result gives slightly higher optical depth than the others, which is primarily driven by the fixed duration of reionization assumed." \\
 

### 2021-03-13 paper/rei.tex
Updates from Overleaf

@@
 \centering
-\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
+\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$. \ch{transpose this table and add a row for taumj, and quote taufid in the caption}}
 \label{tab:PC_stats}
@@
 
+\wh{pick a consistent name for this model and use it for the section title, discussion, figure captions and labels - much of the discussion uses ``two step" which would be fine and may be better than high-z}
+
 \begin{figure}
@@
 \wh{Maybe add to the $\tau$ vs $\tau_{12}$ figure a version that does not have this $\tau_{12}$ cut.
-Discuss more if the difference significant - also mention that even in the direct tanh constraints, the impact of taking $z\gtrsim 6$ has the effect of removing some of the low $\tau$ region.}.
+Mention that even in the direct tanh constraints, the impact of taking $z\gtrsim 6$ also has the effect of removing some of the low $\tau$ region.}
+
+\wh{But it looks like from our conversation (3/12) that this is a small effect on tanh tau with srollv2.
+So instead discuss the impact of our priors - mostly it seems to remove negative tau contributions at high $z$ (from looking at the internal Fig 11) and the shifts relative to tanh tau are mostly just the ability to add a small amount of high z ionization that tanh excludes by its form.
+Rework this discussion below but I think it is compatible with our conclusion that that our conservative inclusion of some unphysical cases is not biasing our model independent results and instead it is the tanh model that carries a small bias due to its assumed form that excludes the small amount of high-z ionization that is still allowed by the data.}.
 
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
-\caption{$\tau_{12,new}$}
-}
+\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\caption{$\tau_{12,new}$ - \wh{this is an internal plot - leave in for now but also make a version for the paper: just $\tau_{12}$ vs $\tau_{\rm PC}$ in our new convention with and without prior either as a single or separate panel whichever is clearer.}}
+\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{/paper/plots/plot_tau12_new_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\caption{$\tau_{12,new}$ (no prior), with prior version is coming}
+\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
+\caption{$\tau$ decomposition (68\% C.L.
+only) \wh{this is a temporary plot for internal purposes - what I'm seeing here is that without the physicality cut the high $z$ side goes unphysically negative.
+I think this is consistent with our description and the $S_1$ curve.}}
 \label{fig:tau12}

### 2021-03-16 paper/rei.tex

@@
 Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
-\begin{equation}
-\tau(z,z_{\rm max}) = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
-\label{eq:cumtau}
+\begin{align}
+\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
+\end{align}
+where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.
+\wh{Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
+\begin{align}
+\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
+\label{eq:cumtauPC}
+\end{align}
+where
+\begin{align}
+    \tau_a(z,z_{\rm max}) &= 
+    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{S_a(z) (1+z)^{2} }{H_{\rm fid}(z)}, \nonumber\\
+    \tau_{\rm fid}(z,z_{\rm max}) &= 
+    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e^{\rm fid}(z) (1+z)^{2} }{H_{\rm fid}(z)},
 \end{equation}
-where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter.  
-\wh{Possibly add a few words on the PC version, list $\tau_a = \tau_a(z_{\rm min},z_{\rm max})$ in an equation or table, and refer to the
-bottom panel of Fig.~\ref{fig:xe}}
-The underlying reason is that CMB observables probe the optical depth and whether it is weighted toward high or low redshift through the amplitude and shape of 
-$C_\ell^{EE}$ -- the higher the redshift, the
-larger the relative contribution at higher $\ell$.
-
-Specific models essentially place an implicit prior on the redshift dependence of the cumulative optical depth.  
+ where ``fid" refers to evaluation in the fiducial cosmology.   When we omit the redshift arguments, we implicitly mean the total range, e.g. $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$. }
+
+ \wh{
+ In Fig.~\ref{fig:xe}, we display $\tau_a(z,z_{\rm max})$.
+ Notice that positive values of the first component $m_1$ mainly provides the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.
+Since these two are the best measured components, the CMB mainly determines the amount of low vs.~high redshift ionization.
+The underlying reason is that in $C_l^{EE}$ the higher the redshift, the
+larger the relative contribution at higher $l$ due to the size of the horizon when the photons scattered.
+For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth.}
+
+\wh{
+In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a model like a near step function reionization and then extracts a single constraint on the total optical depth.}
 %For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
@@
 \end{equation}
+\wh{Here we take $x_e^{\rm max} = 1+ n_{\rm He}/n_{\rm H}$ to account for the contribution of singly ionized helium.}
 In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
@@
  \begin{figure}
-          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau_prior_fac_0.8.pdf}
+          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau12.pdf}
           \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{/paper/plots/plot_tau12_new_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{$\tau_{12,new}$ (no prior), with prior version is coming}
+\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf
+}
+\caption{Internal (with prior version)}
+\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
+\caption{Internal (with prior version)}
+\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{/project2/whu5/chenhe/git/ReiPl18-Results/paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
+\caption{$\tau_{12,new}$ \ch{(took out the cosmology part, only shrunk all posteriors very slightly and shifted them a bit to the left ); with prior version now added, it uses the mean from m3, m4, m5 that are computed from the chains with prior as well.}}
 \label{fig:tau12}

### 2021-03-16 paper/rei.tex

@@
 $m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
-$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$0.018 \\
+$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$ \\
 $m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
@@
 %{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
-{plots/plot_tanh_gaussian_MCMC.png}
+{paper/plots/pl18_tau_posterior_tanh.pdf}
 \caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the full Planck likelihoods (solid), our effective likelihood in the kernel density estimate mode with $f = 0.14$ (dotted), and in the Gaussian approximation mode (dashed) \ch{tune plot and line styles}.
@@
 The truncation at 5 PCs causes wiggles at higher redshifts, but these cause negligible differences at most redshifts.
-}
+\ch{CH: plot is done}}
 \label{fig:plot_taugtz_two_step_contours}
@@
 Using Eqs.~(\ref{eq:mmutoxe}) and (\ref{eq:cumtau}), we can derive constraints on the total optical
-depth $\tau=\tau(0,z_{\rm max})$ with the physicality priors of Eq.~\ref{eq:individualprior}.
-These are shown in Fig.~XX \wh{figure could include $\tau$, $\tau_{12}$ and possibly $\tau_{12}$ without physicality priors?} and correspond to
+depth $\tau=\tau(0,z_{\rm max})$ with the physicality priors of Eq.~\ref{eq:individualprior}.  These are shown in Fig.~XX
+
+\wh{figure could include $\tau_{\rm PC}$, tanh tau and 2 step}
+and correspond to
 \beq
@@
 \eeq  
-\wh{comment on comparison to tanh tau, flexknot etc.}
 Notice that the PC mean, while still compatible with the tanh 
@@
 Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
-for each of its values, e.g. at  left corners (Fig.~\ref{fig:prior_box} solid square or lowest $m_1$) of the range $\tau_{12}=0.018$
-than at $\tau_{12} = 0.125$ \wh{revise numbers} and the implied prior increases linearly from zero between the two.
+for each of its values, e.g. at  left corners (Fig.~\ref{fig:prior_box} solid square or lowest $m_1$) of the range $\tau_{12}=0.023$
+than at $\tau_{12} = 0.130$ and the implied prior increases linearly from zero between the two.
    Ref.~\cite{Millea:2018bko} suggest 
@@
           \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
-           $\tau_{12}$ and \wh{range-bound flat - also if we change $\tau_{12}$ definitions change the numbers in the labels here} in $m_1-m_2$ (rotated rectangle, dashed lines) 
+           $\tau_{12}$ and in $m_1-m_2$ (rotated rectangle, dashed lines) 
            %\ch{CH: [why is this flat in m1 and m2?]} 
@@
 Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
-The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.}
+The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.
+\ch{CH: numbers changed according to new definition with mean of m3, m4, m5 obtained from chains with priors (shifted numbers by about 0.005 in tau.} \ch{CH: this plot is done}}
           \label{fig:prior_box}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf
-}
-\caption{Internal (with prior version)}
+\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\caption{ \ch{CH: this is the same as above but with priors; also tested that $\tau_{12, new}$ with mean of m3, m4, m5 computed from chains with prior applied performs better here (as expected) from chains without prior.}}
 \label{fig:tau12}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
-\caption{Internal (with prior version)}
-\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{/project2/whu5/chenhe/git/ReiPl18-Results/paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
-\caption{$\tau_{12,new}$ \ch{(took out the cosmology part, only shrunk all posteriors very slightly and shifted them a bit to the left ); with prior version now added, it uses the mean from m3, m4, m5 that are computed from the chains with prior as well.}}
+\includegraphics[width=0.5\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
+\caption{$\tau_{12}$ \ch{(CH: Official version used for paper; plot is done.\\
+More details: took out the cosmology part, only shrunk posteriors very slightly and shifted them a bit to the left for the top panel; with prior version now added in the bottom panel, it uses the mean from m3, m4, m5 that are computed from the chains with prior as well.)}}
 \label{fig:tau12}
@@
 
+\begin{figure}[ht]
+\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
+}
+\caption{\ch{Internal for now (with prior version)}}
+\label{fig:tau12}
+\end{figure}
+
+%
+%
+%
 \subsection{Cumulative optical depth}
@@
 The total $\tau$ distribution is broadened only toward the lower $\tau$.
-\ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.}
+\ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.} \ch{CH: plot is done}
 }
@@
 Both yield consistent high-redshift constraints, whereas the \texttt{srollv2} likelihood tightens the lower limit on the low-$z$ contribution, resulting on a tighter lower limit and higher mean for the total $\tau$.
+\ch{CH: plot is done}
 }
@@
 Both the low-redshift constraints and the high-redshift upper limits agree well.
+\ch{CH: plot is done}
 }

### 2021-03-16 paper/rei.tex

@@
  \begin{figure}
-          \includegraphics[width=0.95\columnwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
+          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
           \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
@@
 In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a model like a near step function reionization and then extracts a single constraint on the total optical depth.}
-%For example, the standard steplike model adopted in CAMB uses a tanh function to parameterize the hydrogen and singly ionized helium reionization
+%For example, the standard steplike model adopted in CAMB uses a width=0.48 function to parameterize the hydrogen and singly ionized helium reionization
 % \begin{equation}
-%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \tanh\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
+%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \width=0.48\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
 % \label{eqn:tanh}
@@
 \begin{figure}
-\includegraphics[width=0.40\textwidth]
+\includegraphics[width=0.48\textwidth]
 %{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
@@
 This posterior is symmetric and does not capture the small skewness found in the full and KDE chains.
-\ch{might need to rephrase.}
+\ch{might need to rephrase.
+\ch{This is plot is done.
+(might want to remove text label about zre, or add tanh in legend to mimick legend of Fig.
+5)}}
 }
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.
@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
+\includegraphics[width=0.48\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
 \caption{Example ionization history $x_e(z)$ for the high-$z$ model of \refsec{example2}.
@@
 %\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-\includegraphics[width=0.47\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
 \caption{\ch{Need to fix this plot.}   Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L.
@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
+\includegraphics[width=0.48\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.
@@
 %\begin{figure}
-%\includegraphics[width=0.5\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
+%\includegraphics[\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
 %\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
@@
 %\begin{figure}
-%\includegraphics[width=0.5\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
+%\includegraphics[\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
 %\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
 \caption{$\tau_{12,new}$ - \wh{this is an internal plot - leave in for now but also make a version for the paper: just $\tau_{12}$ vs $\tau_{\rm PC}$ in our new convention with and without prior either as a single or separate panel whichever is clearer.}}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
 \caption{ \ch{CH: this is the same as above but with priors; also tested that $\tau_{12, new}$ with mean of m3, m4, m5 computed from chains with prior applied performs better here (as expected) from chains without prior.}}
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
 \caption{$\tau_{12}$ \ch{(CH: Official version used for paper; plot is done.\\
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
 \caption{$\tau$ decomposition (68\% C.L.
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
 }
@@
 \begin{figure}[ht]
-\includegraphics[width=0.5\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.
@@
 \begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.

### 2021-03-16 paper/rei.tex

@@
 \includegraphics[width=0.48\textwidth]
-%{plots/pl18_pc_zmax30_pliklite_srollv2_1015_tau_posterior_fraccov_1p0_burnin_10000_yes_norm_gaussian0p1_0p12_0p14.pdf}
-{paper/plots/pl18_tau_posterior_tanh.pdf}
+%normalized by area
+%{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_area.pdf}
+%normalized by max
+{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
 \caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the full Planck likelihoods (solid), our effective likelihood in the kernel density estimate mode with $f = 0.14$ (dotted), and in the Gaussian approximation mode (dashed) \ch{tune plot and line styles}.
@@
 \ch{might need to rephrase.
-\ch{This is plot is done.
-(might want to remove text label about zre, or add tanh in legend to mimick legend of Fig.
-5)}}
+\ch{This plot is done.
+Changed convention to normalization by max; to restore figure normalized by area, see comments in latex.}}
 }
 \label{fig:tanh}
+
+\end{figure}
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]
+%normalized by area
+%{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_area.pdf}
+%normalized by max
+{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
+\caption{\ch{P(tau) PC (with and without prior) vs tanh vs tanh highz (normalized by max)}}
+%double check plot, upper and lower limits seem to cross?
+\label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}
 \end{figure}
@@
 The effective likelihoods results oscillate around zero at high-z because of the truncated representation of the models at 5 PCs (enough to represent any observable effects in the data).
-While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth.}
+While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth.
+\ch{add numbers in caption: 95\% (0.039, 0.02), best-fit (0.05290344, 0.005859431)}}
 %double check plot, upper and lower limits seem to cross?

### 2021-03-16 paper/rei.tex
All plots done.

@@
 \begin{figure}
-\includegraphics[width=0.40\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
 \caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane.
@@
 The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper.
+\ch{this plot is done.}
 }
@@
 While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth.
-\ch{add numbers in caption: 95\% (0.039, 0.02), best-fit (0.05290344, 0.005859431)}}
+\ch{this plot is done.}}
 %double check plot, upper and lower limits seem to cross?
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_xez_taulo_0p04_tauhi_0p02.png}
+\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
 \caption{Example ionization history $x_e(z)$ for the high-$z$ model of \refsec{example2}.
@@
 This is a toy model constructed for illustration purposes only.
-Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (XX, XX)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (XX, XX)$ (see also Fig.~\ref{two_parameter_model_2}  \wh{place x's on the models plotted in that plot}).
-\ch{Include 2nd helium in the plot.}
-}
+Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
+$(\taulo, \tauhi) = (0.039, 0.020)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
+\ch{CH: This plot is done}}
+%best-fit (0.05290344, 0.005859431)
+%95%CL (0.039, 0.02) 
 \label{fig:two_step_model}
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
-\caption{\ch{Need to fix this plot.}   Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L.
+\caption{Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L.
 contours for the two-step model  described in \refsec{example2} between different likelihoods: The full Planck likelihoods (shaded), the effective likelihoods in the KDE mode with f = 0.14 (blue solid) and in the Gaussian approximation mode (black dashed).
@@
 As in Fig.~\ref{fig:tanh}, both the KDE and Gaussian modes agree very well with the full likelihood results, with the KDE  contours being slightly broader as expected.
+\ch{To add in caption: best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
+$(\taulo, \tauhi) = (0.039, 0.020)$}  \ch{Otherwise this plot is done.}
 }
@@
 The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
-The ML model is \taulo = 0.0543 and \tauhi = 0.0054 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
+The ML model is \taulo = 0.0543 and \tauhi = 4 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
 %}

### 2021-03-22 paper/rei.tex

@@
 Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB; and will also have implications for inferring cosmic acceleration through the growth of structure.
-[add more refs here]
+\ch{[add more refs here]}
 
@@
 This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}.
-It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}[also cite our inflation papers?].
+It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}\ch{[also cite our inflation papers?]}.
 
@@
 Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019.
-This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from XX to XX.
-[fill in numbers]
+This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from \ch{XX to XX.
+[fill in numbers]}
 
@@
 
+\ch{write this part}
 The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].
@@
 Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release.
-But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots [fill in number and make sure to say its flex knot or PC] in Ref.~\cite{Aghanim:2018eyx}.
+But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots \ch{[fill in number]} in Ref.~\cite{Aghanim:2018eyx}.
 Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
@@
 The paper is structured as follows.
-We first describe in section~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
-Then in section~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature.
-Then in section~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
-Finally, we summarize our results and conclude in section~\ref{sec:conclusion}.
+We first describe in \S~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
+Then in \S~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature.
+Then in \S~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
+Finally, we summarize our results and conclude in \S~\ref{sec:conclusion}.
 
@@
 The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}.
-We now briefly summarize the PC technique in section~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in section~\ref{sec:KDE}.
-We refer the readers to Refs.~\cite{Hu:2003gh, Mortonson:2008rx, Mortonson:2007hq, Heinrich:2018btc}[pick 1-2] and Refs.~\cite{Heinrich:2016ojb} respectively for a more complete description on these topics.
-
-%
-%
- \begin{figure}
-          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
-          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
-We assume fully ionized Universe at $z < 6$.
-\textit{Bottom}: The cumulative optical depth resulting from a unit amplitude PC.
-\ch{plot done}}
-          \label{fig:xe}
-\end{figure}
+We now briefly summarize the PC technique in \S~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S~\ref{sec:KDE}.
+We refer the readers to Refs.~\cite{
+%Hu:2003gh, %Mortonson:2008rx, 
+Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.
+
+%
+%
 
@@
 %
-where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the variances of the PCs.
+where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the expected 
+%
+% added ``expected" to distinguish this from sigma(ma)
+%
+variances of the PCs.
 The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
- Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between the PC functions used in Ref.~\cite{Heinrich:2016ojb} and here.
-
-We then rank-ordered the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
+ Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between our $S_a$ and those of  Ref.~\cite{Heinrich:2016ojb}.
+
+We then rank-order the PCs from low to high variance, so that for the range $z_{\rm min} = 6$ (to be consistent with Ly$\alpha$ forest constraints, e.g.~\cite{Becker:2015lua}) and $z_{\rm max} = 30$, only the first 5 components are needed to describe all the information on $x_e$ carried by $C_\ell^{EE}$ to cosmic variance limit.
 In the work that follows, we therefore truncate the sum at $a = 1 ..
-n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ for the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S XX by  taking
+n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ in the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S \ref{sec:modelindependent} by  taking
  for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}
@@
  %
- \footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
+ % this seems to be in the main text now
+ %\footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
  %
@@
 We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
-For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a transition width $\Delta z = 0.5$ (see also \S \ref{sec:example1}).
+For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a tanh transition width $\Delta z = 0.5$, given the helium fraction
+ \beq
+ f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
+ \eeq
+as the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.
 
@@
 
+ \begin{figure}
+          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
+          \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
+We assume fully ionized Universe at $z < 6$.
+\textit{Bottom}: The cumulative optical depth from $z$ to $z_{\rm max}$ of a unit amplitude PC.
+}
+          \label{fig:xe}
+\end{figure}
+
 %
@@
 The completeness property of the PCs enables us to turn PC chains obtained from a MCMC run into an effective likelihood that can be used to test any reionization model with the CMB without a cumbersome reanalysis at the level of power spectra data.
-In the following, we briefly recap the kernel density estimate (KDE) technique used to build this likelihood and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+In the following, we briefly recap the kernel density estimate (KDE) technique used to build this likelihood, as well as a Gaussian approximation, and refer the readers to Ref.~\cite{Heinrich:2016ojb} for more details.
+
 The PC chains are composed of $N_{\rm sample}$ samples of discrete values of $\mathbf{m}_i = \{m_1, \ldots, m_5\}$ along with multiplicities $w_i$ for $i = 1$...$N_{\rm sample}$.
@@
 \beq
-{\cal L}_{\rm PC}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
+{\cal L}_{\rm KDE}\left({\rm data}|\mathbf{m} \right)  = \sum_{i = 1}^{N_{\rm sample}} w_i K_f(\mathbf{m}-\mathbf{m}_i),
 \eeq
@@
 %Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.
-Evaluating the posterior distribution of model parameters $\bf p$ using the effective likelihood
+
+An even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
+%
+\begin{equation}
+ {\cal L}_{\rm Gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{n_{\rm PC}} | \mathbf{C}| }}.
+ \label{eq:gaussian}
+ \end{equation}
+%
+
+For any set of model parameters ${\bf p}$ with their given prior probability distribution
+$P({\bf p})$, their posterior distribution can be approximated as 
 \begin{equation}
@@
 \end{equation}
-
+with PC\,=\,KDE or Gaussian.
 This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
-For comparison, an even simpler effective likelihood is the Gaussian approximation of the $m_a$ posterior using the PC mean $\bar{\bf m}$ and covariance $\bf C$
-%
-\begin{equation}
- {\cal L}_{\rm gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
- \label{eq:gaussian}
- \end{equation}
-%
-In section~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
+In \S~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
 for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
@@
 \tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
+\label{eq:cumtau}
 \end{align}
 where $n_{\rm H}(0)$ is the hydrogen number density at $z=0$, $\sigma_T$ is the Thomson scattering cross-section and $H(z)$ is the Hubble parameter. 
-\wh{Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
+Given the tight constraints on cosmological parameters in the $\Lambda$CDM model, this quantity is well approximated by
 \begin{align}
@@
 \end{align}
-where
-\begin{align}
-    \tau_a(z,z_{\rm max}) &= 
-    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{S_a(z) (1+z)^{2} }{H_{\rm fid}(z)}, \nonumber\\
-    \tau_{\rm fid}(z,z_{\rm max}) &= 
-    n_{\rm H}^{\rm fid}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e^{\rm fid}(z) (1+z)^{2} }{H_{\rm fid}(z)},
-\end{equation}
- where ``fid" refers to evaluation in the fiducial cosmology.   When we omit the redshift arguments, we implicitly mean the total range, e.g. $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$. }
-
- \wh{
+where $\tau_a$ and $\tau_{\rm fid}$ are defined by employing Eq.~({\ref{eq:mmutoxe}}) in
+Eq.~(\ref{eq:cumtau}) using the fiducial model 
+for $H(z)$ and number densities.
+ When we omit the redshift arguments, we
+implicitly mean the total range, e.g. 
+ $
+ \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
+ $
+ \ch{CH: what about recaling for different cosmological model?}
+
  In Fig.~\ref{fig:xe}, we display $\tau_a(z,z_{\rm max})$.
@@
 larger the relative contribution at higher $l$ due to the size of the horizon when the photons scattered.
-For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth.}
-
-\wh{
-In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a model like a near step function reionization and then extracts a single constraint on the total optical depth.}
-%For example, the standard steplike model adopted in CAMB uses a width=0.48 function to parameterize the hydrogen and singly ionized helium reionization
-% \begin{equation}
-%x_e^{\rm true}(z) = \frac{1+f_{\rm He}}{2}\left\{  1+ \width=0.48\left[ \frac{y(z_*)-y(z)}{\Delta y} \right] \right\},
-% \label{eqn:tanh}
-% \end{equation}
-% where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$, and $\Delta z = 0.5$,
-% excludes finite ionization above its transition redshift by definition.
+For the less well measured PCs with $a>2$, the contributions to the total optical depth rapidly diminish and represent finer distinctions in redshift for the cumulative optical depth and multipole for the underlying $C_l^{EE}$ spectrum.
+
+In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a ``tanh" or near step function reionization (see \S \ref{sec:example1}) and then extracts a single constraint on the total optical depth.
 The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed.
@@
 \end{equation}
-\wh{Here we take $x_e^{\rm max} = 1+ n_{\rm He}/n_{\rm H}$ to account for the contribution of singly ionized helium.}
+Here we take $x_e^{\rm max} = 1+ f_{\rm He} $ to account for the contribution of singly ionized helium.
 In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
@@
 %the Planck 2018 data \wh{and
-our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.
+our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.  We study its impact and robustness in \S \ref{sec:modelindependent}.
 
@@
 In other words, no physical model is excluded by these priors but some unphysical models are included.
-In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against direct constraints for two example models.
+In \S \ref{sec:note_on_priors} we further explore the role of these and other choices of PC priors, and in \S\ref{sec:effective_likelihood} we test PC constraints against exact constraints for two example models.
 
@@
 \includegraphics[width=0.8\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
-
 \caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$.
-We show in red the Planck 2015 and the improved Planck 2018 contours in blue which uses the low-$\ell$ $EE$ likelihood \texttt{srollv2}.
-The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L.
-regions in the $m_a-m_b$ planes in the lower triangle.
-The physicality priors on the ionization history used in this work are delineated by the box boundaries and the unshaded region for $m_1$.
-The black solid line indicates the tanh models with varying $\tau$.
+We show  the 68\% and 95\% $m_a-m_b$ constraints as well as marginal constraints on individual $m_a$ for Planck 2018 in blue,
+using \texttt{srollv2} for the low-$\ell$ $EE$ likelihood, and compare that to the Planck 2015 constraints in red.  The corresponding marginal distributions for individual $m_a$ are also shown.
+%The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle.
+When deriving model independent optical depth constraints we impose an additional physicality prior that excludes the gray region in $m_1$ and regions beyond the box limits in the others.
+ The black solid line indicates the trajectory of tanh models with varying $\tau$.
 The red and black crosses indicate the best-fit tanh models from Planck 2015 and 2018 respectively.
@@
 contours of the constraints.
-\ch{plot done}
 }
@@
 %
-In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$
-%
-\footnote{We The Planck 2015 best-fit is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Plnack 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $dz = 0.015(1+z)$ instead of $dz = 0.5$ for the Planck 2015 best-fit.
-Using $dz = 0.5$ for Planck 2018 moves the cross negligible on this plot.}.
-%
-The shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
-physicality prior from Eq.~(\ref{eq:individualprior}), whereas for the other cases the box itself represents this prior.
+%
+%
+In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
+We also show the trajectory of the tanh model (see \S \ref{sec:example1}) through this space as well as the best fit tanh model points.
+%
+\footnote{The Planck 2015 best-fit for the tanh model is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Planck 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $\Delta z = 0.015(1+z)$ instead of $\Delta z = 0.5$ for the Planck 2015 best-fit.
+Using $\Delta z = 0.5$ for Planck 2018 moves the cross negligibly on this plot.}
+%
+%
+%
+%
+The gray shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
+physicality prior from Eq.~(\ref{eq:individualprior}),
+which is just beyond the border of the displayed regime for all but $m_1$.
 Results using the Planck 2018 data are shown in blue.
 While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained.
-Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth.
-\wh{This preference also increases the impact of placing a physicality prior at low values of $m_1$.
+Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth, especially at high redshift.
+This preference also increases the impact of placing a physicality prior at low values of $m_1$ when deriving model-independent constraints as we shall discuss further in \S \ref{sec:modelindependent}.
 We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}.
-In this excluded region, the optical depth of models are dominated by the low redshift end, but due to the fixed ionization at $z>z_{\rm min}=6$, the desired total optical depth cannot be achieved without an unphysical ionization elsewhere.} \ch{[Not sure if we should say ``In this excluded region" referring to the entire gray band, or specify overlap with data constraints where $m_2 < 0$ mostly?]}
+In the physicality prior excluded region (gray), the ionization and the cumulative optical depth at high redshift becomes negative.
 
@@
 the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
-from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher total optical depth.
-
-\wh{possibly point to a figure below with $\tau_{12}$ in}
+from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher total optical depth (see also Fig.~\ref{fig:prior_box}).
+
+%We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area.
+[Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
+
+%\wh{possibly point to a figure below with $\tau_{12}$ in}
 
@@
 \includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane.
-We compare the Planck 2015 (gray) to the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue).
+\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane. We also 
+the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue).
 The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper.
-\ch{this plot is done.}
-}
+\ch{keeping srollv2 in label only here}
+} \ch{keep same color scheme as Fig. 2, use green maybe for simallEE}
 \label{fig:plot_m1m2_2015_vs_2018}
 \end{figure}
-
-\begin{table}[b]
-\centering
-\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$. \ch{transpose this table and add a row for taumj, and quote taufid in the caption}}
-\label{tab:PC_stats}
-\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
-\hline
-
-			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
-		\\ \hline
-
-$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
-$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
-$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$ \\
-$m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
-$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\
-
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}[b]
-\centering
-\caption{$\tau_a$}
-\label{tab:PC_stats}
-%\begin{tabular}{|r | r r r r r r|}
-\begin{tabular}{|c | c | c | c | c | c|}
-\hline
-$\tau_{\rm fid}$ & $\tau_1$  & $\tau_2$ & $\tau_3$ & $\tau_4$ & $\tau_5$ \\ \hline
-0.08626 & 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
-\hline
-\end{tabular}
-\end{table}
-% all digits without flipping m3, m4:
-%tau_fid 0.08625526676478058
-%taumj [ 0.29403315 -0.11227135 -0.04505736  0.0189787   0.00960184
-\begin{table}[b]
-\centering
-\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
-\label{tab:PC_stats}
-\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
-\hline
-
-			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
-		\\ \hline
-
-$m_1$ 
-	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
-$m_2$ 
-	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
-$m_3$ 
-	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
-$m_4$  
-	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
-$m_5$ 
-	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ 
-\hline
-\end{tabular}
-\end{table}
-
-We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area.
-[Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
 
@@
 However we shall see that a quantitative comparison shows that there is enough skewness in the posteriors that the KDE likelihood
-still performs noticeably better.  \wh{revisit this 
-statement once we firm up the comparison, e.g. what is noticeably better, means, 95\% confidence intervals, far tails?} \ch{yup, will do.}
+still performs noticeably better.  \wh{perhaps shift the tone of this statement since we seem to be concluding that Gaussian is good enough.}
 
@@
 
+\begin{table}[H]
+\centering
+\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$,  and correlation matrix $R_{ab}$. $\tau_a$ is the optical depth contribution for $m_a=1$ relative to 
+$\tau_{\rm fid}=0.08626$.}
+
+\label{tab:PC_stats}
+\begin{tabular}{|r |r r r r r|}
+\hline
+
+			  &
+%			  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 &
+\multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$} 
+		\\ \hline
+
+$m_1$ &
+%$-$0.116 & 0.030 & 
+1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
+$m_2$ & 
+%$-$0.016 & 0.063 & 
+0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
+$m_3$ &
+%$$0.078 & 0.098 & 
+$-$0.208 & 0.048 & 1.000 & 0.064 & $-$0.018 \\
+$m_4$ & 
+%$-$0.077 & 0.131 & 
+$-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
+$m_5$ &
+%0.066 & 0.145 & 
+0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\ \hline
+$\bar m_a$ &
+$-$0.116 & $-$0.016 & $$0.078 & $-$0.077 & 0.066  \\
+$\sigma(m_a)$ & 0.030 & 0.063 & 0.098 & 0.131 & 0.145 \\
+\hline
+$\tau_{a}$ &
+ 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
+\hline
+
+\hline
+\end{tabular}
+\end{table}
+
 \section{Model Testing with PCs}
@@
 We now turn to constraining reionization models with PCs using the
-effective likelihood approach described in~\refsec{KDE}.  The benefit of this approach is that
+effective likelihood approach described in~\refsec.  The benefit of this approach is that
 complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ can now be obtained without reanalyzing the Planck data. 
@@
 
-To use this approach, one simply specifies the functional form for $x_e(z)$ in the code, which then is converted to the PC parameters $m_a$ for the effective likelihood evaluation assuming a \ch{fiducial} $\Lambda$CDM cosmology.
+To use this approach, one simply specifies the functional form for $x_e(z)$ in the code, which then is converted to the PC parameters $m_a$ for the effective likelihood evaluation assuming the fiducial $\Lambda$CDM cosmology.
 To obtain reionization model parameter posteriors, one samples this likelihood with any desired priors.   
 We have also provided an MCMC sampler for models of high dimension (currently in CosmoMC). Note that the model parameters or priors must be arranged to explicitly satisfy 
-full ionization below $z=6$. To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
+full ionization below $z=6$.
+To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
+\wh{Is Gaussian part of this package or a separate package?
+either way should mention this too.}
 
@@
 {results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the full Planck likelihoods (solid), our effective likelihood in the kernel density estimate mode with $f = 0.14$ (dotted), and in the Gaussian approximation mode (dashed) \ch{tune plot and line styles}.
+\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the full Planck likelihoods (shaded), our effective likelihood in the kernel density estimate mode with $f = 0.14$ (blue solid), and in the Gaussian approximation mode (black dashed).
 The KDE mode agree well with the full results, while giving a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
@@
 \ch{might need to rephrase.
-\ch{This plot is done.
-Changed convention to normalization by max; to restore figure normalized by area, see comments in latex.}}
+\wh{drop f=0.14 label? it's the default - add a label that this is tanh} \ch{add text label for tanh, two step ... ; also add x, y axis label with tau tanh etc}
+}
 }
@@
 
-\end{figure}
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]
-%normalized by area
-%{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_area.pdf}
-%normalized by max
-{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{\ch{P(tau) PC (with and without prior) vs tanh vs tanh highz (normalized by max)}}
-%double check plot, upper and lower limits seem to cross?
-\label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}
 \end{figure}
@@
 While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth.
-\ch{this plot is done.}}
+}
 %double check plot, upper and lower limits seem to cross?
@@
  with $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z$.
-Instead of using $\Delta z = 0.5$ as in the standard setting by camb, we adopt a smaller width $\Delta z = 0.015(1+z)$ so as to better capture the required full ionization below $z=6$.
-Here $x_e^{\rm rec}$ is the ionization history from recombination only; the helium fraction
- \beq
- f_{\rm He} = \frac{n_{\rm He}}{n_{H}} = \frac{m_{\rm H}}{m_{\rm He}} \frac{Y_p}{1 - Y_p}, 
- \eeq
- is the ratio of the helium to hydrogen number density, where $Y_p$ is the helium mass fraction, chosen to be consistent with big bang nucleosynthesis for a given baryon density.
-In addition, the doubly ionized helium reionization is parameterized by an additional tanh function in redshift centered at $z_{\rm He} = 3.5$ with width $\Delta z = 0.5$.
-
-The $\tanh$ model is parameterized by its total optical depth $\tau$ for which we take a flat prior 
-with an additional constraint that $z_{\rm re}\ge 6.1$ so as to better satisfy the required $z_{\rm min}=6$ given $\Delta z$.
+Instead of using $\Delta z = 0.5$ as in the standard setting by CAMB, we adopt a smaller width $\Delta z = 0.015(1+z)$ so as to better capture the required full ionization below $z=6$.
+Here $x_e^{\rm rec}$ is the ionization history from recombination only and we follow the fiducial prescription described in \S \ref{sec:background} for the
+ doubly ionized helium transition at $z=3.5$.
+
+The $\tanh$ model is parameterized by its total optical depth for which we take a flat prior 
+with an additional constraint that $z_{\rm re}\ge 6.1$ so as to better satisfy the observationally required full ionization by $z_{\rm min}=6$ given the tanh width.
+%$\Delta z$.
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the trajectory that the $\tanh$ model takes in the PC space.
@@
 
+\ch{quote tau tanh KDE and gauss numbers with labels like $\tau_{tanh} = ... $(KDE).}
 In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which the five other $\Lambda$CDM parameters were also varied using the same priors for the $\tanh$ model.
@@
 This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution.
-As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ if we consider the 68\% lower and upper limits, whereas the full KDE result $\tau = 0.0588_{-0.0072}^{+0.0060}$ is able to capture the skewness in the distribution $\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.
+As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ if we consider the 68\% lower and upper limits, whereas the full KDE result $\tau = 0.0588_{-0.0072}^{+0.0060}$ is able to capture the skewness in the distribution
+\wh{seems like we should quote the direct results to make the claim that it is correctly capturing skewness - also,  is there a reason why we start with symmetric errors instead of just giving these as the main results (maybe also a displayed equation to highlight)?
+if it's to quote the effect of 7\% smoothing we could still do that on the average of + and - errors.}
+$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.
+\ch{Add $\tau(15,\zmax)$ 95\% upper limit for tanh direct MCMC and KDE.} \ch{Change direct MCMC to exact everywhere in text and plots.}
 
@@
 
-[Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.]
+%[Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.]
 
@@
 
-We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integration the exact tanh model in the full likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains.
-[describe more once plot is made].
-From these constraints it is also clear that the tanh model does not allow for high redshift ionization by construction, while the PC chains sampling the entire space of physical models would still allow for some high-$z$ ionization in the Planck 2018 data, as we shall see later in in Section~\ref{sec:results}.
-
-\subsection{Example 2: high-z model}
+We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integrating the tanh model in the exact likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains using Eq.~(\ref{eq:cumtauPC}).
+\wh{
+Notice that because of the truncation of the PC expansion at $a=5$, this representation shows an small oscillatory cumulative $\tau$ at $z \gtrsim 10$ whereas the true model has negligible ionization there.
+This is because the best constrained modes only represent the smooth features in the ionization history and truncation is similar to a low pass filter in Fourier space.
+Though these zero mean, small amplitude oscillations do not bias results when averaged over redshift, this effect should be born in mind when interpreting the model independent results below.
+}
+
+\wh{
+In the direct or in the effective likelihood approaches (ignoring these oscillations),
+we can see that the  tanh model does not allow for high redshift ionization by construction.
+As we shall next see in the next example the data themselves do allow for a small amount of high redshift ionization when the form of the model permits them.
+%As we shall see in \S \ref{sec:modelindependent} this is also captured in our model independent approach to cumulative optical depth constraints. 
+}
+
+\subsection{Example 2: the two-step model}
 \label{sec:example2}
 
-\wh{pick a consistent name for this model and use it for the section title, discussion, figure captions and labels - much of the discussion uses ``two step" which would be fine and may be better than high-z}
+\ch{need to label tanh highz as two step now in plots.}
 
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
-\caption{Example ionization history $x_e(z)$ for the high-$z$ model of \refsec{example2}.
+\caption{Example ionization history $x_e(z)$ for the two-step model of \refsec{example2}.
 In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$.
@@
 Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (0.039, 0.020)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
-\ch{CH: This plot is done}}
+$(\taulo, \tauhi) = (0.039, 0.020)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).}
 %best-fit (0.05290344, 0.005859431)
@@
 As in Fig.~\ref{fig:tanh}, both the KDE and Gaussian modes agree very well with the full likelihood results, with the KDE  contours being slightly broader as expected.
-\ch{To add in caption: best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (0.039, 0.020)$}  \ch{Otherwise this plot is done.}
+{To add in caption: best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
+$(\taulo, \tauhi) = (0.039, 0.020)$}  {Otherwise this plot is done.} \wh{Gaussian looks lumpy in the marginal distributions?
+also in 2D Gaussian may arguably be better than KDE.
+\ch{double check they are all smoothed the same way, and number of chain points}}
 }
@@
 limits on the total optical depth.
-The truncation at 5 PCs causes wiggles at higher redshifts, but these cause negligible differences at most redshifts.
-\ch{CH: plot is done}}
+The truncation at 5 PCs causes wiggles at higher redshifts, but these cause negligible differences at most redshifts.}
 \label{fig:plot_taugtz_two_step_contours}
@@
 We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
-We show an example
+We show examples
 of this two-step model in Fig.~\ref{fig:two_step_model}.
@@
 [double check this too.]
-Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given total optical depth $\tau$.
-For both the full likelihood and effective likelihood runs, we adopt the priors $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$.
+Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$.
+For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$.
 %except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
-We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains. Here we used again $f = 0.14$ in the KDE computation.
-
-In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$. They agree very well. 
-\wh{Not sure if we need this comment - was more important when the width was larger - also not sure if this is the explanation for the visible effect - getdist is smoothing things and that might be more important.}
-Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
+We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model.
+Here we used again $f = 0.14$ in the KDE computation.
+\wh{if we never show anything but $f=0.14$ we could just say it's always this value in the method section}
+
+In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.\wh{The direct, KDE and Gaussian PC treatments again agree very well.} 
+Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors distort near this value.
+The models in Fig.~\ref{fig:two_step_model} are represented by the star and cross symbols for the best fit case and a model allowed at the 95\% CL.
+
+% Not sure we need to get into this subtlety - the caveat being that getdist will smooth the boundary even if it is in reality sharper.
+Not sure we know which effect dominates and it probably doesn't matter much:
+%
+%; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
 This is why the cut on the 2D constraint ellipses at low $\taulo$ is of a curved shape rather than a straight cut, because the $\zre$ prior corresponds to a different value of $\taulo$ for a different $\tauhi$.
@@
 The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$.
-\wh{will be interesting to see if the numbers are more consistent with the $f$ smoothing than the visual plot is - that might help us see if getdist is responsible for extra smoothing that hides things.}
 The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
+\wh{Note also that the model 95\% C.L. allowed model in 
+Fig.~\ref{fig:two_step_model} has
+$\tau(15,30)= ZZ$.}
 
 The Planck 2018 paper~\cite{Aghanim:2018eyx} also constrained $\tau(15, 30)$ in a model-independent way using the FlexKnot method, in which $x_e(z_i)$ at the $i$-th knots are varied as free parameters as well as the number of total knots.
-They obtained $\tau(15, 30) < 0.007$ (95\% C.L.), which is not consistent with the results we obtain for the two-parameter models.
+They obtained $\tau(15, 30) < 0.007$ (95\% C.L.), which would strongly rule out models that are still allowed at 95\% C.L.
+in this class.
+\wh{
  In principle, the FlexKnot method allows for all possible physical ionization histories that increase monotonically with redshifts, and would include the two-parameter model described here.
-But it seems that the FlexKnot method gave more stringent upper limits on high-redshift optical depth in the analysis of Ref.~\cite{} than actually allowed by the data.
-
-In Ref.~\cite{Aghanim:2018eyx}, the authors also suggested that the PC method has limitations due to the fact that it includes some unphysical models in addition to all the physical models.
+But apparently the implicit prior employed by the method disfavors optical depth at high redshift more than a flat prior in
+$\tau_{\rm hi}$ does.
+We shall see in \S \ref{sec:modelindependent} that our model independent PC approach models such a flat prior better and of course the effective likelihood approach allows for any desired prior.}
+
+% Move this somewhere into the prior discussion or the 
+% paper discussion
+%
+%In Ref.~\cite{Aghanim:2018eyx}, the authors also suggested that the PC method has limitations due to the fact that it includes some unphysical models in addition to all the physical models.
 The authors gave this reason for not using PCs to constrain reionization histories.
@@
 In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this 2-parameter model.
-In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ integrated over the exact physical models in the full likelihood chains themselves, as well as those obtained from the 5-PC representation of those physical models in the effective likelihood chains.
-Because the truncated 5 PCs may represent non-physical models (higher PCs can restore the physicality of the model), negative $\tau$ values are allowed at some redshifts, which is of course, unphysical.
+In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ integrated over the exact two-step model in the full likelihood chains themselves, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.
+\wh{As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.
+Notice again the amount of high redshift optical depth that is allowed in the two step model vs the tanh model (cf.~Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).}
+%
+%of those physical models in the effective likelihood chains.
+%Because the truncated 5 PCs may represent non-physical models (higher PCs can restore the physicality of the model), negative $\tau$ values are allowed at some redshifts, which is of course, unphysical.
 Nevertheless, the full vs truncated PC chains give similar 95\% upper limits on the high-redshift optical depth, $\tau(15, 30) < XX$ and $\tau(15, 30) < XX$ respectively.
 We have showed that the discrepancy between the PC-based effective likelihood results and the FlexKnot results are not due to the inclusion of non-physical models.
-
-With this in mind, we will now turn to the model-independent constraints derived from the PC analysis.
+\wh{
+With this in mind, we will now turn to the model-independent constraints derived from the PC analysis to generalize these results on high redshift optical depth beyond the two-step model.}
 
@@
 Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts.
-
-[add plot for $\tau(>z)$ contours for the 2 parameter model for direct MCMC vs KDE chains]
 
@@
 
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]
+%normalized by area
+%{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_area.pdf}
+%normalized by max
+{paper/plots/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
+\caption{Posterior for the total optical depth in the PC analysis (shaded) and for a tanh (black solid line) and two-step model (blue dashed line)}. 
+\label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}
+\label{fig:tauPC}
+\end{figure}
+
 \section{Model-Independent Results for Optical Depth}
+\label{sec:modelindependent}
 
@@
 testing wide classes of models efficiently and with their own physically motivated priors, it can also be employed to extract model-independent constraints on the total and cumulative optical depth. 
-As discussed in \S \ref{sec:cumulative}, since the PC amplitudes are not intrinisically required
-to produce physically sensible ionization histories where $0 \le x_e(z) \le x_e^{\rm max}$, this requires imposing physicality priors using Eq.~\ref{eq:individualprior}.
-In addition these flat range-bound priors on the PC amplitudes $m_a$, imply implicit priors on the optical depth.
-Finally our fiducial analysis has an explicit prior that reionization occured
-at $z<z_{\rm max}=30$.   In this section we present these model-independent results and discuss their robustness to these explicit and implicit priors.
+% this sounds too technical now for an intro to the section:
+%As discussed in \S \ref{sec:cumulative}, since the PC amplitudes are not intrinisically restricted to physically sensible ionization histories where $0 \le x_e(z) \le x_e^{\rm max}$, this requires imposing physicality priors using Eq.~\ref{eq:individualprior}.
+In this section we present these model-independent results, first on the total optical depth and then on the cumulative optical depth at high redshift, and discuss their robustness to priors  on physicality of the ionization history and redshift range of the fiducial analysis.
 }
@@
 \subsection{Total optical depth}
+
 \label{sec:note_on_priors}
 
-Using Eqs.~(\ref{eq:mmutoxe}) and (\ref{eq:cumtau}), we can derive constraints on the total optical
-depth $\tau=\tau(0,z_{\rm max})$ with the physicality priors of Eq.~\ref{eq:individualprior}.  These are shown in Fig.~XX
-
-\wh{figure could include $\tau_{\rm PC}$, tanh tau and 2 step}
-and correspond to
+Using Eq.~(\ref{eq:cumtauPC}), we can derive constraints on the total optical
+depth $\tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max})$ with the additional physicality priors on the PC amplitudes $m_a$ from
+Eq.~\ref{eq:individualprior}.  These 
+correspond to 
 \beq
@@
 \eeq  
-Notice that the PC mean, while still compatible with the tanh 
-tau derived constraints, is almost $0.5\sigma$ higher...
-
-The difference between these results come down to implicit and explicit priors.
-Results based on the tanh model requires that reionization happens suddenly at $z_{\rm re}$ and is maintained thereafter.
+and the full posterior is shown in Fig.~\ref{fig:tauPC}. 
+Notice that the distribution matches that of the two step model from \S \ref{sec:example2} but almost $0.5\sigma$ shifted higher than that of the tanh model from \S \ref{sec:example1}.
+
+The similarity and difference between these results come down to implicit and explicit priors. As we have seen, 
+results based on the tanh model requires that reionization happens suddenly at $z_{\rm re}$ and is maintained thereafter.
 Therefore, the model has a strong implicit prior that a given a total optical depth $\tau$ comes from $z\lesssim z_{\rm re}$ regardless of whether the data allow or prefer
@@
 the PC constraints gave $\tau=XX\pm YY$ which differed substantially from the tanh $\tau=XX\pm YY$.  As we shall discuss
-below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift.
+below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.
 
@@
 
- We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
-\beq
-\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
-\eeq
-where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
+\caption{Posterior of optical depth $\tau_{\rm PC}$ from reionization PCs (blue solid line) vs that of $\tau_{12}$ (shaded) defined in Eq.~\ref{eq:tau12}.
+\ch{$\tau_{\rm PC}$ here is not marginalized over cosmological parameters, in contrast to Figs.~\ref{...}.
+We show the versions without prior (top panel) vs with physicality prior (bottom panel).}\ch{CH: Add more description...}}
+\label{fig:tau12}
+\end{figure}
+
+\wh{
+ We first note that by virtue of Eq.~(\ref{eq:cumtauPC}), the total optical depth receives contribution from all PCs.  However
+ %
+%\beq
+%\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
+%\eeq
+%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
 The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
@@
 \beq
-\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
+\tau_{\rm PC} \approx \tau_{12} \equiv  m_1 \tau_1 + m_2 \tau_2 +
+\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) 
+\label{eq:tau12}
 \eeq
-\wh{possibly redefine
-\beq
-\tau_{12} =  m_1 \tau_1 + m_2 \tau_2 +
-\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) \approx \tau.
-\eeq
-where the term in parentheses is a fixed quantity (refer to eqns/table where tau's and mbars are defined) whose purpose is to remove offsets due to the arbitrary choice of fiducial model.
-}
+where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of fiducial model.
+We explicitly demonstrate this approximate equivalence in Fig.~\ref{fig:tau12} by comparing the
+posterior distributions of $\tau_{\rm PC}$ and $\tau_{12}$ both with and without the physicality prior.  
+}\ch{CH: will add comment on mean from m3, m4, m5 that are computed from the chains with prior as well}
+
 %
@@
 Due to this linearity, flat priors in $m_1$ and $m_2$ correspond to
-flat priors in $\tau_{12}$ for any linear trajectory through the space.   For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015}.   Even without the explicit 
+flat priors in $\tau_{12}$ for any linear trajectory through the space.
+For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015} and is nearly linear through the Planck 2018 allowed region.
+Even without the explicit
 construction of a flat tanh $\tau$ prior in \S \ref{sec:example1}, 
-the implicit prior from $m_1,m_2$ is already nearly flat. \ch{CH: need to ask a question about this.}
+the implicit prior from $m_1,m_2$ is already nearly flat.
 
@@
 for each of its values, e.g. at  left corners (Fig.~\ref{fig:prior_box} solid square or lowest $m_1$) of the range $\tau_{12}=0.023$
-than at $\tau_{12} = 0.130$ and the implied prior increases linearly from zero between the two.
-   Ref.~\cite{Millea:2018bko} suggest 
-inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the \ch{Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}}.
+than at $\tau_{12} = 0.130$ and the implied prior increases linearly from zero between the two.   Ref.~\cite{Millea:2018bko} suggest 
+inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}.
 First notice that this inversion would in fact then produce a non-flat prior on $\tau$ for the tanh model
@@
 removes nearly half of the space that would otherwise be allowed by the
-data at 95\% CL.
-This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).
-\wh{Maybe add to the $\tau$ vs $\tau_{12}$ figure a version that does not have this $\tau_{12}$ cut.
-Mention that even in the direct tanh constraints, the impact of taking $z\gtrsim 6$ also has the effect of removing some of the low $\tau$ region.}
-
-\wh{But it looks like from our conversation (3/12) that this is a small effect on tanh tau with srollv2.
-So instead discuss the impact of our priors - mostly it seems to remove negative tau contributions at high $z$ (from looking at the internal Fig 11) and the shifts relative to tanh tau are mostly just the ability to add a small amount of high z ionization that tanh excludes by its form.
-Rework this discussion below but I think it is compatible with our conclusion that that our conservative inclusion of some unphysical cases is not biasing our model independent results and instead it is the tanh model that carries a small bias due to its assumed form that excludes the small amount of high-z ionization that is still allowed by the data.}.
-
-As discussed in \S \ref{sec:cumulative}, our range-bound priors 
-are conservative in the sense that all the cases that are excluded are definitely unphysical but some of the cases that are included
-may also be unphysical since their physicality depends on higher order PC components that are explicitly truncated in the analysis.
-On the other hand the tanh model provides an explicitly physical example whose trajectory runs close  and nearly parallel to the our physicality prior in the allowed region and so we do not expect the bias incurred by
-including these possibly \wh{unphysical cases} to be large. 
+data at 95\% CL.  This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).  
+This is reflected in the upward shift of the $\tau_{12}$ posterior in Fig.~\ref{fig:tau12} with and without the physicality prior on $m_1$.
+This prior excludes very negative $m_1$ and we can see from Fig.~\ref{fig:xe} that these cases would have an unphysically negative ionization and cumulative optical depth at high $z$.
+More precisely, even though a very positive value
+of $m_2$ can restore physicality at the highest redshifts, the fact that there is a node near $\tau_2(15,z_{\rm max})$ means that no choice can counteract $m_1$ there.  
+Correspondingly, the impact of the $m_1$ prior is to remove
+cases where the high redshift contribution is negative which would otherwise broaden and shift the $\tau_{12}$ posterior to lower values.
+
+\wh{
+This also shows how our range-bound priors 
+are conservative: all the cases that are excluded are definitely unphysical but some of the cases that are included
+may also be unphysical since their physicality depends on higher order PC components.
+The impact on optical depth  however is small at at most contributes a small bias that conservatively {\it lowers} the high redshift contribution which is already a small contribution to the total.
+This can be seen  in the comparison with the two-step model 
+in Fig.~\ref{fig:tauPC} and as we shall return to this point in the cumulative optical depth where the high and low redshift contributions are manifest.}
+%On the other hand the tanh model provides an explicitly physical example whose trajectory runs close  and nearly parallel to the our physicality prior in the allowed region and so we do not expect the bias incurred by
+%including these possibly \wh{unphysical cases} to be large. 
 %\ch{CH: sorry, what do we mean by these models here?}  
@@
 Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
-The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.
-\ch{CH: numbers changed according to new definition with mean of m3, m4, m5 obtained from chains with priors (shifted numbers by about 0.005 in tau.} \ch{CH: this plot is done}}
+The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.}
           \label{fig:prior_box}
@@
 
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{$\tau_{12,new}$ - \wh{this is an internal plot - leave in for now but also make a version for the paper: just $\tau_{12}$ vs $\tau_{\rm PC}$ in our new convention with and without prior either as a single or separate panel whichever is clearer.}}
-\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{ \ch{CH: this is the same as above but with priors; also tested that $\tau_{12, new}$ with mean of m3, m4, m5 computed from chains with prior applied performs better here (as expected) from chains without prior.}}
-\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
-\caption{$\tau_{12}$ \ch{(CH: Official version used for paper; plot is done.\\
-More details: took out the cosmology part, only shrunk posteriors very slightly and shifted them a bit to the left for the top panel; with prior version now added in the bottom panel, it uses the mean from m3, m4, m5 that are computed from the chains with prior as well.)}}
-\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
-\caption{$\tau$ decomposition (68\% C.L.
-only) \wh{this is a temporary plot for internal purposes - what I'm seeing here is that without the physicality cut the high $z$ side goes unphysically negative.
-I think this is consistent with our description and the $S_1$ curve.}}
-\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
-}
-\caption{\ch{Internal for now (with prior version)}}
-\label{fig:tau12}
-\end{figure}
-
 %
@@
 
-The differences between model dependent and model independent approaches are larger for the allowed cumulative optical depth $\tau(z, \zmax)$ at high $z$
-compared to the total optical depth of the previous section.
+The value of a model independent approach is even more apparent for constraints on the
+cumulative optical depth $\tau(z, \zmax)$ at high $z$.
 The main results are again related to the $m_1$ and $m_2$ PC constraints. 
@@
 raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
-Consequently, as shown in Fig.~\ref{fig:prior_box}, the second aspect of PC constraints beyond the well constrained total optical depth, is a positive
+Consequently, as shown in Fig.~\ref{fig:prior_box}, the second aspect of PC constraints beyond the well constrained total optical depth, as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12}), is a positive
 correlation between the two components that represents the high to low $z$ 
-contribution to the total optical depth.  Any finer detail in the ionization history is much more poorly constrained.
-
-\ch{[CH: Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior} needs to go somewhere]}
-% with this prior you still have that $\tau(>z)$ plot includes unphysical models (however, most likely total tau doesn't change, and if any bias from unphysical models, it would lower high-z tau).\\
-
-In Fig.~\ref{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2} top panel, we compare the 68\% and 95\% confidence level contours in $\tau(z, \zmax)$ for the Planck 2015 (red regions) and the Planck 2018 results (solid lines) using all 5 PCs.
-\wh{remind the reader about impact of PC filtering smoothing out features in redshift - refer to the tanh examples to illustrate}
-In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.   Notice first the dramatic improvement in the
-constraint at high-$z$.  Whereas the Planck 2015 allowed and even preferred
-finite optical depth at $z\gtrsim 15$, Planck 2018 no longer does....
-On the other hand, some optical depth is still allowed 
+contribution to the total optical depth.  Any finer redshift detail in the ionization history is much more poorly constrained.
+
+In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
+the 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.
+Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.
+Moreover upper bounds on the optical depth at high
+redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example
 \beq
-\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}),
+\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}).
 \eeq
-in contrast with the tanh model where it is strictly forbidden in the direct analysis and in the PC filtered effective analysis only enters due to the smoothness of the first few PCs in redshift space.
-\wh{contrast this with Flexknot and talk about compatibility with example 2.}
+This should be contrasted with the tanh model 
+ where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
+ On the other hand the bound is comparable to that obtained for the two-step model 
+ \wh{hopefully true, point to the eqn when we get the numbers -- contrast this with Flexknot and talk about compatibility with example 2.}
+
+These results are in contrast to those of the Planck 2015 analysis. 
+In Fig.~\ref{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2} top panel, we compare
+the two.  Whereas the Planck 2018 has only upper limits on the high redshift optical depth, Planck 2015 actually preferred
+finite optical depth at $z\gtrsim 15$.
+
+In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.
+Improvements from srollv2 come mainly from improving the lower bound at low redshift, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.
 
@@
 The total $\tau$ distribution is broadened only toward the lower $\tau$.
-\ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.} \ch{CH: plot is done}
+\ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.}
+\wh{drop the ``with prior" label since with prior is our default approach}
 }
@@
 Both yield consistent high-redshift constraints, whereas the \texttt{srollv2} likelihood tightens the lower limit on the low-$z$ contribution, resulting on a tighter lower limit and higher mean for the total $\tau$.
-\ch{CH: plot is done}
 }
@@
 Both the low-redshift constraints and the high-redshift upper limits agree well.
-\ch{CH: plot is done}
 }
@@
 for the PC method but using the prior inversion procedure
-discussed Section~\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
+discussed \S~\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
 [do we want to check whether the prior inversion procedure could cause a shift at this level?
@@
 Compared to our PC results ($\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\% \mathrm{CL})$), the FlexKnot upper limits are too stringent.
-The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in Section~\ref{sec:example2} with a concrete example of the two-step model.
+The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in \S~\ref{sec:example2} with a concrete example of the two-step model.
 We showed that for this chosen physical model, the $\tau(15, 30)$ upper limits derived from the full Planck likelihood without using any PCs were already not compatible with the FlexKnot upper limits, while the PC results were compatible.
@@
 
+\section{Reference Plots for Internal Checks}
+
+\begin{figure}
+\includegraphics[width=0.48\textwidth]{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
+\caption{\ch{caption to be added, will take away without prior line}}
+\label{fig:tau_posterior_PC_vs_tanh_vs_tanh_highz}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\caption{$\tau_{12,new}$ - \wh{this is an internal plot - leave in for now but also make a version for the paper: just $\tau_{12}$ vs $\tau_{\rm PC}$ in our new convention with and without prior either as a single or separate panel whichever is clearer.}}
+%\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\caption{ \ch{CH: this is the same as above but with priors; also tested that $\tau_{12, new}$ with mean of m3, m4, m5 computed from chains with prior applied performs better here (as expected) from chains without prior.}}
+%\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
+\caption{$\tau$ decomposition (68\% C.L.
+only) \wh{this is a temporary plot for internal purposes - what I'm seeing here is that without the physicality cut the high $z$ side goes unphysically negative.
+I think this is consistent with our description and the $S_1$ curve.}}
+%\label{fig:tau12}
+\end{figure}
+
+\begin{figure}[ht]
+\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
+\caption{\ch{Internal for now (with prior version)}}
+%\label{fig:tau12}
+\end{figure}
+
+\begin{table}[b]
+\centering
+\caption{$\tau_a$}
+%\label{tab:PC_stats}
+%\begin{tabular}{|r | r r r r r r|}
+\begin{tabular}{|c | c | c | c | c | c|}
+\hline
+$\tau_{\rm fid}$ & $\tau_1$  & $\tau_2$ & $\tau_3$ & $\tau_4$ & $\tau_5$ \\ \hline
+0.08626 & 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
+\hline
+\end{tabular}
+\end{table}
+
+\begin{table}[b]
+\centering
+\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$. \ch{transpose this table and add a row for taumj, and quote taufid in the caption}}
+%\label{tab:PC_stats}
+\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
+\hline
+
+			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
+		\\ \hline
+
+$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
+$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
+$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$ \\
+$m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
+$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\
+
+\hline
+\end{tabular}
+\end{table}
+
+% all digits without flipping m3, m4:
+%tau_fid 0.08625526676478058
+%taumj [ 0.29403315 -0.11227135 -0.04505736  0.0189787   0.00960184
+\begin{table}[b]
+\centering
+\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
+%\label{tab:PC_stats}
+\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
+\hline
+
+			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
+		\\ \hline
+
+$m_1$ 
+	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
+$m_2$ 
+	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
+$m_3$ 
+	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
+$m_4$  
+	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
+$m_5$ 
+	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ 
+\hline
+\end{tabular}
+\end{table}
+
 \end{document}

### 2021-03-22 paper/rei.tex

@@
 {paper/plots/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Posterior for the total optical depth in the PC analysis (shaded) and for a tanh (black solid line) and two-step model (blue dashed line)}. 
+\caption{Posterior of the total optical depth for the PC analysis (shaded), the tanh model (black solid line) and the two-step model (blue dashed line)}. 
 \label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}

### 2021-03-22 paper/rei.tex
Updates from Overleaf

@@
 \includegraphics[width=0.48\textwidth]
-%normalized by area
-%{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_area.pdf}
-%normalized by max
-{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the full Planck likelihoods (shaded), our effective likelihood in the kernel density estimate mode with $f = 0.14$ (blue solid), and in the Gaussian approximation mode (black dashed).
-The KDE mode agree well with the full results, while giving a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
+{paper/plots/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
+\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the exact likelihood (shaded) and our effective likelihoods in the kernel density estimate (KDE) mode (blue solid) and the Gaussian approximation mode (black dashed).
+The KDE mode agree well with the exact results, while giving a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
 The Gaussian likelihood works well too, with no extra broadening of the posterior, but a negligible shift in the mean ($\sim XX\%$ of the error in $\tau$).
-This posterior is symmetric and does not capture the small skewness found in the full and KDE chains.
-\ch{might need to rephrase.
-\wh{drop f=0.14 label? it's the default - add a label that this is tanh} \ch{add text label for tanh, two step ... ; also add x, y axis label with tau tanh etc}
-}
+This posterior is symmetric and does not capture the small skewness found in the exact and KDE chains.
+\ch{might need to rephrase.}
 }
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian_dz_auto_zre_prior_6p1.pdf}
+\includegraphics[width=0.48\textwidth]
+{paper/plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.

### 2021-03-22 paper/rei.tex

@@
 \ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.}
-\wh{drop the ``with prior" label since with prior is our default approach}
 }

### 2021-03-22 paper/rei.tex

@@
 constraints on the cumulative optical depth with different Planck likelihoods.
-\textit{Top}: The Planck 2015 vs Planck 2018 likelihoods, the latter including the official low-$\ell$ $EE$ likelihood $\texttt{simallEE}$.
-\textit{Bottom}: The Planck 2018 likelihoods using $\texttt{simallEE}$ for the low-$\ell$ $EE$ likelihood vs \texttt{srollv2} which has better foreground cleaning.
+\textit{Top}: The Planck 2015 vs Planck 2018 likelihoods, the latter including the official low-$\ell$ $EE$ likelihood $\texttt{simall\_EE}$.
+\textit{Bottom}: The Planck 2018 likelihoods using $\texttt{simall\_EE}$ for the low-$\ell$ $EE$ likelihood vs \texttt{srollv2} which has better foreground cleaning.
 Both yield consistent high-redshift constraints, whereas the \texttt{srollv2} likelihood tightens the lower limit on the low-$z$ contribution, resulting on a tighter lower limit and higher mean for the total $\tau$.
@@
 In conclusion, we used constraints on reionization principal components with respect to the large-scale $EE$ data to produce a fast and accurate effective likelihood for the final release of Planck 2018 data.
-In lieu of the official $\texttt{simallEE}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{srollv2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
+In lieu of the official $\texttt{simall\_EE}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{srollv2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
 Our effective likelihood code package is called $\textsc{Relike}$ [to be changed] and is available publicly on Github.

### 2021-03-22 paper/rei.tex

@@
 %: 1) the tanh model with a single transition redshift, and 2) a two-parameter model with two transitions allowing for a high-redshift ionization plateau. 
-which showed good agreement with directly 
+which showed good agreement with
 %The Relike chains converged 300$\times$ and 100$\times$ faster for model 1 and 2 respectively, and showed good agreement compared to directly
-sampling the original Planck likelihoods.
+sampling the exact Planck likelihoods.
 We also find that contrary to the Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation.
@@
 This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
-In \S~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian approaches against a direct implementation
+In \S~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
 for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
@@
 
-Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to a direct MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has one additional parameter modeling a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.
+Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to an exact MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has one additional parameter modeling a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.
 
@@
 if it's to quote the effect of 7\% smoothing we could still do that on the average of + and - errors.}
-$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.
-\ch{Add $\tau(15,\zmax)$ 95\% upper limit for tanh direct MCMC and KDE.} \ch{Change direct MCMC to exact everywhere in text and plots.}
+$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}. \ch{Add $\tau(15,\zmax)$ 95\% upper limit for tanh direct MCMC and KDE.}
 
@@
 \wh{
-In the direct or in the effective likelihood approaches (ignoring these oscillations),
+In the exact or in the effective likelihood approaches (ignoring these oscillations),
 we can see that the  tanh model does not allow for high redshift ionization by construction.
@@
 \label{sec:example2}
-
-\ch{need to label tanh highz as two step now in plots.}
 
@@
 
-In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.\wh{The direct, KDE and Gaussian PC treatments again agree very well.} 
+In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.\wh{The exact, KDE and Gaussian PC treatments again agree very well.} 
 Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors distort near this value.
@@
 
+\ch{Might need to change all simallEE labels to lowE in plots and in text}
+
 %
@@
 To test this effective Planck likelihood code, we used two examples: 1) a canonical tanh model with a single transition redshift, and 2) a two-parameter model consisting of two tanh transitions allowing in addition for a high-redshift ionization plateau compared to the first example.
-We demonstrated accuracy of the Relike code for both examples by comparing the constraints on the reionization model parameters with those obtained from directly sampling the full Planck likelihood and using a modified CAMB code.
+We demonstrated accuracy of the Relike code for both examples by comparing the constraints on the reionization model parameters with those obtained from sampling the exact Planck likelihoods and using a modified CAMB code.
 We also showed great agreement between the confidence levels of the cumulative optical depth evolution between the 5-PC representation and the exact ionization model.

### 2021-03-26 paper/rei.tex
Updates from Overleaf

@@
 \newcommand{\ch}[1]{\textcolor{red}{#1}}
+\definecolor{darkgreen}{rgb}{0.0, 0.2, 0.13}
+\newcommand{\todo}[1]{\textcolor{darkgreen}{\bf{#1}}}
 
@@
 The paper is structured as follows.
-We first describe in \S~\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
-Then in \S~\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature.
-Then in \S~\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
-Finally, we summarize our results and conclude in \S~\ref{sec:conclusion}.
+We first describe in \S\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
+Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature.
+Then in \S\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
+Finally, we summarize our results and conclude in \S\ref{sec:conclusion}.
 
@@
 The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}.
-We now briefly summarize the PC technique in \S~\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S~\ref{sec:KDE}.
+We now briefly summarize the PC technique in \S\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S\ref{sec:KDE}.
 We refer the readers to Refs.~\cite{
@@
 To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples.
-% 1.06 cut w/ physicality prior
-% 1.072 no cut
-For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient.
-%Note also that we employ PC chains without the physicality priors allowing the smoothing kernel to cross over these priors.
+\ch{For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient for most models, and indeed we have tested it to work well for the ones we study in \S~\ref{sec:effective_likelihood}.}
 
@@
 This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
-In \S~\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
+In \S\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
 for two example models: the standard steplike model and a two-step model allowing for high-redshift ionization.
@@
  $
- \ch{CH: what about recaling for different cosmological model?}
+ \ch{CH: what about recaling for different cosmological model? Our tau(z,zmax) plots have cosmological parameters marginalized, while Figs 9-10 (total tau posterio) do not.}
+ \wh{They probably shouldn't, i.e should be constructed directly out of $m_a$, especially if the increased uncertainty is small.
+If it is noticable we'll need to comment on this since our recommended approach is to just use $m_a$ (which has cosmology premarginalized)
+ for other constraints and fix cosmology thereafter.}
 
@@
 The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper.
-\ch{keeping srollv2 in label only here}
 } \ch{keep same color scheme as Fig. 2, use green maybe for simallEE}
@@
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we also show the constraints assuming a Gaussian posterior in $m_a$ using the PC chain covariance and means listed in Table~\ref{tab:PC_stats}.
-The visual agreement with the full posteriors is remarkably good, especially in the well constrained components.
+The visual agreement with the full posteriors is remarkably good, especially in the well-constrained components.
 This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models.
-However we shall see that a quantitative comparison shows that there is enough skewness in the posteriors that the KDE likelihood
-still performs noticeably better.  \wh{perhaps shift the tone of this statement since we seem to be concluding that Gaussian is good enough.}
+\ch{We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences.}
 
@@
 We now turn to constraining reionization models with PCs using the
-effective likelihood approach described in~\refsec.  The benefit of this approach is that
-complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ can now be obtained without reanalyzing the Planck data. 
-This effective likelihood code is available on GitHub at \url{} [give URL].
-The initial release corresponds to our fiducial constraints derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihood combination with $\zmax = 30$.
-
-To use this approach, one simply specifies the functional form for $x_e(z)$ in the code, which then is converted to the PC parameters $m_a$ for the effective likelihood evaluation assuming the fiducial $\Lambda$CDM cosmology.
-To obtain reionization model parameter posteriors, one samples this likelihood with any desired priors.   
-We have also provided an MCMC sampler for models of high dimension (currently in CosmoMC). Note that the model parameters or priors must be arranged to explicitly satisfy 
-full ionization below $z=6$.
-To avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation, we suggest using the default value of $f = 0.14$.
-\wh{Is Gaussian part of this package or a separate package?
-either way should mention this too.}
-
-Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to an exact MCMC analysis for two examples: 1) The tanh model, which is the standard approach used in CAMB; 2) a two-step toy model which has one additional parameter modeling a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.
+effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  \ch{The benefit of this approach is that
+complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function in for example CAMB, each time a new ionization model needs to be tested.
+These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.}
+
+\ch{Our code is available on GitHub at \url{...} [give URL]. 
+The initial release corresponds to the fiducial results in this paper, i.e. derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$. 
+The most important part of the code release is the python package called \texttt{Relike} which implements the Gaussian likelihood.
+The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model.
+The returned likelihood is already marginalized over cosmological parameters.
+To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}}
+
+\ch{For convenience, we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
+This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.
+}
+
+\ch{When using the KDE likelihood, we suggest using the default value of $f = 0.14$ to avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation.
+All our KDE results in this section are computed with $f = 0.14$, which we shall soon show to work well in recovering results from an exact MCMC.
+Note that in both the Gaussian and the KDE modes, the model parameters or priors must be arranged to explicitly satisfy fully-ionized hydrogen and singly-ionized helium for $z\leq 6$.
+}
+
+Next we demonstrate the use and the successful recovery of parameter posteriors using the effective likelihood code by comparing them to an exact MCMC analysis using the original Planck likelihoods for two examples: 1) The tanh model, which is the standard approach adopted in CAMB; 2) a two-step toy model which has one additional parameter allowing for a high-$z$ ionization plateau before the transition to full ionization in the canonical tanh model.
 
@@
 
-\ch{quote tau tanh KDE and gauss numbers with labels like $\tau_{tanh} = ... $(KDE).}
-In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the full likelihood results in which the five other $\Lambda$CDM parameters were also varied using the same priors for the $\tanh$ model.
-%including $\tau$. 
-%Because the PC method cannot accurately represent models that are not fully ionized below $z = 6$, we apply a prior $\zre < 6.1$ to both chains given the transition width of about $0.015(1+z) \approx 0.1$ at $z = 6$.
+In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the exact Planck likelihood results in which the five other $\Lambda$CDM parameters were also varied using their standard priors.
 The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
-We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to a smoothing factor of $1+f = 1.14$ applied to the covariance of the PCs used for smoothing kernel during KDE.
-Indeed, the KDE result $\tau = 0.0588 \pm 0.0066$ has a negligibly lower mean ($0.05\sigma$) and a 6.5\% larger standard deviation (as expected) than the full likelihood result $\tau = 0.0591 \pm 0.0062$.
+We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to the smoothing factor of $1+f = 1.14$ we applied to the PC covariance during KDE.
+Indeed, the KDE result
+%
+\beq
+\tau_{\rm tanh} = 0.0588_{-0.0066}^{+0.0061}\;\; (\mathrm{KDE}) 
+% before extending chains
+%\tau_{\rm tanh} = 0.0588_{-0.0072}^{+0.0061}\;\; (\mathrm{KDE}) 
+\eeq
+%
+has a negligibly lower mean ($0.05\sigma$) and a 4\% larger 68\% confidence region  than the exact likelihood result 
+\beq
+\tau_{\mathrm{tanh}} = 0.0591_{-0.0068}^{+0.0054}\;\; (\mathrm{Exact}).
+\eeq
+}
+%before
+%$\tau = 0.0591 \pm 0.0062$ (exact)
+%$\tau = 0.0588 \pm 0.0066$ (KDE)
+%now (possibly due to longer chains)
+%$\tau = 0.0591 \pm 0.0063$ (exact)
+%$\tau = 0.0588 \pm 0.0064$ (KDE)
 
@@
 [add timing here]
-%and amounts to one evaluation of the multi-variate Gaussian function per sample in the reionization model parameter space. 
-We obtain for the tanh model $\tau = 0.0594 \pm 0.0063$.
+
+\ch{
+We obtain for the tanh model 
+\beq
+\tau_{\rm tanh} = 0.0592_{-0.0062}^{+0.0062}\;\; (\mathrm{Gaussian})
+\eeq
+}
+\ch{CH: need to rewrite the following paragraph. Double check that results do not change varying length of chains ...}
  While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two.
 This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution.
-As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ if we consider the 68\% lower and upper limits, whereas the full KDE result $\tau = 0.0588_{-0.0072}^{+0.0060}$ is able to capture the skewness in the distribution
-\wh{seems like we should quote the direct results to make the claim that it is correctly capturing skewness - also,  is there a reason why we start with symmetric errors instead of just giving these as the main results (maybe also a displayed equation to highlight)?
+As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ \wh{this differs from above}
+ if we consider the 68\% lower and upper limits, whereas the full KDE result 
+\beq
+\tau_{\rm tanh} = 0.0588_{-0.0072}^{+0.0060}\;\; (\mathrm{KDE})
+\eeq
+is able to capture the skewness in the distribution
+%\wh{seems like we should quote the direct results to make the claim that it is correctly capturing skewness - also,  is there a reason why we start with symmetric errors instead of just giving these as the main results (maybe also a displayed equation to highlight)?
 if it's to quote the effect of 7\% smoothing we could still do that on the average of + and - errors.}
-$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}. \ch{Add $\tau(15,\zmax)$ 95\% upper limit for tanh direct MCMC and KDE.}
-
-\ch{[double check all these numbers again (some chains are longer now)]}.
+$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.
+\ch{CH: To turn into a sentence: $\tau_{\rm tanh}(15,\zmax) < 0.002$ for both effectivey likelihoods and $\tau_{\rm tanh}(15,\zmax) = 0$ from exact MCMC.}
+% 95\% UL:
+%tanh Gauss: 1.996574506651048464e-03
+%tanh KDE: 1.996656635642191091e-03
+%tanh exact: 0.000000000000000000e+00
 
@@
 We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integrating the tanh model in the exact likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains using Eq.~(\ref{eq:cumtauPC}).
-\wh{
 Notice that because of the truncation of the PC expansion at $a=5$, this representation shows an small oscillatory cumulative $\tau$ at $z \gtrsim 10$ whereas the true model has negligible ionization there.
 This is because the best constrained modes only represent the smooth features in the ionization history and truncation is similar to a low pass filter in Fourier space.
-Though these zero mean, small amplitude oscillations do not bias results when averaged over redshift, this effect should be born in mind when interpreting the model independent results below.
-}
-
-\wh{
-In the exact or in the effective likelihood approaches (ignoring these oscillations),
+Though these zero mean, small amplitude oscillations do not bias results when averaged over redshift, this effect should be born in mind when interpreting the model-independent results in \S\ref{sec:modelindependent}.
+In the exact or in the effective likelihood approaches (ignoring these oscillations,
 we can see that the tanh model does not allow for high redshift ionization by construction.
 As we shall next see in the next example the data themselves do allow for a small amount of high redshift ionization when the form of the model permits them.
-%As we shall see in \S \ref{sec:modelindependent} this is also captured in our model independent approach to cumulative optical depth constraints. 
 }
@@
 also in 2D Gaussian may arguably be better than KDE.
-\ch{double check they are all smoothed the same way, and number of chain points}}
+\ch{double checked that they are all smoothed with default getdist setting; many more chains for Gaussian, running KDE to longer chains to compare afterward.}}
 }
@@
 We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model.
-Here we used again $f = 0.14$ in the KDE computation.
-\wh{if we never show anything but $f=0.14$ we could just say it's always this value in the method section}
-
-In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$.\wh{The exact, KDE and Gaussian PC treatments again agree very well.} 
+
+In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$. The exact, KDE and Gaussian PC treatments again agree very well. 
 Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors distort near this value.
@@
 
-% Not sure we need to get into this subtlety - the caveat being that getdist will smooth the boundary even if it is in reality sharper.
-Not sure we know which effect dominates and it probably doesn't matter much:
-%
-%; for the two-step model, the correspondence between $\xemin$ and $\zre$ is not one-to-one and depends on the value of $\tauhi$.
-This is why the cut on the 2D constraint ellipses at low $\taulo$ is of a curved shape rather than a straight cut, because the $\zre$ prior corresponds to a different value of $\taulo$ for a different $\tauhi$.
-
-[add timing results here?]
-
-[comment on effective likelihood w/ KDE vs Gaussian likelihood?]
-
-The marginalized 1D constraints from the effective likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$ are consistent with those of the full likelihood $\taulo = XX \pm XX$ and $\tauhi = XX \pm XX$.
-The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < XX$ for the effective likelihood and $\tau(15, 30) < XX$ for the full likelihood.
-\wh{Note also that the model 95\% C.L. allowed model in 
+\todo{add timing results somewhere}
+
+\ch{The marginalized 1D constraints from the KDE and Gaussian likelihoods 
+\beq
+\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (KDE)}
+\eeq 
+and
+\beq
+\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Gaussian)}
+\eeq 
+are consistent with those from the exact likelihood 
+\beq
+\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Exact).}
+\eeq 
+}
+%exact: tauhi < 0.0206
+%KDE: tauhi < 0.0214
+%Gaussian: tauhi < 0.0207
+
+% running longer KDE chains (numbers might be updated)
+
+\ch{The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < 0.017$ for both effective likelihoods and $\tau(15, 30) < 0.016$ for the exact likelihood.}
+\ch{Note also that the 95\% C.L. allowed two-step model in 
 Fig.~\ref{fig:two_step_model} has
-$\tau(15,30)= ZZ$.}
+$\tau(15,30)= 0.013$.}
+%0.01343375 evaluated at fiducial cosmology
 
@@
 in this class.
-\wh{
  In principle, the FlexKnot method allows for all possible physical ionization histories that increase monotonically with redshifts, and would include the two-parameter model described here.
@@
 $\tau_{\rm hi}$ does.
-We shall see in \S \ref{sec:modelindependent} that our model independent PC approach models such a flat prior better and of course the effective likelihood approach allows for any desired prior.}
-
-% Move this somewhere into the prior discussion or the 
-% paper discussion
-%
-%In Ref.~\cite{Aghanim:2018eyx}, the authors also suggested that the PC method has limitations due to the fact that it includes some unphysical models in addition to all the physical models.
-The authors gave this reason for not using PCs to constrain reionization histories.
-We would like to argue that this claim is not well founded.
-
-In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this 2-parameter model.
-In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ integrated over the exact two-step model in the full likelihood chains themselves, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.
-\wh{As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.
-Notice again the amount of high redshift optical depth that is allowed in the two step model vs the tanh model (cf.~Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).}
-%
-%of those physical models in the effective likelihood chains.
-%Because the truncated 5 PCs may represent non-physical models (higher PCs can restore the physicality of the model), negative $\tau$ values are allowed at some redshifts, which is of course, unphysical.
-Nevertheless, the full vs truncated PC chains give similar 95\% upper limits on the high-redshift optical depth, $\tau(15, 30) < XX$ and $\tau(15, 30) < XX$ respectively.
-We have showed that the discrepancy between the PC-based effective likelihood results and the FlexKnot results are not due to the inclusion of non-physical models.
-\wh{
-With this in mind, we will now turn to the model-independent constraints derived from the PC analysis to generalize these results on high redshift optical depth beyond the two-step model.}
-
-%\ch{Could also mention best-fit model below as well.}
-%To test this, we take the best-fit two-parameter model $(\taulo, \tauhi) = (XX, XX)$ and show its cumulative optical depth $\tau(z, \zmax)$ in ~\reffig{plot_taugtz_two_step_best_fit}.
-Compared to the PC constraints (shaded regions), this model is within the 95\% upper limit at all redshifts.
-
-%PC w/ ML model from 2parameter model (tauhi>0.02)
-%\begin{figure}[t]
-%\includegraphics[width=0.450\textwidth]{plots/plot_tau_gtz.pdf}
-%\caption{Same as \reffig{plot_taugtz_PC_vs_tanh}, but superimposed with the cumulative optical depth $\tau(z, \zmax)$ of the best-fit two-step model.
-The two-step model is a toy model that adds a high-$z$ ionization plateau to the standard tanh model with one additional parameter.
-Compared to the cumulative $\tau$ distributions from PCs, the best-fit two-step model is within the 95\% upper limit of the cumulative $\tau$ constraints at all redshifts.
-%}
-%\label{fig:plot_taugtz_two_step_best_fit}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[\textwidth]{results/direct_mcmc/two_parameter_model/tauhi_taulo_chains/pl18_tanh_highz_test2_run1_tri.png}
-%\caption{Direct MCMC chains of the two-parameter model with \tauhi and \taulo: triangle plot of \tauhi and \taulo.
-The marginalized 1D constraints are \taulo = $0.0516 \pm 0.0076$, \tauhi = $0.0100 \pm   0.0066$.
-The ML model is \taulo = 0.0543 and \tauhi = 4 corresponding to $z_{\rm re} = 7.68$ and $x_{e, \mathrm{min}} = 0.021$.
-%}
-%\label{fig:two_parameter_model_2D_plot}
-%\end{figure}
-
-%\begin{figure}
-%\includegraphics[\textwidth]{results/cosmomc_kde/taulo_prior_test/plot_cls_taulo_0p04_tauhi_0p02.png}
-%\caption{Testing \taulo prior -- testing that \taulo $\geq$ 0.04 as \taulo prior is OK, by calculating $\chi^2$ between $C_l$ from full $x_e(z)$ vs PC decomposition of the model (\taulo, \tauhi) = (0.04, 0.02).
-%The $\chi^2 = \sum_{2}^{l_{\rm max}} (\Delta C_l^{EE})^2 / \mathrm{Cov}_l = 0.03$ where $\mathrm{Cov}_l = 2 (C_l^{EE})^2/(2l+1)$. } 
-%\label{fig:taulo_prior_test_cl}
-%\end{figure}
+We shall see in \S \ref{sec:modelindependent} that our model independent PC approach models such a flat prior better and of course the effective likelihood approach allows for any desired prior.
+
+In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this two-step model.
+In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ obtained by integrating over the exact $x_e(z)$ two-step model in the exact likelihood chains, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.
+As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.
+Notice again the amount of high redshift optical depth that is allowed in the two step model vs the tanh model (cf.~Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
+With this in mind, we will now turn to the model-independent constraints derived from the PC analysis to generalize these results on high redshift optical depth beyond the two-step model.
 
@@
 \includegraphics[width=0.48\textwidth]
-%normalized by area
-%{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_area.pdf}
-%normalized by max
 {paper/plots/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
@@
 
-\wh{
 While the PC approach to constraining reionization should mainly be used for 
 testing wide classes of models efficiently and with their own physically motivated priors, it can also be employed to extract model-independent constraints on the total and cumulative optical depth. 
-% this sounds too technical now for an intro to the section:
-%As discussed in \S \ref{sec:cumulative}, since the PC amplitudes are not intrinisically restricted to physically sensible ionization histories where $0 \le x_e(z) \le x_e^{\rm max}$, this requires imposing physicality priors using Eq.~\ref{eq:individualprior}.
 In this section we present these model-independent results, first on the total optical depth and then on the cumulative optical depth at high redshift, and discuss their robustness to priors  on physicality of the ionization history and redshift range of the fiducial analysis.
-}
-
-%So far no priors on the PC chains have been used.
-In constructing the kernel density estimate of Section~\ref{sec:effective_likelihood} we used PC chains without priors; and when using the effective likelihood for model testing, it would always be restricted to a particular set of physical models like in the two examples we showed above.
-To derive model-independent constraints in this section, however, we will now impose the physicality priors of Eq.~\ref{eq:individualprior}.
-
-%Before we show the cumulative optical depth constraints from PCs, let us first make a note about the impact of the physicality prior chosen for this analysis and clarify some wrong claims against it in the literature.
 
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
-\caption{Posterior of optical depth $\tau_{\rm PC}$ from reionization PCs (blue solid line) vs that of $\tau_{12}$ (shaded) defined in Eq.~\ref{eq:tau12}.
-\ch{$\tau_{\rm PC}$ here is not marginalized over cosmological parameters, in contrast to Figs.~\ref{...}.
-We show the versions without prior (top panel) vs with physicality prior (bottom panel).}\ch{CH: Add more description...}}
+\caption{Posterior of total optical depth $\tau_{\rm PC}$ using PC constraints (blue solid line) vs that of $\tau_{12}$ (shaded) defined in Eq.~\ref{eq:tau12}.
+They agree well enough for the purpose of our study here for both without the physicality prior case (top panel) and with prior case (bottom panel).
+Without the prior, the $\tau$ distributions are broadened and shifted low because the prior removes models with negative ionization and optical depth at high-redshifts.}
 \label{fig:tau12}
@@
 
-\wh{
  We first note that by virtue of Eq.~(\ref{eq:cumtauPC}), the total optical depth receives contribution from all PCs.  However
@@
 \eeq
-where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of fiducial model.
+where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of the fiducial \ch{ionization} model.
 We explicitly demonstrate this approximate equivalence in Fig.~\ref{fig:tau12} by comparing the
 posterior distributions of $\tau_{\rm PC}$ and $\tau_{12}$ both with and without the physicality prior.  
-}\ch{CH: will add comment on mean from m3, m4, m5 that are computed from the chains with prior as well}
+\ch{CH: actually plots are made with mean from m3, m4, m5 that are computed from the chains with prior (for slightly better agreement) whereas the table had mean from chains without prior, to be consistent with the correlation matrix reported.
+} \wh{if it is a slight difference use the means from the table.
+also it would be strange if that were the motivation of the no-prior comparison plot which is our start point.}\ch{that's true, will change to no-prior then.}
 
@@
 
-\wh{
 This also shows how our range-bound priors 
@@
 may also be unphysical since their physicality depends on higher order PC components.
-The impact on optical depth  however is small at at most contributes a small bias that conservatively {\it lowers} the high redshift contribution which is already a small contribution to the total.
+The impact on optical depth  however is small and at most contributes a small bias that conservatively {\it lowers} the high redshift contribution which is already a small contribution to the total.
 This can be seen  in the comparison with the two-step model 
-in Fig.~\ref{fig:tauPC} and as we shall return to this point in the cumulative optical depth where the high and low redshift contributions are manifest.}
-%On the other hand the tanh model provides an explicitly physical example whose trajectory runs close  and nearly parallel to the our physicality prior in the allowed region and so we do not expect the bias incurred by
-%including these possibly \wh{unphysical cases} to be large. 
-%\ch{CH: sorry, what do we mean by these models here?}  
+in Fig.~\ref{fig:tauPC} and as we shall return to this point in the cumulative optical depth where the high and low redshift contributions are manifest.
 In any case, using PCs to 
@@
 \eeq
-Reverting the Planck 2018 likelihood to \texttt{lowE} at $z_{\rm max}=30$ would give
-\beq
-\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{lowE}).
+Reverting the Planck 2018 likelihood to \texttt{simall\_EE} at $z_{\rm max}=30$ would give
+\beq
+\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{simall\_EE}).
 \eeq
@@
 
-\ch{Might need to change all simallEE labels to lowE in plots and in text}
-
 %
@@
 %
-%Compared the \texttt{lowE} likelihood results, 
+%Compared the \texttt{simall\_EE} likelihood results, 
 %\beq
-%\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{lowE}),
+%\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{simall\_EE}),
 %\eeq
@@
 %\beq
-%\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\texttt{lowE}).
+%\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\texttt{simall\_EE}).
 %\eeq
@@
 \begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.
@@
 The total $\tau$ distribution is broadened only toward the lower $\tau$.
-\ch{Discuss in text the small difference in upper limits at $\tau(15, \zmax)$ is such that if one could remove unphysical models inside our current prior range, $\tau(15, \zmax)$ upper limit would be only slightly looser.}
 }
@@
 \begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2.pdf}
+\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2_do_rescale_tau_false.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.
@@
 \begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50.pdf}
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
 \caption{Comparison of the 68\% and 95\% C.L.
@@
 for the PC method but using the prior inversion procedure
-discussed \S~\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
+discussed \S\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
 [do we want to check whether the prior inversion procedure could cause a shift at this level?
@@
 Compared to our PC results ($\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\% \mathrm{CL})$), the FlexKnot upper limits are too stringent.
-The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in \S~\ref{sec:example2} with a concrete example of the two-step model.
+The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in \S\ref{sec:example2} with a concrete example of the two-step model.
 We showed that for this chosen physical model, the $\tau(15, 30)$ upper limits derived from the full Planck likelihood without using any PCs were already not compatible with the FlexKnot upper limits, while the PC results were compatible.

### 2021-03-30 paper/rei.tex
Updates from Overleaf

@@
 
-\title{Effective Reionization Likelihood from Planck 2018 Data (?)}
+\title{\ch{\textsc{RELIKE}}: Reionization Effective Likelihood from Planck 2018 Data (?)}
 
@@
 
-We publish Relike, an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC).
-This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is tested.
-We tested the code on two example models
-%: 1) the tanh model with a single transition redshift, and 2) a two-parameter model with two transitions allowing for a high-redshift ionization plateau. 
-which showed good agreement with
-%The Relike chains converged 300$\times$ and 100$\times$ faster for model 1 and 2 respectively, and showed good agreement compared to directly
-sampling the exact Planck likelihoods.
-We also find that contrary to the Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation.
+We publish \textsc{RELIKE} (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC).
+This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested.
+We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
+We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation.
 We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler.
-We expect the code release to dramatically reduce the time for testing individual ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
-%We expect the code release to be useful for individual model testing as the user would not need to modify CAMB every time a different ionization history model is needed; the effective likelihood code also makes it easier to perform joint constraints of the Planck data with other reionization data such as 21cm, luminosity functions, star formation rate measurements, etc.
-Finally, we derive model-independent constraints: $\tau_{\rm PC} = ...$ and $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.).
-The upper limits on high-redshift optical depth is at least a factor of 2 looser those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  $\tau_{\rm PC}(15, 30) < XX $ (95\% C.L.) using a  two-parameter toy model without using PCs.
+We expect the code release to dramatically reduce the time for testing individual \ch{global} ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
+Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.).
+The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  \ch{$\tau_{\rm 2step}(15, 30) < 0.013 $ (95\% C.L.)} using a  two-step toy model without using PCs.}
 
@@
 
+\ch{CH: going through the paper once from beginning to end now.}
+
+\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}.}
+
 The cosmic microwave background (CMB) has entered an era of precision cosmology.
-The measurements from the \textit{Planck} satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution.
-While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces.
+The measurements from the Planck satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution~\cite{Aghanim:2018eyx}.
+While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces (see e.g.~\cite{2016ASSL..423.....M} and references therein).
 Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude.
-Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB; and will also have implications for inferring cosmic acceleration through the growth of structure.
-\ch{[add more refs here]}
-
-Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history, with the step location parameterized by the total Thomson optical depth induced.
-This steplike model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition.
-However, through the shape of the reionization bump induced in the CMB E-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained.
-
-In fact, to extract the most information possible from this E-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum.
+Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.
+
+Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced.
+This steplike \ch{tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition.
+However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.
+
+In fact, to extract the most information possible from this $E$-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum.
 This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}.
-It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, as well as massive neutrinos and gravitational waves in Ref~\cite{Dai:2015dwa}\ch{[also cite our inflation papers?]}.
+It was also adopted for a Planck 2013 analysis for marginalizing ionization history when constraining inflationary parameters in Ref.~\cite{Planck:2013jfk}, for studying inflationary features vs reionization features in Ref.~\cite{Obied:2018qdr}, as well as massive neutrinos and gravitational waves in Ref.~\cite{Dai:2015dwa}.
+%\cite{Obied:2017tpd}
 
@@
 
-Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019.
-This new likelihood, called $\texttt{srollv2}$, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the steplike model is reduced by roughly a factor of two from \ch{XX to XX.
-[fill in numbers]}
+\ch{I'm here}
+Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
+$\texttt{srollv2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
+~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{−0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.
+
+\todo{CH: might need to change all instances of \textt{srollv2} to \textt{SRoll2} as used in the original sroll2 paper; also possibly \textt{simall\_EE} to \textt{SimAll} as used in the Planck 2018 paper.}
 
@@
 The principal component technique for constraining reionization using the large-angle $C_\ell^{EE}$ polarization spectrum was first introduced in Ref.~\cite{Hu:2003gh}.
-We now briefly summarize the PC technique in \S\ref{sec:PC}, as well as the kernel density estimate technique for building the effective likelihood in \S\ref{sec:KDE}.
-We refer the readers to Refs.~\cite{
+Here, we  briefly summarize the PC technique itself in \S\ref{sec:PC}.
+We present  techniques
+%kernel density estimate technique
+for building its effective likelihood which can be used to rapidly explore any reionization model in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{
 %Hu:2003gh, %Mortonson:2008rx, 
@@
 For a Gaussian posterior, increasing the covariance by $1+f$ corresponds to increasing the standard deviation by approximately $1+f/2$.
-To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \approx 1.1 \times 10^{6}$ chain samples.
-\ch{For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient for most models, and indeed we have tested it to work well for the ones we study in \S~\ref{sec:effective_likelihood}.}
+To minimize the amount of smoothing needed while still maintaining good accuracy in the tail of the distribution for any physical models, we oversample the PC distributions by running the PC chains far beyond convergence for about $N_{\sample} \sim 10^6$ chain samples.
+For this $N_{\sample}$, a fraction of $f = 0.14$ should be sufficient for most models, and indeed we have tested it to work well for the ones we study in \S~\ref{sec:effective_likelihood}.
 
@@
 
-Although reionization PCs are mainly a tool for model testing using complete information from CMB polarization, as they do not reconstruct $x_e(z)$ itself, in particular the its rapidly  aspects, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
+Although reionization PCs are mainly a tool for model testing using complete information from large angle CMB polarization, as they do not reconstruct the rapidly varying aspects of $x_e(z)$ itself, they do provide model-independent constraints on redshift integrated quantities such as  the cumulative Thomson optical depth
 \begin{align}
@@
 Eq.~(\ref{eq:cumtau}) using the fiducial model 
-for $H(z)$ and number densities.
+for $H(z)$ and number densities.  This differs from Ref.~\cite{Heinrich:2016ojb} where their cosmological dependence was employed.
  When we omit the redshift arguments, we
@@
  $
- \ch{CH: what about recaling for different cosmological model? Our tau(z,zmax) plots have cosmological parameters marginalized, while Figs 9-10 (total tau posterio) do not.}
- \wh{They probably shouldn't, i.e should be constructed directly out of $m_a$, especially if the increased uncertainty is small.
-If it is noticable we'll need to comment on this since our recommended approach is to just use $m_a$ (which has cosmology premarginalized)
- for other constraints and fix cosmology thereafter.}
 
@@
 %
-Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we do not keep all but the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.
+Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we  keep only the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.
 In other words, no physical model is excluded by these priors but some unphysical models are included.
@@
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the 1D posterior and 2D 68\% and 95\% confidence level contours for the 5 PC amplitudes that describe ionization models up to $\zmax=30$.
-We also show the trajectory of the tanh model (see \S \ref{sec:example1}) through this space as well as the best fit tanh model points.
-%
-\footnote{The Planck 2015 best-fit for the tanh model is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Planck 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $\Delta z = 0.015(1+z)$ instead of $\Delta z = 0.5$ for the Planck 2015 best-fit.
+We also show the trajectory of the tanh model (see \S \ref{sec:example1}) through this space as well as the best fit tanh model points.\footnote{The Planck 2015 best-fit for the tanh model is derived from a chain best-fit model as taken from Ref.~\cite{Heinrich:2016ojb}, whereas for the Planck 2018 point in this paper we used a minimizer to find the best-fit; there are also slight model differences, where in Planck 2018 we used a tanh width of $\Delta z = 0.015(1+z)$ instead of $\Delta z = 0.5$ for the Planck 2015 best-fit.
 Using $\Delta z = 0.5$ for Planck 2018 moves the cross negligibly on this plot.}
@@
 %
-The gray shaded region in the $m_1-m_2$ plane represents the parameter region that would be excluded by an additional 
+The gray shaded region % in the $m_1-m_2$ plane 
+represents the parameter space that would be excluded by an additional 
 physicality prior from Eq.~(\ref{eq:individualprior}),
@@
 While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained.
-Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller amplitude for the first PC, leading to a reduced optical depth, especially at high redshift.
+Contrary to the Planck 2015 results of Ref.~\cite{Heinrich:2016ojb} shown in red, the 2018 data prefers a much smaller $m_1$, leading to a reduced optical depth, especially at high redshift.
 This preference also increases the impact of placing a physicality prior at low values of $m_1$ when deriving model-independent constraints as we shall discuss further in \S \ref{sec:modelindependent}.
@@
 \includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{A zoomed-in of Fig.~\ref{fig:plot_mjs_2018_vs_2015} on the best constrained $m_1-m_2$ PC plane. We also 
-the Planck 2018 constraints using two different low-$\ell$ $EE$ likelihoods: the \texttt{simall\_EE} likelihood (red) and the \texttt{srollv2} likelihood (blue).
-The \texttt{srollv2} likelihood used better foreground removal and is what we use as our fiducial results throughout this paper.
-} \ch{keep same color scheme as Fig. 2, use green maybe for simallEE}
+\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{srollv2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{simall\_EE} likelihood (\wh{green?}).
+With Planck 2018, the \texttt{srollv2} likelihood moderately improves the best constrained direction and raises the optical depth.
+}
 \label{fig:plot_m1m2_2015_vs_2018}
@@
 This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models.
-\ch{We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences.}
+We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences in the shape of their posterior inferences.
 
@@
 We now turn to constraining reionization models with PCs using the
-effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  \ch{The benefit of this approach is that
-complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function in for example CAMB, each time a new ionization model needs to be tested.
-These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.}
+effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  The benefit of this approach is that
+complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function \wh{and jointly sample the other cosmological parameters} in for example CAMB, each time a new ionization model needs to be tested.
+These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.
 
 \ch{Our code is available on GitHub at \url{...} [give URL].
-The initial release corresponds to the fiducial results in this paper, i.e. derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$. 
-The most important part of the code release is the python package called \texttt{Relike} which implements the Gaussian likelihood.
+} The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$.
+The most important part of the code release is the python package called \texttt{Relike} which \wh{provides the PC parameters for a model and also} 
+% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
+implements the Gaussian effective likelihood.
 The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model.
 The returned likelihood is already marginalized over cosmological parameters.
-To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}}
-
-\ch{For convenience, we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
+To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.
+
+For convenience, 
+\wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
+we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
 This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.
-}
-
-\ch{When using the KDE likelihood, we suggest using the default value of $f = 0.14$ to avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation.
+
+When using the KDE likelihood, we suggest using the default value of $f = 0.14$ to avoid over-smoothing parameter posteriors while maintaining accuracy during the KDE operation.
 All our KDE results in this section are computed with $f = 0.14$, which we shall soon show to work well in recovering results from an exact MCMC.
 Note that in both the Gaussian and the KDE modes, the model parameters or priors must be arranged to explicitly satisfy fully-ionized hydrogen and singly-ionized helium for $z\leq 6$.
-}
 
@@
 {paper/plots/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Comparison of the total optical depth posterior for the tanh model in \refsec{example1} between the exact likelihood (shaded) and our effective likelihoods in the kernel density estimate (KDE) mode (blue solid) and the Gaussian approximation mode (black dashed).
-The KDE mode agree well with the exact results, while giving a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing at roughly $\sim$7\% level in the parameter posterior corresponding to $f = 0.14$.
-The Gaussian likelihood works well too, with no extra broadening of the posterior, but a negligible shift in the mean ($\sim XX\%$ of the error in $\tau$).
-This posterior is symmetric and does not capture the small skewness found in the exact and KDE chains.
-\ch{might need to rephrase.}
+\caption{Total optical depth $\tau_{\rm tanh}$ posterior for the tanh model in \refsec{example1} for the exact (Planck 2018) likelihood (shaded), KDE PC 
+(blue solid) and KDE Gaussian (black dashed)
+effective likelihood. The overall agreement is excellent and the KDE results even capture the small skewness of the exact result. \wh{maybe $\tau_{\rm tanh}$ }
 }
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
-\caption{Comparison of the 68\% and 95\% C.L.
-constraints on the cumulative optical depth in the tanh model (described in \refsec{example1}) between the full Planck likelihoods (shaded bands), our effective likelihoods in the KDE mode (blue solid) and Gaussian mode (black dashed).
-The effective likelihoods results oscillate around zero at high-z because of the truncated representation of the models at 5 PCs (enough to represent any observable effects in the data).
-While the Gaussian approximation performs fairly well, the more complete KDE mode gives a slightly more accurate 95\% limits on the total optical depth.
-}
+\caption{Cumulative optical depth in the tanh model as in Fig.~\ref{fig:tanh}
+(68\% and 95\% C.L.
+contours).
+Overall agreement is good in regions where the contribution to $\tau_{\rm tanh}$ is high but the model form forbids high-$z$ contributions.
+The KDE and Gaussian PC results  oscillate around zero at high-z because of effective low pass filter of the truncation to 5 PCs (enough to represent any observable effects in the data).
+\wh{maybe $\tau_{\rm tanh}$} }
 %double check plot, upper and lower limits seem to cross?
@@
 \beq
-\tau_{\rm tanh} = 0.0588_{-0.0066}^{+0.0061}\;\; (\mathrm{KDE}) 
+\tau_{\rm tanh} = 0.0591_{-0.0066}^{+0.0061}\;\; (\mathrm{KDE}) 
 % before extending chains
@@
 %
-has a negligibly lower mean ($0.05\sigma$) and a 4\% larger 68\% confidence region  than the exact likelihood result 
+%has a negligibly lower mean ($0.05\sigma$) and 
+has a 4\% larger 68\% confidence region  than the exact likelihood result 
+%ch: made a mistake, mean is 0.0591, not 0.0588 for KDE.
 \beq
@@
 
-Instead of the effective likelihood, we can alternately use its Gaussian approximation shown in Eq.~(\ref{eq:gaussian}).
+Instead of the KDE likelihood, we can alternatively use the Gaussian approximation given in Eq.~(\ref{eq:gaussian}).
 The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain.
-[add timing here]
-
+\todo{[add timing here]}
+\wh{not sure we need timing results but if you do add them remember you can make Gaussian even faster by precomputing the linear combinations of $m_a$ that diagonalize the covariance (the first one should essentially be $\tau_{12}$ which is interesting pedagogically)} \ch{[CH: got it.]}
 \ch{
@@
 \eeq
-}
-\ch{CH: need to rewrite the following paragraph. Double check that results do not change varying length of chains ...}
- While it is a good approximation (the mean and standard deviation are similar to those from the full likelihood and the effective likelihood using KDE), we see in Fig.~\ref{fig:tanh} that the peak of the distribution for the Gaussian likelihood is offset from the other two.
-This is because the Gaussian likelihood is by definition symmetric in the PC space and so it cannot capture any skewness in the PC distribution.
-As a result the tanh $\tau$ constraints are mostly symmetric too, $\tau = 0.0594_{-0.0064}^{+0.0064}$ \wh{this differs from above}
- if we consider the 68\% lower and upper limits, whereas the full KDE result 
-\beq
-\tau_{\rm tanh} = 0.0588_{-0.0072}^{+0.0060}\;\; (\mathrm{KDE})
-\eeq
-is able to capture the skewness in the distribution
-%\wh{seems like we should quote the direct results to make the claim that it is correctly capturing skewness - also,  is there a reason why we start with symmetric errors instead of just giving these as the main results (maybe also a displayed equation to highlight)?
-if it's to quote the effect of 7\% smoothing we could still do that on the average of + and - errors.}
-$\Delta \tau \sim 0.0012$, a level consistent with shifts in the peaks seen in Fig.~\ref{fig:tanh}.
-\ch{CH: To turn into a sentence: $\tau_{\rm tanh}(15,\zmax) < 0.002$ for both effectivey likelihoods and $\tau_{\rm tanh}(15,\zmax) = 0$ from exact MCMC.}
-% 95\% UL:
+Note that the Gaussian likelihood yields symmetric tanh $\tau$ constraints where as the KDE result 
+is able to capture some of the skewness in the exact distribution.
+The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods is consistent with this level of skewness ($\Delta \tau \sim 0.001$), which is $<0.2\sigma$.} % 95\% UL:
 %tanh Gauss: 1.996574506651048464e-03
@@
 Though these zero mean, small amplitude oscillations do not bias results when averaged over redshift, this effect should be born in mind when interpreting the model-independent results in \S\ref{sec:modelindependent}.
-In the exact or in the effective likelihood approaches (ignoring these oscillations,
+
+To quantify these residual oscillations we consider 
+the 95\% upper limit on the high-redshift optical depth.
+For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while $\tau_{\rm tanh}(15,\zmax) \approx 0$ from the exact likelihood.
+Ignoring this small effect of the oscillations, 
 we can see that the tanh model does not allow for high redshift ionization by construction.
-As we shall next see in the next example the data themselves do allow for a small amount of high redshift ionization when the form of the model permits them.
-}
+As we shall see in the next example the data themselves do allow for a small but significantly larger fraction of the total optical depth to originate from high redshift, when the form of the model permits it.
 
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
-\caption{Example ionization history $x_e(z)$ for the two-step model of \refsec{example2}.
+\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}.
 In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$.
-This is a toy model constructed for illustration purposes only.
 Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (0.039, 0.020)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).}
+$(\taulo, \tauhi) = (0.039, 0.020)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
+\wh{inset label for ``two step"}}
 %best-fit (0.05290344, 0.005859431)
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
-\caption{Comparison of the marginalized posteriors and the 2D 68\% and 95\% C.L.
-contours for the two-step model  described in \refsec{example2} between different likelihoods: The full Planck likelihoods (shaded), the effective likelihoods in the KDE mode with f = 0.14 (blue solid) and in the Gaussian approximation mode (black dashed).
-The results agree well with each other.
-As in Fig.~\ref{fig:tanh}, both the KDE and Gaussian modes agree very well with the full likelihood results, with the KDE  contours being slightly broader as expected.
-{To add in caption: best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (0.039, 0.020)$}  {Otherwise this plot is done.} \wh{Gaussian looks lumpy in the marginal distributions?
-also in 2D Gaussian may arguably be better than KDE.
-\ch{double checked that they are all smoothed with default getdist setting; many more chains for Gaussian, running KDE to longer chains to compare afterward.}}
+\caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L.
+contours).
+The exact, KDE PC and Gaussian PC likelihoods again agree very well with a small contribution from high-$z$ through $\tau_{\rm hi}$ still allowed but not significantly favored.
+Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters
+\wh{which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
+suggested change in model - or find another reason for plotting the currently chosen one}. \wh{add inset ``two step"} 
 }
@@
 {paper/plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
-\caption{Comparison of the 68\% and 95\% C.L.
-constraints on the cumulative optical depth $\tau(z, \zmax)$ for the two-step model, which allows for high-redshift ionization up to $\zmax =30$.
-Note that a physical model with $\tau(15, \zmax) = 0.02$ is still allowed at the 95\% C.L.
-whereas it is ruled with the Planck official results using the FlexKnot method $\tau(15, \zmax) < 0.007$.
-Here again, the KDE and Gaussian likelihoods agree well with the full likelihoods, with better 95\% C.L.
-limits on the total optical depth.
-The truncation at 5 PCs causes wiggles at higher redshifts, but these cause negligible differences at most redshifts.}
+\caption{Cumulative optical depth in the two step model (as in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
+Agreement between the exact, KDE PC and Gaussian PC likelihoods is again excellent and show that a small amount of high-$z$ ionization is allowed.
+\wh{The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_1 \sim 6-7$ especially for $z\gtrsim 15$.} \wh{Maybe $\tau_{\rm 2step}$}  }
 \label{fig:plot_taugtz_two_step_contours}
@@
  \bea
-x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
+x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{1})-y(z)}{\Delta y} \right] \right\} \notag \\
 &+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
@@
  \eea
+ \wh{naming is slightly odd here - maybe $z_{\rm re} \rightarrow z_1$ and $z_{\rm t}\rightarrow z_2$?
+ }\ch{CH: Actually I think it's good to keep zre because it is when the transition to full reionization happens, and is the same nomenclature people are used to with the tanh model.
+We could find a different name for $z_t$ if you'd like?}
 where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{\rm re})$ just as in the first example for a sharper step.
@@
 
-\ch{The 95\% C.L. upper limit on high-redshift optical depth are also consistent: $\tau(15, 30) < 0.017$ for both effective likelihoods and $\tau(15, 30) < 0.016$ for the exact likelihood.}
+\wh{Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$: give $\tau_{2\rm step}$.}
+
+\ch{Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
+\begin{equation}
+    \tau_{2\text{step}}(15, 30) < 0.016 \quad (95\% \text{ C.L., Exact)}
+    \label{eq:2stepbound}
+\end{equation}
+and 
+$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods}
 \ch{Note also that the 95\% C.L. allowed two-step model in 
 Fig.~\ref{fig:two_step_model} has
-$\tau(15,30)= 0.013$.}
+$\tau_{2\text{step}}(15,30)= 0.013$.} 
+\wh{possibly choose a model on the 2D contour at max imum $\tau_{\rm hi}$}
 %0.01343375 evaluated at fiducial cosmology
@@
 {paper/plots/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Posterior of the total optical depth for the PC analysis (shaded), the tanh model (black solid line) and the two-step model (blue dashed line)}. 
-\label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}
+\caption{Total optical depth posterior for the model-independent PC analysis compared with the exact tanh model (black solid line) and the two-step model (blue dashed line) from Figs.~\ref{fig:tanh} and~\ref{fig:two_parameter_model_2D}.
+The PC analysis captures the ability to raise the optical depth through the high-$z$ contributions of models like the two-step case whose form allows it.
+}
+%\label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}
 \label{fig:tauPC}
@@
 This was especially apparent with the Planck 2015 data where
-the PC constraints gave $\tau=XX\pm YY$ which differed substantially from the tanh $\tau=XX\pm YY$.  As we shall discuss
+the PC constraints gave \todo{$\tau=XX\pm YY$ which differed substantially from the tanh $\tau=XX\pm YY$}.  As we shall discuss
 below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
-\caption{Posterior of total optical depth $\tau_{\rm PC}$ using PC constraints (blue solid line) vs that of $\tau_{12}$ (shaded) defined in Eq.~\ref{eq:tau12}.
-They agree well enough for the purpose of our study here for both without the physicality prior case (top panel) and with prior case (bottom panel).
-Without the prior, the $\tau$ distributions are broadened and shifted low because the prior removes models with negative ionization and optical depth at high-redshifts.}
+\caption{Total optical depth $\tau_{\rm PC}$ posterior (shaded) with (top) and without (bottom) the additional physicality prior of Eq.~(\ref{eq:individualprior}).
+The shift corresponds to the elimination of unphysically negative ionization at high $z$ from the $m_1$ PC as can be traced through the approximate $\tau_{12} \approx \tau_{\rm PC}$ built from a linear combination of the first two components defined in Eq.~(\ref{eq:tau12}) (blue lines).
+}
 \label{fig:tau12}
@@
 \eeq
-where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of the fiducial \ch{ionization} model.
+where the term in parentheses is a fixed quantity (see Tab.~\ref{tab:PC_stats}) whose purpose is to remove offsets due to the arbitrary choice of the fiducial ionization model.
 We explicitly demonstrate this approximate equivalence in Fig.~\ref{fig:tau12} by comparing the
 posterior distributions of $\tau_{\rm PC}$ and $\tau_{12}$ both with and without the physicality prior.
-\ch{CH: actually plots are made with mean from m3, m4, m5 that are computed from the chains with prior (for slightly better agreement) whereas the table had mean from chains without prior, to be consistent with the correlation matrix reported.
-} \wh{if it is a slight difference use the means from the table.
-also it would be strange if that were the motivation of the no-prior comparison plot which is our start point.}\ch{that's true, will change to no-prior then.}
 
@@
           \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau12.pdf}
-          \caption {Priors on the $m_1-m_2$ parameter space: the physicality prior (inset square, solid lines) and a flat prior in
-           $\tau_{12}$ and in $m_1-m_2$ (rotated rectangle, dashed lines) 
-           %\ch{CH: [why is this flat in m1 and m2?]} 
-           % flat in 2D or locally flat
+          \caption {Priors on the $m_1-m_2$ parameter space: the default physicality prior (inset square, solid lines) and an alternate prior which is  flat in its marginal
+           $\tau_{12}$ distribution (rotated rectangle, dashed lines) given boundaries
            which are aligned with lines of constant $\tau_{12}$ (light gray lines).
 Note that although the physicality prior allows for more parameter space at higher $\tau_{12}$, that is also the region excluded by the data constraints (ellipses).
-The only difference in the allowed region is that the physicality prior clips the low $m_1$ portion of the allowed region due to the physicality condition and the assumption that reionization occurs at $z\ge 6$.}
+The only difference in the prior across the allowed region is that the physicality prior removes cases where $m_1$ is so small that it requires negative ionization at high-$z$.}
           \label{fig:prior_box}
@@
  where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
- On the other hand the bound is comparable to that obtained for the two-step model 
- \wh{hopefully true, point to the eqn when we get the numbers -- contrast this with Flexknot and talk about compatibility with example 2.}
+ On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~\ref{eq:2stepbound}).
+It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
 
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
-\caption{Comparison of the 68\% and 95\% C.L.
-constraints on the cumulative optical depth $\tau(z, \zmax)$ from PC chains with (shaded) and without (black) applying the physicality prior.
-While the lower limits on $\tau(z, \zmax)$ are lowered when more unphysical models are included, the upper limits do not vary significantly (slightly lowered).
-The total $\tau$ distribution is broadened only toward the lower $\tau$.
+\caption{Cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ from the model-independent PC approach with and without the physicality prior.
+The prior helps exclude models with negative contributions at high-$z$ but conservatively allows some negative values at $z \sim z_{\rm max}=30$ where the oscillations in the PCs no longer cancel due to the cumulative integration.
+Upper limits are nearly unaffected.}
 }
@@
 \includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2_do_rescale_tau_false.pdf}
-\caption{Comparison of the 68\% and 95\% C.L.
-constraints on the cumulative optical depth with different Planck likelihoods.
-\textit{Top}: The Planck 2015 vs Planck 2018 likelihoods, the latter including the official low-$\ell$ $EE$ likelihood $\texttt{simall\_EE}$.
-\textit{Bottom}: The Planck 2018 likelihoods using $\texttt{simall\_EE}$ for the low-$\ell$ $EE$ likelihood vs \texttt{srollv2} which has better foreground cleaning.
-Both yield consistent high-redshift constraints, whereas the \texttt{srollv2} likelihood tightens the lower limit on the low-$z$ contribution, resulting on a tighter lower limit and higher mean for the total $\tau$.
+%previouslky
+%{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2.pdf}
+\caption{Changes in the
+cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ (as in Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}) 
+between Planck 2015 and 2018 (top panel \wh{suggest swap to srollv2}) and between 2018 
+\texttt{srollv2} (default) and $\texttt{simall\_EE}$.
+Planck 2015 favored contributions at high $z$ which are disallowed in 2018 whereas the changes due to \texttt{srollv2} mainly tighten the lower limits at low $z$.
 }
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
-\caption{Comparison of the 68\% and 95\% C.L.
-constraints on the cumulative optical depth from PC chains with $\zmax = 30$ vs $\zmax = 50$ using the Planck 2018 likelihoods.
-The constraints on the cumulative high-redshift optical depth are shown to be robust to $\zmax$.
-Both the low-redshift constraints and the high-redshift upper limits agree well.
+\caption{Robustness of the
+cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.
+\wh{drop the Planck 2018 from label or at least make the $\zmax$ the prominent label}.
+Upper limits at $z \lesssim 30$ are nearly unaffected whereas extending $\zmax$ allows the high-$z$ component to come from even higher-$z$.
 }

### 2021-03-30 paper/rei.tex

@@
 Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
-$\texttt{srollv2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
+$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
 ~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{−0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.
 
-\todo{CH: might need to change all instances of \textt{srollv2} to \textt{SRoll2} as used in the original sroll2 paper; also possibly \textt{simall\_EE} to \textt{SimAll} as used in the Planck 2018 paper.}
-
 Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data.
-In light of the $\texttt{srollv2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
+In light of the $\texttt{SRoll2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
 Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.
@@
 We first describe in \S\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
-Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{srollv2} likelihood, and compare with previous results in literature.
+Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{SRoll2} likelihood, and compare with previous results in literature.
 Then in \S\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
@@
 We have tested that our results do not change if in lieu of \texttt{plik\_lite} we used the full \texttt{plik\_full} likelihood in which the foreground parameters have not been marginalized over.
-For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{srollv2} likelihood [add citation] in our official PC results and the effective likelihood code.
-In comparison with the official Planck-released \texttt{simall\_EE} likelihood, the \texttt{srollv2} likelihood improves foreground cleaning,
+For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{SRoll2} likelihood [add citation] in our official PC results and the effective likelihood code.
+In comparison with the official Planck-released \texttt{SimAll} likelihood, the \texttt{SRoll2} likelihood improves foreground cleaning,
 which enables stronger reionization constraints, especially for the low multipole moments that are important for the low redshift constraints.
@@
 We show  the 68\% and 95\% $m_a-m_b$ constraints as well as marginal constraints on individual $m_a$ for Planck 2018 in blue,
-using \texttt{srollv2} for the low-$\ell$ $EE$ likelihood, and compare that to the Planck 2015 constraints in red.  The corresponding marginal distributions for individual $m_a$ are also shown.
+using \texttt{SRoll2} for the low-$\ell$ $EE$ likelihood, and compare that to the Planck 2015 constraints in red.  The corresponding marginal distributions for individual $m_a$ are also shown.
 %The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle.
@@
 In Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, we also show
-the Planck 2018 constraints using the older \texttt{simall\_EE} likelihood.  The improvement 
-from switching to the \texttt{srollv2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher total optical depth (see also Fig.~\ref{fig:prior_box}).
+the Planck 2018 constraints using the older \texttt{SimAll} likelihood.  The improvement 
+from switching to the \texttt{SRoll2} likelihood mainly strengthens the upper bound in the best constrained direction in the $m_1-m_2$ plane which shifts constraints toward higher total optical depth (see also Fig.~\ref{fig:prior_box}).
 
@@
 \includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{srollv2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{simall\_EE} likelihood (\wh{green?}).
-With Planck 2018, the \texttt{srollv2} likelihood moderately improves the best constrained direction and raises the optical depth.
+\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{SRoll2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{SimAll} likelihood (\wh{green?}).
+With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.
 }
@@
 
-%In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{simall\_EE} and \texttt{srollv2}.
-The \texttt{srollv2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
+%In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{SimAll} and \texttt{SRoll2}.
+The \texttt{SRoll2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
 \wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}.
@@
 \ch{Our code is available on GitHub at \url{...} [give URL].
-} The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{srollv2}$ likelihoods at $\zmax = 30$.
+} The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
 The most important part of the code release is the python package called \texttt{Relike} which \wh{provides the PC parameters for a model and also} 
@@
 \beq
-\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068} \;\; (\texttt{srollv2}),
+\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068} \;\; (\texttt{SRoll2}),
 \label{eq:tau_pc_zmax30}
@@
 \beq
-\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{srollv2}).
-\eeq
-Reverting the Planck 2018 likelihood to \texttt{simall\_EE} at $z_{\rm max}=30$ would give
-\beq
-\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{simall\_EE}).
-\eeq
-As with the tanh model, the \texttt{srollv2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.
+\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
+\eeq
+Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
+\beq
+\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}).
+\eeq
+As with the tanh model, the \texttt{SRoll2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.
 
@@
 \beq
-\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}).
+\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{SRoll2}).
 \eeq
@@
 %\beq
-%\tau_{\rm PC,\, \zmax=30} = 0.0619^{+0.0056}_{-0.0068}\;\; (\texttt{srollv2}),
+%\tau_{\rm PC,\, \zmax=30} = 0.0619^{+0.0056}_{-0.0068}\;\; (\texttt{SRoll2}),
 %\eeq
@@
 %\beq
-%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{srollv2}).
+%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
 %\eeq
 %
-%Compared the \texttt{simall\_EE} likelihood results, 
+%Compared the \texttt{SimAll} likelihood results, 
 %\beq
-%\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{simall\_EE}),
+%\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}),
 %\eeq
@@
 %\beq
-%\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\texttt{simall\_EE}).
+%\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\texttt{SimAll}).
 %\eeq
-%the \texttt{srollv2} error bars on PC $\tau$ was reduced by about 20\%, and the central values rose by about half a sigma. [wanna comment more here?]
+%the \texttt{SRoll2} error bars on PC $\tau$ was reduced by about 20\%, and the central values rose by about half a sigma. [wanna comment more here?]
 %
@@
 %\beq
-%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{srollv2}),
+%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{SRoll2}),
 %\eeq
@@
 \beq
-\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{srollv2}).
+\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{SRoll2}).
 \eeq
@@
 cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ (as in Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}) 
-between Planck 2015 and 2018 (top panel \wh{suggest swap to srollv2}) and between 2018 
-\texttt{srollv2} (default) and $\texttt{simall\_EE}$.
-Planck 2015 favored contributions at high $z$ which are disallowed in 2018 whereas the changes due to \texttt{srollv2} mainly tighten the lower limits at low $z$.
+between Planck 2015 and 2018 (top panel) and between 2018 \texttt{SRoll2} (default) and $\texttt{SimAll}$.
+Planck 2015 favored contributions at high $z$ which are disallowed in 2018 whereas the changes due to \texttt{SRoll2} mainly tighten the lower limits at low $z$.
 }
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
-\caption{Robustness of the
+\caption{Robustness of the \ch{Planck 2018}
 cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.
-\wh{drop the Planck 2018 from label or at least make the $\zmax$ the prominent label}.
 Upper limits at $z \lesssim 30$ are nearly unaffected whereas extending $\zmax$ allows the high-$z$ component to come from even higher-$z$.
@@
 In conclusion, we used constraints on reionization principal components with respect to the large-scale $EE$ data to produce a fast and accurate effective likelihood for the final release of Planck 2018 data.
-In lieu of the official $\texttt{simall\_EE}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{srollv2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
+In lieu of the official $\texttt{SimAll}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{SRoll2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
 Our effective likelihood code package is called $\textsc{Relike}$ [to be changed] and is available publicly on Github.
@@
 
-    \item {The better foreground cleaned \texttt{srollv2} likelihood allowed to tighten the PC constraint on $\tau$ and shifted it toward slightly higher values when compared with the \texttt{plik\_lite\_TTTEEE+lowl+lowE} result  $\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}$ (68\% C.L.).}
+    \item {The better foreground cleaned \texttt{SRoll2} likelihood allowed to tighten the PC constraint on $\tau$ and shifted it toward slightly higher values when compared with the \texttt{plik\_lite\_TTTEEE+lowl+lowE} result  $\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}$ (68\% C.L.).}
 

### 2021-03-30 paper/rei.tex

@@
 
-\title{\ch{\textsc{RELIKE}}: Reionization Effective Likelihood from Planck 2018 Data (?)}
+\title{\ch{RELIKE}: Reionization Effective Likelihood from Planck 2018 Data (?)}
 
@@
 
-We publish \textsc{RELIKE} (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC).
+We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC).
 This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested.
@@
 \includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{SRoll2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{SimAll} likelihood (\wh{green?}).
+\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{SRoll2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{SimAll} likelihood (\wh{green?}) \ch{CH: ok will not label; taking a while since I can't plot a square anymore, figuring out aspect ratio}.
 With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.
@@
 (blue solid) and KDE Gaussian (black dashed)
-effective likelihood. The overall agreement is excellent and the KDE results even capture the small skewness of the exact result. \wh{maybe $\tau_{\rm tanh}$ }
+effective likelihood. The overall agreement is excellent and the KDE results even capture the small skewness of the exact result.
 }
@@
 Agreement between the exact, KDE PC and Gaussian PC likelihoods is again excellent and show that a small amount of high-$z$ ionization is allowed.
-\wh{The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_1 \sim 6-7$ especially for $z\gtrsim 15$.} \wh{Maybe $\tau_{\rm 2step}$}  }
+The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_1 \sim 6-7$ especially for $z\gtrsim 15$.
+}
 \label{fig:plot_taugtz_two_step_contours}
@@
 In lieu of the official $\texttt{SimAll}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{SRoll2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
-Our effective likelihood code package is called $\textsc{Relike}$ [to be changed] and is available publicly on Github.
+Our effective likelihood code package is called Relike and is available publicly on Github.
 This code is beneficial for the community for the purpose of testing \textit{any} global ionization history between $6 < z < 30$ with the Planck data without having to modify CAMB which can be very cumbersome to do.

### 2021-03-30 paper/rei.tex

@@
 
-\ch{Our code is available on GitHub at \url{...} [give URL].
-} The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
-The most important part of the code release is the python package called \texttt{Relike} which \wh{provides the PC parameters for a model and also} 
+Our code is available on GitHub at \todo{give URL\url{...}}.
+The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
+The most important part of the code release is the python package called RELIKE which \wh{provides the PC parameters for a model and also} 
 % added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
@@
 The returned likelihood is already marginalized over cosmological parameters.
-To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \texttt{Relike} within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.
+To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample RELIKE within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.
 
@@
 The KDE and Gaussian PC results  oscillate around zero at high-z because of effective low pass filter of the truncation to 5 PCs (enough to represent any observable effects in the data).
-\wh{maybe $\tau_{\rm tanh}$} }
-%double check plot, upper and lower limits seem to cross?
+}
 \label{fig:plot_taugtz_PC_vs_tanh}

### 2021-04-01 paper/rei.tex

@@
 The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
+\wh{This is also the default analysis for all results that follow except where noted.}
 The most important part of the code release is the python package called RELIKE which \wh{provides the PC parameters for a model and also} 
@@
 Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters
-\wh{which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
-suggested change in model - or find another reason for plotting the currently chosen one}. \wh{add inset ``two step"} 
+which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
+suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot coming up} \wh{add inset ``two step"} 
 }
@@
 Fig.~\ref{fig:two_step_model} has
-$\tau_{2\text{step}}(15,30)= 0.013$.} 
-\wh{possibly choose a model on the 2D contour at max imum $\tau_{\rm hi}$}
-%0.01343375 evaluated at fiducial cosmology
+$\tau_{2\text{step}}(15,30)= 0.017$.}
 
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
-\caption{Robustness of the \ch{Planck 2018}
+\caption{Robustness of the \ch{Planck 2018} \wh{decide on convention - in many places the default analysis isn't labeled, only changes from the default}
 cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.

### 2021-04-02 paper/rei.tex

@@
 Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (0.039, 0.020)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
-\wh{inset label for ``two step"}}
+$(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
+}
 %best-fit (0.05290344, 0.005859431)
-%95%CL (0.039, 0.02) 
+%95%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
 \label{fig:two_step_model}
@@
 %\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
+%
+% longer Gaussian chains only 
+%\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
+%
+% longer KDE and Gaussian chains
+\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
 \caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L.

### 2021-04-02 paper/rei.tex

@@
 \includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (\wh{red}), Planck 2018 (blue, \texttt{SRoll2}  \wh{decide whether to have this labeled}), and Planck 2018 but with the \texttt{SimAll} likelihood (\wh{green?}) \ch{CH: ok will not label; taking a while since I can't plot a square anymore, figuring out aspect ratio}.
+\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green).
 With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.
@@
 effective likelihood approach described in~\S\ref{sec:effective_likelihood}.  The benefit of this approach is that
-complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function \wh{and jointly sample the other cosmological parameters} in for example CAMB, each time a new ionization model needs to be tested.
+complete constraints on any model of the ionization history  between $6 \leq z \leq \zmax$ are obtained ahead of time and compressed into a simpler likelihood, so that when doing model testing one no longer needs to modify the $x_e(z)$ function and jointly sample the other cosmological parameters in for example CAMB, each time a new ionization model needs to be tested.
 These provide savings on human time in addition to the computing speed-ups from the likelihood evaluations.
@@
 The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
-\wh{This is also the default analysis for all results that follow except where noted.}
-The most important part of the code release is the python package called RELIKE which \wh{provides the PC parameters for a model and also} 
+This is also the default analysis for all results that follow except where noted.
+The most important part of the code release is the python package called RELIKE which provides the PC parameters for a model and also
 % added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
@@
 
-For convenience, 
 \wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
@@
 \wh{not sure we need timing results but if you do add them remember you can make Gaussian even faster by precomputing the linear combinations of $m_a$ that diagonalize the covariance (the first one should essentially be $\tau_{12}$ which is interesting pedagogically)} \ch{[CH: got it.]}
-\ch{
+%
 We obtain for the tanh model 
@@
 is able to capture some of the skewness in the exact distribution.
-The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods is consistent with this level of skewness ($\Delta \tau \sim 0.001$), which is $<0.2\sigma$.} % 95\% UL:
+The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods is consistent with this level of skewness ($\Delta \tau \sim 0.001$), which is $<0.2\sigma$.
+% 95\% UL:
 %tanh Gauss: 1.996574506651048464e-03
@@
 
-[Note sure if it is related to this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
+%[Note sure if it is related to this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
 Then transition to the following point]
@@
 which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
-suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot coming up} \wh{add inset ``two step"} 
+suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains coming up}
 }
@@
  \eea
- \wh{naming is slightly odd here - maybe $z_{\rm re} \rightarrow z_1$ and $z_{\rm t}\rightarrow z_2$?
- }\ch{CH: Actually I think it's good to keep zre because it is when the transition to full reionization happens, and is the same nomenclature people are used to with the tanh model.
-We could find a different name for $z_t$ if you'd like?}
-where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{\rm re})$ just as in the first example for a sharper step.
-We choose the second step at $z_{\rm t}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
-In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_t$.
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_1)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{1})$ just as in the first example for a sharper step.
+We choose the second step at $z_{2}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
+In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{1} \lesssim z \lesssim z_2$.
 We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
@@
 
-\ch{The marginalized 1D constraints from the KDE and Gaussian likelihoods 
+The marginalized 1D constraints from the KDE and Gaussian likelihoods 
 \beq
@@
 \beq
-\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Exact).}
+\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Exact).
 \eeq 
@@
 
-\wh{Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$: give $\tau_{2\rm step}$.}
-
-\ch{Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
+Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$: \todo{give $\tau_{2\rm step}$.}
+
+Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
 \begin{equation}
@@
 and 
-$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods}
-\ch{Note also that the 95\% C.L. allowed two-step model in 
+$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods
+Note also that the 95\% C.L. allowed two-step model in 
 Fig.~\ref{fig:two_step_model} has
-$\tau_{2\text{step}}(15,30)= 0.017$.}
+\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}
 
@@
 This was especially apparent with the Planck 2015 data where
-the PC constraints gave \todo{$\tau=XX\pm YY$ which differed substantially from the tanh $\tau=XX\pm YY$}.  As we shall discuss
+the PC constraints gave \ch{$\tau=0.092\pm0.015$ which differed substantially from the tanh $\tau=0.079\pm0.017$~\cite{Heinrich:2016ojb}}.  As we shall discuss
 below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.
@@
 To see this, consider a range-bound prior that is rotated to be oriented along constant $\tau_{12}$ (see Fig.~\ref{fig:prior_box}, blue dashed rectangle).
-\ch{This rotated prior in $m_1-m_2$ space corresponds to a flat prior in $\tau_{12}$ space.}
+This rotated prior in $m_1-m_2$ space corresponds to a flat prior in $\tau_{12}$ space.
 Had the original prior (blue square) been shifted to slightly lower
@@
 95\% CL.   This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would still change the original prior but not the
-rotated one.  \ch{CH: will try rephrasing.}
+rotated one.  
+%will try rephrasing.}
 
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
-\caption{Robustness of the \ch{Planck 2018} \wh{decide on convention - in many places the default analysis isn't labeled, only changes from the default}
+\caption{Robustness of the 
+%\ch{Planck 2018} \wh{decide on convention - in many places the default analysis isn't labeled, only changes from the default}
 cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.
@@
     \item {We checked the robustness of our $\zmax=30$ results by performing a PC analysis of the Planck data with $\zmax = 50$.
-The results are changed negligibly: $\tau(15, 50) < 0.022 $ (95\% C.L.) in comparison with $\tau(15, 30) < 0.023$ (95\% C.L.).}
-
-     \item {Though our high-redshift optical depth constraints are looser than those from the Planck 2018 cosmological parameter paper $\tau(15, 30) < 0.007$ (95\% C.L.) obtained from another model-independent method called FlexKnot, we have tested using explicitly the tanh high-$z$ toy model (without using any PCs) that an $\tau(15, 30) < 0.02$ (95\% C.L.) [get more precise numbers] is still allowed (see Fig.~\ref{fig:plot_taugtz_two_step_contours}), which is more consistent with our PC results.}
+The results are changed negligibly: $\tau_{\rm PC}(15, 50) < 0.022 $ (95\% C.L.) in comparison with $\tau_{\rm PC}(15, 30) < 0.023$ (95\% C.L.).}
+
+     \item {Though our high-redshift optical depth constraints are looser than those from the Planck 2018 cosmological parameter paper $\tau(15, 30) < 0.007$ (95\% C.L.) obtained from another model-independent method called FlexKnot, we have tested using explicitly the tanh high-$z$ toy model (without using any PCs) that an $\tau_{\rm 2step}(15, 30) < 0.017$ (95\% C.L.) is still allowed (see Fig.~\ref{fig:plot_taugtz_two_step_contours}), which is more consistent with our PC results.}
 

### 2021-04-09 paper/rei.tex

@@
 Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.).
-The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering  \ch{$\tau_{\rm 2step}(15, 30) < 0.013 $ (95\% C.L.)} using a  two-step toy model without using PCs.}
+The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering $\tau_{\rm 2step}(15, 30) < 0.017 $ (95\% C.L.) using a  two-step toy model without using PCs.}
 
@@
 Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced.
-This steplike \ch{tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition.
+This \ch{steplike tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition.
 However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.
@@
 $\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
-~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{−0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.
+~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}
 
@@
 
-\ch{write this part}
+\todo{write this part}
 The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].
@@
 Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release.
-But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is XX times less stringent than the Planck official results using Flex Knots \ch{[fill in number]} in Ref.~\cite{Aghanim:2018eyx}.
+But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is about \ch{$\sim 3$ times} less stringent than the Planck official results using Flex Knots in Ref.~\cite{Aghanim:2018eyx}.
 Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
 \caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green).
 With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.
-}
+\todo{add SRollv2.
+convention: put labels only when we do comparisons}}
 \label{fig:plot_m1m2_2015_vs_2018}
@@
 
-Our code is available on GitHub at \todo{give URL\url{...}}.
+Our code is available on GitHub at {give URL\url{...}}.
 The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
@@
 \wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
-we also provide a pre-assembled CosmoMC-Relike code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
-This CosmoMC-Relike code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.
+We also provide a pre-assembled CosmoMC code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
+This CosmoMC code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.
 
@@
 The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain.
-\todo{[add timing here]}
+{[add timing here]}
 \wh{not sure we need timing results but if you do add them remember you can make Gaussian even faster by precomputing the linear combinations of $m_a$ that diagonalize the covariance (the first one should essentially be $\tau_{12}$ which is interesting pedagogically)} \ch{[CH: got it.]}
@@
 $(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
-}
+}\todo{CH: say in text 0.026 is 2D marginalized 95\%, 0.02 is 1D.}
 %best-fit (0.05290344, 0.005859431)
@@
 
+\todo{change z1 z2 to zre and zhi everywhere.}
 The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
@@
 
-\todo{add timing results somewhere}
+{add timing results somewhere}
 
@@
 \beq
-\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (KDE)}
+\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., KDE)}
 \eeq 
@@
 \beq
-\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Gaussian)}
+\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., Gaussian)}
 \eeq 
@@
 \beq
-\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; \mathrm{(95\%\; C.L.)\; (Exact).
+\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., Exact).
 \eeq 
@@
 
-Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$: \todo{give $\tau_{2\rm step}$.}
-
+\ch{comment here about 2D 95\% model picked with tau hi = 0.026 is higher because 0.021 is the 1D marginalized result.}
+Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$:
+\ch{
+\beq
+\tau_{\rm 2step} = 0.0621^{+0.0062}_{-0.0061}.
+\eeq
+}
 Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
@@
 Fig.~\ref{fig:two_step_model} has
-\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}
+\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}\todo{Need to change all 95\% upper limit on tau(15,30) to a one-sided computation instead of 2.}\todo{update plot two-step with longer chain results, run 2 more days exact two-step chains.} \todo{point out in text that 95\% upper limit on tau(15,30) not the same as plot.}
 
@@
 correspond to 
-\beq
-\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068} \;\; (\texttt{SRoll2}),
+%
+\beq
+\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068},
 \label{eq:tau_pc_zmax30}
 \eeq  
+%
 and the full posterior is shown in Fig.~\ref{fig:tauPC}. 
@@
 This was especially apparent with the Planck 2015 data where
-the PC constraints gave \ch{$\tau=0.092\pm0.015$ which differed substantially from the tanh $\tau=0.079\pm0.017$~\cite{Heinrich:2016ojb}}.  As we shall discuss
+the PC constraints gave \ch{$\tau_{\rm PC}=0.092\pm0.015$ which differed substantially from the tanh $\tau_{\rm PC}=0.079\pm0.017$~\cite{Heinrich:2016ojb}}.  As we shall discuss
 below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.
@@
 $z_{\rm max}=50$ which is negligibly different
-\beq
-\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
-\eeq
+\ch{
+\beq
+%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
+\tau_{\rm PC} = 0.0626^{+0.0061}_{-0.0072}\quad (\zmax=50).
+\eeq
+CH: removing SRoll2 label in parenthesis and moving zmax=50 into the parenthesis.}
 Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
@@
 Moreover upper bounds on the optical depth at high
-redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example
-\beq
-\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{SRoll2}).
-\eeq
+redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example,
+\ch{
+\beq
+\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.}).
+\eeq
+}
 This should be contrasted with the tanh model 
@@
 It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
+ \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..}
 
@@
 \beq
-\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.})\;\;(\texttt{SRoll2}).
+\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.}).
 \eeq
@@
 The prior helps exclude models with negative contributions at high-$z$ but conservatively allows some negative values at $z \sim z_{\rm max}=30$ where the oscillations in the PCs no longer cancel due to the cumulative integration.
-Upper limits are nearly unaffected.}
+Upper limits are nearly unaffected.
+\todo{change all tau cumulative plots for PC to $\tau_{\rm PC}$;} \todo{add (with prior)}}
 }

### 2021-04-16 paper/rei.tex

@@
 
-\title{\ch{RELIKE}: Reionization Effective Likelihood from Planck 2018 Data (?)}
+\title{RELIKE: Reionization Effective Likelihood from Planck 2018 Data}
 
@@
 We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler.
-We expect the code release to dramatically reduce the time for testing individual \ch{global} ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as 21cm, luminosity function, star formation history, etc.
+We expect the code release to dramatically reduce the time for testing individual global ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as \ch{line-intensity mapping},
+%21cm,
+luminosity function, star formation history, etc.
 Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.).
@@
 
-\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}.}
+\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}.
+mean free path of ionizing photons  between 5 and 6: \url{https://arxiv.org/pdf/2103.16610.pdf}, $<80\%$ ionized at z = 6.
+}
 
@@
 
-\ch{I'm here}
 Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
@@
 
-\todo{write this part}
-The code is publicly available on GitHub\footnote{[give link]}. [a bit more description here]. [comment on speed]. [comment on joint likelihood].
+The code is publicly available on GitHub at \url{https://github.com/chenheinrich/RELIKE}.
+We publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model.
+If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space; we provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in \texttt{CosmoMC-relike}.
+We expect the code release to dramatically reduce the time for testing individual models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity function, star formation history, 21cm or line-intensity mapping in general, etc.
 
@@
 implements the Gaussian effective likelihood.
-The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \textt{Relike} to return the effective likelihood of the model.
+The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use RELIKE to return the effective likelihood of the model.
 The returned likelihood is already marginalized over cosmological parameters.
@@
 
-\wh{query: since this is the place where KDE is given not sure ``convenience" is the transition we want here.   Do you want to say something about KDE up front?}
 We also provide a pre-assembled CosmoMC code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
@@
 the 95\% upper limit on the high-redshift optical depth.
-For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while $\tau_{\rm tanh}(15,\zmax) \approx 0$ from the exact likelihood.
+For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while \ch{$\tau_{\rm tanh}(15,30) \approx 0$}
+%$\tau_{\rm tanh}(15,\zmax) \approx 0$ 
+from the exact likelihood. \ch{NEW: since both one-sided and two-sided are 0.002 here, I'm doing the one-sided vs two-sided explaination in the two-step section instead.}
 Ignoring this small effect of the oscillations, 
@@
 $(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
-}\todo{CH: say in text 0.026 is 2D marginalized 95\%, 0.02 is 1D.}
+}
 %best-fit (0.05290344, 0.005859431)
@@
 which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
-suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains coming up}
+suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains (still running)}
 }
@@
 
-\ch{comment here about 2D 95\% model picked with tau hi = 0.026 is higher because 0.021 is the 1D marginalized result.}
+\ch{NEW: $\tau_{\rm hi}$ upper limits are all one-sided already.}\\
+
+\ch{Note that the 95\% upper limit on $\tau_{\rm hi} < 0.021$ presented here is a 1D marginalized result; whereas the model shown in Figs~\ref{fig:two_parameter_model_2D} and~\ref{fig:two_parameter_model_2D} which has $\tauhi = 0.026$ is a model picked at the 95\% C.L.
+in the 2D marginalized case, hence the difference from the 1D result here.}
 Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$:
@@
 \end{equation}
+\todo{CH: still missing one-sided $\tau_{2\text{step}}(15, 30)$ for exact.}
 and 
-$\tau_{2\text{step}}(15, 30) < 0.017$ for both effective likelihoods
+\ch{$\tau_{2\text{step}}(15, 30) < 0.016$} for both effective likelihoods. %changed from <0.017--> <0.016 when going from two-sided to one-sided.
 Note also that the 95\% C.L. allowed two-step model in 
 Fig.~\ref{fig:two_step_model} has
-\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}\todo{Need to change all 95\% upper limit on tau(15,30) to a one-sided computation instead of 2.}\todo{update plot two-step with longer chain results, run 2 more days exact two-step chains.} \todo{point out in text that 95\% upper limit on tau(15,30) not the same as plot.}
+\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}
+\ch{Note that the 95\% upper limit on $\tau(15, 30)$ reported in the text are one-sided upper limits, which are slightly different than the two-sided upper limits plotted in the figures of cumulative optical depth (e.g.
+Fig.~\ref{fig:plot_taugtz_two_step_contours})}
 
@@
 In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this two-step model.
-In particular, we show the 68\% and 95\% limits on $\tau(z, \zmax)$ obtained by integrating over the exact $x_e(z)$ two-step model in the exact likelihood chains, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.
+we show the 68\% and 95\% limits on $\tau(z, \zmax)$ obtained by integrating over the exact $x_e(z)$ two-step model in the exact likelihood chains, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.
 As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.
@@
 \tau_{\rm PC} \approx \tau_{12} \equiv  m_1 \tau_1 + m_2 \tau_2 +
-\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) 
+\left(\sum_{a=3}^5 \bar m_a \tau_a + \tau_{\rm fid}\right) ,
 \label{eq:tau12}
@@
 To see this, consider a range-bound prior that is rotated to be oriented along constant $\tau_{12}$ (see Fig.~\ref{fig:prior_box}, blue dashed rectangle).
-This rotated prior in $m_1-m_2$ space corresponds to a flat prior in $\tau_{12}$ space.
-Had the original prior (blue square) been shifted to slightly lower
-$m_1$ and encompassed all of the 95\% CL allowed
-region the two priors would yield  identical constraints at the
-95\% CL.   This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would still change the original prior but not the
-rotated one.  
+This rotated prior in $m_1-m_2$ space corresponds to a constant $P(\tau_{12}$) once the direction orthogonal to $\tau_{12}$ is marginalized.
+Had the  \wh{data constraint been shifted to slightly higher
+$m_1$ so that both prior ranges encompassed all of the 95\% CL allowed
+region they both would yield  the same constraints at that level.}
+% rephrased to flow with the next paragraph of 2015->2018
+This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would  change the original prior but not the
+rotated one and \wh{destroy this desired equivalence.
+More generally the point is that for a multidimensional space, constructing a flat marginal 1D distribution per parameter in the absence of data does not guarantee a locally flat prior in the region supported by the data
+(see \cite{Heinrich:2018btc} for further discussion).}
 %will try rephrasing.}
@@
 $z_{\rm max}=50$ which is negligibly different
-\ch{
+%
 \beq
@@
 \eeq
-CH: removing SRoll2 label in parenthesis and moving zmax=50 into the parenthesis.}
+%
 Reverting the Planck 2018 likelihood to \texttt{SimAll} at $z_{\rm max}=30$ would give
@@
 raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
-Consequently, as shown in Fig.~\ref{fig:prior_box}, the second aspect of PC constraints beyond the well constrained total optical depth, as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12}), is a positive
-correlation between the two components that represents the high to low $z$ 
+Consequently, as shown in Fig.~\ref{fig:prior_box}, beyond the well constrained total optical depth, as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12}), orthogonal to this combination is a somewhat weaker constraint on the the high to low $z$
 contribution to the total optical depth.  Any finer redshift detail in the ionization history is much more poorly constrained.
@@
 In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
-the 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.
+the \ch{two-sided} 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.
 Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.
@@
 \beq
-\tau_{\rm PC}(15, 30) < 0.023 \; (95\%\; \mathrm{C.L.}).
-\eeq
-}
+\tau_{\rm PC}(15, 30) < 0.020 \; (95\%\; \mathrm{C.L.}).
+\label{eq:tauPC15}
+\eeq
+}
+\ch{Recall again that we are reporting the one-sided upper limit here in the text, which is slightly lower than the two-sided limits shown in the figures.}
+%
+\ch{(NEW: changed from 0.023 to 0.020 going from two-sided to one-sided.)}
 This should be contrasted with the tanh model 
@@
 It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
- \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..}
+ \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..} \ch{$\tau_{\rm hi}$ values are confirmed to be one-sided now!
+:) }
 
@@
 In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.
-Improvements from srollv2 come mainly from improving the lower bound at low redshift, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.
+Improvements from \texttt{SRoll2} come mainly from improving the lower bound at low redshift, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.
 
@@
 
-For the high-redshift optical depth, we also obtain consistent upper limits
-%\beq
-%\tau_{\rm PC}(15, 30)< 0.023 \; (95\%\; %\mathrm{C.L.})\;\;(\texttt{SRoll2}),
-%\eeq
-%[double check on full chain (from 10/15) this is from 09/30.]
-%and 
-\beq
-\tau_{\rm PC}(15, 50) < 0.022 \; (95\%\; \mathrm{C.L.}).
-\eeq
+For the cumulative high-redshift optical depth,
+\beq
+\tau_{\rm PC}(15, 50) < 0.019 \; (95\%\; \mathrm{C.L.}),
+\eeq
+we also obtain consistent upper limits to the $\zmax=30$ case of Eq.~(\ref{eq:tauPC15}).
+\ch{NEW: changed 0.022 to 0.019 going from two-sided to one-sided}
 
@@
 
-\begin{itemize}
-    \item other interesting points to discuss and add 
-    \item future outlook (mention joint analyses)
-\end{itemize}
+In the future, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
+Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f.
+adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects.
+[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data?
+More importantly with data in general that wants to vary As during sampling?
+Put prior on As consistent with Planck?].
+
+There has also been recent studies revealing that reionization might not have ended by $z = 6$~\cite{} [cite].
+If this is true, the current PC approach will need be modified to accommodate a range $z_{\rm min} < z < z_{\rm max}$ with a a lower $z_{\rm min} < 6$.
+We leave this for future work as it would require constructing a new set of PC functions.
+
+[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]
+
+In sum, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on specific ionization models, as well as to open the way to jointly constrain legacy CMB large-scale data with a variety of increasingly rich reionization datasets coming up in the next decade.
 

### 2021-04-16 paper/rei.tex
Updates from Overleaf

@@
 $\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
-~\cite{Delouis:2019bub}, allows for improved reionization constraints with its better foreground modeling: The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}
-
-Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data.
-In light of the $\texttt{SRoll2}$ likelihood release, we obtain new reionization PC constraints with this latest likelihood.
-Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.
-
-The code is publicly available on GitHub at \url{https://github.com/chenheinrich/RELIKE}.
-We publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model.
-If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space; we provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in \texttt{CosmoMC-relike}.
-We expect the code release to dramatically reduce the time for testing individual models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity function, star formation history, 21cm or line-intensity mapping in general, etc.
+~\cite{Delouis:2019bub}, adopts better foreground modeling techniques and allowing for improved reionization constraints.
+%
+%The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.
+%\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}
+%
+\ch{Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data.
+In this paper, we obtain new reionization PC constraints with this latest likelihood $\texttt{SRoll2}$.
+Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood code useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.
+}
+
+\ch{We publish this reionization effective likelihood code (RELIKE) which is available on GitHub at \url{https://github.com/chenheinrich/RELIKE}.
+There are two packages that come with the code release: 1) we publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model.
+If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space.
+2) We also provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in the package \texttt{CosmoMC-relike}.
+We expect that this code release would allow much simpler and faster testing of specific models in the Planck data, and to provide consistent and fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity functions, star formation history, 21cm and line-intensity mapping in general, etc.
+}
 
@@
 which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
-suggested change in model - or find another reason for plotting the currently chosen one. \ch{updated plot with longer exact chains (still running)}
+suggested change in model - or find another reason for plotting the currently chosen one.
 }
@@
 adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects.
-[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data?
+\ch{[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data?
 More importantly with data in general that wants to vary As during sampling?
-Put prior on As consistent with Planck?].
+Put prior on As consistent with Planck?]}.
 

### 2021-04-16 paper/rei.tex
Updates from Overleaf

@@
 \caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green).
-With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.
-\todo{add SRollv2.
-convention: put labels only when we do comparisons}}
+With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.}
 \label{fig:plot_m1m2_2015_vs_2018}
@@
 Agreement between the exact, KDE PC and Gaussian PC likelihoods is again excellent and show that a small amount of high-$z$ ionization is allowed.
-The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_1 \sim 6-7$ especially for $z\gtrsim 15$.
+The form of the model requires that the cumulative optical depth steadily declines above the first step at $z_{\rm re} \sim 6-7$ especially for $z\gtrsim 15$.
 }
@@
 
-\todo{change z1 z2 to zre and zhi everywhere.}
 The second example is a two-step model, where an additional step is used to give a high-$z$ ionization plateau: 
  \bea
-x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{1})-y(z)}{\Delta y} \right] \right\} \notag \\
-&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm t}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
+x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
+&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm hi}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
  \label{eqn:tanh_highz}
  \eea
-where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_1)^{1/2}\Delta z_1$ where $\Delta z_1 = 0.015\, (1+z_{1})$ just as in the first example for a sharper step.
-We choose the second step at $z_{2}=28$ with $\Delta z_2 = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
-In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{1} \lesssim z \lesssim z_2$.
+where $y(z)=(1+z)^{3/2}$, $\Delta y=(3/2)(1+z_{\rm re})^{1/2}\Delta z_{\rm re}$ where $\Delta z_{\rm re} = 0.015\, (1+z_{\rm re})$ just as in the first example for a sharper step.
+We choose the second step at $z_{\rm hi}=28$ with $\Delta z_{\rm hi} = 1.0$, since the effective likelihood is limited to models with ionization below $\zmax=30$. 
+In comparison to tanh, a single parameter $\xemin$ is added in this model to control the high-$z$ ionization plateau for $z_{\rm re} \lesssim z \lesssim z_{\rm hi}$.
 We would recover the standard tanh as $\xemin$ approaches the negligible recombination value $x_e^{\rm rec}$.
@@
 In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.
-They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_2$ to be conservative on the preference of $\tauhi$ in the data.
-[double check this too.]
+They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_{\rm re}$ to be conservative on the preference of $\tauhi$ in the data.
 Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$.
@@
 Upper limits are nearly unaffected.
-\todo{change all tau cumulative plots for PC to $\tau_{\rm PC}$;} \todo{add (with prior)}}
 }
@@
 
-In the future, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
+In the future, we expect our code release RELIKE to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
 Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f.
@@
 
-[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]
+\todo{[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]
+}
 

### 2021-04-28 paper/rei.tex
Final copy

@@
 
+\newcommand{\relike}{RELIKE}%\texttt{RELIKE} 
+\newcommand{\cosmomcrelike}{\texttt{CosmoMC-relike}}
+
 \def\mnras{Mon.\ Not.\ R.\ Astron.\ Soc.\ }
@@
 
-\title{RELIKE: Reionization Effective Likelihood from Planck 2018 Data}
+\title{\relike: Reionization Effective Likelihood from Planck 2018 Data}
 
@@
 
-We publish RELIKE (Reionization Effective Likelihood), an fast and accurate effective likelihood code based on constraints from the latest Planck 2018 data on the CMB reionization principal components (PC).
-This code allows one to constrain arbitrary global ionization history $x_e(z)$ between $6 < z < 30$ with Planck without having to modify CAMB each time a different model is being tested.
-We tested the code on two example models which showed good agreement with sampling the exact Planck likelihoods.
-We also find that contrary to Planck 2015, the Planck 2018 PC posteriors are well approximated by a Gaussian in PC amplitude space, this allows us to replace the kernel density estimate (KDE) evaluation of the PC chains in the code with an even faster Gaussian evaluation.
-We supply both the KDE and Gausisan modes of the likelihood in a CosmoMC implementation, and the Gaussian mode alone in an easy-to-use standalone python likelihood package that can be connected to any other sampler.
-We expect the code release to dramatically reduce the time for testing individual global ionization models, and to provide consistent yet fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as \ch{line-intensity mapping},
-%21cm,
-luminosity function, star formation history, etc.
-Finally, we derive model-independent constraints: $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and $\tau_{\rm PC}(15, 30) < 0.023 $ (95\% C.L.).
-The upper limits on high-redshift optical depth is at least a factor of 2 looser than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method; we demonstrate robustness of our results by explicitly recovering $\tau_{\rm 2step}(15, 30) < 0.017 $ (95\% C.L.) using a  two-step toy model without using PCs.}
+We release \relike\  (Reionization Effective Likelihood), a fast and accurate effective likelihood code based on the latest Planck 2018 data 
+that allows one constrain any model for reionization between $6 < z < 30$ using five constraints from the CMB reionization principal components (PC).
+We tested the code on two example models which showed excellent agreement with sampling the exact Planck likelihoods using either a simple Gaussian PC likelihood or its full kernel density estimate.
+This code enables a fast and consistent means for 
+combining Planck constraints with other reionization data sets, such as kinetic Sunyaev-Zeldovich effects, line-intensity mapping, luminosity function, star formation history, quasar spectra, etc, where the redshift dependence of the ionization history is important.
+Since the PC technique tests any reionization history in the given range, we also derive model-independent constraints for the total Thomson optical depth $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and its $15\le z \le 30$ high redshift component $\tau_{\rm PC}(15, 30) < 0.020 $ (95\% C.L.).
+The upper limits on the high-redshift optical depth is a factor of $\sim3$ larger than those reported in the Planck 2018 cosmological parameter paper using the FlexKnot method
+and we validate our results with a direct analysis of a two-step model which permits this small high-$z$ component.
 
@@
 
-\ch{CH: going through the paper once from beginning to end now.}
-
-\ch{CH: may also want to comment on Ref.~\cite{Paoletti:2020ndu} and other papers under paper/refs in overleaf (also just found \url{https://arxiv.org/pdf/2103.14378.pdf}.
-mean free path of ionizing photons  between 5 and 6: \url{https://arxiv.org/pdf/2103.16610.pdf}, $<80\%$ ionized at z = 6.
-}
-
-The cosmic microwave background (CMB) has entered an era of precision cosmology.
-The measurements from the Planck satellite have shown agreement with the standard $\Lambda$CDM model which describes the initial perturbations in the Universe and their evolution~\cite{Aghanim:2018eyx}.
-While many components of the standard cosmology model is well-understood, the details of the process of reionization however, remains one of the most uncertain pieces (see e.g.~\cite{2016ASSL..423.....M} and references therein).
-Its uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude.
+The cosmic microwave background (CMB) has helped establish the $\Lambda$CDM model as the standard cosmological model~\cite{Aghanim:2018eyx}.
+While many components of this standard cosmology model are well-understood, the details of the process of reionization however, remains one of the most uncertain pieces (see e.g.~\cite{2016ASSL..423.....M} and references therein).
+It is usually characterized by a single parameter, the total Thomson optical depth, whose  uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude.
 Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.
 
-Typically, the impact of reionization on the primary CMB fluctuations has been modeled as a steplike transition in the global ionization history \ch{in the form of a tanh function}, with the step location parameterized by the total Thomson optical depth induced.
-This \ch{steplike tanh} model describes a Universe in which all hydrogen becomes fully ionized almost instantaneously at one particular redshift, and assumes, by construction, that there is negligible ionization before the transition.
-However, through the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles, more information on the coarse-grained evolution of the ionization history can be obtained~\cite{Hu:2003gh, Mortonson:2007hq}.
-
-In fact, to extract the most information possible from this $E$-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum.
+Many probes of  the reionization era exist today and with future probes it will only become more important to combine them consistently the full implications of the CMB.
+In the CMB data, free electrons during the reionization epoch Compton scatter with the CMB photons to cause a suppression of the CMB primary anisotropies; on the large scales, the same scattering process of the CMB temperature quadrupole in the frame of the free electrons induces an additional reionization bump in the CMB $E$-mode polarization.
+On the small scales, the kinetic Sunyaev-Zeldovich (kSZ) effects generated by the scattering of the CMB photons off ionized gas in bulk motion~\cite{zeldovich_sunyaev_1969, sunyaev_zeldovich_1980} are used to constrain morphology and duration of reionization~\cite{mcquinn_2005,mesinger_2012_kSZ}.
+Additionally, observations of galaxy luminosity functions, quasars, star formation rate, line-intensity mapping also provide important and complementary information on reionization~\cite{2016ASSL..423.....M}.
+In this paper, we aim to extract all the information in the large-scale $E$-mode observations for constraining the global ionization history of the Universe, and provide a way to consistently analyze them with all of these other probes.
+
+In the standard CMB analysis, reionization has been  modeled as a steplike transition in the global ionization history in the form of a tanh function, with the step location parameterized by the total Thomson optical depth induced.
+This steplike tanh model assumes, by construction, that there is negligible ionization before the transition.
+However, the shape of the reionization bump induced in the CMB $E$-mode polarization at large angles contains more information on the coarse-grained evolution of the ionization history than the total optical depth~\cite{Hu:2003gh, Mortonson:2007hq}.
+
+To extract the most information possible from this $E$-mode bump, Ref.~\cite{Hu:2003gh, Mortonson:2007hq} developed the principal component (PC) method, where a few PCs is sufficient to describe the entire model space of physical ionization histories regarding their observable impact on the large-angle $E$-mode power spectrum.
 This method has been applied to WMAP and Planck data to obtain complete constraints on reionization models~\cite{Mortonson:2008rx, Mortonson:2007hq, Heinrich:2016ojb, Aghanim:2018eyx}.
@@
 %\cite{Obied:2017tpd}
-
-In a re-analysis of the Planck 2015 data with PCs~\cite{Heinrich:2016ojb}, a component of the high-redshift ionization that would have been missed by a simple steplike model~\cite{Heinrich:2016ojb} was uncovered.
-In the latest official Planck 2018 release~\cite{Aghanim:2018eyx}, the PC method was also adopted to probe ionization at high-redshift, whose significance was reduced since the Planck 2015 release, largely due to the reduction of systematics at large-scales~\cite{Aghanim:2018eyx, Heinrich:2018btc}.
-In addition to PCs, the FlexKnot method was employed, which is also able to capture general ionization histories by varying the number of ``knots" in redshifts and the amplitude of the ionization fraction at these knots.
-% Planck intermediate results on reionization history: \cite{Adam:2016hgk}
+In a re-analysis of the Planck 2015 data with PCs~\cite{Heinrich:2016ojb}, a component of the high-redshift ionization that was missed by a simple steplike model was mildly preferred.
+
+In the latest official Planck 2018 release~\cite{Aghanim:2018eyx},  this mild preference for finite high redshift ionization in the Planck 2015 release turned into upper limits, largely due to the reduction of systematics at large-scales~\cite{Aghanim:2018eyx, Millea:2018bko} resulting in reduced $C_l^{EE}$ around $l \lesssim 10$ (see also Ref.~\cite{Heinrich:2018btc}).
+The FlexKnot method was primarily employed, which is also able to capture general ionization histories by varying the number of ``knots" in redshifts and the amplitude of the ionization fraction at these knots.
 
 Since the release of the Planck 2018 official analysis, an improved likelihood for the low-$\ell$ $E$-mode polarization was publicly released in 2019. This new likelihood,
-$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}
-~\cite{Delouis:2019bub}, adopts better foreground modeling techniques and allowing for improved reionization constraints.
-%
-%The error bar on the optical depth $\tau$ in the tanh model is reduced by about $\sim40\%$ from XX to $\tau = 0.0566^{+0.0053}_{-0.0062}$~\cite{}.
-%\ch{doesn't seem to be true anymore, 40\% was compared to the Planck 2018 likelihood paper result, the cosmo parameter paper seems to have similar error bars: $0.0519^{+0.0030}_{−0.0079}$.}
-%
-\ch{Since the Planck final release will likely be the best full-sky survey from space available to us in the next decade, it is important to extract all the information present in the data.
-In this paper, we obtain new reionization PC constraints with this latest likelihood $\texttt{SRoll2}$.
-Enabled by the completeness property of the PCs, we also turn these constraints into an effective likelihood code useful for assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30, following techniques tested in Ref.~\cite{Heinrich:2016ojb}.
-}
-
-\ch{We publish this reionization effective likelihood code (RELIKE) which is available on GitHub at \url{https://github.com/chenheinrich/RELIKE}.
-There are two packages that come with the code release: 1) we publish a python package \texttt{relike}, which can be used to evaluate the Planck likelihood of any reionization model.
-If desired, this package can be connected to a MCMC sampler to sample the posterior of higher-dimensional parameter space.
-2) We also provide the CosmoMC sampler with a fortran implementation of the `relike` likelihood in the package \texttt{CosmoMC-relike}.
-We expect that this code release would allow much simpler and faster testing of specific models in the Planck data, and to provide consistent and fast way of modeling the ionization history in joint analyses of Planck with other reionization datasets such as luminosity functions, star formation history, 21cm and line-intensity mapping in general, etc.
-}
-
-Extending PCs to cover up to $z_{\rm max} = 50$,we verified that there is no hint of ionization beyond $z>30$ in this Planck data release.
-But for the high-z ionization below $z=30$, we did find a less stringent constraint on $\tau(15, 30)$, the optical depth in the redshift range 15 to 30, that is about \ch{$\sim 3$ times} less stringent than the Planck official results using Flex Knots in Ref.~\cite{Aghanim:2018eyx}.
-Moreover, we demonstrate that the PC results of Ref.~\cite{Aghanim:2018eyx} suffers from a logical flaw in the treatment of its priors following Ref.
-~\cite{Millea:2018bko} and ends up penalizing lower optical depth models.
-[quote difference in tau(15, 30) or total tau possibly attributed to this treatment].
-%We follow the appendix of Ref.~\cite{Heinrich:2018btc} for the Planck 2015 data
+$\texttt{SRoll2}$\footnote{SRoll2: \url{http://sroll20.ias.u-psud.fr.}}, adopts better foreground modeling techniques and allowing for improved reionization constraints \cite{Delouis:2019bub}.
+
+Because the Planck data sets the current standard for CMB constraints, it is important to extract all the information present on reionization  so that it may be readily used in future joint analyses with other cosmological probes.
+In this paper, we obtain new reionization PC constraints using legacy Planck 2018 likelihood and this latest likelihood $\texttt{SRoll2}$ for the low-$\ell$ $E$-mode power spectrum.
+
+Enabled by the completeness property of the PCs, we turn these constraints into an effective means of  assessing the CMB likelihood of \textit{any} reionization model out to $\zmax$ = 30. 
+The 95\% C.L.
+upper limit on the cumulative optical depth between $z=15$ and 30 is $\tau(15, 30) < 0.020$  and about $\sim3$ times less stringent than reported using FlexKnot in the official Planck analysis $\tau(15, 30)<0.007$~\cite{Aghanim:2018eyx}.
+We validate our results with an explicit toy model that allows such a small component of high redshift ionization and show that this result is not simply a consequence of principal component priors alone as suggested by \cite{Millea:2018bko} nor should these priors be changed as was adopted in the official Planck analysis~\cite{Aghanim:2018eyx}.
+Extending the range of PCs to cover up to $z_{\rm max} = 50$, we verify that our results are robust to this extension.
+
+With the technique  validated,
+we release the Reionization Effective LIKElihood code (\relike) which is available on GitHub at \url{https://github.com/chenheinrich/RELIKE}.
+There are two packages that come with this code release: 1)  a python package \relike, to quickly evaluate the effective Planck likelihood of any reionization model given $x_e(z)$.
+If desired, this package can be connected to a MCMC sampler to sample the posteriors in a higher-dimensional parameter space.
+2)  the CosmoMC sampler with a fortran implementation of \relike\  code in the package \cosmomcrelike.
+This code enables a  much simpler and faster testing of specific reionization models with the Planck data, and provides a  consistent way of combining this information with other reionization datasets.
 
 The paper is structured as follows.
-We first describe in \S\ref{sec:background} the background on the reionization principal components and the kernel density estimate (KDE) technique used for building the effective likelihood.
-Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{SRoll2} likelihood, and compare with previous results in literature.
-Then in \S\ref{sec:effective_likelihood}, we present the effective likelihood code, and demonstrate its fidelity with two examples, 1) the standard steplike model, and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard one-step model for illustration purposes.
+We first review in \S\ref{sec:background} the methodology of  reionization principal components along with the kernel density estimate (KDE) technique and the Gaussian approximation \cite{Heinrich:2016ojb} used for building the effective likelihoods.
+Then in \S\ref{sec:results}, we describe the PC results using the Planck 2018 + \texttt{SRoll2} likelihoods, and compare with previous results in literature.
+In \S\ref{sec:effective_likelihood}, we test the effective likelihood code and demonstrate its fidelity with two examples: 1) the standard steplike model; and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard tanh model for illustration purposes.
 Finally, we summarize our results and conclude in \S\ref{sec:conclusion}.
@@
 Here, we  briefly summarize the PC technique itself in \S\ref{sec:PC}.
-We present  techniques
-%kernel density estimate technique
-for building its effective likelihood which can be used to rapidly explore any reionization model in \S\ref{sec:KDE}. We refer the readers to Refs.~\cite{
-%Hu:2003gh, %Mortonson:2008rx, 
-Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.
+We present  techniques for building its effective likelihoods which can be used to rapidly explore any reionization model in \S\ref{sec:KDE}.
+We refer the readers to Refs.~\cite{Mortonson:2007hq,Heinrich:2016ojb} respectively for a more complete description of these topics.
 
@@
 %
-where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the expected 
-%
-% added ``expected" to distinguish this from sigma(ma)
-%
-variances of the PCs.
+where we have discretized the redshift space with $\delta z= 0.25$, and where $\sigma_a^2$ are the expected variances of the PCs.
 The Fisher matrix is computed at a fiducial model which we take to be $x_e^{\rm fid}(z)=0.15$ in the parameterized range.
- Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed with Planck 2015 rather than WMAP best-fit 5 remaining $\Lambda$CDM parameters, which results in minor differences between our $S_a$ and those of  Ref.~\cite{Heinrich:2016ojb}.
+ Note that following Ref.~\cite{Heinrich:2018btc}, we have updated the PCs to be computed 
+ with Planck 2015 rather than WMAP best-fit,
+ which results in minor differences between our $S_a$ and those of  Ref.~\cite{Heinrich:2016ojb}.
 
@@
 In the work that follows, we therefore truncate the sum at $a = 1 ..
-n_{\rm{PC}}$, where $n_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ in the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S \ref{sec:modelindependent} by  taking
- for $z_{\rm max} = 50$ where $n_{\rm{PC}} = 7$ is required at the cosmic variance limit.}
-
- In Fig.~\ref{fig:xe}
- %
- % this seems to be in the main text now
- %\footnote{There are very small differences between the PCs here and those shown in Ref.~\cite{Heinrich:2016ojb}, which can be attributed to a chance in the best-fit cosmology (WMAP to Planck 2015) when constructing the PCs.}
- %
- (top), we show the corresponding fiducial basis functions $S_a(z)$.  
+N_{\rm{PC}}$, where $N_{\rm{PC}} = 5$ for $z_{\rm max} = 30$ in the fiducial analysis.\footnote{We also test the robustness of our fiducial analysis in \S \ref{sec:modelindependent} by taking
+ for $z_{\rm max} = 50$ where $N_{\rm{PC}} = 7$ is required at the cosmic variance limit.}
+
+ In Fig.~\ref{fig:xe} (top), we show the corresponding fiducial basis functions $S_a(z)$.  
 Notice that the series resembles a Fourier decomposition of $x_e(z)$ rank ordered from low to high frequency.
 This is because high frequency variations in $x_e$ leave little observable imprint in $C_l^{EE}$.  
-As a consequence  the $n_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
-given any $x_e^{\rm true}(z)$, we can project it onto the $n_{\rm PC}$ PC basis through
+As a consequence  the $N_{\rm PC}$ PCs are a complete representation of the \textit{observable impact} of $x_e(z)$ on the $C_\ell^{EE}$, rather than the ionization history itself. In other words, 
+given any $x_e^{\rm true}(z)$, we can project it onto the $N_{\rm PC}$ PC basis through
 \begin{equation}
@@
 where the reconstructed $x_e(z)$ through Eq.~(\ref{eq:mmutoxe}) with truncated PCs will not reproduce the true ionization history $x_e(z) \neq x_e^{\rm true}(z)$, but rather it is the observed $C_\ell^{EE}$ that is reproduced to cosmic variance precision.
-We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $n_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
+We reiterate therefore, that the PC analysis is not a tool for reconstructing the ionization history from observations, but rather a forward-modeling tool which, by reducing the dimensionality of the model space to $N_{\rm PC}$, allows us to constrain all possible ionization histories between $z_{\rm min}<z<z_{\rm max}$ in a single analysis.
 For the Planck data set, most of the information on the ionization history comes from contraints on the first two modes which probe the amount of low vs.\ high redshift optical depth.
@@
 
-We follow Ref.~\cite{Heinrich:2018btc} to compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh}. 
-For $z<6$, we follow CAMB and assume fully ionized hydrogen and singly ionized helium and for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a tanh transition width $\Delta z = 0.5$, given the helium fraction
+ We  compute the CMB power spectra with PCs using a modified version of CAMB\footnote{CAMB: \url{http://camb.info}}~\cite{Lewis:1999bs, Howlett:2012mh} (see \cite{Heinrich:2016ojb}).
+ %\todo{CH: probably meant \cite{Heinrich:2016ojb}}.
+ We follow CAMB and assume for $z<6$, fully ionized hydrogen and singly ionized helium whereas for $z\leq 3.5$~\cite{Becker:2010cu}, doubly ionized helium~\cite{Becker:2010cu} with a tanh transition width $\Delta z = 0.5$, given the helium fraction
  \beq
@@
 
-We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using COSMOMC\footnote{COSMOMC: \url{http://cosmologist.info/cosmomc}} of the relevant Planck likelihood discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density
+We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using CosmoMC\footnote{CosmoMC: \url{http://cosmologist.info/cosmomc}} \todo{CH:cite} of the relevant Planck likelihoods discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density
 $\Omega_c h^2$, effective angular sound horizon $\theta_{\rm MC}$, scalar power spectrum log amplitude $\ln (10^{10} A_s)$ and
@@
           \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
-We assume fully ionized Universe at $z < 6$.
-\textit{Bottom}: The cumulative optical depth from $z$ to $z_{\rm max}$ of a unit amplitude PC.
-}
+We assume a fiducial model with fully ionized hydrogen and singly ionized helium by $z<6$, and doubly ionized helium at $z = 3.5$.
+          \textit{Bottom}: The cumulative optical depth from $z$ to $z_{\rm max}$ of a unit amplitude PC. } 
           \label{fig:xe}
@@
 \begin{equation}
- {\cal L}_{\rm Gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{n_{\rm PC}} | \mathbf{C}| }}.
+ {\cal L}_{\rm Gauss}\left({\rm data}|\mathbf{m} \right) = \frac{ e^{-\frac{1}{2} ({\bf m}-\bar{\bf m})^T {\bf C}^{-1} ({\bf m}-\bar{\bf m}) } }{\sqrt{(2\pi)^{N_{\rm PC}} | \mathbf{C}| }}.
  \label{eq:gaussian}
@@
 \begin{equation}
-P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p).
+P({\bf p}| {\rm data}) \propto {\cal L}_{\rm PC}\left[ {\rm data}|\mathbf{m}(\bf p) \right] P(\bf p),
 \end{equation}
 with PC\,=\,KDE or Gaussian.
-This approach obviates the need to implement specific reionization models into CAMB and conduct separate COSMOMC sampling for each, thereby significantly reducing both human and computational effort.
+This approach obviates the need to implement specific reionization models into CAMB and conduct separate CosmoMC sampling for each, thereby significantly reducing both human and computational effort.
 In \S\ref{sec:effective_likelihood} we test the effective KDE and Gaussian likelihoods against an exact implementation
@@
 %
-%We will compare its performance to the KDE likelihood for the Planck 2018 data in the two examples of section~\ref{sec:effective_likelihood} as well.
-
+%
 \subsection{Cumulative Optical Depth}
@@
 \begin{align}
-\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)}.
+\tau(z,z_{\rm max}) & = n_{\rm H}(0) \sigma_T \int_z^{z_{\rm max}} dz \frac{x_e(z) (1+z)^{2} }{H(z)},
 \label{eq:cumtau}
@@
 \begin{align}
-\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^5 m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
+\tau_{\rm PC}(z,z_{\rm max})& = \sum_{a=1}^{N_{\rm PC}} m_a \tau_a(z,z_{\rm max}) + \tau_{\rm fid}(z,z_{\rm max}),
 \label{eq:cumtauPC}
@@
 Eq.~(\ref{eq:cumtau}) using the fiducial model 
-for $H(z)$ and number densities.  This differs from Ref.~\cite{Heinrich:2016ojb} where their cosmological dependence was employed.
+for $H(z)$ and number densities.  This differs from Ref.~\cite{Heinrich:2016ojb} where their cosmological dependence was retained in determining $\tau_{\rm PC}$.
  When we omit the redshift arguments, we
@@
  $
- \tau_{\rm PC}=\tau_{\rm PC}(0,z_{\rm max}).
+ \tau_{\rm PC}\equiv \tau_{\rm PC}(0,z_{\rm max}).
  $
 
- In Fig.~\ref{fig:xe}, we display $\tau_a(z,z_{\rm max})$.
- Notice that positive values of the first component $m_1$ mainly provides the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.
+ In Fig.~\ref{fig:xe} (bottom), we display $\tau_a(z,z_{\rm max})$ for the $\zmax=30$ PCs.
+ Notice that positive values of the first component $m_1$ mainly represents the optical depth that is accumulated at high redshift whereas {\it negative} values of $m_2$ provides the same for low redshift.
 Since these two are the best measured components, the CMB mainly determines the amount of low vs.~high redshift ionization.
@@
 In the usual approach to constraining reionization in $\Lambda$CDM, one places an implicit prior on the shape of the cumulative optical depth by choosing a ``tanh" or near step function reionization (see \S \ref{sec:example1}) and then extracts a single constraint on the total optical depth.
-The PCs on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed.
+The PCs, on the other hand, allow for arbitrary values of $x_e(z)$ when no range-bound prior constraints on the mode amplitudes $m_a$ are imposed.
 The one subtlety when placing constraints on the cumulative optical depth is that the analysis also allows
@@
 \begin{equation}
-\sum_{a=1}^5 m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
+\sum_{a=1}^{N_{\rm PC}} m_a^2 \le (x_e^{\rm max}-x_e^{\rm fid})^2,
 \label{eq:prior_sum}
 \end{equation}
-where $x_e^{\rm fid}=0.15$ and $m_a^{-} \le m_a \le m_a^{+}$ with
+and 
+%$x_e^{\rm fid}=0.15$ and 
+$m_a^{-} \le m_a \le m_a^{+}$ with
 \begin{equation}
@@
 Here we take $x_e^{\rm max} = 1+ f_{\rm He} $ to account for the contribution of singly ionized helium.
-In practice the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}) for
-% I think it can't depend on the data but does depend on the truncation
-%the Planck 2018 data \wh{and
-our fiducial analysis.  Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.  We study its impact and robustness in \S \ref{sec:modelindependent}.
-
-%
-Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we  keep only the first $n_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.
+In the fiducial analysis, the prior on the sum of squares in Eq.~(\ref{eq:prior_sum}) is strictly weaker than the individual priors in Eq.~(\ref{eq:individualprior}).
+Henceforth we refer to Eq.~(\ref{eq:individualprior}) as the physicality prior.  We study its impact and robustness in \S \ref{sec:modelindependent}.
+%
+Note that the original $0\le x_e \le x_e^{\rm max}$ condition cannot be strictly enforced here because we  keep only the first $N_{\rm PC}$ PCs, so these priors in $m_a$ are necessary but not sufficient conditions for a physical ionization history.
 In other words, no physical model is excluded by these priors but some unphysical models are included.
@@
 \subsection{Constraints on Principal Components}
-We use the official Planck likelihoods [add citation] \texttt{plik\_lite} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper.
-We have tested that our results do not change if in lieu of \texttt{plik\_lite} we used the full \texttt{plik\_full} likelihood in which the foreground parameters have not been marginalized over.
-For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{SRoll2} likelihood [add citation] in our official PC results and the effective likelihood code.
-In comparison with the official Planck-released \texttt{SimAll} likelihood, the \texttt{SRoll2} likelihood improves foreground cleaning,
-which enables stronger reionization constraints, especially for the low multipole moments that are important for the low redshift constraints.
-
-% can add a comment about impact 
-%d a factor of XX improvement in $\tau$ just in the steplike model [cite].
+We use the official Planck likelihoods~\cite{Aghanim:2019ame} \texttt{plik\_lite\_TTTEEE} for the high-$\ell$ $TT$, $TE$ and $EE$ as well as \texttt{lowl} for the low-$\ell$ $TT$ throughout this paper.
+We have tested that our results do not change if in lieu of \texttt{plik\_lite\_TTTEEE} we used the full \texttt{plik\_full\_TTTEEE} likelihood in which the foreground parameters have not been marginalized over.
+For the low-$\ell$ $EE$ likelihood, we use the third-party released \texttt{SRoll2} likelihood~\cite{Delouis:2019bub} in our official PC results and the effective likelihood code.
+In comparison with the official Planck-released \texttt{SimAll} likelihood, the \texttt{SRoll2} likelihood has improved foreground cleaning,
+which enables stronger reionization constraints, especially from the low multipole moments that are important for the low redshift constraints.
 
 \begin{figure*}
-%\includegraphics[width=0.7\textwidth]{plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML.png}
 \includegraphics[width=0.8\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
@@
 using \texttt{SRoll2} for the low-$\ell$ $EE$ likelihood, and compare that to the Planck 2015 constraints in red.  The corresponding marginal distributions for individual $m_a$ are also shown.
-%The 1D posterior distributions are shown on the diagonal and the 2D 68\% and 95\% C.L. regions in the $m_a-m_b$ planes in the lower triangle.
 When deriving model independent optical depth constraints we impose an additional physicality prior that excludes the gray region in $m_1$ and regions beyond the box limits in the others.
@@
 %
-The gray shaded region % in the $m_1-m_2$ plane 
+The gray shaded region 
 represents the parameter space that would be excluded by an additional 
 physicality prior from Eq.~(\ref{eq:individualprior}),
-which is just beyond the border of the displayed regime for all but $m_1$.
-Results using the Planck 2018 data are shown in blue.
+which is just beyond the border of the displayed regime for all but $m_1$. Results using the Planck 2018 data are shown in blue.
+
 While all five PC amplitudes are  constrained by Planck, the first two are particularly well-constrained.
@@
 
-%We show a zoomed version of the $m_1-m_2$ plane in Fig.~\ref{fig:plot_m1m2_2015_vs_2018}, where we show beyond the physicality prior with the shaded area.
-[Discuss how the ellipse moved in the direction of lower tanh by going to lower m1.]
-
-%\wh{possibly point to a figure below with $\tau_{12}$ in}
-
 %\wh{possibly move the discussion of trajectories into the model section}
@@
 \includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red), Planck 2018 (blue, \texttt{SRoll2}, and Planck 2018 but with the \texttt{SimAll} likelihood (green).
-With Planck 2018, the \texttt{SRoll2} likelihood moderately improves the best constrained direction and raises the optical depth.}
+\caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red) and Planck 2018 (blue,  \texttt{SRoll2}; green, \texttt{SimAll}).
+The default \texttt{SRoll2} likelihood moderately improves the best constrained direction  and raises the optical depth for Planck 2018.}
 \label{fig:plot_m1m2_2015_vs_2018}
@@
 The visual agreement with the full posteriors is remarkably good, especially in the well-constrained components.
-This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) may suffice for rough constraints on models.
+This suggests that the simple Gaussian likelihood of Eq.~(\ref{eq:gaussian}) suffices for most constraints on models.
 We shall see in \S\ref{sec:effective_likelihood} that a quantitative comparison indeed shows good agreement between the KDE and Gaussian likelihoods, with only minor differences in the shape of their posterior inferences.
-
-% This Gaussian approximation agrees well with the actual PC constraints, unlike in the Planck 2015 data where some differences were still visible on top of good agreement.
-The Gaussian likelihood is even faster than the kernel density estimate.
-As shall see in the next section for two explicit examples, the Gaussian likelihood shows generally good agreement with the full KDE result, but is unable to capture, as expected given its symmetric nature, any skewness in the PC distribution that may affect parameter constraints.
-
-%In Fig.~\ref{fig:plot_m1m2_lowE_vs_srollv2} we also compare the Planck 2018 results with two different low-$\ell$ $EE$ likelihoods: \texttt{SimAll} and \texttt{SRoll2}.
-The \texttt{SRoll2} contours are shrunk nearly directly along the lines of constant optical depth $\tau$ ($\tau_{12} = \tau_1 m_1 + \tau_2 m_2$) [wanna show this]?
-\wh{we might make this point in the context of Marius - that $\tau$ is mainly a linear function of $m_1$, $m_2$, and hence flat priors in $m's$ are locally flat in $\tau$}.
 
@@
 \centering
-\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$,  and correlation matrix $R_{ab}$. $\tau_a$ is the optical depth contribution for $m_a=1$ relative to 
-$\tau_{\rm fid}=0.08626$.}
+\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$,  and correlation matrix $R_{ab}={C_{ab}/[\sigma(m_a)\sigma(m_b)]}$. $\tau_a$ is the optical depth contribution for $m_a=1$ 
+which adds to that of the fiducial model, 
+$\tau_{\rm fid}=0.08626$.
+}
 
@@
 			  &
-%			  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 &
 \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$} 
@@
 $m_1$ &
-%$-$0.116 & 0.030 & 
 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
 $m_2$ & 
-%$-$0.016 & 0.063 & 
 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
 $m_3$ &
-%$$0.078 & 0.098 & 
 $-$0.208 & 0.048 & 1.000 & 0.064 & $-$0.018 \\
 $m_4$ & 
-%$-$0.077 & 0.131 & 
 $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
 $m_5$ &
-%0.066 & 0.145 & 
 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\ \hline
@@
 \hline
-
-\hline
 \end{tabular}
@@
 
-Our code is available on GitHub at {give URL\url{...}}.
-The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTTTE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
+Our code is available on GitHub at {\url{https://github.com/chenheinrich/RELIKE}}.
+The initial release corresponds to the fiducial results in this paper, i.e.\ derived from the Planck 2018 $\texttt{plik\_lite\_TTTEEE} + \texttt{lowl} + \texttt{SRoll2}$ likelihoods at $\zmax = 30$.
 This is also the default analysis for all results that follow except where noted.
-The most important part of the code release is the python package called RELIKE which provides the PC parameters for a model and also
-% added this since it now seemed to be emphasizing Gaussian only  - not sure that was intentional.
-implements the Gaussian effective likelihood.
-The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use RELIKE to return the effective likelihood of the model.
+The most important part of the code release is the python package called \relike\  which provides the PC parameters for a model and also implements the Gaussian effective likelihood.
+The user can specify a customized $x_e(z)$ function taking in reionization model parameters and use \relike\  to return the effective likelihood of the model.
 The returned likelihood is already marginalized over cosmological parameters.
-To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample RELIKE within a MCMC sampler such as Cobaya or Cosmosis \ch{[cite]}.
-
-We also provide a pre-assembled CosmoMC code using the CosmoMC sampler in generic mode, with which we have generated the results presented in this section.
-This CosmoMC code has an internal  implementation of the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.
+To obtain marginalized distributions more easily for high-dimensional parameter space, one may choose to sample \relike\ within a MCMC sampler such as Cobaya\footnote{Cobaya: \url{https://cobaya.readthedocs.io/en/latest/}}~\cite{Torrado:2020dgo,2020arXiv200505290T} or CosmoSIS\footnote{CosmoSIS: \url{https://bitbucket.org/joezuntz/cosmosis/wiki/Home}}~\cite{Zuntz:2014csq}.
+
+We also provide a pre-assembled CosmoMC code using the CosmoMC's generic sampler, with which we have generated the results presented in this section.
+This CosmoMC code has an internal  implementation of both the Gaussian and KDE likelihoods in fortran, which is separate from our main python package that only supports the Gaussian likelihood.
 
@@
 {paper/plots/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Total optical depth $\tau_{\rm tanh}$ posterior for the tanh model in \refsec{example1} for the exact (Planck 2018) likelihood (shaded), KDE PC 
-(blue solid) and KDE Gaussian (black dashed)
-effective likelihood. The overall agreement is excellent and the KDE results even capture the small skewness of the exact result.
+\caption{Total optical depth $\tau_{\rm tanh}$ posterior for the tanh model in \refsec{example1} for the exact (Planck 2018) likelihood (shaded), KDE 
+(blue solid) and  Gaussian (black dashed)
+effective PC likelihoods. The overall agreement is excellent and the KDE results even capture the small skewness of the exact result.
 }
@@
 Overall agreement is good in regions where the contribution to $\tau_{\rm tanh}$ is high but the model form forbids high-$z$ contributions.
-The KDE and Gaussian PC results  oscillate around zero at high-z because of effective low pass filter of the truncation to 5 PCs (enough to represent any observable effects in the data).
+Oscillations around zero at high-z in the KDE and Gaussian PC representations are due to the effective low pass filter from the truncation at 5 PCs which does not produce any observable difference in the CMB.
 }
@@
 The $\tanh$ model is parameterized by its total optical depth for which we take a flat prior 
-with an additional constraint that $z_{\rm re}\ge 6.1$ so as to better satisfy the observationally required full ionization by $z_{\rm min}=6$ given the tanh width.
-%$\Delta z$.
+with an additional constraint that $z_{\rm re}\ge 6.1$ so as to  satisfy the  required full ionization by $z_{\rm min}=6$ given the tanh width to good approximation.
+
 In Fig.~\ref{fig:plot_mjs_2018_vs_2015}, we show the trajectory that the $\tanh$ model takes in the PC space.
 Notice that this trajectory passes near the center of all constraints for Planck 2018 but only through the tails for some of the Planck 2015 constraints.
-Correspondingly, whereas the $\tanh$ model was marginally disfavored with Planck 2015 data that is not the case with Planck 2018 data.
+Correspondingly, whereas the $\tanh$ model was marginally disfavored with Planck 2015 data, that is not the case with Planck 2018 data.
 
 In \reffig{tanh} we show the posterior distribution of $\tau$ for the effective likelihood results, in which the only sampled parameter is $\tau$ (all other cosmological parameters are fixed at the Planck best-fit model), vs the exact Planck likelihood results in which the five other $\Lambda$CDM parameters were also varied using their standard priors.
-The posteriors agree well; the effective likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
+More specifically, we find
+%
+\begin{align}
+    \tau_{\mathrm{tanh}} = 0.0591_{-0.0068}^{+0.0054}, & \quad (\mathrm{Exact})\nonumber \\
+    %
+    \tau_{\rm tanh} = 0.0591_{-0.0066}^{+0.0061}, & \quad  (\mathrm{KDE}) \nonumber \\
+    %
+    \tau_{\rm tanh} = 0.0592_{-0.0062}^{+0.0062}, & \quad  (\mathrm{Gaussian}),
+\end{align}
+%
+where our exact result is also consistent with the tanh model constraint in Ref.~\cite{Pagano:2019tci} for the same combination of likelihoods.
+
+The posteriors of all three likelihoods agree well.
+The KDE likelihood gives a slightly broader posterior because the kernel density estimate in the PC space introduces smoothing.
 We expect the smoothing to be roughly at the $\sim$7\% level in the parameter posterior corresponding to the smoothing factor of $1+f = 1.14$ we applied to the PC covariance during KDE.
-Indeed, the KDE result
-%
-\beq
-\tau_{\rm tanh} = 0.0591_{-0.0066}^{+0.0061}\;\; (\mathrm{KDE}) 
-% before extending chains
-%\tau_{\rm tanh} = 0.0588_{-0.0072}^{+0.0061}\;\; (\mathrm{KDE}) 
-\eeq
-%
-%has a negligibly lower mean ($0.05\sigma$) and 
-has a 4\% larger 68\% confidence region  than the exact likelihood result 
-%ch: made a mistake, mean is 0.0591, not 0.0588 for KDE.
-\beq
-\tau_{\mathrm{tanh}} = 0.0591_{-0.0068}^{+0.0054}\;\; (\mathrm{Exact}).
-\eeq
-}
-%before
-%$\tau = 0.0591 \pm 0.0062$ (exact)
-%$\tau = 0.0588 \pm 0.0066$ (KDE)
-%now (possibly due to longer chains)
-%$\tau = 0.0591 \pm 0.0063$ (exact)
-%$\tau = 0.0588 \pm 0.0064$ (KDE)
+Indeed, the KDE result has a 4\% larger 68\% confidence region than the exact likelihood result.
 
@@
 The Gaussian likelihood is faster to evaluate as it does not require looping over the entire $\sim 10^6$ points in the PC chain.
-{[add timing here]}
-\wh{not sure we need timing results but if you do add them remember you can make Gaussian even faster by precomputing the linear combinations of $m_a$ that diagonalize the covariance (the first one should essentially be $\tau_{12}$ which is interesting pedagogically)} \ch{[CH: got it.]}
-%
-We obtain for the tanh model 
-\beq
-\tau_{\rm tanh} = 0.0592_{-0.0062}^{+0.0062}\;\; (\mathrm{Gaussian})
-\eeq
-Note that the Gaussian likelihood yields symmetric tanh $\tau$ constraints where as the KDE result 
-is able to capture some of the skewness in the exact distribution.
-The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods is consistent with this level of skewness ($\Delta \tau \sim 0.001$), which is $<0.2\sigma$.
-% 95\% UL:
-%tanh Gauss: 1.996574506651048464e-03
-%tanh KDE: 1.996656635642191091e-03
-%tanh exact: 0.000000000000000000e+00
-
-%[Make a point that dz = 0.5 for tanh may not be sufficient for experiments after Planck. This was for higher tau, so transition happens at higher redshift. Now tau is much lower.]
-
-%[Note sure if it is related to this statement in Planck 2018 paper: ``The TANH result gives slightly higher optical  depth  than  the  others,  which  is  primarily driven  by  the  fixed  duration  of  reionization assumed.” Is this true?
-Then transition to the following point]
+The Gaussian likelihood result above gives symmetric tanh $\tau$ constraints where as the KDE result is able to capture some of the skewness in the exact distribution.
+The shift in the peaks seen in Fig.~\ref{fig:tanh} between Gaussian and the exact likelihoods, $\Delta \tau \sim 0.001$, is consistent with the level of asymmetric error bars in the latter.
 
 We also show in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh} the cumulative optical depth obtained from directly integrating the tanh model in the exact likelihood chains vs that from the 5-PC representation of the tanh models in the effective likelihood chains using Eq.~(\ref{eq:cumtauPC}).
-Notice that because of the truncation of the PC expansion at $a=5$, this representation shows an small oscillatory cumulative $\tau$ at $z \gtrsim 10$ whereas the true model has negligible ionization there.
+Notice that because of the truncation of the PC expansion at $N_{\rm PC}=5$,
+this representation shows a small oscillatory cumulative $\tau$ at $z \gtrsim 10$ whereas the true model has negligible ionization there.
 This is because the best constrained modes only represent the smooth features in the ionization history and truncation is similar to a low pass filter in Fourier space.
@@
 the 95\% upper limit on the high-redshift optical depth.
-For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while \ch{$\tau_{\rm tanh}(15,30) \approx 0$}
-%$\tau_{\rm tanh}(15,\zmax) \approx 0$ 
-from the exact likelihood. \ch{NEW: since both one-sided and two-sided are 0.002 here, I'm doing the one-sided vs two-sided explaination in the two-step section instead.}
-Ignoring this small effect of the oscillations, 
-we can see that the tanh model does not allow for high redshift ionization by construction.
+For both effective likelihoods  $\tau_{\rm tanh}(15, 30) < 0.002$  while $\tau_{\rm tanh}(15,30) \approx 0$
+from the exact likelihood. 
+By construction, the tanh model does not allow for a significant fraction of the total optical depth to come from high redshift.
 As we shall see in the next example the data themselves do allow for a small but significantly larger fraction of the total optical depth to originate from high redshift, when the form of the model permits it.
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
-\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}.
-In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$.
-Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model at the 95\% C.L.
-$(\taulo, \tauhi) = (0.043, 0.026)$ (see also Fig.~\ref{fig:two_parameter_model_2D}).
-}
-%best-fit (0.05290344, 0.005859431)
-%95%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
+\caption{Two-step ionization history $x_e(z)$ for the model of \refsec{example2}.
+In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_{\rm hi} = 28$ and $\Delta z_{\rm hi} = 1$.
+Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model on the 95\% C.L.
+contour of Fig.~\ref{fig:two_parameter_model_2D} which maximizes $\tauhi$, $(\taulo, \tauhi) = (0.043, 0.026)$.
+}
 \label{fig:two_step_model}
@@
 \begin{figure}[t]
-%\includegraphics[width=0.47\textwidth]{results/cosmomc_kde/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.png}
-%
-% longer Gaussian chains only 
-%\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test5_run1_vs_relike_tanh_highz_test8_run9_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
-%
-% longer KDE and Gaussian chains
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
@@
 contours).
-The exact, KDE PC and Gaussian PC likelihoods again agree very well with a small contribution from high-$z$ through $\tau_{\rm hi}$ still allowed but not significantly favored.
+The exact, KDE PC and Gaussian PC likelihoods again agree very well and a small contribution from high-$z$ through $\tau_{\rm hi}$ is still allowed but not significantly favored.
 Models of Fig.~\ref{fig:two_step_model} are marked and correspond to the best fit parameters and parameters
-which maximize $\tau_{\rm h}$ on the 2D 95\% C.L. contour --
-suggested change in model - or find another reason for plotting the currently chosen one.
+which maximize $\tau_{\rm hi}$ on the 2D 95\% C.L. contour.
 }
@@
 {paper/plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
-\caption{Cumulative optical depth in the two step model (as in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
+\caption{Cumulative optical depth in the two-step model (as in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
 Agreement between the exact, KDE PC and Gaussian PC likelihoods is again excellent and show that a small amount of high-$z$ ionization is allowed.
@@
 x_e&(z)&\,= \frac{1+f_{\rm He} - \xemin}{2}\left\{  1+ \tanh\left[ \frac{y(z_{\rm re})-y(z)}{\Delta y} \right] \right\} \notag \\
-&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm hi}-z}{\Delta z_{2}} \right] \right\} + x_e^{\rm rec},
+&+& \frac{\xemin - x_e^{\rm rec}}{2}\left\{  1+ \tanh\left[ \frac{z_{\rm hi}-z}{\Delta z_{\rm hi}} \right] \right\} + x_e^{\rm rec},
  \label{eqn:tanh_highz}
@@
 
-In practice, we parameterize the model by $\taulo(\zre, \xemin)$ and $\tauhi(\zre, \xemin)$.
-They are defined as in \refeq{cumtau} but with the boundaries $[0, z_{\rm split}]$ and $[\tau(z_{\rm split}, \infty]$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_{\rm re}$ to be conservative on the preference of $\tauhi$ in the data.
+In practice, we parameterize the model by $\taulo$ and $\tauhi$.
+They are defined as in \refeq{cumtau} but with the boundaries $(0, z_{\rm split})$ and $(z_{\rm split}, \infty)$ for $\taulo$ and $\tauhi$ respectively, where the split is chosen as $z_{\rm split} = \zre + \Delta z_{\rm re}$ to be conservative on reducing the preference of $\tauhi$ in the data.
 Given a model $(\taulo, \tauhi)$, we find the corresponding ($\zre, \xemin$) through an iterative search similar to that used in the canonical tanh model for finding $\zre$ given a total optical depth $\tau$.
-For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi, \tau_{\rm tot} \in [0, \tau_{\rm max}]$ where  $\tau_{\rm tot} = \taulo + \tauhi$ is the total optical depth, and $\tau_{\rm max} = 0.35$.
-%except that $\taulo \in [0.03, \tau_{\rm max}]$ for the effective likelihood run. [not sure if this last sentence is needed; since we put a zre prior anyways.]
-We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model.
-
-In ~\reffig{two_parameter_model_2D}, we compare the 1D and 2D marginalized posteriors of $\taulo$ and $\tauhi$. The exact, KDE and Gaussian PC treatments again agree very well. 
-Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors distort near this value.
-The models in Fig.~\ref{fig:two_step_model} are represented by the star and cross symbols for the best fit case and a model allowed at the 95\% CL.
-
-{add timing results somewhere}
-
-The marginalized 1D constraints from the KDE and Gaussian likelihoods 
-\beq
-\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., KDE)}
-\eeq 
-and
-\beq
-\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., Gaussian)}
-\eeq 
-are consistent with those from the exact likelihood 
-\beq
-\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021 \quad (95\% \text{ C.L., Exact).
-\eeq 
-}
-%exact: tauhi < 0.0206
-%KDE: tauhi < 0.0214
-%Gaussian: tauhi < 0.0207
-
-% running longer KDE chains (numbers might be updated)
-
-\ch{NEW: $\tau_{\rm hi}$ upper limits are all one-sided already.}\\
-
-\ch{Note that the 95\% upper limit on $\tau_{\rm hi} < 0.021$ presented here is a 1D marginalized result; whereas the model shown in Figs~\ref{fig:two_parameter_model_2D} and~\ref{fig:two_parameter_model_2D} which has $\tauhi = 0.026$ is a model picked at the 95\% C.L.
-in the 2D marginalized case, hence the difference from the 1D result here.}
-Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints are correspondingly slightly larger than $\tau_{\rm tanh}$:
-\ch{
-\beq
-\tau_{\rm 2step} = 0.0621^{+0.0062}_{-0.0061}.
-\eeq
-}
+For both the full likelihood and effective likelihood runs, we adopt the flat priors on $\taulo, \tauhi \in [0, 0.35]$.
+We further apply a prior cut in $\zre$ space to keep models with $\zre > 6.1$ for both the full and effective likelihood chains as in the tanh model to ensure full reionization by $z = 6$ to good approximation.
+
+In ~\reffig{two_parameter_model_2D}, we display the posteriors of $\taulo$ and $\tauhi$. The exact, KDE and Gaussian PC treatments again agree very well. 
+Note that in the canonical tanh model $\taulo \lesssim 0.04$ roughly corresponds to $\zre < 6$ so that the posteriors are effectively cut off  near this value.
+The models in Fig.~\ref{fig:two_step_model} are represented by the star and cross symbols for the best fit case and a model allowed at the 95\% C.L. respectively.
+
+The marginalized 1D constraints from the exact, KDE and Gaussian likelihoods also agree well
+\begin{align}
+\taulo = 0.0526^{+0.0058}_{-0.0078}, \; \tauhi < 0.021\; (95\% \text{ C.L.}), & \quad({\rm Exact}) \nonumber\\
+\taulo = 0.0525^{+0.0062}_{-0.0079}, \; \tauhi < 0.021\;  (95\% \text{ C.L.}), & \quad({\rm KDE}) \nonumber\\
+\taulo = 0.0528^{+0.0064}_{- 0.0077}, \; \tauhi < 0.021\; (95\% \text{ C.L.}), & \quad({\rm Gaussian}).
+\end{align}
+Note that the upper limit on $\tau_{\rm hi}$ presented here is the one-sided 95\% C.L.
+limit in 1D; whereas the model shown in Figs.~\ref{fig:two_step_model} and~\ref{fig:two_parameter_model_2D} is located on the 95\% C.L.
+contour in 2D and has therefore a larger $\tau_{\rm hi} = 0.026$.
+
+Moreover, we can derive the constraints on the total optical depth, which are also consistent between the three likelihoods as well:
+%
+\begin{align}
+\tau_{\rm 2step} = 0.0621^{+0.0062}_{-0.0061}, & \quad (\text{Exact)} \nonumber\\
+\tau_{\rm 2step} = 0.0619^{+0.0063}_{-0.0065},  & \quad (\text{KDE)} \nonumber\\
+\tau_{\rm 2step} = 0.0618^{+0.0062}_{-0.0062}, & \quad (\text{Gaussian).}
+\end{align}
+%
+Given the small amount of $\tau_{\rm hi}$ still allowed, the total optical depth constraints in the two-step model are correspondingly slightly larger than $\tau_{\rm tanh}$.
+This is consistent with other examples in the literature where extended reionization to high redshifts were allowed (e.g. Refs.~\cite{Ahn:2020btj, Paoletti:2020ndu}). More generally,
+models which are sufficiently flexible at allowing for high-redshift ionization will have a comparable difference in their total optical depth.
+
 Finally the 95\% C.L. upper limit on high-redshift optical depth are also consistent:
-\begin{equation}
-    \tau_{2\text{step}}(15, 30) < 0.016 \quad (95\% \text{ C.L., Exact)}
+%
+\begin{align}
+\tau_{2\text{step}}(15, 30) < 0.014\; (95\% \text{ C.L.}), & \quad (\text{Exact}) \nonumber\\
+\tau_{2\text{step}}(15, 30) < 0.016\; (95\% \text{ C.L.}), & \quad (\text{KDE}) \nonumber\\
+\tau_{2\text{step}}(15, 30) < 0.016\; (95\% \text{ C.L.}), & \quad (\text{Gaussian}).
 \label{eq:2stepbound}
-\end{equation}
-\todo{CH: still missing one-sided $\tau_{2\text{step}}(15, 30)$ for exact.}
-and 
-\ch{$\tau_{2\text{step}}(15, 30) < 0.016$} for both effective likelihoods. %changed from <0.017--> <0.016 when going from two-sided to one-sided.
+\end{align}
+%
+The small difference in PC results vs.~exact reflects the mild oscillatory PC artifact around $z\sim 15$ in Fig.~\ref{fig:plot_taugtz_two_step_contours}, where we show the cumulative optical depth constraints on this two-step model in the exact vs the PC likelihood construction.
+Note that the 95\% upper limit on $\tau(15, 30)$ reported in the text are one-sided upper limits, whereas the contours in the cumulative optical depth figures enclose their respective C.L.
 Note also that the 95\% C.L. allowed two-step model in 
 Fig.~\ref{fig:two_step_model} has
-\ch{$\tau_{2\text{step}}(15,30)= 0.017$.}
-\ch{Note that the 95\% upper limit on $\tau(15, 30)$ reported in the text are one-sided upper limits, which are slightly different than the two-sided upper limits plotted in the figures of cumulative optical depth (e.g.
-Fig.~\ref{fig:plot_taugtz_two_step_contours})}
-
-The Planck 2018 paper~\cite{Aghanim:2018eyx} also constrained $\tau(15, 30)$ in a model-independent way using the FlexKnot method, in which $x_e(z_i)$ at the $i$-th knots are varied as free parameters as well as the number of total knots.
-They obtained $\tau(15, 30) < 0.007$ (95\% C.L.), which would strongly rule out models that are still allowed at 95\% C.L.
+$\tau_{2\text{step}}(15,30)= 0.017$.
+
+The Planck 2018 cosmological parameter paper~\cite{Aghanim:2018eyx} also constrained $\tau(15, 30)$ in a model-independent way using the FlexKnot method, in which $x_e(z_i)$ at the $i$-th knots are varied as free parameters as well as the number of total knots.
+They obtained $\tau(15, 30) < 0.007$ (95\% C.L.), which would strongly rule out models that are still allowed at the 95\% C.L.
 in this class.
@@
 
-In Fig.~\ref{fig:plot_taugtz_two_step_contours} we show the cumulative optical depth constraints on this two-step model.
-we show the 68\% and 95\% limits on $\tau(z, \zmax)$ obtained by integrating over the exact $x_e(z)$ two-step model in the exact likelihood chains, as well as those obtained from its 5-PC representation using the effective KDE and Gaussian likelihoods.
-As in the tanh model, the agreement is excellent apart from the small redshift oscillations due to the low pass filter nature of the PC representation.
-Notice again the amount of high redshift optical depth that is allowed in the two step model vs the tanh model (cf.~Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
-With this in mind, we will now turn to the model-independent constraints derived from the PC analysis to generalize these results on high redshift optical depth beyond the two-step model.
-
 \begin{figure}[ht]
@@
 {paper/plots/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{Total optical depth posterior for the model-independent PC analysis compared with the exact tanh model (black solid line) and the two-step model (blue dashed line) from Figs.~\ref{fig:tanh} and~\ref{fig:two_parameter_model_2D}.
-The PC analysis captures the ability to raise the optical depth through the high-$z$ contributions of models like the two-step case whose form allows it.
-}
-%\label{fig:plot_tau_posterior_PC_vs_tanh_vs_tanh_highz}
+\caption{Total optical depth posterior for the model-independent PC analysis (shaded) compared with the exact tanh model (blue solid line) and the two-step model (black dashed line) from Figs.~\ref{fig:tanh} and~\ref{fig:two_parameter_model_2D}.
+The PC analysis captures the ability to raise the optical depth through the high-$z$ contributions in models whose form allow it like in the two-step case.
+}
 \label{fig:tauPC}
@@
 While the PC approach to constraining reionization should mainly be used for 
-testing wide classes of models efficiently and with their own physically motivated priors, it can also be employed to extract model-independent constraints on the total and cumulative optical depth. 
-In this section we present these model-independent results, first on the total optical depth and then on the cumulative optical depth at high redshift, and discuss their robustness to priors  on physicality of the ionization history and redshift range of the fiducial analysis.
+testing wide classes of models efficiently and with their own physically motivated priors, it can also be employed to extract model-independent constraints on the optical depth. 
+In this section we present these model-independent results, 
+first on the total optical depth and then on the cumulative optical depth at high redshift, and discuss their robustness to priors  on physicality of the ionization history and redshift range of the fiducial analysis.
 
@@
 and the full posterior is shown in Fig.~\ref{fig:tauPC}. 
-Notice that the distribution matches that of the two step model from \S \ref{sec:example2} but almost $0.5\sigma$ shifted higher than that of the tanh model from \S \ref{sec:example1}.
+Notice that the distribution matches that of the two-step model from \S \ref{sec:example2} but almost $0.5\sigma$ shifted higher than that of the tanh model from \S \ref{sec:example1}.
 
@@
 Therefore, the model has a strong implicit prior that a given a total optical depth $\tau$ comes from $z\lesssim z_{\rm re}$ regardless of whether the data allow or prefer
-ionization at higher $z$.  While this prior may be well motivated
-theoretically \cite{XX} (sims that give prompt reionization), it should be distinguished from constraints that are data driven.  
-This was especially apparent with the Planck 2015 data where
-the PC constraints gave \ch{$\tau_{\rm PC}=0.092\pm0.015$ which differed substantially from the tanh $\tau_{\rm PC}=0.079\pm0.017$~\cite{Heinrich:2016ojb}}.  As we shall discuss
-below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.
+ionization at higher $z$. While this prior may be well motivated theoretically, it should be distinguished from constraints that are data driven.  
+This was especially apparent with the Planck 2015 data where the PC constraints gave $\tau_{\rm PC}=0.092\pm0.015$ which differed substantially from the tanh $\tau_{\rm PC}=0.079\pm0.017$~\cite{Heinrich:2016ojb}.
+As we shall discuss below, these differences were even more striking in the implications for the cumulative $\tau$ at high redshift as we have also concretely illustrated in the two-step model of \S \ref{sec:example2}.
 
@@
 They do of course come with their own priors namely the flat
-range-bound ones from Eq.~\eqref{eq:individualprior}.
+range-bound ones from Eq.~\ref{eq:individualprior}.
 These have in fact caused some confusion in the literature when interpreted in terms of the total optical depth \cite{Millea:2018bko}.
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
-\caption{Total optical depth $\tau_{\rm PC}$ posterior (shaded) with (top) and without (bottom) the additional physicality prior of Eq.~(\ref{eq:individualprior}).
-The shift corresponds to the elimination of unphysically negative ionization at high $z$ from the $m_1$ PC as can be traced through the approximate $\tau_{12} \approx \tau_{\rm PC}$ built from a linear combination of the first two components defined in Eq.~(\ref{eq:tau12}) (blue lines).
-}
+\caption{Total optical depth $\tau_{\rm PC}$ posterior (shaded) without (top) and with (bottom) the additional physicality prior of Eq.~(\ref{eq:individualprior}).  
+Imposing the physicality prior eliminates negative ionization at high $z$ and shifts the distribution to a higher total $\tau$.
+This unphysical contribution comes from the $m_1$ PC, as can be traced through the approximate $\tau_{12} \approx \tau_{\rm PC}$ built from a linear combination of the first two components defined in Eq.~(\ref{eq:tau12}) (blue lines).}
 \label{fig:tau12}
@@
 
- We first note that by virtue of Eq.~(\ref{eq:cumtauPC}), the total optical depth receives contribution from all PCs.  However
- %
-%\beq
-%\tau_{\rm PC} = \tau_{\rm fid} + \sum_a m_a \tau_a,
-%\eeq
-%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
-The PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
+ We first note that by virtue of Eq.~(\ref{eq:cumtauPC}), the total optical depth receives contribution from all PCs.
+However the PCs are rank ordered by their expected constraining power and so most of the information on the total optical depth comes from
 the first two components
@@
 
-%
-% Therefore, while retaining all physical models, they also includes some unphysical models.
-The authors of Ref.~\cite{Millea:2018bko} (hereafter MB18) raised a concern that these PC priors implies a non-flat prior in $\tau$ and claimed that it biases the PC result of  Ref.~\cite{Heinrich:2016ojb} and accounts for most of the shift between the tanh and PC $\tau$ posteriors.
-In Ref.~\cite{Heinrich:2018btc}, we showed that the use of our physicality prior does not cause bias for data that is constraining, which includes the Planck 2015 dataset.
-On the other hand, we showed in Ref.~\cite{Heinrich:2016ojb} that if one adopts the method proposed in Ref.~\cite{Millea:2018bko}.
-in an attempt to flatten the $\tau$ prior (by multiplying its inverse at each point in the multidimensional PC space), one would actually end up with a biased $\tau$ constraint for the Planck 2015 data.
-%
-%We briefly recap the argument below.
-Let us use $\tau \approx \tau_{12}$ in order to illustrate the impacts of the priors more easily.
-We first note that by virtue of Eq.~\ref{eq:mmutoxe}, the total optical depth receives contribution from all PCs
-%\beq
-%\tau = \tau_{\rm fid} + \sum_a m_a \tau_a,
-%\eeq
-%where $\tau_a$ is the contribution from a unit amplitude PC and $\tau_{\rm fid}$, that from the fiducial model.
-Because most of the contributions to $\tau$ comes from the first two components (see also Fig.
-7 of Ref.~\cite{Heinrich:2018btc} in the Appendix), we can approximate it as
-%\beq
-%\tau_{12} = \tau_{\rm fid} + m_1 \tau_1 + m_2 \tau_2 \approx \tau.
-%\eeq
 Due to this linearity, flat priors in $m_1$ and $m_2$ correspond to
 flat priors in $\tau_{12}$ for any linear trajectory through the space.
-For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015} and is nearly linear through the Planck 2018 allowed region.
+For example, the tanh model trajectory is shown in Fig.~{\ref{fig:plot_mjs_2018_vs_2015} and is nearly linear through the Planck 2018 allowed region in the $m_1-m_2$ plane.
 Even without the explicit
@@
 $m_2$ axis.
-Therefore, the range-bound flat prior in $m_1,m_2$ (blue square) considered {\it alone} corresponds to a prior that is not flat in $\tau_{12}$ due to the extent  of the allowed parameter space
-for each of its values, e.g. at  left corners (Fig.~\ref{fig:prior_box} solid square or lowest $m_1$) of the range $\tau_{12}=0.023$
-than at $\tau_{12} = 0.130$ and the implied prior increases linearly from zero between the two.   Ref.~\cite{Millea:2018bko} suggest 
-inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$ and this reweighting was implemented in the Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}.
-First notice that this inversion would in fact then produce a non-flat prior on $\tau$ for the tanh model
-that disfavors high values.   This is because
-the additional parameter space allowed by the range-bound $m_a$ prior is excluded by the functional form of the model even though the reweighting occurs globally and includes the tanh trajectory as well.
+Therefore, the range-bound flat prior in $m_1$, $m_2$ (blue square)  is not flat in its {\it marginal} $\tau_{12}$  distribution due to the extent  of the allowed parameter space along lines of constant $\tau_{12}$: At
+the implied lowest $\tau_{12}=0.023$ of the range
+in Fig.~\ref{fig:prior_box},
+this extent vanishes and then increases linearly until $\tau_{12} = 0.130$.
+ Ref.~\cite{Millea:2018bko} suggests 
+inverting this prior point-by-point in the $m_a$ parameter space by dividing by the implied prior $P(\tau_{\rm PC}) [ \approx P(\tau_{12})]$; this reweighting was implemented in the Planck 2018 paper on cosmological parameters~\cite{Aghanim:2018eyx}.
+First, notice that this inversion would 
+produce a non-flat prior on $\tau$ for the tanh model
+that disfavors high values of $\tau$.   This is because
+the additional parameter space allowed by the range-bound $m_a$ prior is excluded by the functional form of the model, even though the reweighting occurs globally and includes the tanh trajectory as well.
 
@@
 This rotated prior in $m_1-m_2$ space corresponds to a constant $P(\tau_{12}$) once the direction orthogonal to $\tau_{12}$ is marginalized.
-Had the  \wh{data constraint been shifted to slightly higher
-$m_1$ so that both prior ranges encompassed all of the 95\% CL allowed
-region they both would yield  the same constraints at that level.}
-% rephrased to flow with the next paragraph of 2015->2018
-This is in spite of the fact that reweighting by $1/P(\tau_{12})$ would  change the original prior but not the
-rotated one and \wh{destroy this desired equivalence.
+Had the data constraint been shifted to slightly higher
+$m_1$ so that both prior ranges encompassed all of the 95\% C.L. allowed
+region they both would yield  the same constraints at that level.
+However the reweighting by $1/P(\tau_{12})$ would change the original prior but not the rotated one and destroy this desired equivalence. 
 More generally the point is that for a multidimensional space, constructing a flat marginal 1D distribution per parameter in the absence of data does not guarantee a locally flat prior in the region supported by the data
-(see \cite{Heinrich:2018btc} for further discussion).}
-%will try rephrasing.}
+(see \cite{Heinrich:2018btc} for further discussion).
 
@@
 removes nearly half of the space that would otherwise be allowed by the
-data at 95\% CL.  This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).  
+data at 95\% C.L.  This is in contrast to the Planck 2015 data where only a very small region is removed (see Fig.~\ref{fig:plot_m1m2_2015_vs_2018}).  
 This is reflected in the upward shift of the $\tau_{12}$ posterior in Fig.~\ref{fig:tau12} with and without the physicality prior on $m_1$.
@@
 are conservative: all the cases that are excluded are definitely unphysical but some of the cases that are included
-may also be unphysical since their physicality depends on higher order PC components.
-The impact on optical depth  however is small and at most contributes a small bias that conservatively {\it lowers} the high redshift contribution which is already a small contribution to the total.
+may also be unphysical since their physicality depends on higher-order PC components.
+The impact on optical depth  however is small and at most, contributes a small bias that conservatively {\it lowers} the high redshift contribution which is already a small contribution to the total.
 This can be seen  in the comparison with the two-step model 
-in Fig.~\ref{fig:tauPC} and as we shall return to this point in the cumulative optical depth where the high and low redshift contributions are manifest.
+in Fig.~\ref{fig:tauPC} and as we shall return to this point in the cumulative optical depth constraints in the next section where the high and low redshift contributions are manifest.
 In any case, using PCs to 
@@
 \beq
-%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
 \tau_{\rm PC} = 0.0626^{+0.0061}_{-0.0072}\quad (\zmax=50).
@@
 As with the tanh model, the \texttt{SRoll2} likelihood provides tighter constraints that shift the distribution toward higher optical depth.
-
-%models that 
-%In Fig.~\ref{fig:prior_box}, we show the Planck 2018 constraints (ellipse) in the $m_1-m_2$ plane along with constant lines of $\tau_{12}$ (light gray).
-The solid blue box corresponds to the physicality prior, whereas the dashed blue box is constructed to be aligned with the directions of the constant $\tau_{12}$ lines, so that when projected onto the $\tau_{12}$ directions, gives a true flat $\tau_{12}$ prior.
-The physicality prior upon projection would correspond to a linearly rising prior in $\tau_{12}$ over the region where the posterior has support.
-One would be eager to conclude that this produces a biased constraint on $\tau_{12}$.
-However, because the Planck data is more constraining than the prior in $m_1$ and $m_2$, (the ellipse occupying a much smaller area in one part of the prior box), the value of the prior at high $\tau_{12}$ becomes irrelevant for the posterior.
-However, if one tries to mitigate for the $\tau_{12}$ prior by multiplying the inverse prior value at each point in the PC space over the area where data is more constraining than the prior, then one would bias the posterior toward lower $\tau_{12}$ in this case.
-
-%What is then the true impact of this physicality prior we have imposed.
-We have established that the orientation of the prior box in the $m_1-m_2$ plane does not matter for this dataset since the data is more constraining than prior.
-However, the location of the data constraint does matter, as the prior box currently cuts off half of the ellipse in the Planck 2018 data (a larger amount than in Planck 2015).
-The consequence of this cut in the $\tau$ constraints essentially amounts to excluding models not fully ionized models by $\zre \sim 6$, as consistent with observational constraints.
-
-%The other impact of using the physicality prior is that some unphysical models are included inside the prior.
-We reiterate that we do not adopt this prior for the KDE and unphysical models do not impact results when actually testing physical models using the effective likelihood code.
-We will show in the next section that their impact on the model-independent $\tau$ results are also minimal.
-
-%[add a sentence to reiterate that the shift between PC and steplike model seen in the Planck 2015 is not due to the use of physicality prior? and some other Marius/Planck paper points?]
-
-%[restore some text about evolution of planck data and HFI systematics causing PC-tanh shifts in Planck 2015 data?
-And address the claim: ``In the Planck 2018 paper on cosmological parameters, the reduction of optical depth from the PC results seen in the Planck 2018 data compared to the Planck 2015 data was attributed to the reduction of HFI systematics, with subdominant contribution from prior."]
-%They also proposed a method to flatten the $\tau$ prior by multiplying its inverse at each point in the multidimensional PC space.
-The authors claim that adopting the physicality priors in Ref.~\cite{Heinrich:2016ojb} contributed to a large fraction of the shift in the $\tau$ posterior between the tanh and the PC results in the Planck 2015 data.
 
@@
 
-The value of a model independent approach is even more apparent for constraints on the
+The value of a model-independent approach is even more apparent for constraints on the
 cumulative optical depth $\tau(z, \zmax)$ at high $z$.
-The main results are again related to the $m_1$ and $m_2$ PC constraints. 
-As discussed in Ref.~\cite{Heinrich:2016ojb}, raising $m_1$ increases the  optical depth  at all $z$ but with a larger weighting at high $z$ whereas
-raising $m_2$ shifts the optical depth from low $z$ to high $z$.  
-Consequently, as shown in Fig.~\ref{fig:prior_box}, beyond the well constrained total optical depth, as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12}), orthogonal to this combination is a somewhat weaker constraint on the the high to low $z$
-contribution to the total optical depth.  Any finer redshift detail in the ionization history is much more poorly constrained.
+The main results are again related to the $m_1$ and $m_2$ PC constraints, both of which are well constrained.  
+Consequently, as shown in Fig.~\ref{fig:prior_box}, beyond the well-constrained total optical depth (as represented by the $\tau_{12}$ linear combination in Eq.~(\ref{eq:tau12})) 
+is a somewhat weaker constraint in the constant $\tau_{12}$ direction that reflects whether that $\tau_{12}$ comes mainly from high or low redshift (see \S \ref{sec:cumulative}).
+Any finer redshift detail in the ionization history is much more poorly constrained.
 
 In Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}, we show the 
-the \ch{two-sided} 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.
-Notice that without the prior, a larger range of unphysically negative optical depth is allowed whereas with the prior it is mostly eliminated except for the highest redshift.
-Moreover upper bounds on the optical depth at high
+the 68\% and 95\% confidence level contours for  $\tau_{\rm PC}(z,z_{\rm max})$ with and without the physicality prior on $m_1$.
+Notice that without the prior, a larger range of unphysically negative optical depth is allowed, whereas with the prior it is mostly eliminated except for the highest redshifts.
+Moreover, upper bounds on the optical depth at high
 redshift are largely unchanged.   These upper limits allow some contribution at high-$z$ but do not require it.   For example,
-\ch{
+%
 \beq
@@
 \eeq
-}
-\ch{Recall again that we are reporting the one-sided upper limit here in the text, which is slightly lower than the two-sided limits shown in the figures.}
-%
-\ch{(NEW: changed from 0.023 to 0.020 going from two-sided to one-sided.)}
+%
+Recall again that we are reporting the one-sided upper limit here in the text, which is slightly lower than values at the contours shown in the figures.
 This should be contrasted with the tanh model 
- where it is  forbidden by form of the ionization history in conjunction with the total optical depth constraint.  
- On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~\ref{eq:2stepbound}).
-It is slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$).
- \wh{I might add a few words about the relationship to $\tau_{\rm hi}$ but first let's make sure that all upper limits are quoted as one sided 95\% since I'm not sure we're comparing apples-to-apples yet..} \ch{$\tau_{\rm hi}$ values are confirmed to be one-sided now!
-:) }
-
-These results are in contrast to those of the Planck 2015 analysis. 
-In Fig.~\ref{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2} top panel, we compare
+where  high-$z$ ionization is forbidden by the functional form of the ionization history.
+On the other hand the bound is comparable to that obtained for the two-step model (see Eq.~(\ref{eq:2stepbound})).
+It is actually slightly larger given that the form of the two-step model still requires a steadily decreasing cumulative optical depth at $z>z_{\rm re}$ with the additional allowed contributions reflected in $\tau_{\rm hi}$.
+
+These results are in contrast to those of the Planck 2015 analysis. In Fig.~\ref{fig:plot_taugtz_2015_vs_2018_simallEE_vs_2018_srollv2} top panel, we compare
 the two. Whereas the Planck 2018 has only upper limits on the high redshift optical depth, Planck 2015 actually preferred
@@
 In the lower panel, we compare the Planck 2018 results for the two different low-$\ell$ $EE$ likelihoods.
-Improvements from \texttt{SRoll2} come mainly from improving the lower bound at low redshift, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.
+Improvements from \texttt{SRoll2} come mainly from improving the lower bound at low redshifts, consistent with the improved constraints at the lowest multipoles of $C_l^{EE}$.
 
@@
 contours of $\tau(z, \zmax)$ are consistent between the $\zmax = 30$ (red regions) and $\zmax = 50$ results (black lines).
-
-%Furthermore, their total optical depth constraints are practically identical: %
-%\beq
-%\tau_{\rm PC,\, \zmax=30} = 0.0619^{+0.0056}_{-0.0068}\;\; (\texttt{SRoll2}),
-%\eeq
-%and
-%\beq
-%\tau_{\rm PC,\, \zmax=50} = 0.0626^{+0.0061}_{-0.0072}\;\; (\texttt{SRoll2}).
-%\eeq
-%
-%Compared the \texttt{SimAll} likelihood results, 
-%\beq
-%\tau_{\rm PC,\, \zmax=30} = 0.0582 ^{+0.0072}_{-0.0083}\;\; (\texttt{SimAll}),
-%\eeq
-%and
-%\beq
-%\tau_{\rm PC,\, \zmax=50} = 0.0582^{+0.0077}_{-0.0088}\;\; (\texttt{SimAll}).
-%\eeq
-%the \texttt{SRoll2} error bars on PC $\tau$ was reduced by about 20\%, and the central values rose by about half a sigma. [wanna comment more here?]
-%
-
 For the cumulative high-redshift optical depth,
@@
 we also obtain consistent upper limits to the $\zmax=30$ case of Eq.~(\ref{eq:tauPC15}).
-\ch{NEW: changed 0.022 to 0.019 going from two-sided to one-sided}
 
@@
 }
-% no prior: 95\% C.L.  = [-0.01935756790924691, 0.021029019067727237]
-% with prior: {'z': 15, 'mean': 0.00694422689213187, 'std': 0.007568921831067864, 'limit68': [-0.0007215571045808687, 0.014684826477572082], 'limit95': [-0.006316156744039184, 0.02284538046347211]}
-%
-%}
 \label{fig:plot_taugtz_2018_with_vs_without_physicality_prior}
@@
 \includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2_do_rescale_tau_false.pdf}
-%previouslky
-%{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2.pdf}
 \caption{Changes in the
@@
 \includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
-\caption{Robustness of the 
-%\ch{Planck 2018} \wh{decide on convention - in many places the default analysis isn't labeled, only changes from the default}
-cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.
+\caption{Robustness of the cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.
 Upper limits at $z \lesssim 30$ are nearly unaffected whereas extending $\zmax$ allows the high-$z$ component to come from even higher-$z$.
@@
 
-In conclusion, we used constraints on reionization principal components with respect to the large-scale $EE$ data to produce a fast and accurate effective likelihood for the final release of Planck 2018 data.
-In lieu of the official $\texttt{SimAll}$ likelihood for low-$\ell$ $EE$ we used the independent later release of the $\texttt{SRoll2}$ likelihood in 2019 which has better foreground cleaning allowing for slightly improved parameter constraints.
-Our effective likelihood code package is called Relike and is available publicly on Github.
-This code is beneficial for the community for the purpose of testing \textit{any} global ionization history between $6 < z < 30$ with the Planck data without having to modify CAMB which can be very cumbersome to do.
-The convergence time to reach the Gilman-Rubin criteria $R_{-1} < 0.01$ is about 300 times less with the Relike code than with a traditional MCMC sampling of the Planck likelihood for the one-parameter canonical tanh model.
-
-To test this effective Planck likelihood code, we used two examples: 1) a canonical tanh model with a single transition redshift, and 2) a two-parameter model consisting of two tanh transitions allowing in addition for a high-redshift ionization plateau compared to the first example.
-We demonstrated accuracy of the Relike code for both examples by comparing the constraints on the reionization model parameters with those obtained from sampling the exact Planck likelihoods and using a modified CAMB code.
-We also showed great agreement between the confidence levels of the cumulative optical depth evolution between the 5-PC representation and the exact ionization model.
-
-Besides conducting explicit model testing with the Relike code, we also obtained model-independent constraints on the reionization history using the PC chains themselves.
-We summarize these results below.
+In conclusion, we have used constraints on reionization principal components of the large scale $EE$ polarization to produce a fast and accurate effective likelihood for the final release of Planck 2018 data.
+We replaced the official $\texttt{SimAll}$ likelihood for low-$\ell$ $EE$ with the $\texttt{SRoll2}$ likelihood independently released in 2019 which uses improved foreground cleaning methods.
+Our effective likelihood code package is called \relike\  and is available publicly on Github.
+We expect the code to facilitate the testing of \textit{any} global ionization history between $6 < z < 30$ with the Planck data and to enable fast and consistent joint analyses with other reionization datasets.
+
+To test this effective likelihood code, we used two examples: 1) a canonical tanh model with a single transition redshift; and 2) a two-parameter toy model consisting of two tanh transitions allowing for an additional high-redshift ionization plateau.
+We demonstrated accuracy of RELIKE for both examples by comparing model parameter constraints with those obtained from sampling the exact Planck likelihoods using a modified CAMB code.
+We also showed excellent agreement between the confidence levels of the cumulative optical depth evolution between those of the exact ionization models and their 5-PC representations.
+
+Besides conducting explicit model testing with  \relike, we also extracted model-independent constraints using the PC chains themselves. We summarize these results below.
 
@@
 
-    \item {We obtained the constraint on the total optical depth $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ (68\% C.L.) 
-    using PCs and the Planck 2018 likelihoods using the \texttt{plik\_lite\_TTTEEE+lowl+srollv2} likelihood combination.}
-
-    \item {The better foreground cleaned \texttt{SRoll2} likelihood allowed to tighten the PC constraint on $\tau$ and shifted it toward slightly higher values when compared with the \texttt{plik\_lite\_TTTEEE+lowl+lowE} result  $\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}$ (68\% C.L.).}
-
-    \item {We checked the robustness of our $\zmax=30$ results by performing a PC analysis of the Planck data with $\zmax = 50$.
-The results are changed negligibly: $\tau_{\rm PC}(15, 50) < 0.022 $ (95\% C.L.) in comparison with $\tau_{\rm PC}(15, 30) < 0.023$ (95\% C.L.).}
-
-     \item {Though our high-redshift optical depth constraints are looser than those from the Planck 2018 cosmological parameter paper $\tau(15, 30) < 0.007$ (95\% C.L.) obtained from another model-independent method called FlexKnot, we have tested using explicitly the tanh high-$z$ toy model (without using any PCs) that an $\tau_{\rm 2step}(15, 30) < 0.017$ (95\% C.L.) is still allowed (see Fig.~\ref{fig:plot_taugtz_two_step_contours}), which is more consistent with our PC results.}
+    \item {We obtained the total optical depth constraint $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ (68\% C.L.) 
+    using PCs in our default likelihoods \texttt{plik\_lite\_TTTEEE+lowl+Sroll2}.}
+
+    \item {Using the \texttt{SRoll2} likelihood tightened the PC constraint on $\tau$ and shifted it toward slightly higher values of $\tau$ compared to the \texttt{plik\_lite\_TTTEEE+lowl+SimAll} result $\tau_{\rm PC} = 0.0582 ^{+0.0072}_{-0.0083}$ (68\% C.L.).}
+
+    \item {We checked the robustness of our $\zmax=30$ results by performing a PC analysis of the Planck data with $\zmax = 50$ where 7 PCs were used for the larger redshift range.
+The results changed negligibly: $\tau_{\rm PC} = 0.0626 ^{+0.0061}_{-0.0072}$ (68\% C.L.).}
+
+    \item{The constraint on the high-redshift optical depth $\tau_{\rm PC}(15, 50) < 0.020$ (95\% C.L.) is also stable when extending $\zmax$ to 50: $\tau_{\rm PC}(15, 30) < 0.019$ (95\% C.L.).}
+
+     \item {Our high-redshift optical depth upper limit is however a factor of $\sim3$ larger
+     than that from the Planck 2018 cosmological parameter paper $\tau(15, 30) < 0.007$ (95\% C.L.) obtained from another model-independent method called FlexKnot.
+Using explicitly a two-step toy model (without using any PCs), we tested that $\tau_{\rm 2step}(15, 30) < 0.016$ (95\% C.L.) is still allowed, which is consistent with our PC results.}
 
@@
 
-We clarified some confusion raised in the literature on the effects of applying a range-bound prior on the PC amplitude during the analysis, and demonstrated that these priors, though they include some unphysical ionization models, do not affect the total $\tau$ constraint and the high-redshift optical depth upper limits significantly.
-In any case, as shown in Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior} removing more unphysical models would only loosen the high-$z$ optical depth upper limit, in contrary to what was claimed in Ref.~\cite{Millea:2018bko}.
-The shift from a higher to lower total $\tau$ from Planck 2015 to Planck 2018 in any model-independent is however due to the use of XX data as demonstrated by Ref.~\cite{Millea:2018bko}, in relation to better systematics reduction on the low multipoles $l<10$ [check number]~\cite{Heinrich:2018btc}.
-[cite also a Planck paper].
-
-In the future, we expect our code release RELIKE to significantly simplify the process for obtaining Planck constraints on particular ionization models. 
-Moreover, the ability of \textt{relike} to fully capture the Planck likelihood is important for accurate and consistent joint analyses with other reionization datasets (c.f.
-adopting a Planck $\tau$ prior inferred in the tanh model.) As the large-scale CMB measurements from Planck has concluded, current and future ground-based CMB measurements such as Simons Observatory and CMB-S4 will continue to refine the small-scale CMB measurements and constrain reionization through the kinetic Sunyaev-Zeldovich (kSZ) effects.
-\ch{[Address how we have marginalized over cosmology, and instruct on how to do joint fits with high-l data?
-More importantly with data in general that wants to vary As during sampling?
-Put prior on As consistent with Planck?]}.
-
-There has also been recent studies revealing that reionization might not have ended by $z = 6$~\cite{} [cite].
-If this is true, the current PC approach will need be modified to accommodate a range $z_{\rm min} < z < z_{\rm max}$ with a a lower $z_{\rm min} < 6$.
-We leave this for future work as it would require constructing a new set of PC functions.
-
-\todo{[Any comments on Ahn et al. \url{https://arxiv.org/pdf/2011.03582.pdf} and \url{https://arxiv.org/pdf/2006.16828.pdf}?]
-}
-
-In sum, we expect our code release \texttt{relike} to significantly simplify the process for obtaining Planck constraints on specific ionization models, as well as to open the way to jointly constrain legacy CMB large-scale data with a variety of increasingly rich reionization datasets coming up in the next decade.
+Finally, we clarified the effects of applying a range-bound prior on the PC amplitude while deriving model-independent constraints on the optical depth (note that this prior does not enter the \relike\  code).
+We demonstrated (see Fig.~\ref{fig:plot_taugtz_2018_with_vs_without_physicality_prior}) that removing more unphysical models would only loosen the high-$z$ optical depth upper limit, in contrary to what was claimed in Ref.~\cite{Millea:2018bko}.
+\\
+
+Given that the CMB intrinsically constrains integrated quantities of the ionization history, we recommend that model-independent constraints be formulated in the cumulative optical depth space rather than the ionization history space.
+There is also a tendancy in the literature to compare a specific model with more general constraints on $x_e(z)$.
+We caution the reader that these constraints are derived with a specific prior assumption and sometimes on a limited set of models.
+So our recommendation for constraining specific models is to use the \relike\  code which properly returns the effective Planck likelihood for any specific $x_e(z)$ function.
+
+We also recommend using \relike\  to properly combine CMB large-scale datasets with other reionization datasets such as kSZ, galaxy luminosity functions, quasars spectra, line-intensity mapping, star formation rate, etc.
+In the past, joint analyses of the reionization history often used the Planck tanh constraint as a $\tau$ prior, which is an approximate proxy for the actual likelihood of the model in question.
+With the release of \relike, one can now easily obtain the full Planck constraint on the specific model and conduct consistent joint analyses.
+
+To do so, one can interface the python \relike\  package with existing MCMC samplers for generic data such as Cobaya and CosmoSIS, or those adapted for specific reionization datasets (e.g.\ 21CMMC~\cite{Greig:2015qca} and CosmoMCReion~\cite{Chatterjee:2021ygm}).
+When the joint dataset is constraining enough (e.g.\ kSZ), one can sample while varying cosmological parameters since the \relike\  likelihood is valid for cosmologies consistent with the Planck best-fit.
+We expect that joint analyses work in the future (see e.g.\ Ref.~\cite{Ahn:2020btj, Qin:2020xrg} for some recent examples) would be made much easier with \relike.
+\\
+
+In the future, \relike\ can be extended in several ways.
+Recent studies of quasars may indicate that the Universe have not been entirely ionized by $z = 6$ ($\sim$80\% ionized at $z = 6$)~\cite{Becker:2021jyx}. If this result is confirmed,
+%in the future, 
+the PC approach can be easily modified for a smaller $\zmin$ in a manner similar to the $\zmax$ study here.
+
+Likewise current and future CMB experiments such as CLASS~\cite{Watts:2018etg} and LiteBIRD~\cite{Hazumi:2021yqq} may also improve on Planck and eventually provide cosmic variance limited measurements of CMB polarization on the largest scales, yielding improved reionization constraints.
+The same approach used here can be easily updated to capture the better constraints from these data in a joint effective likelihood.
+\\
+
+In sum, our release of the \relike\  code significantly reduces the time to constrain specific ionization models with Planck.
+It also opens the way for consistent and fast joint analyses of current and future CMB large-scale polarization data with a variety of increasingly rich reionization datasets in the next decades.
 
@@
 
-\subsection{Comparison to previous results}
-\wh{integrate this into the model sections}
-
-The Planck collaboration has released constraints on the total optical depth $\tau$ with a steplike model parameterized by the tanh function, which has evolved from $\tau = 0.067 \pm 0.022$ (Planck 2015) to $\tau = 0.055 \pm 0.009$ (Planck 2016), then to $\tau = 0.0506 \pm 0.0086$ (Planck 2018) [cite papers].
-The major improvement comes from clean-up of systematics in the High Frequency Instrument (HFI) data.
-
-In Planck 2018 VI, using only lowE data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for the tanh model,
-$\tau = 0.0504^{+0.0050}_{-0.0079}$ 
-for the FlexKnot method (with flat $\tau$ prior) and 
-$\tau = 0.0487^{+0.0038}_{-0.0081}$
-for the PC method but using the prior inversion procedure
-discussed \S\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models.
-[do we want to check whether the prior inversion procedure could cause a shift at this level?
-do we want to evaluate this statement?]
-
-They also found the upper limit on high-redshift optical depth to be: $\tau(15, 30) < 0.006$ for the FlexKnot method with a flat $\tau(15, 30)$ prior or $<0.007$ with a flat knot prior.
-The PC result for the upper limit was not computed.
-
-Compared to our PC results ($\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\% \mathrm{CL})$), the FlexKnot upper limits are too stringent.
-The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in \S\ref{sec:example2} with a concrete example of the two-step model.
-We showed that for this chosen physical model, the $\tau(15, 30)$ upper limits derived from the full Planck likelihood without using any PCs were already not compatible with the FlexKnot upper limits, while the PC results were compatible.
-So the discrepancy cannot be attributed to the prior choice in the PC results.
-It also seems that the inclusion of unphysical model would in any case shift the optical depth at high-redshift toward more negative values, which would decrease rather than increase the upper limit.
-
-[just noticed that our tau(15, 30) $<$ 0.023 result is with srollv2, and flexknot tau(15,30) $<$ 0.007 is with lowE likelihood.
-In our analysis lowE to sroll2 had a 0.004 increase in total tau.
-I should run the tau(15, 30) constraint on the lowE chain as well just to double check.]
-
-[Need to address Planck2018 cosmology paper claim: ``The PCA result is slightly lower, and is partly affected by the imperfect physicality priors that allow unphysical negative ionization fractions."]
-
-[Also, they only used lowE likelihood, fixed all other cosmological parameters (including $A_s e^{-2\tau}$). Wondering if that does anything to the difference between our results]
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\section{Outline (in progress)}
-
-\begin{enumerate}
-    \item{Intro}
-        \begin{enumerate}
-            \item 1st paragraph: CMB, reionization, why important
-            \item discuss modeling of ionization history in CMB inference, tanh vs PCs (fail to capture high-z; PC captures complete parameter space), mention flex knots/other methods; cite previous PC works.
-            \item give context about Planck final results; latest srollv2 likelihood, important to harness all information available, PC allows us to do that and turn into an effective likelihood.
-            \item this is what we did: we obtain PC results for Planck 2018, and turn PC chains into a effective likelihood for ionization history: highlight key points like fast inference, entire model space up to zmax 30, code publicly available.
-            \item we also compared to Planck official results, which is over stringent at high z, which gives additional motivations for using this likelihood; verified no hint of ionization at z>30; state any additional results.
-            \item break down of sections.
-        \end{enumerate}
-
-	\item{Background}
-		\begin{itemize}
-			\item{Reionization Principal Components}	
-			\item{Kernel Density Estimate}
-		\end{itemize}
-	\item{Planck 2018 PC results:\\
-		- can discuss discrepancy here on high-redshift tau($>$15) constraints.\\
-		- compare with our own 2015 PC results
-		- zmax = 30 vs 50}
-	\item{Effective Likelihood}
-		\begin{itemize}
-			\item{Code Description}
-			\item{Examples - one and two parameter models}
-		\end{itemize}
-	\item{Discussion}
-
-\end{enumerate}
-
------- \textbf{Some Numbers} ------ \\
-
-\textbf{Planck official results 2015, 2016, 2018} \\
-
-\begin{itemize}
-
-    \item Planck 2015 results (LFI): $\tau = 0.067 \pm 0.022$ (68\%) \\
-
-    \item Planck 2016  intermediate results (LFI): $\tau = 0.055 \pm 0.009$ (68\%) \\
-
-    \item Planck 2018 results: $\tau = 0.0506 \pm 0.0086$ (68\%) \\
-
-\end{itemize}
-
-\textbf{Planck 2018 paper} \\
-
-Here lowE means lowE data only, fixing all other cosmological parameters including $A_s e^{-2\tau}$\\
-
-\begin{itemize}
-
-\item $\tau = 0.0519+0.0030-0.0079$ (lowE; flat $\tau$ prior; TANH)(68\%); \\
-
-\item $\tau = 0.0504+0.0050
--0.0079$ (lowE; flat $\tau$ prior; FlexKnot)(68\%); \\
-
-\item $\tau = 0.0487+0.0038
--0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).
-
-\end{itemize}
-
-\textbf{Upper limit on high-redshift optical depth} (only given for the FlexKnot method):
-
-$\tau(15, 30) < 0.006$ (lowE, flat $\tau(15, 30)$, FlexKnot); \\
-
-$\tau(15, 30) < 0.007$ (lowE, flat knot, FlexKnot).\\
-
-Statements to address:
-
-\begin{enumerate}
-    \item {\textbf{Statement on prior in Planck 2018}:  ``Heinrich \& Hu (2018) construct a prior that is uniform on $\tau$, but which increases the allowed unphysical parameter space and is chosen a posteriori.
-Here we instead use the flat prior constructed by the procedure described in Millea \& Bouchet (2018) and Handley \& Millea (2019), which does not admit extra unphysical models and gives the most generic prior that leaves the prior on $\tau$ uniform."}
-
-    \item {Is this statement on tanh in Planck 2018 paper correct?} ``The TANH result gives slightly higher optical depth than the others, which is primarily driven by the fixed duration of reionization assumed." \\
-
-    \item{\textbf{Other comments on our PC prior giving higher optical depth}: ``The PCA result is slightly lower, and is partly affected by the imperfect physicality priors that allow unphysical negative ionization fractions." and ``Millea \& Bouchet showed that the majority of this (2$\sigma$) preference disappeared when using the lower-noise Planck HFI SimLow likelihood (intermediate results reduction of systematics), with an additional sub-dominant effect due to the choice of prior."}
-
-\end{enumerate}
-
-How likelihoods and systematics evolved over time:
-
-\begin{enumerate}
-    \item 
-    \item {Planck 2018 paper attribute the 2018 upper limit on $\tau(15, 30)$ being \textbf{about 3 times more} stringent than the intermediate results by Millea \& Bouchet to the $SimLow \rightarrow SimAll$ where there was better control of systematics in HFI polarization.}
-\end{enumerate}
-
-[high redshift going away should be attributed to better control of systematics in HFI polarization data (changes in the SimAll likelihood compared to SimLow) [cite Planck 2018 results cp]]
-
-\section{Reference Plots for Internal Checks}
-
-\begin{figure}
-\includegraphics[width=0.48\textwidth]{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
-\caption{\ch{caption to be added, will take away without prior line}}
-\label{fig:tau_posterior_PC_vs_tanh_vs_tanh_highz}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{$\tau_{12,new}$ - \wh{this is an internal plot - leave in for now but also make a version for the paper: just $\tau_{12}$ vs $\tau_{\rm PC}$ in our new convention with and without prior either as a single or separate panel whichever is clearer.}}
-%\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
-\caption{ \ch{CH: this is the same as above but with priors; also tested that $\tau_{12, new}$ with mean of m3, m4, m5 computed from chains with prior applied performs better here (as expected) from chains without prior.}}
-%\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
-\caption{$\tau$ decomposition (68\% C.L.
-only) \wh{this is a temporary plot for internal purposes - what I'm seeing here is that without the physicality cut the high $z$ side goes unphysically negative.
-I think this is consistent with our description and the $S_1$ curve.}}
-%\label{fig:tau12}
-\end{figure}
-
-\begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
-\caption{\ch{Internal for now (with prior version)}}
-%\label{fig:tau12}
-\end{figure}
-
-\begin{table}[b]
-\centering
-\caption{$\tau_a$}
-%\label{tab:PC_stats}
-%\begin{tabular}{|r | r r r r r r|}
-\begin{tabular}{|c | c | c | c | c | c|}
-\hline
-$\tau_{\rm fid}$ & $\tau_1$  & $\tau_2$ & $\tau_3$ & $\tau_4$ & $\tau_5$ \\ \hline
-0.08626 & 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{table}[b]
-\centering
-\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$. \ch{transpose this table and add a row for taumj, and quote taufid in the caption}}
-%\label{tab:PC_stats}
-\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
-\hline
-
-			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
-		\\ \hline
-
-$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
-$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
-$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$ \\
-$m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
-$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\
-
-\hline
-\end{tabular}
-\end{table}
-
-% all digits without flipping m3, m4:
-%tau_fid 0.08625526676478058
-%taumj [ 0.29403315 -0.11227135 -0.04505736  0.0189787   0.00960184
-\begin{table}[b]
-\centering
-\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
-%\label{tab:PC_stats}
-\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
-\hline
-
-			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$}
-		\\ \hline
-
-$m_1$ 
-	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
-$m_2$ 
-	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
-$m_3$ 
-	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
-$m_4$  
-	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
-$m_5$ 
-	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ 
-\hline
-\end{tabular}
-\end{table}
-
 \end{document}
-
-% Notes:
-% tau1 = (taulo + tauhi) --> (zre, xemin) --> xe(z) ==> tau2 ==> mjs --> tau3(mjs)
-% taulo + tauhi = tau_desired

### 2021-04-28 results/notes_during_paper_writing.tex
Final copy

\documentclass[prd,amsmath,amssymb,floatfix,superscriptaddress,nofootinbib]{revtex4-1}
\usepackage{bm}
\usepackage{amsmath}
\usepackage{epsfig}
\usepackage{color}
\usepackage{natbib}
\usepackage{textcase}
\usepackage{graphicx}
\usepackage{ifthen}
\usepackage{xstring}
\usepackage{graphicx}
\usepackage[utf8]{inputenc} 
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{epstopdf}
\epstopdfsetup{update}
\DeclareGraphicsExtensions{.ps, .png}
\epstopdfDeclareGraphicsRule{.ps}{pdf}{.pdf}{ps2pdf -dEPSCrop -dNOSAFER #1 \OutputFile} 
\usepackage{dcolumn} 
\usepackage{multirow}
\usepackage{appendix}
\usepackage{footnote}
\usepackage{tabularx,ragged2e,booktabs}
\usepackage[normalem]{ulem}
\usepackage{float}
\restylefloat{table}

\newcommand{\tauhi}{$\tau_{\rm hi}\,$}
\newcommand{\taulo}{$\tau_{\rm lo}\,$}

\newcommand{\refsec}[1]{section~\ref{sec:#1}}
\newcommand{\refeq}[1]{Eq.~(\ref{eq:#1})}
\newcommand{\refssec}[1]{section~\ref{subsec:#1}}
\newcommand{\reffig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\refFig}[1]{Fig.~\ref{fig:#1}}
\newcommand{\curv}{{\cal R}}
\newcommand{\xef}{x_e^{\rm fid}}
\newcommand{\xmax}{x_e^{\rm max}}
\newcommand{\zmax}{z_{\rm max}}
\newcommand{\zmin}{z_{\rm min}}
\newcommand{\xemin}{x_e^{\rm min}}

\newcommand{\ra}{\rightarrow}
\def\max{_{\mathrm{max}}}
\def\lsim{\mathrel{\raise.3ex\hbox{$$<$$\kern-.75em\lower1ex\hbox{$\sim$}}}}
\def\gsim{\mathrel{\raise.3ex\hbox{$$>$$\kern-.75em\lower1ex\hbox{$\sim$}}}}

\newcommand{\beq}{\begin{equation}}
\newcommand{\eeq}{\end{equation}}

\newcommand{\bea}{\begin{eqnarray}}
\newcommand{\eea}{\end{eqnarray}}

\newcommand{\wh}[1]{\textcolor{blue}{#1}}
\newcommand{\ch}[1]{\textcolor{red}{#1}}

\def\mnras{Mon.\ Not.\ R.\ Astron.\ Soc.\ }
\definecolor{darkgreen}{cmyk}{0.85,0.2,1.00,0.2} 
\definecolor{purple}{cmyk}{0.5,1.0,0,0} 
\def\physrep{Phys.~Rep.}

\definecolor{ultramarine}{rgb}{0.07, 0.04, 0.56}
\definecolor{cadmiumgreen}{rgb}{0.0, 0.42, 0.24}
\definecolor{indigo(dye)}{rgb}{0.0, 0.25, 0.42}
\usepackage[linktocpage=true]{hyperref}
\hypersetup{
colorlinks=true,
citecolor=ultramarine,
linkcolor=cadmiumgreen,
urlcolor=indigo(dye),
pdfauthor={},
pdftitle={},
pdfsubject={}
}

\begin{document}

\title{Reionization Planck 2018 \\
Notes during paper writing phase}

\author{Chen Heinrich}
\author{Wayne Hu}

\maketitle

\section{Comparison to previous results}

The Planck collaboration has released constraints on the total optical depth $\tau$ with a steplike model parameterized by the tanh function, which has evolved from $\tau = 0.067 \pm 0.022$ (Planck 2015) to $\tau = 0.055 \pm 0.009$ (Planck 2016), then to $\tau = 0.0506 \pm 0.0086$ (Planck 2018) \todo{[cite papers]}. The major improvement comes from clean-up of systematics in the High Frequency Instrument (HFI) data.

In Planck 2018 VI, using only \texttt{SimAll} data and fixing all cosmological parameters including $A_s e^{-2\tau}$, the optical depth constraints was found to be $\tau = 0.0519^{+0.0030}_{-0.0079}$ for the tanh model, 
$\tau = 0.0504^{+0.0050}_{-0.0079}$ 
for the FlexKnot method (with flat $\tau$ prior) and 
$\tau = 0.0487^{+0.0038}_{-0.0081}$
for the PC method but using the prior inversion procedure
discussed \S\ref{sec:note_on_priors} and attributed to the slightly lower central value from the PC method to the fact that the priors contain unphysical models. [do we want to check whether the prior inversion procedure could cause a shift at this level? do we want to evaluate this statement?]

They also found the upper limit on high-redshift optical depth to be: $\tau(15, 30) < 0.006$ for the FlexKnot method with a flat $\tau(15, 30)$ prior or $<0.007$ with a flat knot prior. The PC result for the upper limit was not computed.

Compared to our PC results ($\tau(15, \zmax)_{\rm PC,\, \zmax=30} < 0.023 \; (95\% \mathrm{CL})$), the FlexKnot upper limits are too stringent. The difference is not attributed to the inclusion of unphysical models due to the truncation at 5 PCs as we showed in \S\ref{sec:example2} with a concrete example of the two-step model. We showed that for this chosen physical model, the $\tau(15, 30)$ upper limits derived from the full Planck likelihood without using any PCs were already not compatible with the FlexKnot upper limits, while the PC results were compatible. So the discrepancy cannot be attributed to the prior choice in the PC results. It also seems that the inclusion of unphysical model would in any case shift the optical depth at high-redshift toward more negative values, which would decrease rather than increase the upper limit.

[just noticed that our tau(15, 30) $<$ 0.023 result is with srollv2, and flexknot tau(15,30) $<$ 0.007 is with \texttt{SimAll} likelihood. In our analysis \texttt{SimAll} to sroll2 had a 0.004 increase in total tau. I should run the tau(15, 30) constraint on the \texttt{SimAll} chain as well just to double check.]

[Need to address Planck2018 cosmology paper claim: ``The PCA result is slightly lower, and is partly affected by the imperfect physicality priors that allow unphysical negative ionization fractions."]

[Also, they only used \texttt{SimAll} likelihood, fixed all other cosmological parameters (including $A_s e^{-2\tau}$). Wondering if that does anything to the difference between our results]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Some Numbers}

\textbf{Planck official results 2015, 2016, 2018} \\

\begin{itemize}

    \item Planck 2015 results (LFI): $\tau = 0.067 \pm 0.022$ (68\%) \\

    \item Planck 2016  intermediate results (LFI): $\tau = 0.055 \pm 0.009$ (68\%) \\

    \item Planck 2018 results: $\tau = 0.0506 \pm 0.0086$ (68\%) \\

\end{itemize}

\textbf{Planck 2018 paper} \\

Here lowE means lowE data only, fixing all other cosmological parameters including $A_s e^{-2\tau}$\\

\begin{itemize}

\item $\tau = 0.0519+0.0030-0.0079$ (lowE; flat $\tau$ prior; TANH)(68\%); \\

\item $\tau = 0.0504+0.0050
-0.0079$ (lowE; flat $\tau$ prior; FlexKnot)(68\%); \\

\item $\tau = 0.0487+0.0038
-0.0081$ (lowE; flat $\tau$ prior; PCA)(68\%).

\end{itemize}

\textbf{Upper limit on high-redshift optical depth} (only given for the FlexKnot method):

$\tau(15, 30) < 0.006$ (lowE, flat $\tau(15, 30)$, FlexKnot); \\

$\tau(15, 30) < 0.007$ (lowE, flat knot, FlexKnot).\\

Statements to address:

\begin{enumerate}
    \item {\textbf{Statement on prior in Planck 2018}:  ``Heinrich \& Hu (2018) construct a prior that is uniform on $\tau$, but which increases the allowed unphysical parameter space and is chosen a posteriori. Here we instead use the flat prior constructed by the procedure described in Millea \& Bouchet (2018) and Handley \& Millea (2019), which does not admit extra unphysical models and gives the most generic prior that leaves the prior on $\tau$ uniform."}

    \item {Is this statement on tanh in Planck 2018 paper correct?} ``The TANH result gives slightly higher optical depth than the others, which is primarily driven by the fixed duration of reionization assumed." \\

    \item{\textbf{Other comments on our PC prior giving higher optical depth}: ``The PCA result is slightly lower, and is partly affected by the imperfect physicality priors that allow unphysical negative ionization fractions." and ``Millea \& Bouchet showed that the majority of this (2$\sigma$) preference disappeared when using the lower-noise Planck HFI SimLow likelihood (intermediate results reduction of systematics), with an additional sub-dominant effect due to the choice of prior."}

\end{enumerate}

How likelihoods and systematics evolved over time:

\begin{enumerate}
    \item 
    \item {Planck 2018 paper attribute the 2018 upper limit on $\tau(15, 30)$ being \textbf{about 3 times more} stringent than the intermediate results by Millea \& Bouchet to the $SimLow \rightarrow SimAll$ where there was better control of systematics in HFI polarization.}
\end{enumerate}

[high redshift going away should be attributed to better control of systematics in HFI polarization data (changes in the SimAll likelihood compared to SimLow) [cite Planck 2018 results cp]]

\section{Reference Plots for Internal Checks}

\begin{figure}
\includegraphics[width=0.48\textwidth]{results/tau_posterior_comparisons/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
\caption{\ch{caption to be added, will take away without prior line}}
\label{fig:tau_posterior_PC_vs_tanh_vs_tanh_highz}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_False_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{$\tau_{12,new}$ - this is an internal plot - leave in for now but also make a version for the paper: just $\tau_{12}$ vs $\tau_{\rm PC}$ in our new convention with and without prior either as a single or separate panel whichever is clearer.}
%\label{fig:tau12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/plot_taumj_decomposition_apply_cut_True_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
\caption{this is the same as above but with priors; also tested that $\tau_{12, new}$ with mean of m3, m4, m5 computed from chains with prior applied performs better here (as expected) from chains without prior.}
%\label{fig:tau12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only.pdf}
\caption{$\tau$ decomposition (68\% C.L. only) \wh{this is a temporary plot for internal purposes - what I'm seeing here is that without the physicality cut the high $z$ side goes unphysically negative.  I think this is consistent with our description and the $S_1$ curve.}}
%\label{fig:tau12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.48\textwidth]{results/tau_pc_decomposition/pl18_taugtz_pc_decomposition_68_only_apply_cut_True.pdf}
\caption{\ch{Internal for now (with prior version)}}
%\label{fig:tau12}
\end{figure}

\begin{table}[b]
\centering
\caption{$\tau_a$}
%\label{tab:PC_stats}
%\begin{tabular}{|r | r r r r r r|}
\begin{tabular}{|c | c | c | c | c | c|}
\hline
$\tau_{\rm fid}$ & $\tau_1$  & $\tau_2$ & $\tau_3$ & $\tau_4$ & $\tau_5$ \\ \hline
0.08626 & 0.29403 & $-$0.11227 & 0.04506 & $-$0.01898 & 0.00960 \\
\hline
\end{tabular}
\end{table}

\begin{table}[b]
\centering
\caption{PC chain means $\bar m_a$, standard deviations $\sigma(m_a)$, and correlation matrix $R_{ab}$.}
%\label{tab:PC_stats}
\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
\hline

			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$} 
		\\ \hline

$m_1$ & $-$0.116 & 0.030 & 1.000 & 0.657 & $-$0.208 & $-$0.094 & 0.067 \\
$m_2$ & $-$0.016 & 0.063 & 0.657 & 1.000 & 0.048 & $-$0.273 & 0.142 \\
$m_3$ & $$0.078 & 0.098 & $-$0.208 & 0.048 & 1.000 & 0.064 & $-$ \\
$m_4$ & $-$0.077 & 0.131 & $-$0.094 & $-$0.273 & 0.064 & 1.000 & $-$0.195 \\
$m_5$ & 0.066 & 0.145 & 0.067 & 0.142 & $-$0.018 & $-$0.195 & 1.000 \\

\hline
\end{tabular}
\end{table}

% all digits without flipping m3, m4:
%tau_fid 0.08625526676478058
%taumj [ 0.29403315 -0.11227135 -0.04505736  0.0189787   0.00960184
\begin{table}[b]
\centering
\caption{Planck 2015 results (with Mortonson PCs) [Will not be in the paper, just for us to compare] }
%\label{tab:PC_stats}
\begin{tabular}{|r | r r@{\hskip 0.06in}|r r r r r|}
\hline

			  &  \multicolumn{1}{c}{$\bar m_a$} & \multicolumn{1}{c}{$\sigma(m_a)$}	 & \multicolumn{1}{|c}{$m_1$} & \multicolumn{1}{c}{$m_2$} & \multicolumn{1}{c}{$m_3$} & \multicolumn{1}{c}{$m_4$} & \multicolumn{1}{c|}{$m_5$} 
		\\ \hline

$m_1$ 
	& 0.002 & 0.053 & 1.000 & 0.450 & $-$0.432 & 0.273 & $-$0.073 \\ 
$m_2$ 
	& $-$0.030 &  0.101 & 0.450 & 1.000 & $-$0.262 & 0.055 & 0.072 \\ 
$m_3$ 
	& 0.019 &  0.128 & $-$0.432 & $-$0.262 &1.000 & $-$0.417 & 0.155 \\
$m_4$  
	& $-$0.012 & 0.143 &  0.273 & 0.055 & $-$0.417 & 1.000 & $-$0.428 \\ 
$m_5$ 
	& 0.026 & 0.143 & $-$0.073 & 0.072 & 0.155 & $-$0.428 & 1.000\\ 
\hline
\end{tabular}
\end{table}

% Notes:
% tau1 = (taulo + tauhi) --> (zre, xemin) --> xe(z) ==> tau2 ==> mjs --> tau3(mjs)
% taulo + tauhi = tau_desired

\begin{figure}
\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
\caption{Two step ionization history $x_e(z)$ for the model of \refsec{example2}. In addition to a tanh transition at low-redshift as in \refsec{example1}, there is an additional ionization plateau at high redshifts with a fixed transition at $z_t = 28$ and $\Delta z = 1$. Here we show the best-fit model in the Planck 2018 data $(\taulo, \tauhi) = (0.053, 0.006)$, as well as  a model on the 95\% C.L. contour of Fig.~\ref{fig:two_parameter_model_2D}, $(\taulo, \tauhi) = (0.043, 0.026)$. 
%best-fit (0.05290344, 0.005859431)
%95\%CL (0.0427, 0.0256): zre = 6.4685, xemin = 0.0925 
}
\label{fig:two_step_model}
\end{figure}

\end{document}

%\ch{Should we do a table giving an overview of tau measurement evolution from Planck collaboration and in the literature, for tanh and other model-independent approaches?}
%\todo{e.g.
%Planck 2015 (LFI only)\\
%Planck 2016 XLVI(\cite{Aghanim:2016yuo}) with reduction of systematics for HFI \\
%Planck 2016 XLVII (\cite{Adam:2016hgk}) - intermediate results LFI + HFI, tanh\\
%Ref.~\cite{Millea:2018bko} - intermediate results LFI + HFI generic model. \\
%Planck 2018 cosmological parameters paper LFI + HFI w/ SimAll low-l EE likelihood (FlexKnot)\\
%Srollv2 measurement Ref.~\cite{Pagano:2019tci}
%}
%

### 2021-04-28 submissions/arxiv_v1_20210428/rei.tex
added submissions folder

--- 2021-04-28 paper/rei.tex
@@
-\documentclass[prd,twocolumn,amsmath,amssymb,floatfix,superscriptaddress,nofootinbib]{revtex4-1}
+\documentclass[aps,prd,twocolumn,amsmath,amssymb,showpacs,floatfix,superscriptaddress,nofootinbib]{revtex4-1}
+
+\pdfoutput=1
+\synctex=1
+
+%\usepackage{epsfig}
+\usepackage{natbib}
+%\usepackage{textcase}
+%\usepackage{ifthen}
+%\usepackage{xstring}
+\usepackage[utf8]{inputenc} 
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{bm}
-\usepackage{amsmath}
-\usepackage{epsfig}
-\usepackage{color}
-\usepackage{natbib}
-\usepackage{textcase}
-\usepackage{graphicx}
-\usepackage{ifthen}
-\usepackage{xstring}
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc} 
-\usepackage{amssymb}
 \usepackage{latexsym}
+\usepackage{graphicx,epstopdf}
 \usepackage{epstopdf}
@@
 \usepackage{dcolumn} 
-\usepackage{multirow}
-\usepackage{appendix}
+%\usepackage{multirow}
+%\usepackage{appendix}
 \usepackage{footnote}
 \usepackage{tabularx,ragged2e,booktabs}
+\usepackage{breqn}
 \usepackage[normalem]{ulem}
@@
 \restylefloat{table}
+\usepackage{color}
 
@@
 
-\newcommand{\relike}{RELIKE}%\texttt{RELIKE} 
-\newcommand{\cosmomcrelike}{\texttt{CosmoMC-relike}}
+\newcommand{\relike}{RELIKE}
+\newcommand{\cosmomcrelike}{\texttt{CosmoMC-RELIKE}}
 
@@
 
-\title{\relike: Reionization Effective Likelihood from Planck 2018 Data}
+\title{RELIKE: Reionization Effective Likelihood from Planck 2018 Data}
 
 \author{Chen Heinrich}\email{chenhe@caltech.edu}
-\affiliation{California Institute of Technology, Pasadena, California 91109, USA}
-\affiliation{Jet Propulsion Laboratory, California Institute of Technology, Pasadena, California 91109, USA}
+\affiliation{$California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
 
 \author{Wayne Hu}
-\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago Illinois 60637}
-\affiliation{Department of Astronomy \& Astrophysics,
- University of Chicago, Illinois 60637}
+\affiliation{$Kavli\ Institute\ for\ Cosmological\ Physics,\ Enrico Fermi Institute,\ University of Chicago,\ Chicago\ Illinois\ 60637$}
+\affiliation{$Department\ of\ Astronomy\ \&\ Astrophysics,\ University\ of\ Chicago,\ Illinois\ 60637$}
 
@@
 
-We release \relike\  (Reionization Effective Likelihood), a fast and accurate effective likelihood code based on the latest Planck 2018 data 
+We release RELIKE (Reionization Effective Likelihood), a fast and accurate effective likelihood code based on the latest Planck 2018 data 
 that allows one constrain any model for reionization between $6 < z < 30$ using five constraints from the CMB reionization principal components (PC).
@@
 \end{abstract}
-\pacs{}
+%\pacs{}
 
@@
 It is usually characterized by a single parameter, the total Thomson optical depth, whose  uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude.
-Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.
+Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk},
+%Allison:2015qca
+and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.
 
@@
 In the CMB data, free electrons during the reionization epoch Compton scatter with the CMB photons to cause a suppression of the CMB primary anisotropies; on the large scales, the same scattering process of the CMB temperature quadrupole in the frame of the free electrons induces an additional reionization bump in the CMB $E$-mode polarization.
-On the small scales, the kinetic Sunyaev-Zeldovich (kSZ) effects generated by the scattering of the CMB photons off ionized gas in bulk motion~\cite{zeldovich_sunyaev_1969, sunyaev_zeldovich_1980} are used to constrain morphology and duration of reionization~\cite{mcquinn_2005,mesinger_2012_kSZ}.
+On the small scales, the kinetic Sunyaev-Zeldovich (kSZ) effects generated by the scattering of the CMB photons off ionized gas in bulk motion~\cite{Sunyaev:1970er, Sunyaev:1980vz} are used to constrain morphology and duration of reionization~\cite{mcquinn_2005,mesinger_2012_kSZ}.
 Additionally, observations of galaxy luminosity functions, quasars, star formation rate, line-intensity mapping also provide important and complementary information on reionization~\cite{2016ASSL..423.....M}.
@@
  \begin{figure}
-          \includegraphics[width=0.48\textwidth]{paper/plots/pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
+          \includegraphics[width=0.48\textwidth]{pl18_plot_pub_xe_basis_tau_basis_zmax30_heinrich.pdf}
           \caption{\textit{Top}: Rank-ordered CMB reionization principal components $S_a$ for $a = 1..5$ for $\zmax = 30$.
@@
 \begin{figure*}
-\includegraphics[width=0.8\textwidth]{paper/plots/plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
+\includegraphics[width=0.8\textwidth]{plot_mj_triangle_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_srollv2_1015_wTauTrajectory_pl15_wTanhML_wGaussEllipse.pdf}
 \caption{Constraints from Planck data on the amplitudes of the five reionization principal components that describe all physical ionization models up to $z_{\rm max} = 30$.
@@
 \begin{figure}
-\includegraphics[width=0.4\textwidth]{plots/plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
+\includegraphics[width=0.4\textwidth]{plot_m1_m2_t18_r12_t19_t20_vs_pl18_pc_zmax30_pliklite_0930_vs_pl18_pc_zmax30_pliklite_srollv2_1015.pdf}
 \caption{Changes in the best constrained $m_1-m_2$ PC plane between Planck 2015 (red) and Planck 2018 (blue,  \texttt{SRoll2}; green, \texttt{SimAll}).
@@
 \includegraphics[width=0.48\textwidth]
-{paper/plots/pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
+{pl18_tau_posterior_tanh_vs_tanh_kde_vs_tanh_gaussian_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
 \caption{Total optical depth $\tau_{\rm tanh}$ posterior for the tanh model in \refsec{example1} for the exact (Planck 2018) likelihood (shaded), KDE 
@@
 \begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
+\includegraphics[width=0.48\textwidth]{pl18_taugtz_tanh_direct_vs_kde_vs_gaussian.pdf}
 \caption{Cumulative optical depth in the tanh model as in Fig.~\ref{fig:tanh}
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]{paper/plots/plot_xe_tanh_highz.pdf}
+\includegraphics[width=0.48\textwidth]{plot_xe_tanh_highz.pdf}
 \caption{Two-step ionization history $x_e(z)$ for the model of \refsec{example2}.
@@
 \begin{figure}[t]
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
+\includegraphics[width=0.48\textwidth]{pl18_tanh_highz_test7_run1_vs_relike_tanh_highz_test9_run5_f0p14_taulo_prior_0p03_zre_prior_6p1_taulo_prior_0p0_tri.pdf}
 \caption{Two-step parameter constraints (marginalized posteriors and 68\% and 95\% C.L.
@@
 \begin{figure}
-\includegraphics[width=0.48\textwidth]
-{paper/plots/pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
+\includegraphics[width=0.48\textwidth]{pl18_taugtz_tanh_highz_direct_vs_kde_vs_gaussian.pdf}
 \caption{Cumulative optical depth in the two-step model (as in Fig.~\ref{fig:plot_taugtz_PC_vs_tanh}).
@@
 \includegraphics[width=0.48\textwidth]
-{paper/plots/pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
+{pl18_tau_posterior_tanh_vs_tanh_highz_vs_pc_dz_auto_zre_prior_6p1_normalized_by_max.pdf}
 \caption{Total optical depth posterior for the model-independent PC analysis (shaded) compared with the exact tanh model (blue solid line) and the two-step model (black dashed line) from Figs.~\ref{fig:tanh} and~\ref{fig:two_parameter_model_2D}.
@@
 \begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
+\includegraphics[width=0.48\textwidth]{plot_tau12_new_apply_cut_False_vs_True_pl18_pc_zmax30_pliklite_srollv2_1015_two_panels.pdf}
 \caption{Total optical depth $\tau_{\rm PC}$ posterior (shaded) without (top) and with (bottom) the additional physicality prior of Eq.~(\ref{eq:individualprior}).  
@@
  \begin{figure}
-          \includegraphics[width=0.9\columnwidth]{paper/plots/pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau12.pdf}
+          \includegraphics[width=0.9\columnwidth]{pl18_pc_zmax30_pliklite_srollv2_1015_plot_rotated_box_flat_tau12.pdf}
           \caption {Priors on the $m_1-m_2$ parameter space: the default physicality prior (inset square, solid lines) and an alternate prior which is  flat in its marginal
@@
 \begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
+\includegraphics[width=0.48\textwidth]{pl18_taugtz_pc_zmax30_pl18_srollv2_with_and_without_physicality_prior_do_rescale_tau_false.pdf}
 \caption{Cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ from the model-independent PC approach with and without the physicality prior.
@@
 \begin{figure}[ht]
-\includegraphics[width=0.45\textwidth]{paper/plots/pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2_do_rescale_tau_false.pdf}
+\includegraphics[width=0.45\textwidth]{pl18_taugtz_pc_zmax30_pl15_vs_pl18_simallEE_vs_pl18_srollv2_do_rescale_tau_false.pdf}
 \caption{Changes in the
@@
 \begin{figure}[ht]
-\includegraphics[width=0.48\textwidth]{paper/plots/pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
+\includegraphics[width=0.48\textwidth]{pl18_taugtz_pl18_srollv2_pc_zmax30_vs_zmax50_do_rescale_tau_false.pdf}
 \caption{Robustness of the cumulative optical depth $\tau_{\rm PC}(z, \zmax)$ to changes from the default $\zmax = 30$ to $\zmax = 50$.
@@
 
+\begin{acknowledgements}
+We thank Tzu-Ching Chang, GuoChao Sun, Lluis Mas-Aribas and Olivier Dor\'{e} for useful discussions. W.H. was supported by U.S. Dept. of Energy contract DE-FG02-13ER41958 and the Simons Foundation.
+\end{acknowledgements}
+
 \bibliography{rei.bib}

### 2021-04-28 paper/rei.tex
Updated paper/ to mirror latest arxiv submission (v1) + additionally restored Allison et al. reference; moved intermediate plots to results/plots_during_paper_writing/;

@@
 It is usually characterized by a single parameter, the total Thomson optical depth, whose  uncertainty propagates into the inferences of other important parameters such as the primordial power spectrum amplitude.
-Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk,Allison:2015qca}; and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.
+Through it, the uncertainty in reionization will become one of the major sources of uncertainty for measuring the sum of neutrino masses from future gravitational lensing measurements of the CMB~ \cite{Smith:2006nk, Allison:2015qca} and will also have implications for inferring cosmic acceleration through the growth of structure~\cite{Hu:2003pt}.
 

### 2021-05-11 paper/rei.tex
fixed typo in taupc(15,30) <-->taupc(15,50) in conclusion; fixed error in .bib file.

@@
 
-    \item{The constraint on the high-redshift optical depth $\tau_{\rm PC}(15, 50) < 0.020$ (95\% C.L.) is also stable when extending $\zmax$ to 50: $\tau_{\rm PC}(15, 30) < 0.019$ (95\% C.L.).}
+    \item{The constraint on the high-redshift optical depth $\tau_{\rm PC}(15, 30) < 0.020$ (95\% C.L.) is also stable when extending $\zmax$ to 50: $\tau_{\rm PC}(15, 50) < 0.019$ (95\% C.L.).}
 

### 2021-05-11 paper/rei.tex
finalized version for PRD submission v1

@@
 In the future, \relike\ can be extended in several ways.
-Recent studies of quasars may indicate that the Universe have not been entirely ionized by $z = 6$ ($\sim$80\% ionized at $z = 6$)~\cite{Becker:2021jyx}. If this result is confirmed,
-%in the future, 
-the PC approach can be easily modified for a smaller $\zmin$ in a manner similar to the $\zmax$ study here.
+Recent studies of quasars may indicate that the Universe have not been entirely ionized by $z = 6$ ($\sim$80\% ionized at $z = 6$)~\cite{Becker:2021jyx}.
+If this result is confirmed, the PC approach can be easily modified for a smaller $\zmin$ in a manner similar to the $\zmax$ study here.
 
-Likewise current and future CMB experiments such as CLASS~\cite{Watts:2018etg} and LiteBIRD~\cite{Hazumi:2021yqq} may also improve on Planck and eventually provide cosmic variance limited measurements of CMB polarization on the largest scales, yielding improved reionization constraints.
+Likewise current and future CMB experiments such as CLASS~\cite{Watts:2018etg, Watts:2019uvq} and LiteBIRD~\cite{Hazumi:2021yqq} may also improve on Planck and eventually provide cosmic variance limited measurements of CMB polarization on the largest scales, yielding improved reionization constraints.
 The same approach used here can be easily updated to capture the better constraints from these data in a joint effective likelihood.

### 2021-05-11 submissions/PRD/PRD_submission_v1/rei.tex
added rei.tex file for PRD submission

--- 2021-05-11 paper/rei.tex
@@
 \author{Wayne Hu}
-\affiliation{$Kavli\ Institute\ for\ Cosmological\ Physics,\ Enrico Fermi Institute,\ University of Chicago,\ Chicago\ Illinois\ 60637$}
+\affiliation{$Kavli\ Institute\ for\ Cosmological\ Physics,\ Enrico\ Fermi\ Institute,\ University\ of\ Chicago,\ Chicago\ Illinois\ 60637$}
 \affiliation{$Department\ of\ Astronomy\ \&\ Astrophysics,\ University\ of\ Chicago,\ Illinois\ 60637$}
@@
 We release RELIKE (Reionization Effective Likelihood), a fast and accurate effective likelihood code based on the latest Planck 2018 data 
-that allows one constrain any model for reionization between $6 < z < 30$ using five constraints from the CMB reionization principal components (PC).
+that allows one to constrain any model for reionization between $6 < z < 30$ using five constraints from the CMB reionization principal components (PC).
 We tested the code on two example models which showed excellent agreement with sampling the exact Planck likelihoods using either a simple Gaussian PC likelihood or its full kernel density estimate.
 This code enables a fast and consistent means for 
-combining Planck constraints with other reionization data sets, such as kinetic Sunyaev-Zeldovich effects, line-intensity mapping, luminosity function, star formation history, quasar spectra, etc, where the redshift dependence of the ionization history is important.
+combining Planck constraints with other reionization data sets, such as kinetic Sunyaev-Zeldovich effects, line-intensity mapping, luminosity function, star formation history, quasar spectra, etc., where the redshift dependence of the ionization history is important.
 Since the PC technique tests any reionization history in the given range, we also derive model-independent constraints for the total Thomson optical depth $\tau_{\rm PC} = 0.0619^{+0.0056}_{-0.0068}$ and its $15\le z \le 30$ high redshift component $\tau_{\rm PC}(15, 30) < 0.020 $ (95\% C.L.).

### 2021-05-11 paper/rei.tex
fixed affiliation problem by removing \usepackage{breqn}

@@
 \usepackage{tabularx,ragged2e,booktabs}
-\usepackage{breqn}
+%\usepackage{breqn}
 \usepackage[normalem]{ulem}
@@
 \author{Chen Heinrich}\email{chenhe@caltech.edu}
-\affiliation{$California\ Institute\ of\ Technology,\ Pasadena,\ California\ 91109,\ USA$}
+\affiliation{California Institute of Technology, Pasadena, California 91109,USA}
 
 \author{Wayne Hu}
-\affiliation{$Kavli\ Institute\ for\ Cosmological\ Physics,\ Enrico Fermi Institute,\ University of Chicago,\ Chicago\ Illinois\ 60637$}
-\affiliation{$Department\ of\ Astronomy\ \&\ Astrophysics,\ University\ of\ Chicago,\ Illinois\ 60637$}
+\affiliation{Kavli Institute for Cosmological Physics, Enrico Fermi Institute, University of Chicago, Chicago Illinois 60637}
+\affiliation{Department of Astronomy \& Astrophysics,
+ University of Chicago, Illinois 60637}
 

### 2021-07-08 submissions/PRD/response_to_referee.tex
added response_to_referee.tex

\documentclass[10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[OT1]{fontenc}
\usepackage{amsfonts, amsmath, amsthm, amssymb}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{listings}
\usepackage[margin=1in]{geometry}
\usepackage{color}

\newcommand{\wh}[1]{\textcolor{blue}{#1}}

\title{Response to Referee Report 1}
\author{Chen Heinrich, Wayne Hu}
\date{\today}
\begin{document}
\maketitle

We thank the referee for the thoughtful comments and suggestions. We address each of them below, and summarize changes made in the manuscript.

\begin{enumerate}
    \item{
        \textit{This paper presents a new measurement of the optical depth likelihood from the Planck 2018 package, as well as a public package for connecting reionization models to the Planck data. The paper provides a very nice overview of the constraining power of the Planck measurements in this space and a “gold standard” measurement of the optical depth. The paper is an application (with extensive testing) of earlier techniques, so I do not have any major concerns. I therefore believe it should be published, although I have a couple of questions/comments for the authors.}

        \textit{My most important comment is about the assumption that reionization is complete by z=6. As noted briefly in the conclusions, measurements of absorption toward high-z quasars over the last few years have cast considerable doubt on this assumption, even though it was conventional wisdom for many years. The data do not require reionization continuing to lower redshifts, but they certainly allow it (and some would say favor it). As such, I would strongly recommend providing some more quantitative guidance to the reader about how loosening the z$>$6 prior would affect the optical depth measurements.}\\
        \\
        The prior $z>6$ should not affect the optical depth measurement very much. One can see this by looking at the posterior plot on optical depth in the tanh model. $z=6$ corresponds to $\tau \approx 0.039$, which is disfavored by current Planck data. That being said, allowing for $z<6$ in the PC analysis could be useful when combining with other data sets. We plan to update our code release to include $z>5$ in the future.
    }

    \item{
        \textit{Second, the authors have demonstrated that the commonly-used tanh model places important implicit priors on reionization history models. While the PC approach is very nice in a mathematical sense, the tanh model is easy to visualize so offers advantages in developing intuition. I am just curious if the authors have any recommendations for more flexible functional forms that may be useful for developing intuition while allowing more, e.g., high-z contributions. This is probably beyond the scope but would be a useful service if available!
        }\\
        \\
        We think that the two-step toy model in this paper could be useful for developing intuition. It maximizes the high-redshift ionization contributions with the ionization fraction being a plateau rather than a decaying tail. In fact, Watts et al. 2019 (https://arxiv.org/abs/1910.00590) used this toy model to evaluate the potential of future large-scale CMB experiments to constrain reionization. These future experiments are able to better constrain the $10\leq l\leq20$ region, which has higher sensitivity to the high-redshift component of reionization (e.g. $x_e^{\rm min}$ in the two-step model) than $l<10$ does.
        }

    \item{
        \textit{
        Also, two very minor comments: (1) the paper roadmap at the end of the introduction ignores section V, and in the last paragraph of II.A. there appears to be a missing reference.
        }\\
        \\
        Thank you for catching that! We added the following sentence in the roadmap: ``We also derive in \S V model-independent results on the optical depth using the PC analysis."

        We also fixed the citation at the end of section II.A. Thank you again!
        }

\end{enumerate}

\end{document}

### 2021-07-08 paper/rei.tex

@@
 In \S\ref{sec:effective_likelihood}, we test the effective likelihood code and demonstrate its fidelity with two examples: 1) the standard steplike model; and 2) a two-parameter model where a plateau of ionization at high-$z$ is added to the standard tanh model for illustration purposes.
+We also derive in \S\ref{sec:modelindependent} model-independent results on the optical depth using the PC analysis.
 Finally, we summarize our results and conclude in \S\ref{sec:conclusion}.
@@
 
-We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using CosmoMC\footnote{CosmoMC: \url{http://cosmologist.info/cosmomc}} \todo{CH:cite} of the relevant Planck likelihoods discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density
+We obtain posterior constraints on the PC parameters $m_a$ from Markov Chain Monte Carlo sampling using CosmoMC\footnote{CosmoMC: \url{http://cosmologist.info/cosmomc}} \cite{Lewis:2013hha, Lewis:2002ah} of the relevant Planck likelihoods discussed in \S \ref{sec:results} in an otherwise fiducial $\Lambda$CDM cosmology with the standard 5 parameters: the baryon density $\Omega_b h^2$, cold dark matter density
 $\Omega_c h^2$, effective angular sound horizon $\theta_{\rm MC}$, scalar power spectrum log amplitude $\ln (10^{10} A_s)$ and
